---
title: "Before You Start Learning R"
author: "Benjamin Apam (PhD)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE
)
```

# Purpose of this document

This document prepares the learner's computer and working habits before the first R lesson. Completing these checks prevents the most common beginner problems, including missing packages, incorrect working directories, missing datasets and unsaved code.

# Required software

The course requires R and RStudio Desktop. R performs the calculations. RStudio provides the interface in which scripts and R Markdown documents are written and executed.

The minimum R packages are `rmarkdown` and `knitr`. Install them once by running:

```{r install-packages, eval=FALSE}
install.packages(c("rmarkdown", "knitr"))
```

Installation is different from loading. Installation places a package on the computer. Loading makes an installed package available in the current R session.

# Open the project correctly

Always open the course by double-clicking:

```text
learning-r.Rproj
```

This establishes the repository folder as the project root. The lesson files can then use stable relative paths such as:

```r
read.csv("data/student_scores.csv")
```

Avoid absolute paths such as:

```r
read.csv("C:/Users/Name/Desktop/learning-r/data/student_scores.csv")
```

Absolute paths normally work only on the computer on which they were written.

# Check the working directory

```{r working-directory}
getwd()
```

The result should end with the folder name `learning-r` or `learning-r-lesson1`, depending on where the repository was extracted.

Confirm that the first lesson is visible:

```{r check-main-files}
file.exists("01-understanding-r.Rmd")
file.exists("data/student_scores.csv")
```

Both results should be `TRUE`.

# Run the automated setup check

Open `00-setup-check.R` and click **Source**, or run:

```{r run-check, eval=FALSE}
source("00-setup-check.R")
```

The script checks the required packages, project files and the first teaching dataset.

# Understand the lesson format

Each lesson is a standalone `.Rmd` file. It contains explanatory text, equations, R code, generated output, practical work and assessment questions.

A code chunk has the following form:

````text
```{r example-name}
x <- 10
x * 2
```
````

The label after `r` identifies the chunk. Every label within a document must be unique.

# How to run code

You can run code in four main ways:

1. Place the cursor on a line and press `Ctrl + Enter`.
2. Select several lines and press `Ctrl + Enter`.
3. Click the green triangle on a code chunk.
4. Click **Knit** to execute the complete lesson and create a report.

During learning, run small sections first. After completing a lesson, restart R and knit the entire document to verify that the work is reproducible.

# Recommended file discipline

Do not rename or move the datasets before completing the lessons that use them. Do not manually modify the CSV files. When a task requires a change, create a new R object and perform the change in code.

# First diagnostic calculation

```{r first-diagnostic}
values <- c(4, 8, 12, 16)
mean(values)
```

The expected result is 10. If it appears, R is performing calculations correctly.

# First dataset import

```{r first-import}
student_scores <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

head(student_scores)
```

The output should display the first six student records.

# Session information

```{r session-information}
sessionInfo()
```

This information is useful when diagnosing package or version-related problems.

# Readiness checklist

Before beginning Lesson 1, confirm that:

- R is installed;
- RStudio Desktop is installed;
- `rmarkdown` and `knitr` are installed;
- the project was opened through `learning-r.Rproj`;
- `00-setup-check.R` completes successfully;
- `student_scores.csv` imports successfully; and
- this document knits to HTML without an error.
