---
title: "Lesson 3: The First R Session and Systematic Command Execution"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260713)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```

# Lesson overview

The first active session in R is where the learner begins to convert written
questions into executable instructions. At this stage, it is tempting to treat
the Console as a calculator and to judge success only by whether a number is
printed. That approach is too narrow.

A disciplined R session involves several connected activities:

1. reading the prompt correctly;
2. writing syntactically complete expressions;
3. executing code from the appropriate source;
4. distinguishing printed output from stored results;
5. inspecting the structure of objects;
6. checking warnings and errors;
7. preserving code in a script or R Markdown document; and
8. rerunning the work from a clean session.

This lesson develops these habits through detailed examples. The commands are
elementary, but the working practices are professional. A learner who develops
these practices early will find later work in data cleaning, statistical
modelling and automation much easier.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. interpret the ordinary R prompt and the continuation prompt;
2. distinguish the Console, Source pane and rendered document;
3. execute one line, several lines and a complete script;
4. explain how R parses and evaluates an expression;
5. use arithmetic, comparison and logical operators;
6. apply operator precedence correctly;
7. call functions with positional and named arguments;
8. create sequences and repeated patterns;
9. understand vectorised command execution;
10. distinguish automatic printing from assignment;
11. inspect objects using several complementary functions;
12. use the R help system;
13. identify incomplete expressions and common syntax errors;
14. diagnose unexpected output systematically;
15. write a clearly structured first R script; and
16. conduct a short analysis from a clean R session.

# 1. Beginning an R session

An R session begins when the R engine starts. In RStudio, the Console normally
displays an opening message followed by the ordinary prompt:

```text
>
```

The prompt indicates that R is ready to receive an instruction.

A session has a current state. This state may include:

- objects stored in the global environment;
- packages attached to the search path;
- options that have been changed;
- the current working directory;
- open graphics devices; and
- random-number-generator state.

The state changes as commands are executed.

## 1.1 Confirming that R is responsive

Enter a simple expression:

```{r confirm-r-responsive}
1 + 1
```

R returns:

```text
#> [1] 2
```

This confirms that the expression was parsed and evaluated.

The `[1]` indicates that the first value printed on that line is the first
element of the result. It does not form part of the numerical answer.

## 1.2 Inspecting the current session

```{r inspect-current-session}
R.version.string
getwd()
search()
```

These commands report:

- the current R version;
- the current working directory; and
- the current search path.

They describe different aspects of the session.

# 2. Console, Source pane and rendered output

## 2.1 The Console

The Console executes commands immediately. It is suitable for:

- testing a short expression;
- inspecting an object;
- checking a function argument;
- experimenting with an unfamiliar command; and
- reading an error message.

It is not the best place to keep important analytical work because Console
commands may not remain organised or reproducible.

## 2.2 The Source pane

The Source pane contains saved `.R` or `.Rmd` files.

Code written in a source file can be:

- saved;
- reviewed;
- corrected;
- shared;
- version controlled; and
- rerun from the beginning.

A professional workflow uses the Source pane as the permanent analytical
record and the Console as an execution and diagnostic area.

## 2.3 Rendered output

When an R Markdown document is knitted, the code chunks are executed in order
and the results are inserted into the output document.

Rendered output may contain:

- explanatory prose;
- code;
- printed results;
- tables;
- figures;
- equations; and
- automatically generated values.

The rendered document is a report. The `.Rmd` file remains the source.

# 3. The ordinary prompt and continuation prompt

## 3.1 Ordinary prompt

```text
>
```

The ordinary prompt means that R is ready for a new expression.

## 3.2 Continuation prompt

```text
+
```

The continuation prompt means that the previous expression is incomplete.

For example, this command has no closing parenthesis:

```r
mean(c(10, 20, 30)
```

R waits because it cannot determine whether the expression has ended.

Other causes include:

- an unclosed quotation mark;
- an unclosed bracket;
- an unfinished function call;
- a trailing arithmetic operator; and
- an incomplete block of code.

## 3.3 Completing an expression

The incomplete command:

```r
mean(c(10, 20, 30)
```

can be completed by entering:

```r
)
```

However, beginners should generally cancel and rewrite the complete command
clearly.

In RStudio, press:

```text
Esc
```

to cancel the unfinished expression.

## 3.4 Unclosed quotation example

```r
programme <- "Statistics
```

R waits because the text has not been closed with another quotation mark.

Correct:

```{r complete-character-expression}
programme <- "Statistics"
programme
```

# 4. Expressions and evaluation

An expression is something R can evaluate.

Examples include:

```{r expression-examples}
25
4 + 7
sqrt(81)
c(10, 20, 30)
```

R evaluates each expression and returns a result.

## 4.1 Literal values

```{r literal-values}
42
3.5
"Bolgatanga"
TRUE
```

These are literal values because they are written directly.

## 4.2 Compound expression

```{r compound-expression}
(12 + 8) / 4
```

R evaluates the parenthesised addition first and then divides the result.

\[
\frac{12+8}{4}
=
\frac{20}{4}
=
5.
\]

## 4.3 Function call as an expression

```{r function-call-expression}
mean(c(12, 15, 18, 21))
```

The inner function `c()` creates a vector. The outer function `mean()`
calculates its arithmetic mean.

# 5. Running code from the Source pane

## 5.1 Running one line

Place the cursor on a line and press:

```text
Ctrl + Enter
```

RStudio sends the current line to the Console.

## 5.2 Running selected lines

Highlight several lines and press:

```text
Ctrl + Enter
```

Only the selected code is executed.

## 5.3 Running a complete chunk

Inside an R Markdown document, use the green Run button on the code chunk or
the relevant keyboard shortcut.

## 5.4 Running all earlier chunks

RStudio provides commands to run all chunks above the current position. This is
useful when a later chunk depends on objects created earlier.

However, a well-designed lesson should also knit successfully from the
beginning without requiring manual intervention.

## 5.5 Running a complete R script

```{r source-complete-script, eval=FALSE}
source("my_analysis.R")
```

`source()` reads and executes the commands in the specified `.R` file.

# 6. Arithmetic operators in practical contexts

## 6.1 Addition

Suppose a training programme has 48 participants in the morning session and 37
in the afternoon session.

```{r arithmetic-addition-context}
morning_participants <- 48
afternoon_participants <- 37

total_participants <- morning_participants + afternoon_participants
total_participants
```

\[
48+37=85.
\]

## 6.2 Subtraction

If 6 participants leave before completion:

```{r arithmetic-subtraction-context}
completed_participants <- total_participants - 6
completed_participants
```

## 6.3 Multiplication

Suppose each completed participant receives a GH₵25 meal allowance.

```{r arithmetic-multiplication-context}
meal_allowance_per_person <- 25

total_meal_allowance <-
  completed_participants * meal_allowance_per_person

total_meal_allowance
```

## 6.4 Division

Share the total allowance equally across five budget lines:

```{r arithmetic-division-context}
number_of_budget_lines <- 5

allowance_per_budget_line <-
  total_meal_allowance / number_of_budget_lines

allowance_per_budget_line
```

## 6.5 Exponentiation

Suppose a rectangular plot is 8 metres long and 8 metres wide.

```{r exponentiation-context}
plot_side <- 8
plot_area <- plot_side ^ 2

plot_area
```

## 6.6 Integer division

Suppose 103 questionnaires are packed into bundles of 20.

```{r integer-division-context}
questionnaires <- 103
bundle_size <- 20

complete_bundles <- questionnaires %/% bundle_size
complete_bundles
```

## 6.7 Remainder

```{r remainder-context}
unbundled_questionnaires <- questionnaires %% bundle_size
unbundled_questionnaires
```

The relationship is:

\[
103=(5\times20)+3.
\]

# 7. Operator precedence

R follows an order of operations.

A simplified order is:

1. parentheses;
2. exponentiation;
3. multiplication and division;
4. addition and subtraction;
5. comparisons;
6. logical operations.

## 7.1 Multiplication before addition

```{r precedence-multiplication-before-addition}
10 + 4 * 3
```

\[
10 + (4\times3)=22.
\]

## 7.2 Parentheses change the result

```{r precedence-parentheses}
(10 + 4) * 3
```

\[
(10+4)\times3=42.
\]

## 7.3 Exponentiation

```{r precedence-exponentiation}
2 + 3 ^ 2
```

\[
2+3^2=2+9=11.
\]

## 7.4 Clearer code with parentheses

Even when parentheses are not mathematically required, they may improve
readability:

```{r precedence-readable-expression}
weighted_result <-
  (0.40 * 75) +
  (0.60 * 68)

weighted_result
```

# 8. Comparison operators

Comparison operators return logical values.

| Operator | Meaning |
|---|---|
| `>` | greater than |
| `<` | less than |
| `>=` | greater than or equal to |
| `<=` | less than or equal to |
| `==` | equal to |
| `!=` | not equal to |

## 8.1 Single comparisons

```{r single-comparisons}
75 > 50
45 >= 50
10 == 10
10 != 8
```

## 8.2 Pass-status context

```{r pass-status-context}
final_score <- 64
pass_mark <- 50

final_score >= pass_mark
```

## 8.3 Exact equality

Use:

```r
==
```

to test equality.

```{r equality-comparison}
recorded_total <- 82
calculated_total <- 32 + 50

recorded_total == calculated_total
```

A single equals sign is not the standard comparison operator.

# 9. Logical operators

Logical operators combine or reverse logical conditions.

| Operator | Meaning |
|---|---|
| `&` | both conditions are true |
| `|` | at least one condition is true |
| `!` | reverse a logical value |

## 9.1 Both conditions

A student qualifies for a scholarship when:

- the final score is at least 80; and
- attendance is at least 90 per cent.

```{r logical-and-context}
score <- 84
attendance_rate <- 0.94

scholarship_eligible <-
  (score >= 80) &
  (attendance_rate >= 0.90)

scholarship_eligible
```

## 9.2 At least one condition

A student requires academic counselling when the score is below 50 or
attendance is below 75 per cent.

```{r logical-or-context}
requires_counselling <-
  (score < 50) |
  (attendance_rate < 0.75)

requires_counselling
```

## 9.3 Negation

```{r logical-negation-context}
!scholarship_eligible
```

# 10. Functions and their arguments

A function performs a defined operation.

## 10.1 One argument

```{r one-argument-function}
sqrt(144)
```

## 10.2 Several positional arguments

```{r positional-arguments}
round(3.14159265, 3)
```

The second argument is interpreted according to its position.

## 10.3 Named arguments

```{r named-arguments}
round(
  x = 3.14159265,
  digits = 3
)
```

Named arguments make the instruction clearer.

## 10.4 Default arguments

```{r default-argument-behaviour}
round(3.14159265)
```

When `digits` is omitted, the function uses its default.

## 10.5 Inspecting arguments

```{r inspect-function-arguments}
args(round)
args(seq)
args(mean)
```

## 10.6 Nested functions

```{r nested-function-context}
round(
  mean(
    c(65, 72, 81, 76, 89)
  ),
  digits = 1
)
```

Read nested calls from the inside outward.

# 11. Frequently used mathematical functions

## 11.1 Absolute value

```{r absolute-value-function}
abs(-15)
```

## 11.2 Square root

```{r square-root-function}
sqrt(225)
```

## 11.3 Natural logarithm

```{r natural-logarithm-function}
log(10)
```

## 11.4 Base-10 logarithm

```{r base-ten-logarithm-function}
log10(1000)
```

## 11.5 Exponential function

```{r exponential-function}
exp(1)
```

## 11.6 Ceiling and floor

```{r ceiling-floor-functions}
ceiling(4.2)
floor(4.8)
```

`ceiling()` rounds upward to the next integer. `floor()` rounds downward.

## 11.7 Sign

```{r sign-function}
sign(-10)
sign(0)
sign(24)
```

# 12. Creating sequences

## 12.1 Colon operator

```{r colon-sequence}
1:10
```

This creates consecutive integers.

A descending sequence is also possible:

```{r descending-colon-sequence}
10:1
```

## 12.2 `seq()` with an interval

```{r seq-by-interval}
seq(
  from = 0,
  to = 100,
  by = 10
)
```

## 12.3 `seq()` with a required length

```{r seq-by-length}
seq(
  from = 0,
  to = 1,
  length.out = 6
)
```

## 12.4 Monthly identifiers

```{r month-identifiers}
month_number <- seq(
  from = 1,
  to = 12,
  by = 1
)

month_number
```

# 13. Repeating values

## 13.1 Repeat one value

```{r repeat-one-value}
rep(
  "Control",
  times = 5
)
```

## 13.2 Repeat a pattern

```{r repeat-pattern}
rep(
  c("Control", "Treatment"),
  times = 3
)
```

## 13.3 Repeat each element

```{r repeat-each-element}
rep(
  c("Control", "Treatment"),
  each = 3
)
```

The difference between `times` and `each` is important.

## 13.4 Experimental group context

```{r experimental-groups}
plot_id <- 1:8

treatment <- rep(
  c("A", "B"),
  each = 4
)

plot_id
treatment
```

# 14. Combining values

The `c()` function combines values.

```{r combine-numeric-values}
scores <- c(
  68,
  74,
  81,
  59,
  90
)

scores
```

## 14.1 Length

```{r vector-length}
length(scores)
```

## 14.2 Total

```{r vector-total}
sum(scores)
```

## 14.3 Mean

```{r vector-mean}
mean(scores)
```

## 14.4 Minimum and maximum

```{r vector-minimum-maximum}
min(scores)
max(scores)
```

## 14.5 Summary

```{r vector-summary}
summary(scores)
```

# 15. Vectorised execution

R ordinarily applies operations element by element.

## 15.1 Add two vectors

```{r vectorised-addition}
coursework <- c(28, 32, 25, 36)
examination <- c(45, 50, 39, 52)

final_scores <- coursework + examination
final_scores
```

The operation is:

\[
(28+45,\ 32+50,\ 25+39,\ 36+52).
\]

## 15.2 Multiply a vector by a constant

```{r vectorised-multiplication}
proportions <- c(0.45, 0.62, 0.80, 0.91)
percentages <- proportions * 100

percentages
```

## 15.3 Compare all elements

```{r vectorised-comparison}
final_scores >= 50
```

## 15.4 Count logical results

```{r count-logical-results}
sum(final_scores >= 50)
```

## 15.5 Proportion satisfying a condition

```{r proportion-logical-results}
mean(final_scores >= 50)
```

# 16. Recycling

When vectors differ in length, R may repeat the shorter vector.

```{r exact-recycling}
c(10, 20, 30, 40) + c(1, 2)
```

The shorter vector is treated as:

\[
(1,2,1,2).
\]

## 16.1 Recycling warning

```{r recycling-warning, warning=TRUE}
c(10, 20, 30, 40, 50) + c(1, 2)
```

Because 5 is not an exact multiple of 2, R warns that the lengths are not
compatible.

A warning does not necessarily stop execution. It signals that the result
should be examined.

# 17. Automatic printing and invisible results

## 17.1 Automatic printing

```{r automatic-printing}
scores
```

Entering an object name at the top level prints its value.

## 17.2 Assignment suppresses printing

```{r assignment-no-print}
mean_score <- mean(scores)
```

The result is stored but not automatically printed.

```{r print-assigned-result}
mean_score
```

## 17.3 Explicit printing

```{r explicit-print-function}
print(mean_score)
```

Explicit printing becomes important inside loops and functions.

## 17.4 Parenthesised assignment

```{r parenthesised-assignment}
(rounded_mean <- round(mean_score, digits = 1))
```

Parentheses cause the assigned result to be printed.

# 18. Inspecting objects systematically

No single inspection function answers every question.

## 18.1 Class

```{r inspect-class}
class(scores)
```

## 18.2 Internal type

```{r inspect-typeof}
typeof(scores)
```

## 18.3 Length

```{r inspect-length}
length(scores)
```

## 18.4 Structure

```{r inspect-structure}
str(scores)
```

## 18.5 Attributes

```{r inspect-attributes}
attributes(scores)
```

A simple unnamed numeric vector may have no additional attributes.

## 18.6 First and last values

```{r inspect-head-tail}
head(scores)
tail(scores)
```

# 19. Commands that inspect the environment

## 19.1 List objects

```{r inspect-ls}
ls()
```

## 19.2 Check existence

```{r inspect-exists}
exists("scores")
exists("unknown_object")
```

## 19.3 Remove an object

```{r remove-object}
temporary_result <- 999
rm(temporary_result)
```

## 19.4 Confirm removal

```{r confirm-object-removal}
exists("temporary_result")
```

# 20. The R help system

## 20.1 Direct help

```{r direct-help, eval=FALSE}
?mean
```

Equivalent:

```{r help-function, eval=FALSE}
help(mean)
```

## 20.2 Search help

```{r search-help, eval=FALSE}
help.search("standard deviation")
```

## 20.3 View examples

```{r view-examples, eval=FALSE}
example(mean)
```

## 20.4 Inspect function arguments

```{r inspect-mean-args}
args(mean)
```

## 20.5 Search the web from R help

```{r rsite-search, eval=FALSE}
RSiteSearch("vector recycling")
```

# 21. Syntax errors

A syntax error means that R cannot parse the instruction.

## 21.1 Missing parenthesis

```r
mean(c(10, 20, 30)
```

## 21.2 Missing comma

```r
round(3.14159 digits = 2)
```

## 21.3 Unclosed quotation

```r
institution <- "Bolgatanga Technical University
```

## 21.4 Invalid object name

```r
2nd_score <- 80
```

## 21.5 Correct forms

```{r corrected-syntax-examples}
mean(c(10, 20, 30))

round(
  3.14159,
  digits = 2
)

institution <-
  "Bolgatanga Technical University"

second_score <- 80
```

# 22. Runtime errors

A runtime error occurs after parsing, while R attempts to evaluate the
expression.

## 22.1 Object not found

```{r object-not-found-demo, error=TRUE}
mean(nonexistent_scores)
```

## 22.2 Inappropriate type

```{r inappropriate-type-demo, error=TRUE}
sqrt("sixteen")
```

## 22.3 Missing file

```{r missing-file-demo, error=TRUE}
read.csv("data/file_that_does_not_exist.csv")
```

# 23. Warnings

A warning indicates a potential problem but often allows execution to continue.

## 23.1 Mean of non-numeric text

```{r warning-nonnumeric-mean, warning=TRUE}
mean(c("High", "Medium", "Low"))
```

## 23.2 Recycling warning

```{r warning-recycling-repeat, warning=TRUE}
1:5 + 1:2
```

Warnings must be read and interpreted. They should not be suppressed merely to
make output look cleaner.

# 24. Messages

Some functions provide messages rather than warnings or errors.

Package attachment often produces messages about:

- attached packages;
- masked functions;
- startup notes.

In a teaching document, messages may be hidden with:

```r
message = FALSE
```

only when the underlying information is already understood.

# 25. Debugging unexpected output

Not every problem produces an error.

Suppose the intention is to calculate a percentage:

```{r unexpected-proportion}
passed <- 18
total <- 25

passed / total
```

The result is a proportion, 0.72.

To express it as a percentage:

```{r expected-percentage}
(passed / total) * 100
```

The first result was not erroneous. It simply answered the question on a
different scale.

## 25.1 Diagnostic questions

When output is unexpected, ask:

1. Did I use the correct object?
2. Does the object contain the values I expect?
3. Is the object of the correct type?
4. Are missing values present?
5. Did operator precedence change the calculation?
6. Did vector recycling occur?
7. Is the result a proportion rather than a percentage?
8. Was the correct function used?
9. Did a previous object remain in memory?
10. Does the result have plausible units?

# 26. A structured first script

A clear script can contain the following sections:

```r
# ==============================================================================
# Project information
# ==============================================================================

# ==============================================================================
# Import data
# ==============================================================================

# ==============================================================================
# Validate data
# ==============================================================================

# ==============================================================================
# Analyse data
# ==============================================================================

# ==============================================================================
# Export results
# ==============================================================================
```

## 26.1 Worked script

```{r structured-script-example}
# Import the supplied student dataset.
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

# Validate that final scores lie within 0 and 100.
scores_are_valid <- all(
  student_data$final_score >= 0 &
    student_data$final_score <= 100
)

# Calculate summaries.
number_of_students <- nrow(student_data)
mean_final_score <- mean(student_data$final_score)
median_final_score <- median(student_data$final_score)
pass_percentage <-
  mean(student_data$final_score >= 50) * 100

# Display results.
scores_are_valid
number_of_students
mean_final_score
median_final_score
pass_percentage
```

# 27. Running code from a clean session

A clean-session test is performed as follows:

1. save the source file;
2. restart R;
3. open the project;
4. run the file from the first line;
5. inspect warnings and errors;
6. confirm that all outputs are recreated.

This test helps identify:

- objects created manually but not documented;
- packages loaded earlier but not declared;
- incorrect working-directory assumptions;
- missing input files; and
- code executed in the wrong order.

# 28. Integrated analysis: clinic waiting times

## 28.1 Import the data

```{r integrated-clinic-import}
clinic <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)
```

## 28.2 Inspect the data

```{r integrated-clinic-inspect}
head(clinic)
str(clinic)
dim(clinic)
names(clinic)
```

## 28.3 Calculate summary statistics

```{r integrated-clinic-summary}
mean_waiting_time <- mean(
  clinic$waiting_time,
  na.rm = TRUE
)

median_waiting_time <- median(
  clinic$waiting_time,
  na.rm = TRUE
)

maximum_waiting_time <- max(
  clinic$waiting_time,
  na.rm = TRUE
)

mean_waiting_time
median_waiting_time
maximum_waiting_time
```

## 28.4 Evaluate a service target

Assume that the service target is 30 minutes.

```{r integrated-clinic-target}
waiting_target <- 30

target_met <- clinic$waiting_time <= waiting_target
target_met
```

## 28.5 Count and percentage

```{r integrated-clinic-target-summary}
number_meeting_target <- sum(
  target_met,
  na.rm = TRUE
)

percentage_meeting_target <- mean(
  target_met,
  na.rm = TRUE
) * 100

number_meeting_target
percentage_meeting_target
```

## 28.6 Interpretation

The mean summarises the arithmetic centre of waiting times. The median
identifies the middle waiting time and may be less sensitive to unusually long
waits. The target percentage reports the share of observed visits completed
within 30 minutes.

# 29. Guided practice

## 29.1 Education

```{r guided-education}
scores <- c(48, 55, 62, 71, 39, 84, 66)

mean(scores)
median(scores)
sum(scores >= 50)
mean(scores >= 50) * 100
```

Explain each output.

## 29.2 Agriculture

```{r guided-agriculture}
plot_yield <- c(2.4, 2.9, 3.1, 2.7, 3.5, 2.2)

mean(plot_yield)
min(plot_yield)
max(plot_yield)
plot_yield >= 3
```

## 29.3 Procurement

```{r guided-procurement}
delivery_days <- c(4, 7, 5, 3, 9, 6)
agreed_days <- 5

delivery_days <= agreed_days
sum(delivery_days <= agreed_days)
mean(delivery_days <= agreed_days) * 100
```

## 29.4 Business

```{r guided-business}
monthly_revenue <- c(
  12500,
  13800,
  14250,
  13100,
  15500,
  16000
)

sum(monthly_revenue)
mean(monthly_revenue)
monthly_revenue > 14000
```

# 30. Independent exercises

## Exercise 1

Explain the difference among the Console, Source pane and rendered output.

## Exercise 2

Give three examples of incomplete expressions that cause the continuation
prompt.

## Exercise 3

Calculate the total cost of 45 items costing GH₵18.50 each. Store each input
and the result in clearly named objects.

## Exercise 4

Use `%/%` and `%%` to express 289 minutes as complete hours and remaining
minutes.

## Exercise 5

Demonstrate the difference between:

```r
8 + 2 * 5
```

and:

```r
(8 + 2) * 5
```

Explain the result.

## Exercise 6

Create a sequence from 5 to 50 in intervals of 5.

## Exercise 7

Use `rep()` to create:

```text
A A A B B B C C C
```

## Exercise 8

Create coursework and examination vectors for six students. Add them
element-wise and calculate the pass percentage.

## Exercise 9

Create one example each of:

- a syntax error;
- a runtime error;
- a warning; and
- unexpected but valid output.

Explain the difference.

## Exercise 10

Import `rainfall_records.csv`, inspect its structure and calculate the observed
mean rainfall after accounting for missing values.

## Exercise 11

Write a complete R script that imports `monthly_sales.csv`, validates positive
revenue, calculates total revenue and reports the percentage of records with
revenue above the mean.

## Exercise 12

Restart R and run the complete script from Exercise 11. Record any dependency
that was missing and correct it.

# 31. Lesson assessment

## Part A: Conceptual questions

1. What does the ordinary R prompt mean?
2. What causes the continuation prompt?
3. Why is the Console not sufficient as the permanent analytical record?
4. What is an expression?
5. How does assignment affect automatic printing?
6. What is operator precedence?
7. What is the difference between `==` and `<-`?
8. What do `&`, `|` and `!` mean?
9. What is a positional argument?
10. What is a named argument?
11. What is vectorisation?
12. What is recycling?
13. Why can recycling be dangerous?
14. What is the difference between an error and a warning?
15. What is hidden session state?
16. Why should a script be tested after restarting R?
17. What information does `str()` provide?
18. Why should several inspection functions be used?
19. What is the role of `source()`?
20. Why can valid code still produce an inappropriate answer?

## Part B: Practical assessment

Create a standalone R Markdown report that:

1. imports `clinic_visits.csv`;
2. inspects the data;
3. calculates mean, median, minimum and maximum waiting time;
4. determines the number and percentage meeting a 30-minute target;
5. demonstrates one arithmetic operator of each major type;
6. demonstrates one comparison and one combined logical condition;
7. creates one sequence and one repeated pattern;
8. contains one intentionally displayed warning with an explanation;
9. contains no unresolved error;
10. explains every result in complete sentences; and
11. knits successfully from a clean R session.

# 32. Key terms

**Automatic printing:** Display of an expression's value when it is evaluated
at the top level without assignment.

**Console:** The interface through which R receives commands directly.

**Continuation prompt:** The `+` prompt displayed when an expression is
incomplete.

**Evaluation:** The process by which R calculates the meaning and result of an
expression.

**Expression:** Code that R can parse and evaluate.

**Logical operator:** An operator that combines or reverses logical conditions.

**Named argument:** A function input supplied using its argument name.

**Operator precedence:** Rules determining the order in which parts of an
expression are evaluated.

**Positional argument:** A function input interpreted according to its
position.

**Recycling:** Repetition of a shorter vector during an operation involving
vectors of unequal length.

**Runtime error:** An error arising while R evaluates syntactically valid code.

**Source file:** A saved file containing R or R Markdown code.

**Syntax error:** An error preventing R from parsing an expression.

**Vectorisation:** Element-by-element application of an operation across a
vector.

**Warning:** A message indicating a possible problem while often allowing the
operation to continue.

# 33. Completion checklist

Before proceeding, confirm that you can:

- identify the ordinary prompt;
- identify and cancel the continuation prompt;
- execute one line and several selected lines;
- explain expressions and evaluation;
- use arithmetic operators;
- apply operator precedence;
- use comparison operators;
- combine logical conditions;
- call functions with named arguments;
- inspect function arguments;
- create sequences;
- repeat patterns;
- combine values into vectors;
- explain vectorisation;
- recognise recycling;
- distinguish assignment from automatic printing;
- inspect objects with `class()`, `typeof()`, `length()` and `str()`;
- use the help system;
- identify syntax errors;
- identify runtime errors;
- interpret warnings;
- debug unexpected output;
- structure an R script;
- run code from a clean session; and
- conduct a small contextual analysis.

# Lesson conclusion

The first R session should not be approached as a sequence of unrelated
commands. It should be treated as a disciplined computational process.

The learner writes an expression, R parses and evaluates it, and the learner
then inspects, validates and interprets the result. Important code is preserved
in a source file and tested from a clean session.

The working sequence is:

\[
\boxed{
\text{write}
\rightarrow
\text{execute}
\rightarrow
\text{inspect}
\rightarrow
\text{validate}
\rightarrow
\text{interpret}
\rightarrow
\text{save}
}
\]

The next lesson develops the object system in greater depth, including
assignment, naming conventions, object inspection, copying, updating,
attributes, coercion, deletion and safe object-management practices.
