---
title: "Lesson 4: Objects, Assignment and Naming Conventions"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260713)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```

# Lesson overview

Objects are central to R. Data, results, models, functions and graphs are all
represented as objects. A learner who understands only commands but not objects
will struggle to build reliable analytical workflows.

An object has at least three important features:

1. a name through which it can be referenced;
2. a value or structure stored in memory; and
3. properties such as type, class, length or attributes.

Assignment connects a name to a value. Naming conventions make code readable.
Inspection functions reveal what an object contains. Updating and deletion
change the state of the R session.

This lesson develops object management carefully because later lessons depend
on it. The aim is not merely to create `x <- 10`. The aim is to understand what
has been created, how it behaves, how it can be verified and how naming choices
affect the reliability of a complete analysis.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain what an R object is;
2. use the preferred assignment operator;
3. distinguish assignment from comparison and named arguments;
4. create objects containing numeric, character and logical values;
5. choose valid and meaningful object names;
6. explain case sensitivity;
7. inspect type, class, length and structure;
8. copy and update objects;
9. understand object independence after ordinary assignment;
10. use names and attributes;
11. list, search for and remove objects;
12. avoid accidental overwriting;
13. recognise reserved words and invalid names;
14. use backticks only when unavoidable;
15. understand basic coercion caused by mixed values;
16. validate objects before analysis;
17. organise related objects consistently; and
18. develop a safe object-management workflow.

# 1. What is an object?

An object is a named structure stored in R's memory.

Objects may contain:

- one number;
- several numbers;
- text;
- logical values;
- dates;
- tables;
- functions;
- statistical models;
- plots; or
- other objects.

Create a simple object:

```{r first-object}
student_count <- 120
```

This instruction has three components:

- `student_count` is the object name;
- `<-` is the assignment operator;
- `120` is the value assigned.

The instruction can be read as:

> Store the value 120 under the name `student_count`.

Display the object:

```{r display-first-object}
student_count
```

# 2. Assignment

## 2.1 Preferred assignment operator

```{r preferred-assignment}
coursework_score <- 34
```

The operator `<-` is widely preferred for ordinary R assignment because it
visually separates assignment from other uses of `=`.

## 2.2 Equals assignment

R also permits:

```{r equals-assignment}
examination_score = 52
```

This creates the object. However, the course will use `<-` for ordinary
assignment.

## 2.3 Named arguments use `=`

```{r named-argument-equals}
round(
  x = 3.14159,
  digits = 2
)
```

Here, `=` supplies values to named function arguments. It does not create
objects named `x` or `digits` in the global environment.

## 2.4 Right assignment

R also supports:

```{r right-assignment}
75 -> final_score
final_score
```

This is valid but less common. Left assignment is generally easier to read in
long analytical workflows.

## 2.5 Assignment does not print automatically

```{r assignment-invisible-result}
total_score <- coursework_score + examination_score
```

The value has been stored but is not automatically displayed.

```{r inspect-total-score}
total_score
```

## 2.6 Parenthesised assignment

```{r parenthesised-assignment}
(adjusted_score <- total_score + 2)
```

The object is created and the value is printed.

# 3. Assignment versus equality

In mathematics:

\[
x=10
\]

may express equality.

In R:

```r
x <- 10
```

assigns a value.

To test whether `x` equals 10, use:

```{r equality-test}
x <- 10

x == 10
```

The distinction is:

| Purpose | Operator |
|---|---|
| assign a value | `<-` |
| compare for equality | `==` |
| supply a named argument | `=` |

# 4. Creating objects of different basic kinds

## 4.1 Numeric object

```{r numeric-object}
student_age <- 23
```

## 4.2 Character object

```{r character-object}
student_programme <- "Applied Statistics"
```

Text must be enclosed in quotation marks.

## 4.3 Logical object

```{r logical-object}
fees_paid <- TRUE
```

`TRUE` and `FALSE` are logical constants and are not enclosed in quotation
marks.

## 4.4 Missing object value

```{r missing-object}
middle_name <- NA
```

`NA` represents a missing value.

## 4.5 Several values

```{r vector-object}
assessment_scores <- c(72, 68, 81, 75)
```

The object contains a numeric vector of length four.

# 5. Valid object names

## 5.1 Recommended names

```{r recommended-object-names}
number_of_students <- 120
mean_final_score <- 67.5
programme_name <- "Statistics"
```

These names communicate meaning.

## 5.2 `snake_case`

The course recommends `snake_case`, in which words are separated by
underscores.

Examples:

```text
monthly_revenue
student_identifier
mean_waiting_time
number_of_records
```

## 5.3 Names may contain numbers

```{r names-containing-numbers}
score_1 <- 68
score_2 <- 74
```

However, a name should not begin with a number.

Invalid:

```r
1st_score <- 68
```

Valid:

```{r valid-first-score}
first_score <- 68
```

## 5.4 Dots in names

```{r dot-object-name}
student.score <- 70
```

This is valid, but mixing dots and underscores inconsistently can reduce
readability. The course uses underscores.

## 5.5 Avoid spaces

Invalid:

```r
student score <- 70
```

R interprets the space as a separation between tokens.

Use:

```r
student_score <- 70
```

# 6. Case sensitivity

R distinguishes upper-case and lower-case letters.

```{r case-sensitive-objects}
score <- 60
Score <- 75
SCORE <- 90

score
Score
SCORE
```

These are three separate objects.

A consistent lower-case naming style reduces mistakes.

## 6.1 Common case error

```{r common-case-error, error=TRUE}
mean(Assessment_Scores)
```

The existing object is named `assessment_scores`, not
`Assessment_Scores`.

Correct:

```{r correct-case-reference}
mean(assessment_scores)
```

# 7. Reserved words

Some words have special meanings in R and cannot be used as ordinary syntactic
object names.

Examples include:

```text
if
else
repeat
while
function
for
in
next
break
TRUE
FALSE
NULL
Inf
NaN
NA
```

Attempting to use a reserved word may produce a syntax error.

Poor naming choice:

```r
function <- 10
```

Use a descriptive alternative:

```{r reserved-word-alternative}
function_count <- 10
```

# 8. Non-standard names and backticks

R can sometimes work with non-standard names by enclosing them in backticks.

```{r backtick-object-name}
`student score` <- 78
`student score`
```

Although valid, this style is inconvenient.

Prefer:

```{r standard-object-name}
student_score <- 78
```

Backticks are more commonly encountered when importing datasets containing
spaces or punctuation in variable names.

# 9. Inspecting objects

Object inspection should become automatic.

## 9.1 Display value

```{r inspect-value}
student_programme
```

## 9.2 Internal type

```{r inspect-object-type}
typeof(student_age)
typeof(student_programme)
typeof(fees_paid)
```

## 9.3 Class

```{r inspect-object-class}
class(student_age)
class(student_programme)
class(fees_paid)
```

## 9.4 Length

```{r inspect-object-length}
length(student_age)
length(assessment_scores)
```

A scalar-like value in R is ordinarily a vector of length one.

## 9.5 Structure

```{r inspect-object-structure}
str(student_age)
str(student_programme)
str(assessment_scores)
```

## 9.6 Attributes

```{r inspect-object-attributes}
attributes(assessment_scores)
```

An unnamed numeric vector may have no attributes.

## 9.7 Combined inspection

```{r combined-object-inspection}
object_report <- c(
  type = typeof(assessment_scores),
  class = class(assessment_scores),
  length = length(assessment_scores)
)

object_report
```

# 10. Copying objects

```{r copy-object}
original_scores <- c(65, 72, 80)
copied_scores <- original_scores

original_scores
copied_scores
```

At the level required in this course, the two objects can be treated as
independent values after assignment.

Change the copy:

```{r change-copied-object}
copied_scores[1] <- 90

original_scores
copied_scores
```

The original remains unchanged.

# 11. Updating objects

## 11.1 Replace a value completely

```{r replace-object-value}
registration_count <- 100
registration_count <- 112

registration_count
```

## 11.2 Update using the current value

```{r update-using-current-value}
account_balance <- 1500
account_balance <- account_balance + 350

account_balance
```

R evaluates the right-hand side using the old value and then assigns the new
result.

## 11.3 Percentage increase

```{r percentage-increase-object}
initial_salary <- 3200
increase_rate <- 0.10

new_salary <-
  initial_salary *
  (1 + increase_rate)

new_salary
```

## 11.4 Accumulation

```{r accumulated-total}
running_total <- 0

running_total <- running_total + 125
running_total <- running_total + 210
running_total <- running_total + 95

running_total
```

This pattern later appears in loops.

# 12. Accidental overwriting

An existing object is replaced without a warning when the same name is assigned
again.

```{r accidental-overwriting}
survey_data <- c(1, 2, 3)

survey_data <- "This replaced the data"

survey_data
```

R assumes that the replacement was intentional.

## 12.1 Prevention strategies

- use descriptive names;
- inspect the Environment pane;
- avoid generic names such as `data`, `test` and `temp`;
- review code before running blocks;
- keep raw and cleaned data under different names;
- rerun scripts from the beginning; and
- use version control for important projects.

## 12.2 Better naming

```{r better-data-names}
raw_student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

clean_student_data <- raw_student_data
```

The names communicate the workflow stage.

# 13. Object existence

## 13.1 Check one object

```{r check-object-existence}
exists("raw_student_data")
exists("unknown_data")
```

## 13.2 Conditional creation

```{r conditional-object-creation}
if (!exists("analysis_title")) {
  analysis_title <-
    "Student Performance Analysis"
}

analysis_title
```

## 13.3 Avoid relying on existence checks to hide incomplete scripts

A script should normally create the objects it requires. Excessive conditional
creation can make the workflow harder to understand.

# 14. Listing objects

## 14.1 `ls()`

```{r list-objects-ls}
ls()
```

## 14.2 `objects()`

```{r list-objects-objects}
objects()
```

`objects()` is an alias for `ls()`.

## 14.3 Filter names by pattern

```{r filter-object-names}
ls(
  pattern = "score"
)
```

## 14.4 Include hidden names

```{r include-hidden-object-names}
ls(
  all.names = TRUE
)
```

# 15. Removing objects

## 15.1 Remove one object

```{r remove-one-object}
temporary_score <- 99
rm(temporary_score)
```

Confirm:

```{r confirm-one-object-removal}
exists("temporary_score")
```

## 15.2 Remove several objects

```{r create-removal-objects}
temp_a <- 1
temp_b <- 2
temp_c <- 3
```

```{r remove-several-objects}
rm(
  temp_a,
  temp_b,
  temp_c
)
```

## 15.3 Remove by character name

```{r remove-by-character-name}
temporary_name <- "temporary_result"
temporary_result <- 500

rm(
  list = temporary_name
)
```

## 15.4 Clearing the whole environment

```r
rm(list = ls())
```

This command removes all ordinary objects in the global environment.

It should not be placed casually in teaching notes or run without considering
unsaved work. Restarting R is usually a clearer way to obtain a clean session.

# 16. Naming objects by analytical role

Names should distinguish inputs, intermediate values and outputs.

## 16.1 Input objects

```{r role-input-objects}
coursework_marks <- c(30, 28, 35, 24)
examination_marks <- c(48, 42, 51, 39)
```

## 16.2 Derived object

```{r role-derived-object}
final_marks <-
  coursework_marks +
  examination_marks
```

## 16.3 Summary objects

```{r role-summary-objects}
mean_final_mark <- mean(final_marks)
pass_rate <- mean(final_marks >= 50)
```

## 16.4 Reporting object

```{r role-reporting-object}
pass_percentage <- pass_rate * 100
```

The names communicate the sequence.

# 17. Names attached to vector elements

A vector may have element names.

```{r named-vector}
programme_counts <- c(
  Statistics = 35,
  Procurement = 42,
  Hospitality = 28,
  Agriculture = 31
)

programme_counts
```

## 17.1 Inspect names

```{r inspect-vector-names}
names(programme_counts)
```

## 17.2 Extract by name

```{r extract-named-element}
programme_counts["Statistics"]
```

## 17.3 Assign names after creation

```{r assign-vector-names}
department_budget <- c(
  5000,
  7200,
  6100
)

names(department_budget) <- c(
  "Teaching",
  "Research",
  "Outreach"
)

department_budget
```

# 18. Attributes

Attributes provide additional information about an object.

```{r inspect-named-vector-attributes}
attributes(programme_counts)
```

The `names` attribute stores the element names.

## 18.1 Add a custom attribute

```{r custom-attribute}
temperature <- c(29.5, 31.2, 30.1)

attr(
  temperature,
  "unit"
) <- "degrees Celsius"

temperature
attributes(temperature)
```

## 18.2 Retrieve a custom attribute

```{r retrieve-custom-attribute}
attr(
  temperature,
  "unit"
)
```

Custom attributes can be lost through some operations. They should not be used
as a substitute for a well-designed data structure.

# 19. Object classes and behaviour

Class influences how generic functions behave.

```{r numeric-class-example}
class(assessment_scores)
```

Later, classes such as:

- `factor`;
- `Date`;
- `data.frame`;
- `lm`; and
- `ggplot`

will determine how functions such as `print()`, `summary()` and `plot()` behave.

The important principle is:

\[
\text{the same generic function may behave differently for different classes}.
\]

# 20. Basic coercion

A simple vector ordinarily contains one atomic type. When different types are
combined, R may coerce them to a common type.

## 20.1 Numeric and character

```{r coercion-numeric-character}
mixed_values <- c(
  10,
  20,
  "Thirty"
)

mixed_values
typeof(mixed_values)
```

All values become character because text cannot be represented as numeric.

## 20.2 Logical and numeric

```{r coercion-logical-numeric}
logical_numeric <- c(
  TRUE,
  FALSE,
  5
)

logical_numeric
typeof(logical_numeric)
```

Logical values are converted to numeric:

\[
TRUE=1,\qquad FALSE=0.
\]

## 20.3 Why coercion matters

Unexpected coercion can cause later functions to fail.

```{r coercion-mean-warning, warning=TRUE}
mean(mixed_values)
```

Always inspect imported or constructed objects.

# 21. Explicit conversion

R provides functions for explicit conversion.

## 21.1 Character to numeric

```{r explicit-character-to-numeric}
character_numbers <- c(
  "10",
  "20",
  "30"
)

numeric_numbers <-
  as.numeric(character_numbers)

numeric_numbers
```

## 21.2 Invalid conversion

```{r invalid-explicit-conversion, warning=TRUE}
as.numeric(
  c("10", "Twenty", "30")
)
```

The unconvertible value becomes `NA`, and R issues a warning.

## 21.3 Numeric to character

```{r explicit-numeric-to-character}
as.character(
  c(10, 20, 30)
)
```

## 21.4 Numeric to logical

```{r explicit-numeric-to-logical}
as.logical(
  c(0, 1, 2, -1)
)
```

Zero becomes `FALSE`; non-zero finite values become `TRUE`.

# 22. Validating objects

Before using an object, verify that it meets expected conditions.

## 22.1 Numeric check

```{r validate-numeric-object}
is.numeric(assessment_scores)
```

## 22.2 Character check

```{r validate-character-object}
is.character(student_programme)
```

## 22.3 Logical check

```{r validate-logical-object}
is.logical(fees_paid)
```

## 22.4 Length check

```{r validate-object-length}
length(assessment_scores) == 4
```

## 22.5 Range check

```{r validate-score-range}
all(
  assessment_scores >= 0 &
    assessment_scores <= 100
)
```

## 22.6 Missing-value check

```{r validate-missing-values}
anyNA(assessment_scores)
```

`anyNA()` is an efficient way to ask whether any missing value is present.

# 23. Stop when an object is invalid

```{r stop-invalid-object}
if (!is.numeric(assessment_scores)) {
  stop(
    "`assessment_scores` must be numeric.",
    call. = FALSE
  )
}
```

## 23.1 Several validation conditions

```{r several-validation-conditions}
if (anyNA(assessment_scores)) {
  stop(
    "`assessment_scores` contains missing values.",
    call. = FALSE
  )
}

if (!all(
  assessment_scores >= 0 &
    assessment_scores <= 100
)) {
  stop(
    "Scores must lie between 0 and 100.",
    call. = FALSE
  )
}
```

# 24. Objects created from data

Import the student dataset:

```{r import-student-object}
students <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)
```

Inspect:

```{r inspect-student-object}
class(students)
typeof(students)
dim(students)
str(students)
```

`students` is a data-frame object. It contains several column objects.

## 24.1 Extract a column object

```{r extract-column-object}
final_scores <- students$final_score

class(final_scores)
typeof(final_scores)
length(final_scores)
```

## 24.2 Create analytical objects

```{r create-analytical-objects}
pass_mark <- 50

pass_status <-
  final_scores >= pass_mark

number_passed <-
  sum(pass_status)

pass_percentage <-
  mean(pass_status) * 100

number_passed
pass_percentage
```

# 25. Avoid generic object names

Names such as the following are weak:

```text
data
x
temp
test
new
final
result
```

They may be acceptable for a very small mathematical demonstration, but they
become ambiguous in applied analysis.

Compare:

```r
result <- mean(x)
```

with:

```r
mean_final_score <- mean(student_final_scores)
```

The second expression is easier to understand without additional context.

# 26. Avoid misleading names

An object name should match its content.

Poor:

```r
mean_score <- c(60, 70, 80)
```

The object contains three scores, not a mean.

Better:

```{r truthful-object-name}
student_scores <- c(60, 70, 80)
mean_score <- mean(student_scores)
```

# 27. Units in object names

Including units can prevent mistakes.

```{r units-in-object-names}
waiting_time_minutes <- c(15, 22, 31)
distance_kilometres <- 12.5
weight_kilograms <- 68
```

Compare with ambiguous names:

```text
time
distance
weight
```

The unit-aware names improve interpretation.

# 28. Boolean naming conventions

Logical objects are often clearer when named as questions or states.

```{r boolean-naming}
is_registered <- TRUE
has_paid_fees <- FALSE
scores_are_valid <- TRUE
file_exists <- file.exists(
  "data/student_scores.csv"
)
```

These names read naturally inside conditions:

```{r boolean-condition-reading}
if (file_exists) {
  message(
    "The student dataset is available."
  )
}
```

# 29. Organising objects by stage

A data-analysis workflow may use:

```{r stage-object-names}
raw_clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)

analysis_clinic_data <-
  raw_clinic_data

clinic_waiting_time <-
  analysis_clinic_data$waiting_time

mean_clinic_waiting_time <-
  mean(
    clinic_waiting_time,
    na.rm = TRUE
  )
```

The names indicate the transition from raw input to analytical result.

# 30. Object dependencies

An object may depend on earlier objects.

```{r object-dependencies}
unit_price <- 12.5
quantity <- 40

total_cost <-
  unit_price * quantity
```

If `unit_price` changes, `total_cost` does not update automatically.

```{r object-dependency-update}
unit_price <- 15

total_cost
```

The stored result remains based on the earlier value.

Recalculate:

```{r recalculate-dependent-object}
total_cost <-
  unit_price * quantity

total_cost
```

R objects are not spreadsheet formulas that recalculate automatically.

# 31. Object order in scripts

Dependencies determine execution order.

Incorrect order:

```r
total_cost <- unit_price * quantity

unit_price <- 15
quantity <- 40
```

Correct order:

```{r correct-object-order}
unit_price <- 15
quantity <- 40

total_cost <-
  unit_price * quantity

total_cost
```

A script should create inputs before derived objects.

# 32. A small object registry

For a large project, a simple summary can clarify important objects.

```{r object-registry}
object_registry <- data.frame(
  object_name = c(
    "students",
    "final_scores",
    "pass_status",
    "pass_percentage"
  ),
  purpose = c(
    "Imported student data",
    "Final-score vector",
    "Logical pass indicator",
    "Percentage of students who passed"
  ),
  stringsAsFactors = FALSE
)

object_registry
```

This is especially useful in teaching, documentation and collaborative work.

# 33. Integrated worked example: monthly sales objects

## 33.1 Import the data

```{r integrated-sales-import}
raw_sales_data <- read.csv(
  "data/monthly_sales.csv",
  stringsAsFactors = FALSE
)
```

## 33.2 Inspect the object

```{r integrated-sales-inspect}
class(raw_sales_data)
typeof(raw_sales_data)
dim(raw_sales_data)
str(raw_sales_data)
```

## 33.3 Create meaningful column objects

```{r integrated-sales-columns}
units_sold <- raw_sales_data$units_sold
sales_revenue <- raw_sales_data$revenue
```

## 33.4 Validate objects

```{r integrated-sales-validation}
units_are_numeric <-
  is.numeric(units_sold)

revenue_is_numeric <-
  is.numeric(sales_revenue)

units_are_nonnegative <-
  all(units_sold >= 0)

revenue_is_positive <-
  all(sales_revenue > 0)

units_are_numeric
revenue_is_numeric
units_are_nonnegative
revenue_is_positive
```

## 33.5 Create summary objects

```{r integrated-sales-summaries}
number_of_sales_records <-
  length(sales_revenue)

total_sales_revenue <-
  sum(sales_revenue)

mean_sales_revenue <-
  mean(sales_revenue)

maximum_sales_revenue <-
  max(sales_revenue)

number_of_sales_records
total_sales_revenue
mean_sales_revenue
maximum_sales_revenue
```

## 33.6 Create logical performance object

```{r integrated-sales-logical}
above_average_revenue <-
  sales_revenue >
  mean_sales_revenue

above_average_revenue
```

## 33.7 Summarise the logical object

```{r integrated-sales-logical-summary}
number_above_average <-
  sum(above_average_revenue)

percentage_above_average <-
  mean(above_average_revenue) * 100

number_above_average
percentage_above_average
```

# 34. Guided practice

## 34.1 Education

```{r guided-education-objects}
assignment_marks <- c(18, 16, 20, 14)
exam_marks <- c(60, 55, 68, 47)

final_marks <-
  assignment_marks +
  exam_marks

mean_final_mark <-
  mean(final_marks)

pass_status <-
  final_marks >= 50

mean_final_mark
pass_status
```

Explain the role of every object.

## 34.2 Health

```{r guided-health-objects}
waiting_time_minutes <- c(
  15,
  24,
  31,
  18,
  42
)

target_minutes <- 30

target_met <-
  waiting_time_minutes <=
  target_minutes

target_percentage <-
  mean(target_met) * 100

target_percentage
```

## 34.3 Agriculture

```{r guided-agriculture-objects}
plot_yield_tonnes <- c(
  2.4,
  2.9,
  3.2,
  2.7
)

mean_yield_tonnes <-
  mean(plot_yield_tonnes)

yield_above_mean <-
  plot_yield_tonnes >
  mean_yield_tonnes

mean_yield_tonnes
yield_above_mean
```

# 35. Common mistakes

## 35.1 Using spaces in names

Incorrect:

```r
mean score <- 70
```

Correct:

```r
mean_score <- 70
```

## 35.2 Beginning with a number

Incorrect:

```r
2026_sales <- 5000
```

Better:

```r
sales_2026 <- 5000
```

## 35.3 Inconsistent case

```r
student_score <- 70
Student_Score
```

The second name does not refer to the first object.

## 35.4 Misleading names

```r
total <- mean(scores)
```

The object contains a mean, not a total.

## 35.5 Accidental overwrite

```r
student_data <- read.csv("data/student_scores.csv")
student_data <- 50
```

The imported table is replaced.

## 35.6 Using `=` when comparison is intended

Incorrect:

```r
score = 50
```

when the intention is to ask whether the score equals 50.

Correct:

```r
score == 50
```

# 36. Independent exercises

## Exercise 1

Create objects representing a student's name, age, programme, final score and
registration status. Use appropriate types and meaningful names.

## Exercise 2

Demonstrate the difference among `<-`, `==` and `=` in a named function
argument.

## Exercise 3

Create three objects whose names differ only by case. Explain why this is poor
practice even though R allows it.

## Exercise 4

Create a numeric vector of six scores. Inspect its value, type, class, length,
structure and attributes.

## Exercise 5

Copy the vector, change the first element of the copy and confirm that the
original remains unchanged.

## Exercise 6

Create a named vector of budgets for four departments. Extract one budget by
name.

## Exercise 7

Combine logical, numeric and character values in separate demonstrations.
Explain the coercion that occurs.

## Exercise 8

Convert a character vector of numeric text to numeric. Include one invalid text
value and explain the warning.

## Exercise 9

Import `clinic_visits.csv`. Create meaningful objects for waiting time and
consultation duration. Validate that the variables are numeric and nonnegative.

## Exercise 10

Create a sequence of dependent objects for a procurement calculation:
unit price, quantity, subtotal, tax rate, tax amount and total cost. Change the
unit price and demonstrate that dependent objects do not update automatically.

## Exercise 11

Create an object registry describing at least six objects from Exercise 10.

## Exercise 12

Write a script that imports `student_scores.csv`, creates pass-status and
distinction-status objects, validates them and reports percentages.

# 37. Lesson assessment

## Part A: Conceptual questions

1. What is an R object?
2. What does assignment do?
3. Why is `<-` preferred in this course?
4. What is the difference between assignment and comparison?
5. What is `snake_case`?
6. Why is case sensitivity important?
7. What are reserved words?
8. Why should backtick names usually be avoided?
9. What is the difference between `typeof()` and `class()`?
10. What does `str()` provide?
11. What happens when an existing object name is assigned again?
12. What is coercion?
13. Why can coercion create analytical problems?
14. What does `anyNA()` test?
15. Why should objects be validated before analysis?
16. Why are generic names risky?
17. Why should units sometimes be included in object names?
18. What is an object dependency?
19. Why do dependent objects not update automatically?
20. Why should script order reflect object dependencies?

## Part B: Practical assessment

Create a standalone R Markdown report that:

1. imports `monthly_sales.csv`;
2. assigns the imported table to a meaningful object;
3. creates separate objects for units and revenue;
4. inspects their types, classes and lengths;
5. validates that units are nonnegative and revenue is positive;
6. calculates total, mean, minimum and maximum revenue;
7. creates a logical above-average object;
8. reports the number and percentage above average;
9. creates a named summary vector;
10. demonstrates copying and updating an object;
11. demonstrates one coercion example;
12. explains one warning caused by invalid conversion;
13. includes an object registry; and
14. knits successfully from a clean session.

# 38. Key terms

**Assignment:** Binding a name to a value or structure.

**Attribute:** Additional metadata attached to an object.

**Case-sensitive:** Treating upper-case and lower-case letters as distinct.

**Class:** A property influencing how functions behave for an object.

**Coercion:** Conversion of values to a common or requested type.

**Dependency:** A relationship in which one object is calculated from another.

**Global environment:** The ordinary workspace in which interactive objects are
stored.

**Object:** A named structure stored in R's memory.

**Reserved word:** A word with a special grammatical meaning in R.

**Syntactic name:** A name that can be used directly without backticks.

**Type:** The internal storage type of an object.

**Validation:** Checking whether an object satisfies expected rules.

# 39. Completion checklist

Before proceeding, confirm that you can:

- define an R object;
- assign values with `<-`;
- distinguish assignment from equality testing;
- create numeric, character and logical objects;
- select meaningful `snake_case` names;
- identify invalid names;
- explain case sensitivity;
- avoid reserved words;
- inspect type, class, length and structure;
- copy an object;
- update an object;
- recognise accidental overwriting;
- list objects;
- check object existence;
- remove objects safely;
- create and use named vectors;
- inspect attributes;
- explain basic class behaviour;
- recognise implicit coercion;
- perform explicit conversion;
- interpret a coercion warning;
- validate object type, length, range and missingness;
- stop execution when an object is invalid;
- name objects by analytical role;
- include units where appropriate;
- explain object dependencies; and
- organise object creation in a logical script order.

# Lesson conclusion

Objects provide the structure through which R stores and manipulates
information. Assignment creates the relationship between a name and a value.
Good naming communicates purpose. Inspection reveals what has been created.
Validation protects the analysis from inappropriate inputs.

The safe object workflow is:

\[
\boxed{
\text{name clearly}
\rightarrow
\text{assign}
\rightarrow
\text{inspect}
\rightarrow
\text{validate}
\rightarrow
\text{transform}
\rightarrow
\text{preserve}
}
\]

The next lesson will examine atomic data types and special values in greater
depth, including doubles, integers, characters, logical values, complex
numbers, raw values, missing values, infinities and undefined numeric results.
