---
title: "Lesson 5: Atomic Data Types and Special Values"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(digits = 4, scipen = 999, width = 90)
set.seed(20260713)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Every value stored in R has an underlying type. The type determines what the
value means, how it is stored and which operations are valid. A number can be
added, text can be joined, and a logical value can be used in a condition.
However, applying a numerical function to text usually fails because the
operation is incompatible with the object's type.

This lesson introduces R's atomic data types and the special values that appear
in real datasets. Particular attention is given to:

- doubles and integers;
- character values;
- logical values;
- complex values;
- raw values;
- missing values;
- infinite values;
- undefined numerical results;
- type testing;
- coercion; and
- explicit conversion.

The aim is to develop the habit of checking data types before analysis.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain the meaning of an atomic data type;
2. distinguish double and integer storage;
3. create character and logical values correctly;
4. identify complex and raw values;
5. use `typeof()`, `class()` and type-testing functions;
6. explain implicit coercion;
7. convert values explicitly;
8. recognise invalid conversions;
9. distinguish `NA`, `NaN`, `Inf`, `-Inf` and `NULL`;
10. test for missing, finite and infinite values;
11. explain typed missing values;
12. handle missing values in summaries;
13. identify type problems in imported data; and
14. validate variables before analysis.

# 1. Atomic values

An atomic value is one of the simplest values R can store. Atomic vectors
contain values that share one common atomic type.

The principal atomic types are:

- logical;
- integer;
- double;
- complex;
- character; and
- raw.

In ordinary statistical work, logical, integer, double and character values are
encountered most often.

# 2. Double values

R normally stores ordinary decimal numbers as doubles.

```{r double-value}
x <- 25
typeof(x)
class(x)
```

Although `25` looks like a whole number, its internal type is usually
`double`.

```{r decimal-double}
temperature_celsius <- 31.7
typeof(temperature_celsius)
```

Doubles can represent:

- whole-looking values;
- decimals;
- very large values;
- very small values; and
- scientific notation.

```{r scientific-notation}
large_value <- 1.5e6
small_value <- 2.5e-4

large_value
small_value
```

# 3. Integer values

Append `L` to create an integer literal.

```{r integer-value}
student_count <- 120L
typeof(student_count)
class(student_count)
```

The `L` tells R to store the value as an integer.

```{r integer-versus-double}
double_count <- 120
integer_count <- 120L

typeof(double_count)
typeof(integer_count)
```

The values are numerically equal:

```{r integer-double-equality}
double_count == integer_count
```

Their storage types differ.

## 3.1 Why integers matter

Integers are appropriate for counts such as:

- number of students;
- number of farms;
- number of visits;
- number of items; and
- number of completed tasks.

However, R frequently converts integers to doubles during calculations.

```{r integer-arithmetic}
integer_count / 3L
typeof(integer_count / 3L)
```

# 4. Character values

Character values represent text.

```{r character-values}
student_name <- "Abena"
programme_name <- "Applied Statistics"

typeof(student_name)
class(programme_name)
```

Text must be enclosed in quotation marks.

```{r quotation-marks}
single_quotes <- 'R programming'
double_quotes <- "R programming"

single_quotes == double_quotes
```

Both forms create character values.

## 4.1 Numbers stored as text

```{r numeric-looking-character}
student_id <- "00125"
typeof(student_id)
```

The leading zeros are preserved because the identifier is stored as text.

Student identifiers, telephone numbers and codes should often be character
rather than numeric because arithmetic is not meaningful for them.

# 5. Logical values

Logical values are:

```text
TRUE
FALSE
```

```{r logical-values}
is_registered <- TRUE
has_completed <- FALSE

typeof(is_registered)
class(has_completed)
```

Logical values are produced by comparisons.

```{r comparison-produces-logical}
score <- 68
score >= 50
```

## 5.1 Logical values in arithmetic

```{r logical-arithmetic}
sum(c(TRUE, FALSE, TRUE, TRUE))
mean(c(TRUE, FALSE, TRUE, TRUE))
```

R treats:

\[
TRUE=1,\qquad FALSE=0.
\]

This makes logical vectors useful for counts and proportions.

# 6. Complex values

Complex numbers contain real and imaginary components.

```{r complex-value}
z <- 3 + 4i
typeof(z)
class(z)
```

Inspect components:

```{r complex-components}
Re(z)
Im(z)
Mod(z)
Arg(z)
```

Complex numbers are important in some areas of engineering, signal processing
and advanced mathematics, but they are less common in introductory applied
statistics.

# 7. Raw values

Raw values store bytes.

```{r raw-value}
raw_example <- charToRaw("R")
raw_example
typeof(raw_example)
```

Convert back:

```{r raw-to-character}
rawToChar(raw_example)
```

Raw values are relevant when handling binary files, encodings and low-level
data transfer.

# 8. Inspecting types

## 8.1 `typeof()`

```{r typeof-examples}
typeof(25)
typeof(25L)
typeof("25")
typeof(TRUE)
typeof(3 + 2i)
```

## 8.2 `class()`

```{r class-examples}
class(25)
class(25L)
class("25")
class(TRUE)
```

For simple atomic values, `class()` often resembles the conceptual data class,
while `typeof()` reports internal storage.

## 8.3 `mode()` and `storage.mode()`

```{r mode-storage-mode}
mode(25)
storage.mode(25)
```

These functions exist for historical and technical reasons. For most learning
tasks, `typeof()` and `class()` are clearer.

# 9. Type-testing functions

```{r type-tests}
is.numeric(25)
is.integer(25L)
is.double(25)
is.character("25")
is.logical(TRUE)
is.complex(3 + 2i)
is.raw(charToRaw("R"))
```

## 9.1 Numeric is broader than double

```{r numeric-broader}
is.numeric(25)
is.numeric(25L)
```

Both doubles and integers are considered numeric.

# 10. Implicit coercion

Atomic vectors must use one common type. When values of different types are
combined, R finds a compatible type.

A simplified coercion hierarchy is:

\[
logical
\rightarrow
integer
\rightarrow
double
\rightarrow
complex
\rightarrow
character.
\]

## 10.1 Logical and integer

```{r coercion-logical-integer}
x <- c(TRUE, FALSE, 3L)
x
typeof(x)
```

Logical values become integers.

## 10.2 Integer and double

```{r coercion-integer-double}
x <- c(1L, 2.5)
x
typeof(x)
```

Integers become doubles.

## 10.3 Numeric and character

```{r coercion-numeric-character}
x <- c(10, 20, "Thirty")
x
typeof(x)
```

Everything becomes character.

## 10.4 Why coercion matters

```{r coercion-summary-warning, warning=TRUE}
mean(x)
```

A numerical summary fails because the vector is character.

# 11. Explicit conversion

Use conversion functions when the intended type is known.

## 11.1 To numeric

```{r convert-to-numeric}
x <- c("10", "20", "30")
as.numeric(x)
```

## 11.2 Invalid numeric conversion

```{r invalid-numeric-conversion, warning=TRUE}
as.numeric(c("10", "Twenty", "30"))
```

The invalid value becomes `NA`.

## 11.3 To integer

```{r convert-to-integer}
as.integer(c(3.8, 4.2, -2.9))
```

The decimal part is truncated rather than conventionally rounded.

## 11.4 To character

```{r convert-to-character}
as.character(c(10, 20, 30))
```

## 11.5 To logical

```{r convert-to-logical}
as.logical(c(0, 1, 2, -4))
```

Zero becomes `FALSE`; non-zero values become `TRUE`.

# 12. Missing values

R represents missing values with `NA`.

```{r missing-value}
score <- NA
score
```

A missing value is not the same as zero.

```{r missing-versus-zero}
is.na(NA)
is.na(0)
```

## 12.1 Typed missing values

```{r typed-missing-values}
typeof(NA)
typeof(NA_integer_)
typeof(NA_real_)
typeof(NA_character_)
typeof(NA_complex_)
```

Typed missing values are useful in advanced programming and function design.

# 13. Detecting missing values

```{r detect-missing}
scores <- c(65, NA, 72, 81, NA)

is.na(scores)
sum(is.na(scores))
anyNA(scores)
which(is.na(scores))
```

## 13.1 Why `x == NA` is incorrect

```{r compare-with-na}
scores == NA
```

Use:

```{r correct-na-test}
is.na(scores)
```

A missing value has unknown content, so ordinary equality cannot determine
whether it equals something.

# 14. Missing values in summaries

```{r summary-with-missing}
mean(scores)
sum(scores)
```

By default, missingness propagates.

```{r remove-missing-summary}
mean(scores, na.rm = TRUE)
sum(scores, na.rm = TRUE)
```

The argument `na.rm = TRUE` removes missing values only for that calculation.

## 14.1 Observed sample size

```{r observed-sample-size}
sum(!is.na(scores))
```

This count should often be reported because the effective sample size may differ
from the total vector length.

# 15. Infinite values

```{r positive-infinity}
1 / 0
```

```{r negative-infinity}
-1 / 0
```

Test:

```{r test-infinite}
is.infinite(c(1, Inf, -Inf))
```

# 16. Undefined numerical values

```{r nan-value}
0 / 0
```

The result is `NaN`, meaning “not a number”.

```{r test-nan}
is.nan(0 / 0)
```

`NaN` is also treated as missing:

```{r nan-is-na}
is.na(NaN)
```

# 17. Finite values

```{r finite-values}
x <- c(10, NA, NaN, Inf, -Inf, 25)

is.finite(x)
```

A finite value is an ordinary numerical value that is not missing or infinite.

# 18. `NULL`

`NULL` represents the absence of an object or component.

```{r null-value}
x <- NULL
x
is.null(x)
length(x)
```

`NULL` differs from `NA`.

- `NA` is a missing value occupying a position.
- `NULL` often represents no value or no component.

```{r null-versus-na}
length(NA)
length(NULL)
```

# 19. Special numerical constants

```{r special-constants}
pi
Inf
-Inf
NaN
```

R also recognises:

```{r boolean-shortcuts}
T
F
```

However, use `TRUE` and `FALSE` because `T` and `F` can be overwritten.

```{r overwrite-t-example}
T <- 100
T
TRUE
```

Avoid this practice.

# 20. Type problems in imported data

Import the rainfall dataset.

```{r import-rainfall}
rainfall <- read.csv(
  "data/rainfall_records.csv",
  stringsAsFactors = FALSE
)

str(rainfall)
```

Inspect variable classes:

```{r rainfall-classes}
vapply(
  rainfall,
  class,
  FUN.VALUE = character(1)
)
```

Check missing values:

```{r rainfall-missing-counts}
colSums(is.na(rainfall))
```

## 20.1 Character numbers from messy data

```{r messy-character-numbers}
messy_values <- c("10", "15", "missing", "21")
```

Convert carefully:

```{r convert-messy-values, warning=TRUE}
converted_values <- as.numeric(messy_values)
converted_values
```

Identify failed conversions:

```{r identify-failed-conversions}
messy_values[
  is.na(converted_values) &
    !is.na(messy_values)
]
```

# 21. Safer conversion workflow

```{r safer-conversion-workflow}
raw_values <- c("12", "18", "Twenty", "25")

converted <- suppressWarnings(
  as.numeric(raw_values)
)

conversion_failed <-
  is.na(converted) &
  !is.na(raw_values)

data.frame(
  raw_value = raw_values,
  converted_value = converted,
  failed = conversion_failed,
  stringsAsFactors = FALSE
)
```

Suppressing a warning is acceptable here because the failure is explicitly
captured and inspected.

# 22. Validating type and range

```{r validate-numeric-scores}
scores <- c(65, 72, 81, 59)

is.numeric(scores)
all(scores >= 0 & scores <= 100)
anyNA(scores)
```

Create a reusable validation sequence:

```{r validation-sequence}
if (!is.numeric(scores)) {
  stop("`scores` must be numeric.", call. = FALSE)
}

if (anyNA(scores)) {
  stop("`scores` contains missing values.", call. = FALSE)
}

if (!all(scores >= 0 & scores <= 100)) {
  stop("Scores must lie between 0 and 100.", call. = FALSE)
}
```

# 23. Integrated example: clinic data types

```{r import-clinic}
clinic <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)

str(clinic)
```

Inspect types:

```{r clinic-types}
vapply(
  clinic,
  typeof,
  FUN.VALUE = character(1)
)
```

Inspect classes:

```{r clinic-classes}
vapply(
  clinic,
  class,
  FUN.VALUE = character(1)
)
```

Check waiting time:

```{r clinic-waiting-validation}
is.numeric(clinic$waiting_time)
anyNA(clinic$waiting_time)
all(
  clinic$waiting_time >= 0,
  na.rm = TRUE
)
```

Calculate observed mean:

```{r clinic-observed-mean}
mean(
  clinic$waiting_time,
  na.rm = TRUE
)
```

# 24. Guided practice

## 24.1 Education

```{r guided-education-types}
student_id <- "00041"
age <- 21L
score <- 74.5
passed <- score >= 50

typeof(student_id)
typeof(age)
typeof(score)
typeof(passed)
```

## 24.2 Procurement

```{r guided-procurement-types}
supplier_code <- "SUP-014"
delivery_days <- 5L
invoice_value <- 18500.75
delivered_on_time <- TRUE
```

Inspect each object's type and class.

## 24.3 Missing-data context

```{r guided-missing-data}
waiting_times <- c(18, 25, NA, 32, 21)

length(waiting_times)
sum(is.na(waiting_times))
sum(!is.na(waiting_times))
mean(waiting_times, na.rm = TRUE)
```

# 25. Common mistakes

## 25.1 Treating identifiers as numbers

```r
student_id <- 00125
```

Leading zeros are lost. Use:

```r
student_id <- "00125"
```

## 25.2 Assuming `as.integer()` rounds

```{r integer-truncation}
as.integer(4.9)
```

The result is 4, not 5.

## 25.3 Ignoring failed conversion

```{r failed-conversion-warning, warning=TRUE}
as.numeric("forty")
```

## 25.4 Confusing `NULL` and `NA`

They represent different concepts and have different lengths.

## 25.5 Removing missing values without explanation

`na.rm = TRUE` is a computational instruction, not a complete missing-data
strategy.

# 26. Independent exercises

1. Create one object of each principal atomic type and inspect it.
2. Demonstrate the difference between `25` and `25L`.
3. Create an identifier that preserves leading zeros.
4. Demonstrate logical-to-numeric coercion.
5. Demonstrate numeric-to-character coercion.
6. Convert character numbers to numeric and identify an invalid value.
7. Create a vector containing `NA`, `NaN`, `Inf`, `-Inf` and ordinary values.
8. Test each special value with appropriate functions.
9. Explain the difference between `NULL` and `NA`.
10. Import `rainfall_records.csv` and report missing values by column.
11. Calculate observed rainfall summaries.
12. Write a validation routine for a percentage variable ranging from 0 to 100.

# 27. Lesson assessment

Create a standalone R Markdown report that:

1. defines and explains each atomic type;
2. demonstrates implicit coercion;
3. demonstrates explicit conversion;
4. includes one invalid conversion and interprets the warning;
5. distinguishes `NA`, `NaN`, `Inf`, `-Inf` and `NULL`;
6. imports `rainfall_records.csv`;
7. inspects variable types;
8. reports missingness;
9. calculates observed summaries; and
10. validates at least one numeric variable.

# 28. Completion checklist

Confirm that you can:

- distinguish double and integer;
- create character and logical values;
- identify complex and raw values;
- inspect type and class;
- use type-testing functions;
- explain coercion;
- convert types explicitly;
- identify failed conversion;
- distinguish special values;
- detect missing values;
- calculate summaries with missing data;
- identify finite values;
- explain `NULL`;
- inspect imported variable types; and
- validate a variable before analysis.

# Lesson conclusion

Data type is not a minor technical detail. It determines which operations are
valid and how R interprets a value. Reliable analysis therefore begins with
inspection and validation.

The recommended sequence is:

\[
\boxed{
\text{inspect type}
\rightarrow
\text{identify special values}
\rightarrow
\text{convert deliberately}
\rightarrow
\text{validate}
\rightarrow
\text{analyse}
}
\]
