---
title: "Lesson 6: Vectors and Vectorised Operations"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(digits = 4, scipen = 999, width = 90)
set.seed(20260713)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Vectors are the foundation of R's data model. A column in a dataset is commonly
a vector, and most statistical functions operate on vectors. Understanding
vectors therefore has direct consequences for data cleaning, summaries,
logical conditions, simulation and modelling.

This lesson covers vector creation, naming, indexing, arithmetic, comparison,
recycling, missing values, ordering, set operations and common vector
functions.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. create vectors using several methods;
2. explain vector length and type;
3. name vector elements;
4. perform vectorised arithmetic;
5. apply comparisons and logical operations;
6. understand recycling;
7. generate sequences and repeated patterns;
8. summarise vectors;
9. sort and rank values;
10. use set operations;
11. handle missing values;
12. identify duplicates;
13. combine and append vectors; and
14. validate vector compatibility.

# 1. Creating vectors with `c()`

```{r create-vector-c}
scores <- c(68, 74, 81, 59, 90)
scores
```

Inspect:

```{r inspect-vector}
length(scores)
typeof(scores)
class(scores)
str(scores)
```

# 2. Creating sequences

```{r create-sequence-colon}
1:10
```

```{r create-sequence-seq}
seq(from = 0, to = 100, by = 10)
```

```{r create-sequence-length}
seq(from = 0, to = 1, length.out = 6)
```

# 3. Repeating values

```{r repeat-values-times}
rep(c("A", "B"), times = 3)
```

```{r repeat-values-each}
rep(c("A", "B"), each = 3)
```

# 4. Named vectors

```{r named-vector}
programme_counts <- c(
  Statistics = 35,
  Procurement = 42,
  Hospitality = 28,
  Agriculture = 31
)

programme_counts
```

```{r inspect-vector-names}
names(programme_counts)
programme_counts["Statistics"]
```

# 5. Vectorised arithmetic

```{r vectorised-addition}
coursework <- c(30, 28, 35, 24)
examination <- c(48, 42, 51, 39)

final_scores <- coursework + examination
final_scores
```

```{r vectorised-subtraction}
remaining_marks <- 100 - final_scores
remaining_marks
```

```{r vectorised-multiplication}
proportions <- c(0.45, 0.72, 0.81)
percentages <- proportions * 100
percentages
```

```{r vectorised-division}
revenue <- c(1000, 1500, 2100)
units <- c(20, 25, 30)

revenue_per_unit <- revenue / units
revenue_per_unit
```

# 6. Comparison vectors

```{r vector-comparisons}
final_scores >= 50
final_scores < 70
final_scores == 78
final_scores != 78
```

Count and proportion:

```{r logical-count-proportion}
sum(final_scores >= 50)
mean(final_scores >= 50)
mean(final_scores >= 50) * 100
```

# 7. Combined logical conditions

```{r combined-logical-vector}
scores <- c(45, 58, 72, 84, 91)
attendance <- c(0.80, 0.92, 0.88, 0.95, 0.70)

eligible <-
  (scores >= 70) &
  (attendance >= 0.85)

eligible
```

```{r logical-or-vector}
requires_follow_up <-
  (scores < 50) |
  (attendance < 0.75)

requires_follow_up
```

# 8. Membership

```{r membership-vector}
programmes <- c(
  "Statistics",
  "Hospitality",
  "Agriculture",
  "Procurement"
)

programmes %in% c("Statistics", "Procurement")
```

# 9. Recycling

```{r exact-recycling-vector}
c(10, 20, 30, 40) + c(1, 2)
```

```{r recycling-warning-vector, warning=TRUE}
c(10, 20, 30, 40, 50) + c(1, 2)
```

Validate lengths:

```{r validate-vector-lengths}
x <- c(10, 20, 30)
y <- c(1, 2, 3)

length(x) == length(y)
```

# 10. Summary functions

```{r vector-summary-functions}
scores <- c(65, 72, 81, 59, 90)

length(scores)
sum(scores)
mean(scores)
median(scores)
min(scores)
max(scores)
range(scores)
sd(scores)
var(scores)
summary(scores)
```

# 11. Quantiles

```{r vector-quantiles}
quantile(scores)
```

```{r custom-quantiles}
quantile(
  scores,
  probs = c(0.10, 0.25, 0.50, 0.75, 0.90)
)
```

# 12. Sorting and ordering

```{r sort-ascending}
sort(scores)
```

```{r sort-descending}
sort(scores, decreasing = TRUE)
```

```{r order-indices}
order(scores)
```

```{r rank-values}
rank(scores)
```

# 13. Minimum and maximum positions

```{r which-min-max}
which.min(scores)
which.max(scores)
```

Retrieve the values:

```{r retrieve-min-max}
scores[which.min(scores)]
scores[which.max(scores)]
```

# 14. Duplicates

```{r duplicate-vector}
student_ids <- c(
  "S001",
  "S002",
  "S003",
  "S002",
  "S004"
)

duplicated(student_ids)
anyDuplicated(student_ids)
unique(student_ids)
```

Count unique values:

```{r count-unique}
length(unique(student_ids))
```

# 15. Missing values in vectors

```{r vector-missing-values}
rainfall <- c(115.2, 98.4, NA, 121.7, 110.5)

is.na(rainfall)
sum(is.na(rainfall))
anyNA(rainfall)
mean(rainfall, na.rm = TRUE)
```

Complete cases:

```{r complete-cases-vector}
complete.cases(rainfall)
rainfall[complete.cases(rainfall)]
```

# 16. Appending vectors

```{r append-vector-c}
first_batch <- c(10, 20, 30)
second_batch <- c(40, 50)

all_values <- c(first_batch, second_batch)
all_values
```

```{r append-function}
append(
  first_batch,
  values = 25,
  after = 2
)
```

# 17. Set operations

```{r set-operation-data}
group_a <- c("Ama", "Kojo", "Esi", "Yaw")
group_b <- c("Esi", "Yaw", "Akua", "Kofi")
```

```{r union-operation}
union(group_a, group_b)
```

```{r intersection-operation}
intersect(group_a, group_b)
```

```{r set-difference}
setdiff(group_a, group_b)
setdiff(group_b, group_a)
```

```{r set-equality}
setequal(
  c("A", "B", "C"),
  c("C", "B", "A")
)
```

# 18. Pairwise functions

```{r pairwise-minimum}
pmin(
  c(10, 30, 20),
  c(15, 25, 22)
)
```

```{r pairwise-maximum}
pmax(
  c(10, 30, 20),
  c(15, 25, 22)
)
```

# 19. Cumulative functions

```{r cumulative-sum}
monthly_sales <- c(100, 120, 90, 150)
cumsum(monthly_sales)
```

```{r cumulative-product}
cumprod(c(1.05, 1.03, 0.98, 1.04))
```

```{r cumulative-min-max}
cummin(c(10, 8, 12, 5, 9))
cummax(c(10, 8, 12, 5, 9))
```

# 20. Differences and lags

```{r vector-differences}
values <- c(100, 120, 115, 140)
diff(values)
```

The differences are:

\[
120-100,\quad115-120,\quad140-115.
\]

# 21. Sampling vectors

```{r sample-values}
sample(
  x = scores,
  size = 3,
  replace = FALSE
)
```

Sampling with replacement:

```{r sample-with-replacement}
sample(
  x = scores,
  size = 8,
  replace = TRUE
)
```

# 22. Random vectors

```{r random-normal-vector}
set.seed(20260713)

simulated_scores <- rnorm(
  n = 10,
  mean = 70,
  sd = 8
)

round(simulated_scores, 1)
```

# 23. Vector validation

```{r vector-validation}
validate_scores <- c(65, 72, 81, 59)

is.numeric(validate_scores)
length(validate_scores) > 0
!anyNA(validate_scores)
all(validate_scores >= 0 & validate_scores <= 100)
```

# 24. Integrated example: rainfall vector

```{r import-rainfall-vector}
rainfall_data <- read.csv(
  "data/rainfall_records.csv",
  stringsAsFactors = FALSE
)

rainfall_mm <- rainfall_data$rainfall_mm
```

Inspect:

```{r inspect-rainfall-vector}
str(rainfall_mm)
length(rainfall_mm)
sum(is.na(rainfall_mm))
```

Summarise:

```{r summarise-rainfall-vector}
mean(rainfall_mm, na.rm = TRUE)
median(rainfall_mm, na.rm = TRUE)
min(rainfall_mm, na.rm = TRUE)
max(rainfall_mm, na.rm = TRUE)
quantile(rainfall_mm, na.rm = TRUE)
```

# 25. Common mistakes

- forgetting that vector operations are element-wise;
- relying on unintended recycling;
- confusing `sort()` with `order()`;
- ignoring duplicates;
- removing missing values without explanation;
- comparing with `NA` using `==`;
- assuming different-length vectors are compatible;
- overlooking type coercion.

# 26. Independent exercises

1. Create a numeric vector using `c()`.
2. Create sequences using `:`, `seq()` and `rep()`.
3. Create a named vector and extract values by name.
4. Perform element-wise arithmetic on two compatible vectors.
5. Demonstrate exact and inexact recycling.
6. Calculate logical counts and percentages.
7. Sort, order and rank a vector.
8. Identify duplicates and unique values.
9. Demonstrate set operations.
10. Calculate cumulative sums and first differences.
11. Sample values with and without replacement.
12. Import rainfall data and summarise the rainfall vector.

# 27. Lesson assessment

Create a standalone R Markdown report that demonstrates:

- vector creation;
- naming;
- vectorised arithmetic;
- comparisons;
- logical summaries;
- recycling;
- missing values;
- sorting;
- duplicates;
- set operations;
- cumulative functions;
- sampling; and
- validation.

# 28. Completion checklist

Confirm that you can:

- create vectors;
- name vector elements;
- calculate vector summaries;
- perform vectorised arithmetic;
- count logical conditions;
- explain recycling;
- sort and order;
- identify duplicates;
- use set operations;
- handle missing values;
- generate cumulative values;
- sample values; and
- validate vector compatibility.

# Lesson conclusion

Vectors are the primary computational structure in R. Their element-wise
behaviour makes statistical operations concise, but it also requires careful
attention to type, length, missingness and recycling.
