---
title: "Lesson 7: Matrices, Arrays and Multidimensional Structures"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(digits = 4, scipen = 999, width = 90)
set.seed(20260713)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Vectors are one-dimensional. Matrices and arrays extend this idea to two or
more dimensions. They are useful for mathematical computation, experimental
layouts, repeated measurements, image-like data and simulation.

This lesson develops matrix and array creation, dimension naming, indexing,
arithmetic, matrix algebra, summaries and reshaping.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. create matrices by row and by column;
2. inspect dimensions and names;
3. extract rows, columns and cells;
4. perform element-wise operations;
5. perform matrix multiplication;
6. transpose matrices;
7. calculate determinants and inverses;
8. solve linear systems;
9. bind vectors into matrices;
10. apply functions over margins;
11. create arrays;
12. index multidimensional arrays; and
13. validate dimension compatibility.

# 1. Creating a matrix

```{r create-matrix-columnwise}
m <- matrix(
  1:12,
  nrow = 3,
  ncol = 4
)

m
```

R fills matrices column by column by default.

```{r create-matrix-rowwise}
m_row <- matrix(
  1:12,
  nrow = 3,
  ncol = 4,
  byrow = TRUE
)

m_row
```

# 2. Inspecting a matrix

```{r inspect-matrix}
class(m)
typeof(m)
dim(m)
nrow(m)
ncol(m)
length(m)
str(m)
```

# 3. Naming dimensions

```{r name-matrix-dimensions}
student_scores <- matrix(
  c(
    30, 48,
    28, 42,
    35, 51
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("Ama", "Kojo", "Esi"),
    c("Coursework", "Examination")
  )
)

student_scores
```

# 4. Matrix indexing

```{r matrix-single-cell}
student_scores[2, 1]
```

```{r matrix-row}
student_scores[1, ]
```

```{r matrix-column}
student_scores[, 2]
```

```{r matrix-named-index}
student_scores["Esi", "Examination"]
```

Preserve matrix dimensions:

```{r preserve-matrix-dimensions}
student_scores[
  1,
  ,
  drop = FALSE
]
```

# 5. Replacing matrix values

```{r replace-matrix-cell}
student_scores["Kojo", "Coursework"] <- 31
student_scores
```

# 6. Matrix arithmetic

```{r matrix-arithmetic-data}
a <- matrix(1:6, nrow = 2)
b <- matrix(7:12, nrow = 2)
```

```{r matrix-elementwise-addition}
a + b
```

```{r matrix-elementwise-multiplication}
a * b
```

Element-wise multiplication is not matrix multiplication.

# 7. Matrix multiplication

Use `%*%`.

```{r matrix-multiplication}
x <- matrix(
  c(1, 2, 3, 4),
  nrow = 2,
  byrow = TRUE
)

y <- matrix(
  c(5, 6, 7, 8),
  nrow = 2,
  byrow = TRUE
)

x %*% y
```

For multiplication:

\[
A_{m\times n}B_{n\times p}
\]

the inner dimensions must agree.

# 8. Transpose

```{r matrix-transpose}
t(student_scores)
```

Rows become columns and columns become rows.

# 9. Row and column summaries

```{r row-sums}
rowSums(student_scores)
```

```{r column-sums}
colSums(student_scores)
```

```{r row-means}
rowMeans(student_scores)
```

```{r column-means}
colMeans(student_scores)
```

# 10. Binding vectors

```{r cbind-example}
coursework <- c(30, 28, 35)
examination <- c(48, 42, 51)

score_matrix <- cbind(
  coursework,
  examination
)

score_matrix
```

```{r rbind-example}
first_semester <- c(65, 72, 81)
second_semester <- c(68, 75, 84)

semester_matrix <- rbind(
  first_semester,
  second_semester
)

semester_matrix
```

# 11. Diagonal matrices

```{r identity-matrix}
diag(4)
```

```{r diagonal-values}
diag(c(2, 4, 6))
```

Extract diagonal:

```{r extract-diagonal}
diag(x)
```

# 12. Determinant

```{r matrix-determinant}
det(x)
```

A square matrix with determinant zero is singular and has no ordinary inverse.

# 13. Matrix inverse

```{r matrix-inverse}
solve(x)
```

Check:

```{r inverse-check}
x %*% solve(x)
```

The result is approximately the identity matrix.

# 14. Solving linear systems

Consider:

\[
2x+y=8
\]

\[
x+3y=9.
\]

```{r solve-linear-system}
coefficient_matrix <- matrix(
  c(
    2, 1,
    1, 3
  ),
  nrow = 2,
  byrow = TRUE
)

constant_vector <- c(8, 9)

solution <- solve(
  coefficient_matrix,
  constant_vector
)

solution
```

# 15. Apply functions over margins

```{r apply-matrix}
apply(
  student_scores,
  MARGIN = 1,
  FUN = mean
)
```

`MARGIN = 1` means rows.

```{r apply-matrix-columns}
apply(
  student_scores,
  MARGIN = 2,
  FUN = mean
)
```

`MARGIN = 2` means columns.

# 16. Missing values in matrices

```{r matrix-with-missing}
m_missing <- matrix(
  c(1, 2, NA, 4, 5, 6),
  nrow = 2
)

m_missing
```

```{r matrix-missing-summary}
rowMeans(
  m_missing,
  na.rm = TRUE
)

colMeans(
  m_missing,
  na.rm = TRUE
)
```

# 17. Creating arrays

An array can have more than two dimensions.

```{r create-array}
experiment_array <- array(
  1:24,
  dim = c(3, 4, 2)
)

experiment_array
```

The dimensions may represent:

- 3 treatments;
- 4 blocks;
- 2 seasons.

# 18. Naming array dimensions

```{r named-array}
yield_array <- array(
  1:24,
  dim = c(3, 4, 2),
  dimnames = list(
    Treatment = c("A", "B", "C"),
    Block = paste0("B", 1:4),
    Season = c("Dry", "Wet")
  )
)

yield_array
```

# 19. Array indexing

```{r array-single-cell}
yield_array[
  "B",
  "B3",
  "Wet"
]
```

```{r array-one-layer}
yield_array[
  ,
  ,
  "Dry"
]
```

# 20. Apply over array margins

Treatment means:

```{r array-treatment-means}
apply(
  yield_array,
  MARGIN = 1,
  FUN = mean
)
```

Season means:

```{r array-season-means}
apply(
  yield_array,
  MARGIN = 3,
  FUN = mean
)
```

Treatment-by-season means:

```{r array-treatment-season-means}
apply(
  yield_array,
  MARGIN = c(1, 3),
  FUN = mean
)
```

# 21. Reshaping vectors into matrices

```{r reshape-vector-matrix}
values <- 1:12

matrix(
  values,
  nrow = 3,
  byrow = TRUE
)
```

The data are not changed; only the dimension attribute is added.

# 22. Flattening structures

```{r flatten-matrix}
as.vector(student_scores)
```

```{r flatten-array}
as.vector(yield_array)
```

# 23. Converting matrices to data frames

```{r matrix-to-data-frame}
score_data_frame <- as.data.frame(student_scores)
score_data_frame
```

A data frame can contain columns of different types, while a matrix has one
atomic type.

# 24. Dimension compatibility

```{r incompatible-matrix-multiplication, error=TRUE}
matrix(1:6, nrow = 2) %*%
  matrix(1:8, nrow = 4)
```

Inspect dimensions before multiplication:

```{r inspect-multiplication-dimensions}
dim(matrix(1:6, nrow = 2))
dim(matrix(1:8, nrow = 4))
```

# 25. Integrated example: assessment matrix

```{r integrated-assessment-matrix}
assessment_matrix <- matrix(
  c(
    28, 48,
    30, 44,
    35, 51,
    24, 39,
    32, 54
  ),
  nrow = 5,
  byrow = TRUE,
  dimnames = list(
    paste0("Student_", 1:5),
    c("Coursework", "Examination")
  )
)

assessment_matrix
```

Student totals:

```{r assessment-totals}
rowSums(assessment_matrix)
```

Component means:

```{r assessment-component-means}
colMeans(assessment_matrix)
```

Weighted score:

```{r assessment-weights}
weights <- c(0.40, 0.60)

assessment_matrix %*% weights
```

# 26. Common mistakes

- assuming matrices fill by row;
- confusing `*` with `%*%`;
- dropping dimensions unintentionally;
- multiplying incompatible matrices;
- calculating an inverse for a singular matrix;
- forgetting missing-value handling;
- assuming matrices can contain independent column types.

# 27. Independent exercises

1. Create a 4-by-3 matrix filled by row.
2. Add row and column names.
3. Extract cells, rows and columns.
4. Replace one value.
5. Calculate row and column summaries.
6. Demonstrate element-wise and matrix multiplication.
7. Transpose a matrix.
8. Calculate determinant and inverse where valid.
9. Solve a two-equation linear system.
10. Create a three-dimensional array.
11. Calculate means over different margins.
12. Convert a matrix to a data frame.

# 28. Lesson assessment

Create a standalone R Markdown report that:

- creates and names a matrix;
- demonstrates indexing;
- calculates row and column summaries;
- performs element-wise and matrix multiplication;
- transposes a matrix;
- solves a linear system;
- creates a named three-dimensional array;
- summarises array margins; and
- explains dimension compatibility.

# 29. Completion checklist

Confirm that you can:

- create matrices;
- control filling direction;
- name dimensions;
- index cells, rows and columns;
- preserve dimensions;
- perform matrix arithmetic;
- use `%*%`;
- transpose;
- calculate row and column summaries;
- bind vectors;
- work with diagonal matrices;
- calculate determinant and inverse;
- solve linear systems;
- create and index arrays; and
- apply functions over margins.

# Lesson conclusion

Matrices and arrays extend vectors into multidimensional structures. They are
especially valuable when the analytical problem has a natural row-column or
higher-dimensional organisation.
