---
title: "Lesson 8: Lists, Data Frames and Tibbles"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(digits = 4, scipen = 999, width = 90)
set.seed(20260713)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Lists and data frames allow R to organise heterogeneous information. A list can
contain objects of entirely different forms. A data frame is a list of
equal-length columns and is the principal structure for rectangular datasets.

Tibbles are modern data-frame variants supplied by the `tibble` package. This
lesson begins with base R lists and data frames, then introduces tibbles as an
optional extension.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. create and inspect lists;
2. extract list components;
3. modify lists;
4. understand nested lists;
5. create data frames;
6. inspect rows, columns and dimensions;
7. add, remove and transform columns;
8. append rows carefully;
9. understand recycling in data-frame creation;
10. distinguish matrices and data frames;
11. import CSV data as a data frame;
12. understand tibble behaviour; and
13. validate rectangular data structures.

# 1. Creating a list

```{r create-list}
student_record <- list(
  name = "Ama",
  age = 22L,
  programme = "Statistics",
  scores = c(68, 74, 81),
  registered = TRUE
)

student_record
```

A list can contain different object types and lengths.

# 2. Inspecting a list

```{r inspect-list}
class(student_record)
typeof(student_record)
length(student_record)
names(student_record)
str(student_record)
```

# 3. Extracting list components

## 3.1 `$`

```{r list-dollar-extraction}
student_record$name
student_record$scores
```

## 3.2 Double brackets

```{r list-double-bracket}
student_record[["programme"]]
student_record[[4]]
```

Double brackets extract the component itself.

## 3.3 Single brackets

```{r list-single-bracket}
student_record["programme"]
student_record[4]
```

Single brackets return a smaller list.

# 4. Modifying lists

```{r modify-list-component}
student_record$age <- 23L
student_record
```

Add a component:

```{r add-list-component}
student_record$email <-
  "ama@example.com"

student_record
```

Remove a component:

```{r remove-list-component}
student_record$email <- NULL
student_record
```

# 5. Nested lists

```{r nested-list}
course <- list(
  title = "Learning R",
  lecturer = list(
    name = "Dr. Apam",
    role = "Lecturer"
  ),
  lessons = c(
    "Foundations",
    "Data structures",
    "Analysis"
  )
)

course
```

Extract nested values:

```{r extract-nested-list}
course$lecturer$name
course[["lecturer"]][["role"]]
```

# 6. Lists as function results

Many R functions return lists.

```{r list-function-result}
model_information <- list(
  coefficients = c(
    intercept = 2.5,
    slope = 1.2
  ),
  residual_sd = 0.8,
  converged = TRUE
)

model_information$coefficients
```

Later, fitted statistical models will use list-like structures.

# 7. Creating a data frame

```{r create-data-frame}
students <- data.frame(
  student_id = c(
    "S001",
    "S002",
    "S003",
    "S004"
  ),
  programme = c(
    "Statistics",
    "Procurement",
    "Hospitality",
    "Agriculture"
  ),
  age = c(21L, 23L, 22L, 24L),
  final_score = c(72, 68, 81, 75),
  passed = c(TRUE, TRUE, TRUE, TRUE),
  stringsAsFactors = FALSE
)

students
```

# 8. Inspecting a data frame

```{r inspect-data-frame}
class(students)
typeof(students)
dim(students)
nrow(students)
ncol(students)
names(students)
str(students)
summary(students)
```

A data frame is internally a list of equal-length columns.

```{r data-frame-is-list}
is.list(students)
```

# 9. Extracting columns

```{r data-frame-dollar}
students$final_score
```

```{r data-frame-double-bracket}
students[["final_score"]]
```

```{r data-frame-single-bracket}
students["final_score"]
```

The first two return a vector. The third returns a one-column data frame.

# 10. Extracting rows and cells

```{r data-frame-cell}
students[2, 4]
```

```{r data-frame-row}
students[2, ]
```

```{r data-frame-multiple-rows}
students[c(1, 3), ]
```

# 11. Selecting columns

```{r select-columns-name}
students[
  c(
    "student_id",
    "programme",
    "final_score"
  )
]
```

```{r select-columns-position}
students[, c(1, 2, 4)]
```

Preserve data-frame structure:

```{r preserve-data-frame-column}
students[
  ,
  "final_score",
  drop = FALSE
]
```

# 12. Adding columns

```{r add-grade-column}
students$grade <- c(
  "B",
  "B",
  "A",
  "B"
)

students
```

Calculate from existing columns:

```{r add-distinction-column}
students$distinction <-
  students$final_score >= 80

students
```

# 13. Updating columns

```{r update-programme-column}
students$programme[
  students$student_id == "S004"
] <- "Ecological Agriculture"

students
```

# 14. Removing columns

```{r remove-column-null}
students$grade <- NULL
students
```

Alternative:

```{r remove-column-subset}
students_without_distinction <-
  students[
    ,
    names(students) != "distinction"
  ]

students_without_distinction
```

# 15. Adding rows

```{r add-row-rbind}
new_student <- data.frame(
  student_id = "S005",
  programme = "Statistics",
  age = 20L,
  final_score = 77,
  passed = TRUE,
  distinction = FALSE,
  stringsAsFactors = FALSE
)

students <- rbind(
  students,
  new_student
)

students
```

Column names and compatible types should match.

# 16. Data-frame length rules

All columns must have equal lengths or be recyclable.

```{r invalid-data-frame-length, error=TRUE}
data.frame(
  x = 1:3,
  y = 1:2
)
```

A length-one value can be recycled:

```{r data-frame-length-one-recycling}
data.frame(
  student_id = c("S1", "S2", "S3"),
  institution = "BTU",
  stringsAsFactors = FALSE
)
```

# 17. Matrices versus data frames

A matrix has one atomic type.

```{r mixed-matrix}
mixed_matrix <- matrix(
  c(1, 2, "A", "B"),
  nrow = 2
)

typeof(mixed_matrix)
```

Everything becomes character.

A data frame can preserve different column types.

```{r mixed-data-frame}
mixed_data_frame <- data.frame(
  number = c(1, 2),
  category = c("A", "B"),
  stringsAsFactors = FALSE
)

str(mixed_data_frame)
```

# 18. Importing CSV data

```{r import-student-data-frame}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

head(student_data)
str(student_data)
```

# 19. Validating a data frame

```{r validate-data-frame}
is.data.frame(student_data)
nrow(student_data) > 0
ncol(student_data) > 0
```

Required columns:

```{r validate-required-columns}
required_columns <- c(
  "student_id",
  "programme",
  "coursework_score",
  "examination_score",
  "final_score"
)

all(
  required_columns %in%
    names(student_data)
)
```

Unique identifiers:

```{r validate-unique-identifiers}
!anyDuplicated(
  student_data$student_id
)
```

Score arithmetic:

```{r validate-score-arithmetic}
all(
  student_data$final_score ==
    student_data$coursework_score +
    student_data$examination_score
)
```

# 20. Column-wise summaries

```{r data-frame-column-summary}
mean(student_data$final_score)
median(student_data$final_score)
table(student_data$programme)
```

Apply over numeric columns:

```{r numeric-columns}
numeric_columns <- vapply(
  student_data,
  is.numeric,
  FUN.VALUE = logical(1)
)

numeric_columns
```

```{r summarise-numeric-columns}
vapply(
  student_data[numeric_columns],
  mean,
  na.rm = TRUE,
  FUN.VALUE = numeric(1)
)
```

# 21. `lapply()`, `sapply()` and `vapply()`

## 21.1 `lapply()`

```{r lapply-data-frame}
lapply(
  student_data,
  class
)
```

Always returns a list.

## 21.2 `sapply()`

```{r sapply-data-frame}
sapply(
  student_data,
  class
)
```

Attempts to simplify.

## 21.3 `vapply()`

```{r vapply-data-frame}
vapply(
  student_data,
  class,
  FUN.VALUE = character(1)
)
```

Requires an expected output type and is safer in programming.

# 22. Tibbles

A tibble is a modern data-frame variant.

Check package availability:

```{r check-tibble-package}
requireNamespace(
  "tibble",
  quietly = TRUE
)
```

Create a tibble if installed:

```{r create-tibble, eval=FALSE}
student_tibble <- tibble::tibble(
  student_id = c("S001", "S002"),
  score = c(72, 81)
)

student_tibble
```

## 22.1 Tibble printing

Tibbles print a limited number of rows and show column types.

## 22.2 Tibble extraction

In a tibble:

```r
student_tibble["score"]
```

returns a tibble, while:

```r
student_tibble[["score"]]
```

returns the vector.

## 22.3 Convert between forms

```{r convert-data-frame-tibble, eval=FALSE}
student_tibble <-
  tibble::as_tibble(student_data)

base_data_frame <-
  as.data.frame(student_tibble)
```

# 23. Nested data structures

A list can store several data frames.

```{r list-of-data-frames}
department_data <- list(
  statistics = student_data[
    student_data$programme == "Statistics",
  ],
  procurement = student_data[
    student_data$programme == "Procurement",
  ]
)

names(department_data)
```

Inspect:

```{r inspect-list-data-frames}
lapply(
  department_data,
  dim
)
```

# 24. Integrated example: clinic data frame

```{r import-clinic-data-frame}
clinic <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)

head(clinic)
str(clinic)
```

Validate required columns:

```{r clinic-required-columns}
clinic_required <- c(
  "visit_id",
  "sex",
  "age",
  "visit_type",
  "waiting_minutes",
  "consultation_minutes"
)

all(
  clinic_required %in%
    names(clinic)
)
```

Add target status:

```{r clinic-target-column}
clinic$target_met <-
  clinic$waiting_minutes <= 30

clinic
```

Create summary list:

```{r clinic-summary-list}
clinic_summary <- list(
  number_of_visits = nrow(clinic),
  mean_waiting_time = mean(
    clinic$waiting_minutes,
    na.rm = TRUE
  ),
  median_waiting_time = median(
    clinic$waiting_minutes,
    na.rm = TRUE
  ),
  target_percentage = mean(
    clinic$target_met,
    na.rm = TRUE
  ) * 100
)

clinic_summary
```

# 25. Common mistakes

- using single brackets when a vector is required;
- using double brackets when a one-column data frame is required;
- adding rows with incompatible columns;
- assuming matrices and data frames behave identically;
- relying on silent recycling;
- overwriting a column unintentionally;
- failing to validate identifiers;
- assuming tibbles are base R objects without installing `tibble`.

# 26. Independent exercises

1. Create a list containing mixed objects.
2. Extract components with `$`, `[[` and `[`.
3. Add and remove list components.
4. Create a nested list.
5. Create a data frame with mixed column types.
6. Inspect dimensions, names and structure.
7. Add calculated columns.
8. Remove columns.
9. Add a compatible row.
10. Compare matrix and data-frame coercion.
11. Import and validate student data.
12. Use `lapply()`, `sapply()` and `vapply()`.
13. Create a tibble if the package is installed.
14. Create a list of programme-specific data frames.

# 27. Lesson assessment

Create a standalone R Markdown report that:

- creates and modifies a list;
- demonstrates three extraction methods;
- creates a data frame;
- adds and removes columns;
- appends a row;
- validates required columns and unique identifiers;
- compares matrix and data-frame behaviour;
- summarises numeric columns;
- demonstrates `lapply()`, `sapply()` and `vapply()`; and
- optionally creates a tibble.

# 28. Completion checklist

Confirm that you can:

- create lists;
- inspect and extract list components;
- modify and nest lists;
- create data frames;
- inspect rows, columns and structure;
- extract columns correctly;
- add and remove columns;
- append compatible rows;
- explain data-frame length rules;
- distinguish matrices and data frames;
- import CSV data;
- validate data frames;
- apply functions across columns;
- explain tibbles; and
- convert between tibbles and data frames.

# Lesson conclusion

Lists provide flexible heterogeneous storage. Data frames organise equal-length
vectors into rectangular datasets. Together, they form the structural basis of
most practical R analysis.
