---
title: "Lesson 9: Indexing, Subsetting and Extracting Values"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

A dataset becomes useful only when the analyst can locate and extract the
observations and variables required for a particular question. In R, this is
done through indexing and subsetting.

Indexing answers questions such as:

- What is the third score?
- Which students passed?
- Which rows belong to the Statistics programme?
- Which columns contain assessment information?
- Which observations have missing rainfall values?
- Which records satisfy several conditions at the same time?

Subsetting is not merely a technical operation. It is a translation of an
analytical condition into code. An incorrect subset may produce a convincing
but invalid result. This lesson therefore treats indexing as both a programming
skill and a data-validation responsibility.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain one-dimensional and two-dimensional indexing;
2. use positive, negative and logical indices;
3. extract values by position and by name;
4. understand the difference between `[`, `[[` and `$`;
5. subset vectors, matrices, lists and data frames;
6. preserve or drop dimensions deliberately;
7. select rows using logical conditions;
8. select columns using names and positions;
9. combine row and column conditions;
10. use `which()`, `which.min()` and `which.max()`;
11. identify and extract missing or duplicate observations;
12. modify selected elements safely;
13. avoid common indexing errors; and
14. construct reproducible analytical subsets.

# 1. The meaning of an index

An index identifies the position or name of an element.

Consider:

```{r create-score-vector}
scores <- c(
  68,
  74,
  81,
  59,
  90
)

scores
```

The five positions are:

\[
1,\ 2,\ 3,\ 4,\ 5.
\]

The third position contains 81.

```{r extract-third-score}
scores[3]
```

The general structure is:

```text
object[index]
```

# 2. Positive positional indexing

Positive indices select positions.

```{r positive-single-index}
scores[1]
```

```{r positive-multiple-index}
scores[
  c(
    1,
    3,
    5
  )
]
```

Repeated indices are allowed:

```{r repeated-positive-index}
scores[
  c(
    2,
    2,
    4
  )
]
```

The selected value at position 2 is returned twice.

## 2.1 Ordering follows the index

```{r positive-index-order}
scores[
  c(
    5,
    1,
    3
  )
]
```

R returns values in the order specified by the index.

# 3. Negative indexing

Negative indices exclude positions.

```{r negative-single-index}
scores[-2]
```

This returns every score except the second.

```{r negative-multiple-index}
scores[
  -c(
    1,
    5
  )
]
```

This excludes the first and fifth values.

## 3.1 Do not mix positive and negative indices

```{r mixed-index-error, error=TRUE}
scores[
  c(
    1,
    -2
  )
]
```

Except for zero, positive and negative indices cannot be mixed in the same
selection.

# 4. Zero and out-of-range indices

```{r zero-index}
scores[0]
```

Position zero selects nothing.

```{r out-of-range-index}
scores[10]
```

A positive position beyond the vector length returns `NA`.

Negative exclusion beyond the vector length has no practical effect:

```{r negative-out-of-range}
scores[-10]
```

# 5. Logical indexing

A logical index selects positions corresponding to `TRUE`.

```{r logical-index-explicit}
logical_index <- c(
  TRUE,
  FALSE,
  TRUE,
  FALSE,
  TRUE
)

scores[logical_index]
```

Logical indexing is particularly useful when the index is generated from a
condition.

```{r logical-index-condition}
scores[
  scores >= 70
]
```

The expression:

```r
scores >= 70
```

produces:

```{r inspect-logical-condition}
scores >= 70
```

R then extracts values corresponding to `TRUE`.

# 6. Compound conditions

Select scores from 70 through 85:

```{r compound-range-condition}
scores[
  scores >= 70 &
    scores <= 85
]
```

Select scores below 60 or above 85:

```{r compound-or-condition}
scores[
  scores < 60 |
    scores > 85
]
```

Exclude scores equal to 74:

```{r compound-not-condition}
scores[
  scores != 74
]
```

# 7. Named-vector indexing

```{r create-named-vector}
programme_counts <- c(
  Statistics = 35,
  Procurement = 42,
  Hospitality = 28,
  Agriculture = 31
)

programme_counts
```

Extract by name:

```{r extract-named-vector}
programme_counts[
  "Statistics"
]
```

Extract several names:

```{r extract-multiple-names}
programme_counts[
  c(
    "Statistics",
    "Agriculture"
  )
]
```

Unknown names return `NA`:

```{r unknown-vector-name}
programme_counts[
  "Engineering"
]
```

# 8. `which()`

`which()` returns the positions where a logical condition is `TRUE`.

```{r which-example}
which(
  scores >= 70
)
```

Use positions to extract values:

```{r which-extraction}
high_score_positions <- which(
  scores >= 70
)

scores[
  high_score_positions
]
```

Direct logical indexing is often shorter, but `which()` is useful when the
positions themselves are required.

# 9. Minimum and maximum positions

```{r which-minimum-maximum}
which.min(scores)
which.max(scores)
```

Extract corresponding values:

```{r extract-minimum-maximum}
scores[
  which.min(scores)
]

scores[
  which.max(scores)
]
```

# 10. Missing-value indexing

```{r vector-with-missing}
rainfall <- c(
  115.2,
  98.4,
  NA,
  121.7,
  110.5
)
```

Select missing positions:

```{r select-missing-values}
rainfall[
  is.na(rainfall)
]
```

Select observed values:

```{r select-observed-values}
rainfall[
  !is.na(rainfall)
]
```

Locate missing positions:

```{r locate-missing-values}
which(
  is.na(rainfall)
)
```

# 11. Duplicate indexing

```{r duplicate-identifiers}
student_ids <- c(
  "S001",
  "S002",
  "S003",
  "S002",
  "S004",
  "S003"
)
```

Identify repeated occurrences:

```{r duplicate-status}
duplicated(student_ids)
```

Extract repeated values:

```{r extract-duplicates}
student_ids[
  duplicated(student_ids)
]
```

Extract all observations participating in duplication:

```{r extract-all-duplicate-members}
student_ids[
  duplicated(student_ids) |
    duplicated(
      student_ids,
      fromLast = TRUE
    )
]
```

# 12. Replacing selected vector elements

```{r replace-vector-element}
corrected_scores <- scores

corrected_scores[4] <- 62

corrected_scores
```

Replace several positions:

```{r replace-multiple-elements}
corrected_scores[
  c(
    1,
    2
  )
] <- c(
  70,
  76
)

corrected_scores
```

Replace using a condition:

```{r replace-by-condition}
capped_scores <- c(
  45,
  72,
  108,
  95
)

capped_scores[
  capped_scores > 100
] <- 100

capped_scores
```

# 13. Matrix indexing

Create a matrix:

```{r create-indexing-matrix}
score_matrix <- matrix(
  c(
    30,
    48,
    28,
    42,
    35,
    51
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c(
      "Ama",
      "Kojo",
      "Esi"
    ),
    c(
      "Coursework",
      "Examination"
    )
  )
)

score_matrix
```

The general matrix structure is:

```text
matrix[row, column]
```

## 13.1 Extract one cell

```{r extract-matrix-cell}
score_matrix[
  2,
  1
]
```

## 13.2 Extract one row

```{r extract-matrix-row}
score_matrix[
  1,
]
```

## 13.3 Extract one column

```{r extract-matrix-column}
score_matrix[
  ,
  2
]
```

## 13.4 Extract by names

```{r extract-matrix-names}
score_matrix[
  "Esi",
  "Examination"
]
```

# 14. Dimension dropping

When one matrix row or column is selected, R may simplify the result to a
vector.

```{r matrix-dimension-drop}
class(
  score_matrix[
    1,
  ]
)
```

Preserve matrix form:

```{r preserve-matrix-dimension}
score_matrix[
  1,
  ,
  drop = FALSE
]
```

```{r inspect-preserved-matrix}
class(
  score_matrix[
    1,
    ,
    drop = FALSE
  ]
)
```

# 15. Logical indexing in matrices

```{r matrix-logical-selection}
score_matrix[
  score_matrix >= 40
]
```

This returns a vector of cells satisfying the condition.

Replace selected values:

```{r matrix-logical-replacement}
adjusted_matrix <- score_matrix

adjusted_matrix[
  adjusted_matrix < 30
] <- 30

adjusted_matrix
```

# 16. List indexing

```{r create-indexing-list}
student_record <- list(
  name = "Ama",
  age = 22L,
  programme = "Statistics",
  scores = c(
    68,
    74,
    81
  )
)
```

## 16.1 Single brackets

```{r list-single-bracket}
student_record[
  "programme"
]
```

The result is a smaller list.

## 16.2 Double brackets

```{r list-double-bracket}
student_record[[
  "programme"
]]
```

The result is the component itself.

## 16.3 Dollar operator

```{r list-dollar-operator}
student_record$programme
```

## 16.4 Positional extraction

```{r list-positional-extraction}
student_record[[4]]
```

# 17. Data-frame indexing

Import the student dataset:

```{r import-student-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

head(student_data)
```

The general structure is:

```text
data_frame[row, column]
```

# 18. Extracting data-frame cells and rows

```{r data-frame-cell}
student_data[
  1,
  1
]
```

```{r data-frame-first-row}
student_data[
  1,
]
```

```{r data-frame-multiple-rows}
student_data[
  c(
    1,
    3,
    5
  ),
]
```

# 19. Extracting data-frame columns

```{r data-frame-column-position}
student_data[
  ,
  2
]
```

```{r data-frame-column-name}
student_data[
  ,
  "programme"
]
```

Both may return a vector.

Preserve a one-column data frame:

```{r preserve-data-frame-column}
student_data[
  ,
  "programme",
  drop = FALSE
]
```

# 20. `$`, `[[` and `[` for data frames

```{r data-frame-dollar}
student_data$final_score
```

```{r data-frame-double-bracket}
student_data[[
  "final_score"
]]
```

```{r data-frame-single-bracket}
student_data[
  "final_score"
]
```

The first two return a vector. The third returns a one-column data frame.

# 21. Selecting several columns

```{r select-several-columns}
student_data[
  c(
    "student_id",
    "programme",
    "final_score"
  )
]
```

Select by positions:

```{r select-several-column-positions}
student_data[
  ,
  c(
    1,
    2,
    6
  )
]
```

# 22. Selecting rows with conditions

Students who passed:

```{r select-passing-students}
passing_students <- student_data[
  student_data$final_score >= 50,
]

passing_students
```

Students with distinctions:

```{r select-distinction-students}
distinction_students <- student_data[
  student_data$final_score >= 80,
]

distinction_students
```

# 23. Combining row and column selection

```{r combine-row-column-selection}
student_data[
  student_data$final_score >= 80,
  c(
    "student_id",
    "programme",
    "final_score"
  )
]
```

This selects only distinction records and only the named columns.

# 24. Multiple conditions in data frames

```{r multiple-data-frame-conditions}
selected_students <- student_data[
  student_data$final_score >= 70 &
    student_data$programme == "Statistics",
]

selected_students
```

Use `%in%` for several categories:

```{r membership-data-frame-condition}
selected_programmes <- student_data[
  student_data$programme %in%
    c(
      "Statistics",
      "Procurement"
    ),
]

selected_programmes
```

# 25. Excluding rows and columns

Exclude the first row:

```{r exclude-first-row}
student_data[
  -1,
]
```

Exclude selected columns:

```{r exclude-selected-columns}
student_data[
  ,
  !names(student_data) %in%
    c(
      "coursework_score",
      "examination_score"
    )
]
```

# 26. `subset()`

```{r subset-function}
subset(
  student_data,
  final_score >= 70,
  select = c(
    student_id,
    programme,
    final_score
  )
)
```

`subset()` can be convenient interactively, but bracket indexing is more
explicit and is generally preferred inside functions.

# 27. Selecting complete cases

```{r create-incomplete-data-frame}
incomplete_data <- data.frame(
  id = 1:5,
  score = c(
    65,
    NA,
    72,
    81,
    NA
  ),
  attendance = c(
    0.90,
    0.85,
    NA,
    0.95,
    0.70
  )
)

incomplete_data
```

Rows complete across all columns:

```{r complete-case-rows}
incomplete_data[
  complete.cases(
    incomplete_data
  ),
]
```

Rows complete for one variable:

```{r complete-specific-variable}
incomplete_data[
  !is.na(
    incomplete_data$score
  ),
]
```

# 28. Row numbers

```{r row-number-selection}
row_numbers <- seq_len(
  nrow(student_data)
)

row_numbers
```

Use a condition to retrieve row numbers:

```{r row-number-condition}
which(
  student_data$final_score >= 80
)
```

# 29. Ordering and indexing

Order students from highest to lowest score:

```{r order-data-frame}
score_order <- order(
  student_data$final_score,
  decreasing = TRUE
)

student_data[
  score_order,
]
```

Order by programme and then descending score:

```{r multi-variable-order}
programme_score_order <- order(
  student_data$programme,
  -student_data$final_score
)

student_data[
  programme_score_order,
]
```

# 30. Modifying selected data-frame values

```{r modify-data-frame-copy}
corrected_student_data <- student_data
```

Correct one cell:

```{r correct-one-cell}
corrected_student_data[
  1,
  "coursework_score"
] <- 35
```

Recalculate the corresponding final score:

```{r recalculate-one-final-score}
corrected_student_data[
  1,
  "final_score"
] <-
  corrected_student_data[
    1,
    "coursework_score"
  ] +
  corrected_student_data[
    1,
    "examination_score"
  ]
```

Inspect:

```{r inspect-corrected-row}
corrected_student_data[
  1,
]
```

# 31. Safe modification with validation

```{r validate-corrected-scores}
all(
  corrected_student_data$final_score ==
    corrected_student_data$coursework_score +
    corrected_student_data$examination_score
)
```

Never alter a derived variable without checking whether related variables must
also be updated.

# 32. Integrated example: clinic subsetting

```{r import-clinic-data}
clinic <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)

head(clinic)
names(clinic)
```

Visits with waiting times greater than 30 minutes:

```{r clinic-long-wait}
long_wait_records <- clinic[
  clinic$waiting_minutes > 30,
]

long_wait_records
```

Long waits for new visits:

```{r clinic-multiple-condition}
selected_clinic_records <- clinic[
  clinic$waiting_minutes > 30 &
    clinic$visit_type == "New",
  c(
    "visit_id",
    "sex",
    "age",
    "waiting_minutes",
    "visit_type"
  )
]

selected_clinic_records
```

The record with the longest waiting time:

```{r clinic-longest-wait}
clinic[
  which.max(
    clinic$waiting_minutes
  ),
]
```

# 33. Common mistakes

- using zero-based indexing as in some other languages;
- mixing positive and negative indices;
- forgetting the comma in data-frame indexing;
- confusing `[` and `[[`;
- allowing dimensions to drop unintentionally;
- comparing missing values with `== NA`;
- using `&` where `|` is intended;
- forgetting parentheses around complex conditions;
- modifying derived values without recalculating dependencies;
- assuming an unknown name produces an error rather than `NA`.

# 34. Independent exercises

1. Extract selected positions from a numeric vector.
2. Exclude positions using negative indices.
3. Select values with logical conditions.
4. Create and index a named vector.
5. locate minimum, maximum and missing positions.
6. extract duplicates.
7. modify selected vector values.
8. index a matrix by positions and names.
9. preserve matrix dimensions.
10. demonstrate `[`, `[[` and `$` on a list.
11. import student data and select passing students.
12. select distinction students from specified programmes.
13. identify complete cases in rainfall or clinic data.
14. order records by one and then two variables.
15. modify a copied dataset and validate the result.

# 35. Lesson assessment

Create a standalone R Markdown report that:

1. demonstrates positive, negative and logical vector indexing;
2. uses named indexing;
3. identifies missing and duplicate values;
4. indexes a matrix by row, column and names;
5. demonstrates `[`, `[[` and `$`;
6. imports `student_scores.csv`;
7. selects rows using one and multiple conditions;
8. selects specified columns;
9. orders records;
10. modifies a copied value safely; and
11. validates the modified dataset.

# 36. Completion checklist

Confirm that you can:

- index vectors by position;
- exclude positions;
- use logical indices;
- select by name;
- locate positions with `which()`;
- identify minimum and maximum positions;
- extract missing and duplicate observations;
- modify selected elements;
- index matrices;
- preserve dimensions;
- distinguish `[`, `[[` and `$`;
- subset data-frame rows and columns;
- combine conditions;
- select complete cases;
- order records; and
- validate modifications.

# Lesson conclusion

Indexing translates analytical conditions into selected observations and
variables. Reliable subsetting requires careful attention to positions, names,
logical conditions, dimensions and missing values.

The working sequence is:

\[
\boxed{
\text{define condition}
\rightarrow
\text{inspect index}
\rightarrow
\text{subset}
\rightarrow
\text{verify result}
\rightarrow
\text{analyse}
}
\]
