---
title: "Lesson 10: Operators, Comparisons and Logical Conditions"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Operators allow R to combine values, compare observations, test membership and
construct logical conditions. These operations form the foundation of data
validation, recoding, filtering and decision-making.

This lesson develops arithmetic, relational, logical, membership and special
operators. It also explains precedence, vectorised versus short-circuit logic,
floating-point comparison and missing-value behaviour.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. use arithmetic operators correctly;
2. distinguish assignment and comparison;
3. apply relational operators;
4. combine conditions using logical operators;
5. distinguish `&` from `&&` and `|` from `||`;
6. use membership and sequence operators;
7. understand operator precedence;
8. test approximate numerical equality;
9. handle missing logical values;
10. create validation conditions;
11. translate verbal criteria into code; and
12. inspect complex logical expressions systematically.

# 1. Arithmetic operators

| Operator | Meaning |
|---|---|
| `+` | addition |
| `-` | subtraction |
| `*` | multiplication |
| `/` | division |
| `^` | exponentiation |
| `%/%` | integer division |
| `%%` | remainder |

```{r arithmetic-operators}
10 + 3
10 - 3
10 * 3
10 / 3
10 ^ 3
10 %/% 3
10 %% 3
```

# 2. Unary operators

```{r unary-operators}
+5
-5
```

The minus sign can negate a value.

```{r negate-vector}
values <- c(
  5,
  -3,
  8
)

-values
```

# 3. Relational operators

| Operator | Meaning |
|---|---|
| `>` | greater than |
| `<` | less than |
| `>=` | greater than or equal to |
| `<=` | less than or equal to |
| `==` | equal to |
| `!=` | not equal to |

```{r relational-operators}
75 > 50
75 < 50
75 >= 75
75 <= 74
75 == 75
75 != 70
```

# 4. Assignment is not comparison

```{r assignment-comparison}
score <- 75

score == 75
```

`<-` stores a value. `==` tests equality.

# 5. Vectorised comparisons

```{r vectorised-comparisons}
scores <- c(
  45,
  58,
  72,
  84,
  91
)

scores >= 50
scores < 70
scores == 72
scores != 72
```

# 6. Logical AND

Use `&` when both conditions must be true.

```{r logical-and}
attendance <- c(
  0.80,
  0.92,
  0.88,
  0.95,
  0.70
)

eligible <-
  (scores >= 70) &
  (attendance >= 0.85)

eligible
```

# 7. Logical OR

Use `|` when at least one condition must be true.

```{r logical-or}
requires_support <-
  (scores < 50) |
  (attendance < 0.75)

requires_support
```

# 8. Logical NOT

```{r logical-not}
!eligible
```

Negation reverses `TRUE` and `FALSE`.

# 9. Exclusive OR

```{r exclusive-or}
xor(
  TRUE,
  FALSE
)

xor(
  TRUE,
  TRUE
)
```

`xor()` is true when exactly one condition is true.

# 10. Vectorised and short-circuit operators

`&` and `|` operate element by element.

```{r vectorised-logical-operators}
c(
  TRUE,
  FALSE,
  TRUE
) & c(
  TRUE,
  TRUE,
  FALSE
)
```

`&&` and `||` examine only the first element and are mainly used in control
flow.

```{r short-circuit-operators}
first_condition <- TRUE
second_condition <- FALSE

first_condition &&
  second_condition

first_condition ||
  second_condition
```

The operands supplied to `&&` and `||` must each be a single logical value.

To show the contrast with vectorised operators:

```{r vectorised-versus-short-circuit}
left_vector <- c(
  TRUE,
  FALSE
)

right_vector <- c(
  TRUE,
  TRUE
)

left_vector &
  right_vector

left_vector |
  right_vector

left_vector[1] &&
  right_vector[1]
```

Use:

- `&` and `|` for vectors;
- `&&` and `||` for single decision conditions.

# 11. Membership operator

```{r membership-operator}
programme <- c(
  "Statistics",
  "Hospitality",
  "Agriculture",
  "Procurement"
)

programme %in% c(
  "Statistics",
  "Procurement"
)
```

Negated membership:

```{r negated-membership}
!programme %in% c(
  "Statistics",
  "Procurement"
)
```

For readability, parentheses are often helpful:

```{r parenthesised-negated-membership}
!(
  programme %in% c(
    "Statistics",
    "Procurement"
  )
)
```

# 12. Sequence operator

```{r sequence-operator}
1:10
```

The colon operator has relatively high precedence.

```{r sequence-precedence}
1:3 + 1
```

This is interpreted as:

```r
(1:3) + 1
```

# 13. Matrix multiplication and outer product

Matrix multiplication:

```{r matrix-multiplication-operator}
a <- matrix(
  1:4,
  nrow = 2
)

b <- matrix(
  5:8,
  nrow = 2
)

a %*% b
```

Outer product:

```{r outer-product-operator}
1:3 %o% 1:4
```

# 14. Set and model-related operators

The formula operator `~` is used later in statistical models.

```{r formula-operator}
model_formula <-
  final_score ~
  coursework_score +
  examination_score

model_formula
```

The namespace operator identifies a package function:

```{r namespace-operator}
rmarkdown::pandoc_available()
```

# 15. Operator precedence

Consider:

```{r precedence-example-one}
10 + 2 * 5
```

Multiplication occurs before addition.

```{r precedence-example-two}
(10 + 2) * 5
```

Parentheses change the order.

Logical example:

```{r logical-precedence}
TRUE | FALSE & FALSE
```

`&` is evaluated before `|`.

Prefer explicit parentheses:

```{r explicit-logical-precedence}
TRUE | (
  FALSE & FALSE
)
```

# 16. Chained comparisons are invalid logic

In mathematics one may write:

\[
50 \leq x \leq 100.
\]

Do not translate this as:

```r
50 <= x <= 100
```

Correct:

```{r correct-range-comparison}
x <- 75

x >= 50 &
  x <= 100
```

# 17. Floating-point comparison

Some decimal calculations cannot be represented exactly in binary floating
point.

```{r floating-point-example}
0.1 + 0.2
```

Direct comparison:

```{r floating-point-direct-comparison}
0.1 + 0.2 == 0.3
```

Use `all.equal()`:

```{r floating-point-all-equal}
all.equal(
  0.1 + 0.2,
  0.3
)
```

Or use a tolerance:

```{r floating-point-tolerance}
abs(
  (0.1 + 0.2) - 0.3
) < 1e-8
```

# 18. Missing values in comparisons

```{r missing-comparison}
x <- c(
  10,
  NA,
  20
)

x > 15
```

The result for the missing value is `NA`.

```{r missing-logical-summary}
sum(
  x > 15
)
```

This returns `NA`.

Use:

```{r missing-logical-summary-remove}
sum(
  x > 15,
  na.rm = TRUE
)
```

# 19. Three-valued logic

R logical expressions may produce:

- `TRUE`;
- `FALSE`;
- `NA`.

```{r three-valued-logic}
TRUE & NA
FALSE & NA
TRUE | NA
FALSE | NA
```

Interpretation:

- `FALSE & NA` is definitely false;
- `TRUE | NA` is definitely true;
- other results remain unknown.

# 20. `isTRUE()` and `isFALSE()`

```{r istrue-isfalse}
isTRUE(TRUE)
isTRUE(NA)
isFALSE(FALSE)
isFALSE(NA)
```

These functions are useful when a control-flow condition must be exactly true
or false.

# 21. `identical()`

```{r identical-examples}
identical(
  1,
  1
)

identical(
  1,
  1L
)
```

The second result is false because type is considered.

```{r equality-versus-identical}
1 == 1L
```

Numerical equality and complete identity are different concepts.

# 22. Validation conditions

```{r validation-condition}
scores <- c(
  65,
  72,
  81,
  59
)

scores_are_numeric <-
  is.numeric(scores)

scores_have_no_missing_values <-
  !anyNA(scores)

scores_are_in_range <-
  all(
    scores >= 0 &
      scores <= 100
  )

scores_are_numeric
scores_have_no_missing_values
scores_are_in_range
```

Combine:

```{r combined-validation-condition}
scores_are_valid <-
  scores_are_numeric &&
  scores_have_no_missing_values &&
  scores_are_in_range

scores_are_valid
```

The component conditions are single logical values, so `&&` is appropriate.

# 23. Translating verbal rules

Rule:

> A student progresses when the final score is at least 50 and attendance is at
> least 75 per cent.

```{r translate-progression-rule}
final_score <- 64
attendance_rate <- 0.82

progresses <-
  final_score >= 50 &&
  attendance_rate >= 0.75

progresses
```

Rule:

> A procurement record requires review when the value exceeds GH₵50,000 or
> delivery is more than seven days late.

```{r translate-procurement-rule}
contract_value <- 62000
delay_days <- 4

requires_review <-
  contract_value > 50000 ||
  delay_days > 7

requires_review
```

# 24. Vectorised rule application

```{r vectorised-rule-data}
final_scores <- c(
  45,
  58,
  72,
  84
)

attendance_rates <- c(
  0.90,
  0.72,
  0.80,
  0.95
)

progress_status <-
  (final_scores >= 50) &
  (attendance_rates >= 0.75)

progress_status
```

# 25. Logical counts and proportions

```{r logical-counts-proportions}
sum(progress_status)
mean(progress_status)
mean(progress_status) * 100
```

# 26. De Morgan's laws

Negation of “A and B”:

\[
\neg(A\land B)
=
(\neg A)\lor(\neg B).
\]

```{r de-morgan-and}
a <- c(
  TRUE,
  TRUE,
  FALSE,
  FALSE
)

b <- c(
  TRUE,
  FALSE,
  TRUE,
  FALSE
)

!(a & b)

(!a) | (!b)
```

Negation of “A or B”:

\[
\neg(A\lor B)
=
(\neg A)\land(\neg B).
\]

```{r de-morgan-or}
!(a | b)

(!a) & (!b)
```

# 27. Integrated example: student eligibility

```{r import-student-eligibility}
students <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

head(students)
```

Create conditions:

```{r eligibility-conditions}
pass_condition <-
  students$final_score >= 50

distinction_condition <-
  students$final_score >= 80

programme_condition <-
  students$programme %in%
  c(
    "Statistics",
    "Procurement"
  )
```

Select distinction students in selected programmes:

```{r combined-eligibility-subset}
selected_students <- students[
  distinction_condition &
    programme_condition,
  c(
    "student_id",
    "programme",
    "final_score"
  )
]

selected_students
```

# 28. Common mistakes

- confusing `<-` and `==`;
- using `=` when comparison is intended;
- using `&&` on vectors;
- writing chained comparisons;
- ignoring missing logical values;
- comparing floating-point values with exact equality;
- omitting parentheses in complex expressions;
- using `&` when the rule requires `|`;
- failing to test each component condition separately.

# 29. Independent exercises

1. Demonstrate all arithmetic operators.
2. Create relational comparisons on a vector.
3. Combine two logical conditions with `&`.
4. Combine two conditions with `|`.
5. Demonstrate negation.
6. Compare vectorised and short-circuit operators.
7. Use `%in%` and negated membership.
8. demonstrate precedence with and without parentheses.
9. test approximate numerical equality.
10. demonstrate three-valued logic.
11. create a multi-part validation condition.
12. translate three verbal decision rules into R.
13. apply a rule vector-wise to imported data.
14. verify De Morgan's laws.

# 30. Lesson assessment

Create a standalone R Markdown report that:

- demonstrates arithmetic and relational operators;
- distinguishes assignment from comparison;
- uses `&`, `|`, `!`, `&&` and `||`;
- uses `%in%`;
- demonstrates precedence;
- explains floating-point comparison;
- handles missing logical results;
- creates and combines validation conditions; and
- applies a multi-condition rule to a supplied dataset.

# 31. Completion checklist

Confirm that you can:

- use arithmetic operators;
- use relational operators;
- distinguish assignment and comparison;
- combine logical conditions;
- choose vectorised or short-circuit operators;
- test membership;
- apply precedence correctly;
- avoid chained comparisons;
- compare decimals approximately;
- interpret three-valued logic;
- create validation conditions; and
- translate verbal rules into code.

# Lesson conclusion

Operators are the grammar of computational decisions. Reliable analytical
conditions require correct operators, explicit parentheses, appropriate
missing-value handling and careful translation of verbal rules.

\[
\boxed{
\text{state rule}
\rightarrow
\text{build component conditions}
\rightarrow
\text{combine}
\rightarrow
\text{inspect}
\rightarrow
\text{apply}
}
\]
