---
title: "Lesson 11: Conditional Execution with if, else and ifelse()"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Conditional execution allows a program to choose what to do based on the
current data or situation. Instead of running every command unconditionally, R
can evaluate a logical test and execute only the appropriate branch.

This lesson develops:

- `if`;
- `else`;
- `else if`;
- nested conditions;
- vectorised `ifelse()`;
- `switch()`;
- defensive validation;
- early stopping;
- conditional messages; and
- decision rules in applied contexts.

The central distinction is that `if` makes one decision, while `ifelse()`
usually makes element-wise decisions across a vector.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. write a valid `if` statement;
2. use `if` with `else`;
3. construct several branches using `else if`;
4. understand scalar logical conditions;
5. distinguish `if` from `ifelse()`;
6. create vectorised classifications;
7. handle missing conditions safely;
8. use `switch()` for named choices;
9. stop execution when validation fails;
10. issue warnings and messages conditionally;
11. avoid deeply nested and ambiguous logic;
12. translate decision rules into control flow; and
13. test conditional code systematically.

# 1. Basic `if`

The general structure is:

```text
if (condition) {
  code_to_run
}
```

Example:

```{r basic-if}
score <- 72

if (score >= 50) {
  print(
    "The student passed."
  )
}
```

The body executes only when the condition is `TRUE`.

# 2. `if` requires one logical decision

```{r if-scalar-condition}
condition <- score >= 50

condition
length(condition)
```

The condition should be one logical value.

Using a vector condition is inappropriate:

```{r if-vector-condition, error=TRUE}
scores <- c(
  45,
  72,
  81
)

if (scores >= 50) {
  print(
    "At least one student passed."
  )
}
```

Create a single summary condition:

```{r if-any-condition}
if (any(scores >= 50)) {
  print(
    "At least one student passed."
  )
}
```

Or:

```{r if-all-condition}
if (all(scores >= 50)) {
  print(
    "Every student passed."
  )
}
```

# 3. `if` and `else`

```{r if-else-example}
score <- 42

if (score >= 50) {
  result <- "Pass"
} else {
  result <- "Fail"
}

result
```

Exactly one branch runs.

# 4. Bracing and line placement

In R, `else` should normally appear on the same line as the closing brace:

```r
if (condition) {
  ...
} else {
  ...
}
```

This avoids parsing problems in interactive execution.

# 5. `else if`

```{r else-if-grading}
score <- 76

if (score >= 80) {
  grade <- "A"
} else if (score >= 70) {
  grade <- "B"
} else if (score >= 60) {
  grade <- "C"
} else if (score >= 50) {
  grade <- "D"
} else {
  grade <- "F"
}

grade
```

Conditions are checked from top to bottom. The first true branch is selected.

# 6. Order matters

Incorrect ordering:

```{r incorrect-ordering}
score <- 85

if (score >= 50) {
  classification <- "Pass"
} else if (score >= 80) {
  classification <- "Distinction"
} else {
  classification <- "Fail"
}

classification
```

The distinction branch is never reached because every distinction score also
satisfies `score >= 50`.

Correct:

```{r correct-ordering}
if (score >= 80) {
  classification <- "Distinction"
} else if (score >= 50) {
  classification <- "Pass"
} else {
  classification <- "Fail"
}

classification
```

# 7. Compound conditions inside `if`

```{r compound-if-condition}
score <- 84
attendance <- 0.93

if (
  score >= 80 &&
  attendance >= 0.90
) {
  award_status <- "Eligible"
} else {
  award_status <- "Not eligible"
}

award_status
```

Because both components are scalar logical values, `&&` is appropriate.

# 8. Nested `if` statements

```{r nested-if}
registered <- TRUE
fees_paid <- FALSE

if (registered) {
  if (fees_paid) {
    access_status <- "Full access"
  } else {
    access_status <-
      "Registration confirmed, fees outstanding"
  }
} else {
  access_status <- "Not registered"
}

access_status
```

Nested conditions can be useful, but excessive nesting reduces readability.

# 9. Flattening nested logic

The previous example can be written with `else if`:

```{r flattened-if}
if (!registered) {
  access_status <- "Not registered"
} else if (!fees_paid) {
  access_status <-
    "Registration confirmed, fees outstanding"
} else {
  access_status <- "Full access"
}

access_status
```

# 10. Returning values from `if`

`if` is an expression and can return a value.

```{r if-return-value}
score <- 68

result <- if (score >= 50) {
  "Pass"
} else {
  "Fail"
}

result
```

This form can be concise and readable.

# 11. Missing conditions

```{r missing-if-condition, error=TRUE}
score <- NA

if (score >= 50) {
  print(
    "Pass"
  )
}
```

A missing condition is neither true nor false.

Handle it explicitly:

```{r handle-missing-if}
if (is.na(score)) {
  result <- "Score missing"
} else if (score >= 50) {
  result <- "Pass"
} else {
  result <- "Fail"
}

result
```

# 12. `isTRUE()` in control flow

```{r istrue-control-flow}
condition <- NA

if (isTRUE(condition)) {
  status <- "Condition confirmed"
} else {
  status <- "Condition not confirmed"
}

status
```

`isTRUE()` returns `TRUE` only when the input is exactly a single true value.

# 13. Vectorised `ifelse()`

The general form is:

```text
ifelse(test, yes, no)
```

```{r ifelse-basic}
scores <- c(
  45,
  58,
  72,
  84
)

pass_status <- ifelse(
  scores >= 50,
  "Pass",
  "Fail"
)

pass_status
```

`ifelse()` makes one decision for each vector element.

# 14. Nested `ifelse()`

```{r nested-ifelse}
grade <- ifelse(
  scores >= 80,
  "A",
  ifelse(
    scores >= 70,
    "B",
    ifelse(
      scores >= 60,
      "C",
      ifelse(
        scores >= 50,
        "D",
        "F"
      )
    )
  )
)

grade
```

Nested `ifelse()` works, but deeply nested code becomes difficult to read.

# 15. `if` versus `ifelse()`

Use `if` when:

- one decision is required;
- a block of commands may run;
- validation or control flow is involved.

Use `ifelse()` when:

- element-wise vector classification is required;
- the output should have the same length as the test.

Example:

```{r compare-if-ifelse}
single_score <- 72

single_result <- if (
  single_score >= 50
) {
  "Pass"
} else {
  "Fail"
}

vector_results <- ifelse(
  scores >= 50,
  "Pass",
  "Fail"
)

single_result
vector_results
```

# 16. Type coercion in `ifelse()`

```{r ifelse-type-coercion}
ifelse(
  c(
    TRUE,
    FALSE
  ),
  1,
  "No"
)
```

The result becomes character because the `yes` and `no` values require a common
type.

# 17. Missing tests in `ifelse()`

```{r ifelse-missing-test}
test_values <- c(
  TRUE,
  FALSE,
  NA
)

ifelse(
  test_values,
  "Yes",
  "No"
)
```

The result corresponding to a missing test is `NA`.

Handle explicitly:

```{r ifelse-handle-missing}
ifelse(
  is.na(test_values),
  "Unknown",
  ifelse(
    test_values,
    "Yes",
    "No"
  )
)
```

# 18. Conditional replacement

```{r conditional-replacement}
scores <- c(
  45,
  58,
  108,
  84
)

corrected_scores <- ifelse(
  scores > 100,
  100,
  scores
)

corrected_scores
```

A direct indexing replacement may be clearer:

```{r direct-conditional-replacement}
corrected_scores <- scores

corrected_scores[
  corrected_scores > 100
] <- 100

corrected_scores
```

# 19. `switch()`

`switch()` is useful when selecting among named alternatives.

```{r switch-example}
output_format <- "html"

selected_description <- switch(
  output_format,
  html = "Web document",
  word = "Microsoft Word document",
  pdf = "PDF document",
  "Unknown format"
)

selected_description
```

# 20. Numeric `switch()`

```{r numeric-switch}
choice <- 2

switch(
  choice,
  "First option",
  "Second option",
  "Third option"
)
```

Named forms are generally clearer.

# 21. Conditional messages

```{r conditional-message}
file_path <- "data/student_scores.csv"

if (file.exists(file_path)) {
  message(
    "The student dataset is available."
  )
}
```

# 22. Conditional warnings

```{r conditional-warning}
scores <- c(
  65,
  72,
  105
)

if (any(scores > 100)) {
  warning(
    "At least one score exceeds 100.",
    call. = FALSE
  )
}
```

# 23. Stopping execution

```{r stop-on-invalid-score, error=TRUE}
invalid_scores <- c(
  65,
  72,
  -5
)

if (any(
  invalid_scores < 0 |
    invalid_scores > 100
)) {
  stop(
    "Scores must lie between 0 and 100.",
    call. = FALSE
  )
}
```

`stop()` is appropriate when continuing would invalidate the analysis.

# 24. Defensive file validation

```{r defensive-file-validation}
student_path <-
  "data/student_scores.csv"

if (!file.exists(student_path)) {
  stop(
    paste(
      "Required file not found:",
      student_path
    ),
    call. = FALSE
  )
}

students <- read.csv(
  student_path,
  stringsAsFactors = FALSE
)
```

# 25. Defensive variable validation

```{r defensive-variable-validation}
required_variables <- c(
  "student_id",
  "programme",
  "final_score"
)

missing_variables <- setdiff(
  required_variables,
  names(students)
)

if (length(missing_variables) > 0L) {
  stop(
    paste(
      "Missing variables:",
      paste(
        missing_variables,
        collapse = ", "
      )
    ),
    call. = FALSE
  )
}
```

# 26. Decision table thinking

Before writing code, state the decision rules.

| Condition | Result |
|---|---|
| score is missing | Missing |
| score at least 80 | Distinction |
| score at least 50 | Pass |
| otherwise | Fail |

Implementation:

```{r decision-table-implementation}
score <- 76

if (is.na(score)) {
  result <- "Missing"
} else if (score >= 80) {
  result <- "Distinction"
} else if (score >= 50) {
  result <- "Pass"
} else {
  result <- "Fail"
}

result
```

# 27. Applied example: procurement approval

Rule:

- value below GH₵10,000: departmental approval;
- GH₵10,000 through GH₵49,999: director approval;
- GH₵50,000 or more: tender committee review.

```{r procurement-approval}
procurement_value <- 42000

if (procurement_value < 0) {
  approval_level <- "Invalid value"
} else if (procurement_value < 10000) {
  approval_level <- "Departmental approval"
} else if (procurement_value < 50000) {
  approval_level <- "Director approval"
} else {
  approval_level <-
    "Tender committee review"
}

approval_level
```

# 28. Applied example: clinic triage

```{r clinic-triage}
waiting_time <- 55
severity <- "High"

if (
  severity == "High" ||
  waiting_time > 60
) {
  triage_status <- "Immediate attention"
} else if (waiting_time > 30) {
  triage_status <- "Priority queue"
} else {
  triage_status <- "Standard queue"
}

triage_status
```

# 29. Applied example: agricultural advisory

```{r agricultural-advisory}
rainfall_mm <- 72
soil_moisture <- "Low"

if (
  rainfall_mm < 50 &&
  soil_moisture == "Low"
) {
  advice <- "Irrigation strongly recommended"
} else if (
  rainfall_mm < 80
) {
  advice <- "Monitor soil moisture"
} else {
  advice <- "Rainfall currently adequate"
}

advice
```

# 30. Vectorised classification of imported data

```{r import-student-classification}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_data$result <- ifelse(
  student_data$final_score >= 50,
  "Pass",
  "Fail"
)

student_data$distinction <- ifelse(
  student_data$final_score >= 80,
  "Yes",
  "No"
)

head(student_data)
```

# 31. Safer multi-category vector classification

Use `cut()` when categories are based on intervals:

```{r cut-grade-classification}
student_data$grade <- cut(
  student_data$final_score,
  breaks = c(
    -Inf,
    49,
    59,
    69,
    79,
    Inf
  ),
  labels = c(
    "F",
    "D",
    "C",
    "B",
    "A"
  ),
  right = TRUE
)

head(
  student_data[
    c(
      "final_score",
      "grade"
    )
  ]
)
```

This is often clearer than deeply nested `ifelse()`.

# 32. Testing branches

A conditional statement should be tested with values that exercise every
branch.

```{r branch-test-values}
test_scores <- c(
  NA,
  45,
  65,
  85
)

classify_one_score <- function(score) {
  if (is.na(score)) {
    "Missing"
  } else if (score >= 80) {
    "Distinction"
  } else if (score >= 50) {
    "Pass"
  } else {
    "Fail"
  }
}

vapply(
  test_scores,
  classify_one_score,
  FUN.VALUE = character(1)
)
```

The function syntax will be treated fully in a later lesson. It is used here
only to demonstrate systematic branch testing.

# 33. Common mistakes

- supplying a vector condition to `if`;
- confusing `if` and `ifelse()`;
- placing broad conditions before specific conditions;
- failing to handle `NA`;
- using `&` instead of `&&` in scalar control flow;
- creating deeply nested code;
- continuing after invalid input;
- using `ifelse()` where direct indexing is clearer;
- allowing `ifelse()` to coerce output unexpectedly;
- failing to test every branch.

# 34. Independent exercises

1. Write a basic `if` statement.
2. Write an `if`-`else` pass rule.
3. Create an `else if` grading system.
4. demonstrate why branch order matters.
5. combine two scalar conditions.
6. handle a missing input.
7. classify a vector using `ifelse()`.
8. demonstrate type coercion in `ifelse()`.
9. use `switch()` for three named choices.
10. stop execution when a file is missing.
11. issue a warning for invalid values.
12. create decision rules for education, health or procurement.
13. classify imported student data.
14. test every branch of a rule.

# 35. Lesson assessment

Create a standalone R Markdown report that:

1. demonstrates `if`;
2. demonstrates `if` with `else`;
3. uses several `else if` branches;
4. handles a missing condition;
5. distinguishes `if` and `ifelse()`;
6. performs vectorised classification;
7. uses `switch()`;
8. validates a file and required variables;
9. issues a warning and an error condition appropriately;
10. applies a contextual decision rule; and
11. tests every possible branch.

# 36. Completion checklist

Confirm that you can:

- write an `if` statement;
- add an `else` branch;
- use `else if`;
- order conditions correctly;
- combine scalar conditions;
- handle missing conditions;
- distinguish `if` and `ifelse()`;
- classify vectors;
- recognise `ifelse()` coercion;
- use `switch()`;
- issue messages and warnings;
- stop invalid execution;
- validate files and variables;
- translate decision tables into code; and
- test all branches.

# Lesson conclusion

Conditional execution converts rules into program behaviour. Reliable
conditions require clear decision tables, correct branch ordering, explicit
missing-value handling and systematic testing.

\[
\boxed{
\text{state decision rule}
\rightarrow
\text{order branches}
\rightarrow
\text{handle missingness}
\rightarrow
\text{execute}
\rightarrow
\text{test every branch}
}
\]
