---
title: "Lesson 12: Repetition with for, while and repeat Loops"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Repetition is necessary when the same operation must be performed several
times. An analyst may need to:

- calculate a statistic for several variables;
- inspect every file in a folder;
- apply a rule to several programmes;
- generate repeated simulation results;
- create several plots;
- validate multiple datasets; or
- continue a process until a condition is met.

R supports repetition through vectorised operations, the apply family and
explicit loops. This lesson focuses on `for`, `while` and `repeat` loops while
also explaining when a loop is unnecessary.

The purpose is not merely to make code repeat. The learner must also understand
loop state, indexing, preallocation, stopping conditions, result storage and
error prevention.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain why repetition is used in programming;
2. distinguish vectorisation from explicit loops;
3. write a valid `for` loop;
4. iterate over positions, values and names;
5. store loop results safely;
6. preallocate output objects;
7. use nested loops;
8. use `next` and `break`;
9. write a valid `while` loop;
10. prevent infinite loops;
11. use `repeat` with an explicit stopping condition;
12. process files and data columns iteratively;
13. understand loop-created state;
14. compare loops with `lapply()`, `sapply()` and `vapply()`; and
15. debug repetitive code systematically.

# 1. Why repetition is needed

Suppose a learner has five scores:

```{r create-loop-scores}
scores <- c(
  65,
  72,
  81,
  59,
  90
)

scores
```

The square of each score could be written manually:

```{r manual-repetition}
scores[1] ^ 2
scores[2] ^ 2
scores[3] ^ 2
scores[4] ^ 2
scores[5] ^ 2
```

This is repetitive and difficult to maintain.

A vectorised solution is:

```{r vectorised-square}
scores ^ 2
```

A loop solution is:

```{r loop-square-display}
for (score in scores) {
  print(
    score ^ 2
  )
}
```

The vectorised form is preferable for this simple task. The loop is useful for
learning control flow and for more complex operations.

# 2. General structure of a `for` loop

```text
for (variable in sequence) {
  code_to_repeat
}
```

Example:

```{r basic-for-loop}
for (number in 1:5) {
  print(number)
}
```

During each iteration:

1. `number` receives the next value;
2. the body executes;
3. the process continues until the sequence is exhausted.

# 3. Iterating over values

```{r iterate-over-values}
programmes <- c(
  "Statistics",
  "Procurement",
  "Hospitality"
)

for (programme in programmes) {
  print(
    programme
  )
}
```

The loop variable contains the actual value.

# 4. Iterating over positions

```{r iterate-over-positions}
for (i in seq_along(scores)) {
  print(
    scores[i]
  )
}
```

`seq_along(scores)` safely creates positions from 1 through the vector length.

## 4.1 Why `seq_along()` is preferred

```{r empty-vector-sequence}
empty_vector <- numeric(0)

1:length(empty_vector)
seq_along(empty_vector)
```

`1:length(empty_vector)` produces an unintended sequence when the vector is
empty. `seq_along()` correctly returns an empty integer vector.

# 5. Iterating over names

```{r named-vector-loop}
programme_counts <- c(
  Statistics = 35,
  Procurement = 42,
  Hospitality = 28
)

for (programme_name in names(programme_counts)) {
  current_count <-
    programme_counts[
      programme_name
    ]

  print(
    paste(
      programme_name,
      current_count
    )
  )
}
```

# 6. Storing loop results

Printing inside a loop does not create a reusable result object.

Preallocate a vector:

```{r preallocate-loop-result}
squared_scores <- numeric(
  length(scores)
)

squared_scores
```

Fill it:

```{r store-loop-results}
for (i in seq_along(scores)) {
  squared_scores[i] <-
    scores[i] ^ 2
}

squared_scores
```

# 7. Why preallocation matters

A weak pattern repeatedly grows an object:

```{r growing-vector-example}
growing_result <- numeric(0)

for (score in scores) {
  growing_result <- c(
    growing_result,
    score ^ 2
  )
}

growing_result
```

This works, but repeated growth can be inefficient for large loops.

Preferred:

```{r preallocated-result-example}
efficient_result <- numeric(
  length(scores)
)

for (i in seq_along(scores)) {
  efficient_result[i] <-
    scores[i] ^ 2
}

efficient_result
```

# 8. Preallocating character and logical results

```{r character-preallocation}
result_labels <- character(
  length(scores)
)

for (i in seq_along(scores)) {
  result_labels[i] <- if (
    scores[i] >= 50
  ) {
    "Pass"
  } else {
    "Fail"
  }
}

result_labels
```

Logical output:

```{r logical-preallocation}
distinction_status <- logical(
  length(scores)
)

for (i in seq_along(scores)) {
  distinction_status[i] <-
    scores[i] >= 80
}

distinction_status
```

# 9. Looping over data-frame columns

Import student data:

```{r import-student-loop-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)
```

Identify numeric columns:

```{r identify-numeric-columns}
numeric_status <- vapply(
  student_data,
  is.numeric,
  FUN.VALUE = logical(1)
)

numeric_names <- names(
  student_data
)[
  numeric_status
]

numeric_names
```

Calculate means using a loop:

```{r loop-column-means}
column_means <- numeric(
  length(numeric_names)
)

names(column_means) <-
  numeric_names

for (variable_name in numeric_names) {
  column_means[
    variable_name
  ] <- mean(
    student_data[[
      variable_name
    ]],
    na.rm = TRUE
  )
}

column_means
```

# 10. Looping over rows

```{r row-loop-example}
student_totals <- numeric(
  nrow(student_data)
)

for (i in seq_len(
  nrow(student_data)
)) {
  student_totals[i] <-
    student_data$coursework_score[i] +
    student_data$examination_score[i]
}

student_totals
```

Compare with the stored final score:

```{r validate-row-loop}
all(
  student_totals ==
    student_data$final_score
)
```

# 11. `seq_len()` for row iteration

```{r seq-len-example}
seq_len(
  nrow(student_data)
)
```

`seq_len(n)` safely creates integers from 1 through `n`.

# 12. Conditional logic inside loops

```{r conditional-inside-loop}
performance_label <- character(
  length(scores)
)

for (i in seq_along(scores)) {
  if (scores[i] >= 80) {
    performance_label[i] <-
      "Distinction"
  } else if (scores[i] >= 50) {
    performance_label[i] <-
      "Pass"
  } else {
    performance_label[i] <-
      "Fail"
  }
}

performance_label
```

# 13. Using `next`

`next` skips the remainder of the current iteration.

```{r next-example}
values <- c(
  10,
  NA,
  20,
  30
)

observed_doubled <- numeric(0)

for (value in values) {
  if (is.na(value)) {
    next
  }

  observed_doubled <- c(
    observed_doubled,
    value * 2
  )
}

observed_doubled
```

# 14. Using `break`

`break` stops the loop completely.

```{r break-example}
for (value in c(
  10,
  20,
  -5,
  30
)) {
  if (value < 0) {
    print(
      "Negative value found. Stopping."
    )

    break
  }

  print(value)
}
```

# 15. Nested loops

```{r nested-loop-example}
rows <- 1:3
columns <- 1:2

for (row_index in rows) {
  for (column_index in columns) {
    print(
      paste(
        "Row",
        row_index,
        "Column",
        column_index
      )
    )
  }
}
```

Nested loops should be used carefully because the number of iterations grows
quickly.

# 16. Building a multiplication table

```{r multiplication-table-loop}
multiplication_table <- matrix(
  0,
  nrow = 5,
  ncol = 5
)

for (i in 1:5) {
  for (j in 1:5) {
    multiplication_table[
      i,
      j
    ] <- i * j
  }
}

multiplication_table
```

# 17. The `while` loop

General structure:

```text
while (condition) {
  repeated_code
}
```

Example:

```{r basic-while-loop}
counter <- 1

while (counter <= 5) {
  print(counter)

  counter <- counter + 1
}
```

The condition is tested before each iteration.

# 18. Avoiding infinite `while` loops

Incorrect:

```r
counter <- 1

while (counter <= 5) {
  print(counter)
}
```

The counter never changes, so the condition remains true.

Correct:

```{r safe-while-loop}
counter <- 1

while (counter <= 5) {
  print(counter)

  counter <- counter + 1
}
```

# 19. Accumulation with `while`

```{r while-accumulation}
total <- 0
current_value <- 1

while (current_value <= 10) {
  total <- total + current_value

  current_value <-
    current_value + 1
}

total
```

# 20. Convergence-style loop

```{r while-convergence}
estimate <- 100
target <- 10
iteration <- 0
maximum_iterations <- 100

while (
  abs(estimate - target) > 0.01 &&
  iteration < maximum_iterations
) {
  estimate <-
    (estimate + target) / 2

  iteration <-
    iteration + 1
}

estimate
iteration
```

A maximum-iteration safeguard helps prevent unintended infinite loops.

# 21. The `repeat` loop

`repeat` has no condition in its header.

```{r basic-repeat-loop}
counter <- 1

repeat {
  print(counter)

  counter <- counter + 1

  if (counter > 5) {
    break
  }
}
```

The loop must contain an explicit stopping condition.

# 22. Input-validation pattern with `repeat`

```{r repeat-validation-pattern}
values_to_check <- c(
  -2,
  0,
  5
)

position <- 1

repeat {
  current_value <-
    values_to_check[
      position
    ]

  if (current_value > 0) {
    valid_value <- current_value

    break
  }

  position <- position + 1
}

valid_value
```

# 23. Processing several files

List CSV files:

```{r list-csv-files-loop}
csv_files <- list.files(
  path = "data",
  pattern = "\\.csv$",
  full.names = TRUE
)

csv_files
```

Store dimensions:

```{r loop-file-dimensions}
file_dimensions <- vector(
  mode = "list",
  length = length(csv_files)
)

names(file_dimensions) <-
  basename(csv_files)

for (i in seq_along(csv_files)) {
  current_data <- read.csv(
    csv_files[i],
    stringsAsFactors = FALSE
  )

  file_dimensions[[i]] <-
    dim(current_data)
}

file_dimensions
```

# 24. Error handling inside loops

Use `tryCatch()` when one failure should not terminate every iteration.

```{r trycatch-loop-example}
file_results <- vector(
  mode = "list",
  length = length(csv_files)
)

names(file_results) <-
  basename(csv_files)

for (i in seq_along(csv_files)) {
  file_results[[i]] <- tryCatch(
    {
      current_data <- read.csv(
        csv_files[i],
        stringsAsFactors = FALSE
      )

      list(
        success = TRUE,
        rows = nrow(current_data),
        columns = ncol(current_data)
      )
    },
    error = function(error_object) {
      list(
        success = FALSE,
        message = conditionMessage(
          error_object
        )
      )
    }
  )
}

file_results
```

# 25. Looping over programmes

```{r programme-loop}
programme_names <- unique(
  student_data$programme
)

programme_summary <- data.frame(
  programme = programme_names,
  number_of_students = integer(
    length(programme_names)
  ),
  mean_final_score = numeric(
    length(programme_names)
  ),
  stringsAsFactors = FALSE
)

for (i in seq_along(
  programme_names
)) {
  current_programme <-
    programme_names[i]

  current_subset <- student_data[
    student_data$programme ==
      current_programme,
  ]

  programme_summary$number_of_students[i] <-
    nrow(current_subset)

  programme_summary$mean_final_score[i] <-
    mean(
      current_subset$final_score
    )
}

programme_summary
```

# 26. Vectorisation versus loops

Vectorised:

```{r vectorised-total}
coursework <-
  student_data$coursework_score

examination <-
  student_data$examination_score

vectorised_total <-
  coursework + examination
```

Loop:

```{r loop-total}
loop_total <- numeric(
  nrow(student_data)
)

for (i in seq_len(
  nrow(student_data)
)) {
  loop_total[i] <-
    coursework[i] +
    examination[i]
}
```

Compare:

```{r compare-vector-loop}
identical(
  vectorised_total,
  loop_total
)
```

Use vectorisation when it is clear and natural. Use loops when the operation
requires sequential state, complex branching or side effects.

# 27. Apply-family alternatives

```{r lapply-alternative}
lapply(
  student_data[
    numeric_names
  ],
  mean,
  na.rm = TRUE
)
```

```{r sapply-alternative}
sapply(
  student_data[
    numeric_names
  ],
  mean,
  na.rm = TRUE
)
```

```{r vapply-alternative}
vapply(
  student_data[
    numeric_names
  ],
  mean,
  na.rm = TRUE,
  FUN.VALUE = numeric(1)
)
```

# 28. Simulation loop

```{r simulation-loop}
number_of_simulations <- 100
simulated_means <- numeric(
  number_of_simulations
)

for (i in seq_len(
  number_of_simulations
)) {
  simulated_sample <- rnorm(
    n = 30,
    mean = 70,
    sd = 10
  )

  simulated_means[i] <-
    mean(simulated_sample)
}

summary(simulated_means)
```

# 29. Progress messages

```{r progress-message-loop}
for (i in 1:5) {
  message(
    "Processing iteration ",
    i,
    " of 5"
  )
}
```

For large loops, progress messages should not be printed excessively.

# 30. Common mistakes

- forgetting to preallocate;
- growing objects repeatedly;
- using `1:length(x)` for possibly empty vectors;
- failing to update a `while` condition;
- omitting `break` from `repeat`;
- confusing values and positions;
- overwriting loop results;
- using `print()` instead of storing output;
- nesting loops unnecessarily;
- hiding errors inside loops without recording them;
- using a loop where vectorisation is clearer.

# 31. Independent exercises

1. Print the numbers 1 through 10 with a `for` loop.
2. Iterate over programme names.
3. square a vector and store the result using preallocation.
4. classify scores with conditional logic inside a loop.
5. skip missing values using `next`.
6. stop when an invalid value is found using `break`.
7. create a nested-loop multiplication table.
8. write a safe `while` loop.
9. use `repeat` with a stopping condition.
10. calculate means for all numeric columns in a data frame.
11. process every CSV file in the data folder.
12. create programme summaries using a loop.
13. compare a loop with vectorised code.
14. simulate 500 sample means.

# 32. Lesson assessment

Create a standalone R Markdown report that:

1. demonstrates a basic `for` loop;
2. iterates over positions and names;
3. preallocates and stores results;
4. uses conditional logic in a loop;
5. demonstrates `next` and `break`;
6. creates a nested loop;
7. uses a safe `while` loop;
8. uses `repeat`;
9. processes all supplied CSV files;
10. creates a grouped summary with a loop;
11. compares loop and vectorised results; and
12. includes a simulation loop.

# 33. Completion checklist

Confirm that you can:

- write a `for` loop;
- iterate over values and positions;
- use `seq_along()` and `seq_len()`;
- preallocate results;
- store character and logical outputs;
- loop over rows and columns;
- use `next` and `break`;
- write nested loops;
- write safe `while` loops;
- use `repeat`;
- process files iteratively;
- handle loop errors;
- compare loops and vectorisation;
- use apply-family alternatives; and
- run a simulation loop.

# Lesson conclusion

Repetition is reliable only when the sequence, state, stopping condition and
result storage are all clear.

\[
\boxed{
\text{define sequence}
\rightarrow
\text{preallocate}
\rightarrow
\text{iterate}
\rightarrow
\text{store}
\rightarrow
\text{validate}
}
\]
