---
title: "Lesson 13: Writing and Using Functions"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

A function packages a reusable operation under a meaningful name. Functions
reduce duplication, clarify analytical intent and make code easier to test.

A well-designed function should:

- have one clear purpose;
- accept explicit inputs;
- validate important assumptions;
- return a predictable result;
- avoid unnecessary side effects;
- use meaningful argument names; and
- provide useful error messages.

This lesson moves from simple functions to functions that work with vectors and
data frames.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain the purpose of functions;
2. write a function using `function()`;
3. define formal arguments;
4. use default arguments;
5. return values explicitly and implicitly;
6. validate arguments;
7. write functions for vectors and data frames;
8. use `...` appropriately;
9. distinguish pure functions from side-effect functions;
10. document a function;
11. test functions with representative inputs;
12. apply functions repeatedly; and
13. debug common function errors.

# 1. Why functions are useful

Suppose the same pass percentage is calculated several times:

```{r repeated-pass-code}
scores_a <- c(
  45,
  58,
  72,
  84
)

scores_b <- c(
  65,
  71,
  49,
  90
)

mean(scores_a >= 50) * 100
mean(scores_b >= 50) * 100
```

Create a function:

```{r create-pass-function}
calculate_pass_percentage <- function(
  scores,
  pass_mark = 50
) {
  mean(
    scores >= pass_mark,
    na.rm = TRUE
  ) * 100
}
```

Use it:

```{r use-pass-function}
calculate_pass_percentage(
  scores_a
)

calculate_pass_percentage(
  scores_b
)
```

# 2. Function anatomy

```text
function_name <- function(arguments) {
  body
}
```

In:

```r
calculate_pass_percentage <- function(
  scores,
  pass_mark = 50
) {
  ...
}
```

- `calculate_pass_percentage` is the object name;
- `function` creates a function object;
- `scores` is a required argument;
- `pass_mark` has a default;
- braces contain the body.

# 3. A function with no arguments

```{r no-argument-function}
greet_learner <- function() {
  "Welcome to Learning R."
}

greet_learner()
```

The function returns the final expression.

# 4. One required argument

```{r one-argument-function}
square_value <- function(x) {
  x ^ 2
}

square_value(5)
square_value(-3)
```

# 5. Several arguments

```{r several-argument-function}
calculate_weighted_score <- function(
  coursework,
  examination,
  coursework_weight,
  examination_weight
) {
  coursework * coursework_weight +
    examination * examination_weight
}

calculate_weighted_score(
  coursework = 78,
  examination = 65,
  coursework_weight = 0.40,
  examination_weight = 0.60
)
```

# 6. Default arguments

```{r default-argument-function}
calculate_weighted_score <- function(
  coursework,
  examination,
  coursework_weight = 0.40,
  examination_weight = 0.60
) {
  coursework * coursework_weight +
    examination * examination_weight
}

calculate_weighted_score(
  coursework = 78,
  examination = 65
)
```

Override defaults:

```{r override-function-default}
calculate_weighted_score(
  coursework = 78,
  examination = 65,
  coursework_weight = 0.30,
  examination_weight = 0.70
)
```

# 7. Positional and named matching

```{r positional-function-call}
calculate_weighted_score(
  78,
  65
)
```

Named calls are clearer:

```{r named-function-call}
calculate_weighted_score(
  coursework = 78,
  examination = 65
)
```

# 8. Return values

The final evaluated expression is returned automatically.

```{r implicit-return}
add_values <- function(
  x,
  y
) {
  x + y
}

add_values(
  5,
  7
)
```

Explicit return:

```{r explicit-return}
add_values_explicit <- function(
  x,
  y
) {
  result <- x + y

  return(result)
}

add_values_explicit(
  5,
  7
)
```

# 9. Early return

```{r early-return-function}
classify_score <- function(
  score
) {
  if (is.na(score)) {
    return("Missing")
  }

  if (score >= 80) {
    return("Distinction")
  }

  if (score >= 50) {
    return("Pass")
  }

  "Fail"
}

classify_score(85)
classify_score(65)
classify_score(45)
classify_score(NA)
```

# 10. Argument validation

```{r validate-function-argument}
calculate_mean_score <- function(
  scores
) {
  if (!is.numeric(scores)) {
    stop(
      "`scores` must be numeric.",
      call. = FALSE
    )
  }

  mean(
    scores,
    na.rm = TRUE
  )
}

calculate_mean_score(
  c(
    65,
    72,
    81
  )
)
```

Invalid call:

```{r invalid-function-call, error=TRUE}
calculate_mean_score(
  c(
    "High",
    "Low"
  )
)
```

# 11. Range validation

```{r score-validation-function}
validate_scores <- function(
  scores
) {
  if (!is.numeric(scores)) {
    stop(
      "`scores` must be numeric.",
      call. = FALSE
    )
  }

  if (length(scores) == 0L) {
    stop(
      "`scores` must not be empty.",
      call. = FALSE
    )
  }

  if (!all(
    scores >= 0 &
      scores <= 100,
    na.rm = TRUE
  )) {
    stop(
      "Scores must lie between 0 and 100.",
      call. = FALSE
    )
  }

  TRUE
}

validate_scores(
  c(
    65,
    72,
    81
  )
)
```

# 12. Returning several values

Use a named list:

```{r multi-result-function}
summarise_scores <- function(
  scores
) {
  validate_scores(scores)

  list(
    number = length(scores),
    missing = sum(
      is.na(scores)
    ),
    mean = mean(
      scores,
      na.rm = TRUE
    ),
    median = median(
      scores,
      na.rm = TRUE
    ),
    minimum = min(
      scores,
      na.rm = TRUE
    ),
    maximum = max(
      scores,
      na.rm = TRUE
    )
  )
}

score_summary <- summarise_scores(
  c(
    65,
    72,
    NA,
    81
  )
)

score_summary
```

Extract:

```{r extract-function-result}
score_summary$mean
score_summary$missing
```

# 13. Returning a data frame

```{r data-frame-return-function}
summarise_numeric_vector <- function(
  x,
  variable_name = "x"
) {
  if (!is.numeric(x)) {
    stop(
      "`x` must be numeric.",
      call. = FALSE
    )
  }

  data.frame(
    variable = variable_name,
    observed = sum(
      !is.na(x)
    ),
    missing = sum(
      is.na(x)
    ),
    mean = mean(
      x,
      na.rm = TRUE
    ),
    median = median(
      x,
      na.rm = TRUE
    ),
    minimum = min(
      x,
      na.rm = TRUE
    ),
    maximum = max(
      x,
      na.rm = TRUE
    ),
    stringsAsFactors = FALSE
  )
}

summarise_numeric_vector(
  c(
    10,
    20,
    NA,
    30
  ),
  variable_name = "waiting_time"
)
```

# 14. Vectorised functions

```{r vectorised-function}
convert_to_percentage <- function(
  proportion
) {
  proportion * 100
}

convert_to_percentage(
  c(
    0.45,
    0.72,
    0.91
  )
)
```

The function works vector-wise because multiplication is vectorised.

# 15. Functions using conditions

```{r conditional-function}
assign_grade <- function(
  score
) {
  ifelse(
    is.na(score),
    NA_character_,
    ifelse(
      score >= 80,
      "A",
      ifelse(
        score >= 70,
        "B",
        ifelse(
          score >= 60,
          "C",
          ifelse(
            score >= 50,
            "D",
            "F"
          )
        )
      )
    )
  )
}

assign_grade(
  c(
    45,
    58,
    72,
    84,
    NA
  )
)
```

# 16. Functions for data frames

```{r data-frame-function}
add_result_columns <- function(
  data,
  score_variable = "final_score",
  pass_mark = 50,
  distinction_mark = 80
) {
  if (!is.data.frame(data)) {
    stop(
      "`data` must be a data frame.",
      call. = FALSE
    )
  }

  if (!score_variable %in%
      names(data)) {
    stop(
      paste(
        "Variable not found:",
        score_variable
      ),
      call. = FALSE
    )
  }

  score_values <-
    data[[
      score_variable
    ]]

  data$pass_status <- ifelse(
    score_values >= pass_mark,
    "Pass",
    "Fail"
  )

  data$distinction_status <- ifelse(
    score_values >= distinction_mark,
    "Yes",
    "No"
  )

  data
}
```

Use it:

```{r use-data-frame-function}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_results <- add_result_columns(
  student_data
)

head(student_results)
```

# 17. Dynamic variable selection

```{r dynamic-column-function}
calculate_variable_mean <- function(
  data,
  variable
) {
  if (!variable %in%
      names(data)) {
    stop(
      paste(
        "Variable not found:",
        variable
      ),
      call. = FALSE
    )
  }

  values <- data[[
    variable
  ]]

  if (!is.numeric(values)) {
    stop(
      paste(
        variable,
        "must be numeric."
      ),
      call. = FALSE
    )
  }

  mean(
    values,
    na.rm = TRUE
  )
}

calculate_variable_mean(
  student_data,
  "final_score"
)
```

# 18. The `...` argument

`...` allows additional arguments to be passed onward.

```{r dots-function}
calculate_mean <- function(
  x,
  ...
) {
  mean(
    x,
    ...
  )
}

calculate_mean(
  c(
    10,
    NA,
    20
  ),
  na.rm = TRUE
)
```

# 19. Function side effects

A pure function returns a value without altering external state.

```{r pure-function}
double_value <- function(x) {
  x * 2
}

double_value(5)
```

A side-effect function may print, write a file or alter graphics.

```{r side-effect-function}
announce_result <- function(
  message_text
) {
  message(
    message_text
  )

  invisible(NULL)
}

announce_result(
  "Analysis completed."
)
```

# 20. Invisible return values

```{r invisible-return-function}
store_and_report <- function(
  x
) {
  message(
    "Value processed."
  )

  invisible(x)
}

processed_value <- store_and_report(
  25
)

processed_value
```

# 21. Functions calling functions

```{r function-calling-function}
calculate_pass_count <- function(
  scores,
  pass_mark = 50
) {
  validate_scores(scores)

  sum(
    scores >= pass_mark,
    na.rm = TRUE
  )
}

calculate_pass_percentage <- function(
  scores,
  pass_mark = 50
) {
  pass_count <- calculate_pass_count(
    scores,
    pass_mark
  )

  observed_count <- sum(
    !is.na(scores)
  )

  pass_count /
    observed_count *
    100
}

calculate_pass_percentage(
  c(
    45,
    58,
    72,
    84
  )
)
```

# 22. Anonymous functions

```{r anonymous-function}
vapply(
  student_data[
    vapply(
      student_data,
      is.numeric,
      FUN.VALUE = logical(1)
    )
  ],
  function(x) {
    mean(
      x,
      na.rm = TRUE
    )
  },
  FUN.VALUE = numeric(1)
)
```

# 23. Applying a custom function

```{r apply-custom-function}
numeric_variables <- names(
  student_data
)[
  vapply(
    student_data,
    is.numeric,
    FUN.VALUE = logical(1)
  )
]

variable_summaries <- lapply(
  numeric_variables,
  function(variable_name) {
    summarise_numeric_vector(
      student_data[[
        variable_name
      ]],
      variable_name
    )
  }
)

variable_summaries
```

Combine:

```{r combine-function-results}
combined_summaries <- do.call(
  rbind,
  variable_summaries
)

combined_summaries
```

# 24. Function documentation

A simple documentation block:

```r
# Calculate the percentage of observed scores meeting a pass mark.
#
# Arguments:
# scores: Numeric vector of scores.
# pass_mark: Numeric pass threshold. Default is 50.
#
# Returns:
# A single numeric percentage.
```

Good documentation states:

- purpose;
- arguments;
- expected types;
- returned value;
- missing-value behaviour;
- important assumptions;
- examples.

# 25. Testing functions

Test typical input:

```{r test-typical-function-input}
calculate_pass_percentage(
  c(
    45,
    58,
    72,
    84
  )
)
```

Boundary input:

```{r test-boundary-function-input}
calculate_pass_percentage(
  c(
    49,
    50,
    51
  )
)
```

Missing input:

```{r test-missing-function-input}
calculate_pass_percentage(
  c(
    45,
    NA,
    72
  )
)
```

Invalid input:

```{r test-invalid-function-input, error=TRUE}
calculate_pass_percentage(
  c(
    "Pass",
    "Fail"
  )
)
```

# 26. `stopifnot()`

```{r stopifnot-function}
calculate_ratio <- function(
  numerator,
  denominator
) {
  stopifnot(
    is.numeric(numerator),
    is.numeric(denominator),
    length(numerator) == 1L,
    length(denominator) == 1L,
    denominator != 0
  )

  numerator / denominator
}

calculate_ratio(
  10,
  2
)
```

`stopifnot()` is concise, but custom `stop()` messages are often more helpful
for learners and users.

# 27. Naming functions

Good function names describe an action:

```text
calculate_mean_score
validate_student_data
create_summary_table
classify_waiting_time
import_monthly_sales
```

Avoid vague names such as:

```text
do_it
run
test
fun
process
```

# 28. Integrated example: clinic summary function

```{r clinic-summary-function}
summarise_clinic_waiting <- function(
  data,
  waiting_variable = "waiting_minutes",
  target_minutes = 30
) {
  if (!is.data.frame(data)) {
    stop(
      "`data` must be a data frame.",
      call. = FALSE
    )
  }

  if (!waiting_variable %in%
      names(data)) {
    stop(
      paste(
        "Variable not found:",
        waiting_variable
      ),
      call. = FALSE
    )
  }

  waiting_values <-
    data[[
      waiting_variable
    ]]

  if (!is.numeric(waiting_values)) {
    stop(
      "Waiting time must be numeric.",
      call. = FALSE
    )
  }

  data.frame(
    observed_visits = sum(
      !is.na(waiting_values)
    ),
    missing_visits = sum(
      is.na(waiting_values)
    ),
    mean_waiting_minutes = mean(
      waiting_values,
      na.rm = TRUE
    ),
    median_waiting_minutes = median(
      waiting_values,
      na.rm = TRUE
    ),
    percentage_meeting_target = mean(
      waiting_values <=
        target_minutes,
      na.rm = TRUE
    ) * 100
  )
}
```

Use it:

```{r use-clinic-summary-function}
clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)

summarise_clinic_waiting(
  clinic_data
)
```

# 29. Common mistakes

- forgetting parentheses when calling a function;
- confusing function creation and function execution;
- relying on global objects instead of arguments;
- omitting validation;
- returning inconsistent types;
- using vague argument names;
- hiding errors with defaults;
- using `...` without understanding forwarded arguments;
- creating unnecessary side effects;
- failing to test boundary and invalid inputs.

# 30. Independent exercises

1. Write a function with no arguments.
2. write a one-argument mathematical function.
3. create a function with default arguments.
4. use explicit and implicit returns.
5. validate a numeric argument.
6. return several results in a list.
7. return a summary data frame.
8. create a vectorised classification function.
9. create a function that modifies a data frame copy.
10. use dynamic variable selection.
11. pass arguments using `...`.
12. write a function that calls another function.
13. document a function.
14. test typical, boundary, missing and invalid inputs.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. defines at least five custom functions;
2. includes required and default arguments;
3. demonstrates explicit and implicit returns;
4. validates inputs;
5. returns a list and a data frame;
6. includes a vectorised function;
7. includes a data-frame function;
8. uses `...`;
9. documents every function;
10. tests representative and invalid inputs; and
11. applies a function to supplied data.

# 32. Completion checklist

Confirm that you can:

- define a function;
- call a function;
- use required and default arguments;
- use named arguments;
- return values;
- use early return;
- validate inputs;
- return lists and data frames;
- write vectorised functions;
- write data-frame functions;
- select variables dynamically;
- use `...`;
- distinguish return values and side effects;
- use anonymous functions;
- document functions; and
- test functions systematically.

# Lesson conclusion

Functions convert repeated analytical logic into reusable, testable components.

\[
\boxed{
\text{define purpose}
\rightarrow
\text{specify arguments}
\rightarrow
\text{validate}
\rightarrow
\text{compute}
\rightarrow
\text{return}
\rightarrow
\text{test}
}
\]
