---
title: "Lesson 14: Function Scope, Environments and Argument Behaviour"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Functions do not evaluate code in an empty space. They operate within
environments that determine where names are found and where assignments are
stored.

Understanding scope helps explain why:

- a function can use an object defined outside it;
- an object created inside a function may disappear afterwards;
- a local variable can hide a global variable;
- default arguments may depend on earlier arguments;
- `...` can pass arguments to another function;
- modifying an object inside a function usually does not alter the original;
- closures can remember values; and
- careless use of global assignment creates hidden state.

This lesson introduces lexical scoping, local and global environments, lazy
argument evaluation, closures and safe state management.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain an environment in R;
2. distinguish local and global scope;
3. understand lexical scoping;
4. inspect function environments;
5. explain name masking;
6. understand ordinary assignment inside functions;
7. recognise the risks of superassignment;
8. understand argument matching;
9. understand lazy evaluation;
10. define defaults based on earlier arguments;
11. use `missing()` and `match.arg()`;
12. create closures;
13. understand copy-on-modify behaviour conceptually;
14. avoid hidden global dependencies; and
15. design functions with predictable scope.

# 1. What is an environment?

An environment is a structure that stores bindings between names and objects
and has a reference to a parent environment.

The global environment is:

```{r global-environment}
globalenv()
```

List its current names:

```{r global-environment-names}
ls(
  envir = globalenv()
)
```

# 2. Function-local objects

```{r local-object-function}
create_local_value <- function() {
  local_value <- 25

  local_value
}

create_local_value()
```

After the function returns:

```{r local-object-does-not-exist}
exists(
  "local_value",
  envir = globalenv()
)
```

The local object is not created in the global environment.

# 3. Local and global names

```{r create-global-name}
value <- 100
```

Create a local name with the same spelling:

```{r local-name-masking}
use_local_value <- function() {
  value <- 25

  value
}

use_local_value()
value
```

The local assignment masks the global value inside the function.

# 4. Lexical scoping

```{r lexical-scope-example}
tax_rate <- 0.15

calculate_total_with_tax <- function(
  subtotal
) {
  subtotal +
    subtotal * tax_rate
}

calculate_total_with_tax(
  1000
)
```

The function finds `tax_rate` in the environment where the function was
defined.

This works, but it creates an external dependency.

Preferred:

```{r explicit-tax-argument}
calculate_total_with_tax <- function(
  subtotal,
  tax_rate = 0.15
) {
  subtotal +
    subtotal * tax_rate
}

calculate_total_with_tax(
  1000
)
```

The dependency is explicit.

# 5. Inspecting a function environment

```{r inspect-function-environment}
environment(
  calculate_total_with_tax
)
```

Most functions created at the Console or top level of a document are associated
with the global environment.

# 6. Environment search

```{r environment-parent}
parent.env(
  globalenv()
)
```

Inspect the search path:

```{r scope-search-path}
search()
```

R searches for names according to environment relationships and attached
packages.

# 7. Ordinary assignment inside functions

```{r ordinary-local-assignment}
global_counter <- 0

increment_locally <- function() {
  global_counter <- 1

  global_counter
}

increment_locally()
global_counter
```

The function creates a local `global_counter`; the global object remains zero.

# 8. Superassignment

```{r superassignment-example}
global_counter <- 0

increment_globally <- function() {
  global_counter <<-
    global_counter + 1
}

increment_globally()
global_counter
```

`<<-` searches parent environments and modifies an existing binding or creates
one outside the local function.

This creates side effects and should be used sparingly.

# 9. Why superassignment is risky

```{r superassignment-hidden-state}
analysis_total <- 0

add_to_analysis_total <- function(
  value
) {
  analysis_total <<-
    analysis_total + value
}

add_to_analysis_total(10)
add_to_analysis_total(25)

analysis_total
```

The result depends on previous calls and hidden external state.

Preferred pure form:

```{r pure-accumulation-function}
add_values <- function(
  current_total,
  value
) {
  current_total + value
}

analysis_total <- 0
analysis_total <- add_values(
  analysis_total,
  10
)
analysis_total <- add_values(
  analysis_total,
  25
)

analysis_total
```

# 10. Argument matching

R can match arguments by:

- exact name;
- partial name;
- position.

```{r exact-argument-matching}
round(
  x = 3.14159,
  digits = 2
)
```

Positional:

```{r positional-argument-matching}
round(
  3.14159,
  2
)
```

Partial matching may work:

```{r partial-argument-matching}
round(
  3.14159,
  dig = 2
)
```

Avoid relying on partial matching because it can be ambiguous or fragile.

# 11. Required and optional arguments

```{r required-optional-function}
calculate_growth <- function(
  initial_value,
  final_value,
  periods = 1
) {
  (
    final_value /
      initial_value
  ) ^ (
    1 / periods
  ) - 1
}

calculate_growth(
  initial_value = 100,
  final_value = 121,
  periods = 2
)
```

# 12. Defaults depending on earlier arguments

```{r dependent-default-argument}
centre_values <- function(
  x,
  centre = mean(
    x,
    na.rm = TRUE
  )
) {
  x - centre
}

centre_values(
  c(
    10,
    20,
    30
  )
)
```

The default `centre` is calculated from `x`.

# 13. Lazy evaluation

Arguments are generally evaluated only when used.

```{r lazy-evaluation-example}
ignore_second_argument <- function(
  x,
  y
) {
  x
}

ignore_second_argument(
  x = 10,
  y = stop(
    "This argument was evaluated."
  )
)
```

No error occurs because `y` is never used.

# 14. Forcing evaluation

```{r force-argument-example}
force_argument <- function(
  x
) {
  force(x)

  function() {
    x
  }
}

remembered_value <-
  force_argument(25)

remembered_value()
```

`force()` ensures evaluation at a specific point.

# 15. Detecting omitted arguments

```{r missing-argument-function}
describe_input <- function(
  x
) {
  if (missing(x)) {
    return(
      "No input supplied."
    )
  }

  paste(
    "Input:",
    x
  )
}

describe_input()
describe_input(25)
```

# 16. Matching choices with `match.arg()`

```{r match-arg-function}
summarise_value <- function(
  x,
  statistic = c(
    "mean",
    "median",
    "sum"
  )
) {
  statistic <- match.arg(
    statistic
  )

  switch(
    statistic,
    mean = mean(
      x,
      na.rm = TRUE
    ),
    median = median(
      x,
      na.rm = TRUE
    ),
    sum = sum(
      x,
      na.rm = TRUE
    )
  )
}

summarise_value(
  c(
    10,
    20,
    30
  ),
  statistic = "median"
)
```

`match.arg()` validates and standardises a character choice.

# 17. The `...` argument and forwarding

```{r forwarding-dots}
calculate_summary <- function(
  x,
  summary_function,
  ...
) {
  summary_function(
    x,
    ...
  )
}

calculate_summary(
  c(
    10,
    NA,
    20
  ),
  summary_function = mean,
  na.rm = TRUE
)
```

# 18. Capturing `...`

```{r capture-dots}
inspect_dots <- function(
  ...
) {
  list(
    ...
  )
}

inspect_dots(
  first = 10,
  second = "A",
  third = TRUE
)
```

# 19. Function factories and closures

A function can create another function.

```{r multiplier-factory}
create_multiplier <- function(
  multiplier
) {
  function(x) {
    x * multiplier
  }
}

double_value <-
  create_multiplier(2)

triple_value <-
  create_multiplier(3)

double_value(10)
triple_value(10)
```

The returned functions remember `multiplier`. This remembered environment is a
closure.

# 20. Threshold-classification factory

```{r threshold-function-factory}
create_threshold_classifier <- function(
  threshold,
  yes_label = "Meets threshold",
  no_label = "Below threshold"
) {
  function(x) {
    ifelse(
      x >= threshold,
      yes_label,
      no_label
    )
  }
}

classify_pass <-
  create_threshold_classifier(
    threshold = 50,
    yes_label = "Pass",
    no_label = "Fail"
  )

classify_distinction <-
  create_threshold_classifier(
    threshold = 80,
    yes_label = "Distinction",
    no_label = "Not distinction"
  )

classify_pass(
  c(
    45,
    58,
    84
  )
)

classify_distinction(
  c(
    45,
    58,
    84
  )
)
```

# 21. Modifying function inputs

```{r modify-input-copy}
modify_first_value <- function(
  x
) {
  x[1] <- 999

  x
}

original_values <- c(
  10,
  20,
  30
)

modified_values <-
  modify_first_value(
    original_values
  )

original_values
modified_values
```

The original vector remains unchanged in ordinary use.

# 22. Data-frame input modification

```{r modify-data-frame-input}
add_total_score <- function(
  data
) {
  data$total_score <-
    data$coursework_score +
    data$examination_score

  data
}

student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_with_total <-
  add_total_score(
    student_data
  )

"total_score" %in%
  names(student_data)

"total_score" %in%
  names(student_with_total)
```

# 23. Hidden global dependency

Poor design:

```{r hidden-global-dependency}
pass_mark <- 50

calculate_pass_status <- function(
  scores
) {
  scores >= pass_mark
}

calculate_pass_status(
  c(
    45,
    58,
    72
  )
)
```

Preferred:

```{r explicit-function-dependency}
calculate_pass_status <- function(
  scores,
  pass_mark = 50
) {
  scores >= pass_mark
}

calculate_pass_status(
  c(
    45,
    58,
    72
  )
)
```

# 24. Name masking inside functions

```{r function-name-masking}
mean <- 100

calculate_average <- function(
  x
) {
  base::mean(x)
}

calculate_average(
  c(
    10,
    20,
    30
  )
)
```

Using `base::mean()` avoids confusion caused by an object named `mean`.

Remove the masking object:

```{r remove-masking-object}
rm(mean)
```

# 25. Local helper functions

```{r local-helper-function}
summarise_scores <- function(
  scores
) {
  validate_input <- function(
    x
  ) {
    if (!is.numeric(x)) {
      stop(
        "`scores` must be numeric.",
        call. = FALSE
      )
    }

    TRUE
  }

  validate_input(scores)

  c(
    Mean = mean(
      scores,
      na.rm = TRUE
    ),
    Median = median(
      scores,
      na.rm = TRUE
    )
  )
}

summarise_scores(
  c(
    65,
    72,
    81
  )
)
```

The helper function exists only during the outer function call.

# 26. Environment-based state

```{r environment-state-example}
counter_environment <-
  new.env(
    parent = emptyenv()
  )

counter_environment$count <- 0

increment_environment_counter <- function() {
  counter_environment$count <-
    counter_environment$count + 1

  counter_environment$count
}

increment_environment_counter()
increment_environment_counter()
```

Environment-based mutable state is advanced and should be used deliberately.

# 27. Reproducible random functions

```{r random-function}
simulate_scores <- function(
  n,
  mean_score = 70,
  standard_deviation = 10,
  seed = NULL
) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  rnorm(
    n = n,
    mean = mean_score,
    sd = standard_deviation
  )
}

first_simulation <- simulate_scores(
  n = 5,
  seed = 123
)

second_simulation <- simulate_scores(
  n = 5,
  seed = 123
)

first_simulation
second_simulation

identical(
  first_simulation,
  second_simulation
)
```

The seed argument makes reproducibility explicit.

# 28. Integrated example: configurable summary function

```{r configurable-summary-function}
summarise_data_variable <- function(
  data,
  variable,
  statistic = c(
    "mean",
    "median",
    "sum"
  ),
  na_rm = TRUE
) {
  statistic <- match.arg(
    statistic
  )

  if (!is.data.frame(data)) {
    stop(
      "`data` must be a data frame.",
      call. = FALSE
    )
  }

  if (!variable %in%
      names(data)) {
    stop(
      paste(
        "Variable not found:",
        variable
      ),
      call. = FALSE
    )
  }

  values <-
    data[[
      variable
    ]]

  if (!is.numeric(values)) {
    stop(
      paste(
        variable,
        "must be numeric."
      ),
      call. = FALSE
    )
  }

  switch(
    statistic,
    mean = mean(
      values,
      na.rm = na_rm
    ),
    median = median(
      values,
      na.rm = na_rm
    ),
    sum = sum(
      values,
      na.rm = na_rm
    )
  )
}
```

Use:

```{r use-configurable-summary}
summarise_data_variable(
  student_data,
  variable = "final_score",
  statistic = "mean"
)

summarise_data_variable(
  student_data,
  variable = "final_score",
  statistic = "median"
)
```

# 29. Common mistakes

- depending on undeclared global objects;
- using `<<-` unnecessarily;
- relying on partial argument matching;
- forgetting that arguments are lazily evaluated;
- using `...` without validation;
- creating accidental name masking;
- assuming local assignments modify global objects;
- assuming input objects are modified in place;
- failing to make randomness reproducible;
- creating stateful functions without documenting state.

# 30. Independent exercises

1. demonstrate local and global scope.
2. show lexical scoping with an external object.
3. rewrite the function to use an explicit argument.
4. demonstrate ordinary assignment and superassignment.
5. demonstrate positional, exact and partial matching.
6. create a default depending on an earlier argument.
7. demonstrate lazy evaluation.
8. use `missing()`.
9. use `match.arg()`.
10. forward arguments with `...`.
11. create a closure.
12. show that modifying a function input does not alter the original.
13. remove a hidden global dependency.
14. create a reproducible simulation function.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. demonstrates local and global objects;
2. explains lexical scoping;
3. compares `<-` and `<<-`;
4. demonstrates argument matching;
5. demonstrates lazy evaluation;
6. uses `missing()` and `match.arg()`;
7. forwards arguments with `...`;
8. creates a closure;
9. demonstrates safe input modification;
10. removes a global dependency;
11. creates a reproducible random function; and
12. explains all scope decisions.

# 32. Completion checklist

Confirm that you can:

- explain environments;
- distinguish local and global scope;
- explain lexical scoping;
- inspect a function environment;
- recognise name masking;
- explain ordinary local assignment;
- explain superassignment risks;
- understand argument matching;
- explain lazy evaluation;
- define dependent defaults;
- detect omitted arguments;
- validate choices with `match.arg()`;
- forward and capture `...`;
- create closures;
- understand ordinary copy-on-modify behaviour;
- avoid hidden dependencies; and
- design reproducible random functions.

# Lesson conclusion

Function behaviour depends not only on the code inside the body but also on the
environment in which names are resolved and arguments are evaluated.

\[
\boxed{
\text{make dependencies explicit}
\rightarrow
\text{limit side effects}
\rightarrow
\text{validate arguments}
\rightarrow
\text{return predictably}
\rightarrow
\text{test scope}
}
\]
