---
title: "Lesson 15: Importing CSV, Text and Delimited Files"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Most real analyses begin with data stored outside R. Comma-separated values
files, tab-delimited files and other text-based formats are common because they
are portable, transparent and supported by many systems.

Importing data is not simply a matter of running `read.csv()`. A reliable
workflow must consider:

- file location;
- delimiter;
- header structure;
- decimal marks;
- quotation marks;
- missing-value codes;
- character encoding;
- variable names;
- column types;
- row counts;
- duplicated records;
- and post-import validation.

This lesson introduces base R import functions and the `readr` package where
available. It emphasises inspection and verification after import.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain common delimited-text formats;
2. distinguish CSV, TSV and custom-delimited files;
3. use `read.csv()`, `read.table()` and `read.delim()`;
4. specify headers, separators and missing-value codes;
5. import character data without unintended conversion;
6. inspect imported column types;
7. handle quotation marks and comments;
8. work with decimal commas and custom delimiters;
9. use relative paths safely;
10. check file existence before import;
11. use `readr` functions when installed;
12. compare base R and `readr` behaviour;
13. diagnose common import failures;
14. validate imported records; and
15. export cleaned delimited files reproducibly.

# 1. Delimited text files

A delimited file stores one record per row and separates fields using a
specific character.

Common delimiters include:

| Format | Typical delimiter |
|---|---|
| CSV | comma |
| TSV | tab |
| Semicolon-delimited | semicolon |
| Pipe-delimited | vertical bar |
| Space-delimited | one or more spaces |

A CSV example:

```text
student_id,programme,final_score
S001,Statistics,72
S002,Procurement,68
S003,Hospitality,81
```

# 2. Confirming the file path

```{r confirm-student-file}
student_path <- "data/student_scores.csv"

file.exists(
  student_path
)
```

Stop if the file is missing:

```{r stop-if-file-missing}
if (!file.exists(student_path)) {
  stop(
    paste(
      "Required file not found:",
      student_path
    ),
    call. = FALSE
  )
}
```

# 3. Importing CSV with `read.csv()`

```{r import-csv-basic}
student_data <- read.csv(
  student_path,
  stringsAsFactors = FALSE
)

head(student_data)
```

Inspect:

```{r inspect-imported-csv}
str(student_data)
dim(student_data)
names(student_data)
summary(student_data)
```

# 4. `read.csv()` and `read.table()`

`read.csv()` is a convenience wrapper around `read.table()`.

Equivalent form:

```{r read-table-equivalent}
student_data_table <- read.table(
  file = student_path,
  header = TRUE,
  sep = ",",
  stringsAsFactors = FALSE
)

head(student_data_table)
```

Compare:

```{r compare-import-methods}
identical(
  student_data,
  student_data_table
)
```

# 5. Importing tab-delimited files

Use `read.delim()`:

```{r read-delim-example, eval=FALSE}
tab_data <- read.delim(
  "data/example_data.tsv",
  stringsAsFactors = FALSE
)
```

Equivalent:

```{r read-table-tab-example, eval=FALSE}
tab_data <- read.table(
  "data/example_data.tsv",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)
```

# 6. Custom delimiters

Semicolon-delimited:

```{r semicolon-import-example, eval=FALSE}
semicolon_data <- read.table(
  "data/example_semicolon.txt",
  header = TRUE,
  sep = ";",
  stringsAsFactors = FALSE
)
```

Pipe-delimited:

```{r pipe-import-example, eval=FALSE}
pipe_data <- read.table(
  "data/example_pipe.txt",
  header = TRUE,
  sep = "|",
  stringsAsFactors = FALSE
)
```

# 7. Header control

When the first row contains variable names:

```r
header = TRUE
```

When no header is present:

```{r no-header-example, eval=FALSE}
raw_data <- read.table(
  "data/no_header.txt",
  header = FALSE,
  sep = ",",
  stringsAsFactors = FALSE
)

names(raw_data) <- c(
  "student_id",
  "programme",
  "final_score"
)
```

# 8. Missing-value codes

Real files may use:

```text
NA
N/A
.
-
999
Missing
```

Specify recognised missing codes:

```{r missing-codes-example, eval=FALSE}
survey_data <- read.csv(
  "data/survey.csv",
  na.strings = c(
    "",
    "NA",
    "N/A",
    ".",
    "Missing"
  ),
  stringsAsFactors = FALSE
)
```

Do not treat legitimate numeric values such as 999 as missing unless the
codebook explicitly says so.

# 9. Character columns and factors

Historically, older R versions converted character columns to factors by
default. Use:

```r
stringsAsFactors = FALSE
```

for predictable character import.

```{r inspect-character-columns}
vapply(
  student_data,
  class,
  FUN.VALUE = character(1)
)
```

# 10. Quotation marks

CSV files may use quotation marks to protect values containing delimiters.

Example:

```text
institution,location
"Bolgatanga Technical University","Bolgatanga, Ghana"
```

Base R handles standard quotes automatically.

Control quotation characters with:

```{r quote-argument-example, eval=FALSE}
quoted_data <- read.table(
  "data/quoted_data.csv",
  header = TRUE,
  sep = ",",
  quote = "\"",
  stringsAsFactors = FALSE
)
```

# 11. Comments

`read.table()` may treat `#` as a comment character.

Disable comment interpretation:

```{r comment-character-example, eval=FALSE}
data_with_hash <- read.table(
  "data/hash_values.txt",
  header = TRUE,
  sep = ",",
  comment.char = "",
  stringsAsFactors = FALSE
)
```

# 12. White space

Use:

```r
strip.white = TRUE
```

to remove leading and trailing white space in unquoted character fields when
using `read.table()`.

```{r strip-white-example, eval=FALSE}
trimmed_data <- read.table(
  "data/spaced_data.csv",
  header = TRUE,
  sep = ",",
  strip.white = TRUE,
  stringsAsFactors = FALSE
)
```

# 13. Decimal marks

Some files use commas as decimal marks and semicolons as field separators.

```{r decimal-comma-example, eval=FALSE}
decimal_comma_data <- read.table(
  "data/decimal_comma.csv",
  header = TRUE,
  sep = ";",
  dec = ",",
  stringsAsFactors = FALSE
)
```

# 14. Skipping lines

```{r skip-lines-example, eval=FALSE}
data_after_notes <- read.table(
  "data/file_with_notes.txt",
  header = TRUE,
  sep = ",",
  skip = 3,
  stringsAsFactors = FALSE
)
```

This skips the first three lines.

# 15. Reading a limited number of rows

```{r limited-row-import}
student_preview <- read.csv(
  student_path,
  nrows = 5,
  stringsAsFactors = FALSE
)

student_preview
```

This is useful for inspecting large files.

# 16. Selecting columns during import

Base R can control classes:

```{r col-classes-example, eval=FALSE}
selected_data <- read.csv(
  "data/large_file.csv",
  colClasses = c(
    "character",
    "NULL",
    "numeric",
    "numeric"
  ),
  stringsAsFactors = FALSE
)
```

`"NULL"` skips a column.

# 17. Encoding

Text encoding determines how characters are represented.

```{r encoding-example, eval=FALSE}
encoded_data <- read.csv(
  "data/encoded_file.csv",
  fileEncoding = "UTF-8",
  stringsAsFactors = FALSE
)
```

Common encodings include:

- UTF-8;
- Latin-1;
- Windows-1252.

Incorrect encoding may produce corrupted characters.

# 18. File connections

```{r file-connection-example}
connection <- file(
  student_path,
  open = "r"
)

close(
  connection
)
```

Connections are useful for advanced streaming and compressed-file workflows.

# 19. Reading compressed CSV files

Base R can read gzip-compressed files:

```{r gzip-import-example, eval=FALSE}
compressed_data <- read.csv(
  gzfile(
    "data/large_data.csv.gz"
  ),
  stringsAsFactors = FALSE
)
```

# 20. `readr` package

Check availability:

```{r check-readr}
requireNamespace(
  "readr",
  quietly = TRUE
)
```

Import CSV:

```{r readr-csv-example, eval=FALSE}
student_readr <- readr::read_csv(
  student_path,
  show_col_types = FALSE
)
```

Import TSV:

```{r readr-tsv-example, eval=FALSE}
tab_data <- readr::read_tsv(
  "data/example_data.tsv",
  show_col_types = FALSE
)
```

Custom delimiter:

```{r readr-delim-example, eval=FALSE}
custom_data <- readr::read_delim(
  "data/example_pipe.txt",
  delim = "|",
  show_col_types = FALSE
)
```

# 21. Base R versus `readr`

Base R:

- returns a data frame;
- is always available;
- uses arguments such as `sep`, `dec` and `na.strings`.

`readr`:

- returns a tibble;
- reports column specifications;
- often imports large files faster;
- provides `problems()` for parsing diagnostics.

# 22. Parsing problems in `readr`

```{r readr-problems-example, eval=FALSE}
imported_data <- readr::read_csv(
  "data/example.csv",
  show_col_types = FALSE
)

readr::problems(
  imported_data
)
```

Parsing problems should be inspected rather than ignored.

# 23. Explicit `readr` column types

```{r readr-col-types-example, eval=FALSE}
typed_data <- readr::read_csv(
  "data/example.csv",
  col_types = readr::cols(
    student_id = readr::col_character(),
    programme = readr::col_character(),
    final_score = readr::col_double()
  )
)
```

Explicit types reduce ambiguity.

# 24. Post-import validation

Immediately after import, inspect:

```{r post-import-validation}
nrow(student_data)
ncol(student_data)
names(student_data)
str(student_data)
colSums(is.na(student_data))
```

Check required variables:

```{r validate-imported-columns}
required_columns <- c(
  "student_id",
  "programme",
  "coursework_score",
  "examination_score",
  "final_score"
)

all(
  required_columns %in%
    names(student_data)
)
```

Check unique identifiers:

```{r validate-imported-identifiers}
!anyDuplicated(
  student_data$student_id
)
```

# 25. Type correction after import

Suppose numeric values were imported as character:

```{r character-number-example}
raw_score <- c(
  "65",
  "72",
  "Missing",
  "81"
)
```

Convert carefully:

```{r character-number-conversion}
converted_score <- suppressWarnings(
  as.numeric(
    raw_score
  )
)

conversion_report <- data.frame(
  raw = raw_score,
  converted = converted_score,
  failed =
    is.na(converted_score) &
    !is.na(raw_score),
  stringsAsFactors = FALSE
)

conversion_report
```

# 26. Cleaning imported names

Base R may alter non-syntactic names.

```{r check-names-example}
raw_names <- c(
  "Student ID",
  "Final Score",
  "Programme Name"
)

make.names(
  raw_names
)
```

A simple custom cleaning approach:

```{r clean-names-function}
clean_names_base <- function(
  names_vector
) {
  cleaned <- tolower(
    names_vector
  )

  cleaned <- gsub(
    "[^a-z0-9]+",
    "_",
    cleaned
  )

  cleaned <- gsub(
    "^_|_$",
    "",
    cleaned
  )

  make.unique(
    cleaned
  )
}

clean_names_base(
  raw_names
)
```

# 27. Exporting CSV

```{r export-csv-example}
output_directory <- "lesson_15_output"

if (!dir.exists(
  output_directory
)) {
  dir.create(
    output_directory
  )
}

output_path <- file.path(
  output_directory,
  "student_scores_copy.csv"
)

write.csv(
  student_data,
  output_path,
  row.names = FALSE
)

file.exists(
  output_path
)
```

Verify:

```{r verify-exported-csv}
verified_export <- read.csv(
  output_path,
  stringsAsFactors = FALSE
)

identical(
  student_data,
  verified_export
)
```

Cleanup:

```{r cleanup-exported-csv}
unlink(
  output_directory,
  recursive = TRUE,
  force = TRUE
)
```

# 28. Exporting tab-delimited files

```{r export-tab-example, eval=FALSE}
write.table(
  student_data,
  "student_scores.tsv",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

# 29. Integrated example: rainfall import

```{r import-rainfall-delimited}
rainfall_data <- read.csv(
  "data/rainfall_records.csv",
  na.strings = c(
    "",
    "NA",
    "N/A",
    "."
  ),
  stringsAsFactors = FALSE
)

head(rainfall_data)
str(rainfall_data)
```

Validate:

```{r validate-rainfall-import}
nrow(rainfall_data)
names(rainfall_data)
colSums(is.na(rainfall_data))
is.numeric(rainfall_data$rainfall_mm)
all(
  rainfall_data$rainfall_mm >= 0,
  na.rm = TRUE
)
```

# 30. Common import failures

## 30.1 Wrong working directory

```text
cannot open file: No such file or directory
```

Check:

```r
getwd()
file.exists("data/file.csv")
```

## 30.2 Wrong delimiter

All values may appear in one column.

Inspect the first lines and specify the correct `sep`.

## 30.3 Wrong header setting

The first record may become variable names or generic names may be created.

## 30.4 Unexpected types

Numeric columns may become character because of text codes or symbols.

## 30.5 Encoding problems

Accented characters or symbols may display incorrectly.

## 30.6 Row-name column

Some exported files contain an unwanted first column of row numbers.

Prevent this with:

```r
row.names = FALSE
```

during export.

# 31. Independent exercises

1. Import `student_scores.csv` using `read.csv()`.
2. Re-import it using `read.table()`.
3. Confirm whether the objects are identical.
4. Import only the first five rows.
5. inspect column classes and missing values.
6. validate required variables and unique identifiers.
7. demonstrate custom missing-value codes.
8. demonstrate import code for tab, semicolon and pipe delimiters.
9. create a safe character-to-numeric conversion report.
10. clean a set of non-standard variable names.
11. export and verify a CSV copy.
12. import and validate rainfall data.
13. compare base R and `readr` import code.
14. explain five common import failures.

# 32. Lesson assessment

Create a standalone R Markdown report that:

1. explains delimited-file formats;
2. imports supplied CSV data;
3. validates file existence;
4. inspects dimensions, names and types;
5. checks missing values and duplicates;
6. demonstrates custom separators and missing codes;
7. demonstrates safe type conversion;
8. cleans variable names;
9. exports and verifies a file; and
10. compares base R with `readr`.

# 33. Completion checklist

Confirm that you can:

- identify common delimiters;
- check file paths;
- use `read.csv()`;
- use `read.table()`;
- use `read.delim()`;
- specify headers and separators;
- define missing-value codes;
- control character import;
- handle encoding;
- import limited rows;
- use `readr`;
- inspect parsing problems;
- validate imported data;
- correct types safely;
- clean names; and
- export reproducible delimited files.

# Lesson conclusion

Import is the first analytical quality-control stage. A successful file read is
not enough. The imported object must be inspected, validated and compared with
the expected data structure.

\[
\boxed{
\text{locate file}
\rightarrow
\text{specify format}
\rightarrow
\text{import}
\rightarrow
\text{inspect}
\rightarrow
\text{validate}
\rightarrow
\text{document}
}
\]
