---
title: "Lesson 16: Importing Excel, SPSS, Stata and SAS Files"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Many institutions store data in Excel workbooks or statistical-software
formats such as SPSS, Stata and SAS. These formats can preserve information
that plain text files do not, including:

- multiple worksheets;
- variable labels;
- value labels;
- date formats;
- missing-value definitions;
- and software-specific metadata.

Importing such files requires external packages. This lesson introduces
`readxl`, `haven` and selected alternatives. Because package availability and
file formats vary, the examples distinguish executable checks from template
code.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain why proprietary and workbook formats require special tools;
2. inspect package availability;
3. import Excel worksheets with `readxl`;
4. list sheets and choose ranges;
5. control column names and types;
6. import SPSS, Stata and SAS files with `haven`;
7. understand labelled variables;
8. convert labelled values appropriately;
9. preserve or remove metadata deliberately;
10. inspect imported dates and missing values;
11. diagnose common cross-software import problems;
12. validate imported proprietary data;
13. export compatible files where supported; and
14. document package and format dependencies.

# 1. Package requirements

Excel import commonly uses:

```text
readxl
openxlsx
writexl
```

SPSS, Stata and SAS import commonly uses:

```text
haven
foreign
```

Check packages:

```{r check-import-packages}
packages <- c(
  "readxl",
  "haven",
  "writexl",
  "openxlsx"
)

package_status <- vapply(
  packages,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
)

package_status
```

# 2. Installing packages

Run once when needed:

```{r install-import-packages, eval=FALSE}
install.packages(
  c(
    "readxl",
    "haven",
    "writexl",
    "openxlsx"
  )
)
```

Do not place package installation inside routine analysis scripts.

# 3. Excel workbooks

An Excel workbook may contain:

- several worksheets;
- title rows;
- merged cells;
- formulas;
- formatting;
- hidden rows or columns;
- notes above the actual table.

Import requires understanding the workbook structure.

# 4. Listing Excel sheets

```{r list-excel-sheets, eval=FALSE}
readxl::excel_sheets(
  "data/example_workbook.xlsx"
)
```

This should be done before assuming the worksheet name.

# 5. Importing the first worksheet

```{r import-first-excel-sheet, eval=FALSE}
excel_data <- readxl::read_excel(
  "data/example_workbook.xlsx"
)
```

# 6. Importing a named worksheet

```{r import-named-excel-sheet, eval=FALSE}
student_data <- readxl::read_excel(
  "data/example_workbook.xlsx",
  sheet = "Students"
)
```

# 7. Importing by sheet position

```{r import-excel-sheet-position, eval=FALSE}
student_data <- readxl::read_excel(
  "data/example_workbook.xlsx",
  sheet = 2
)
```

Named sheets are usually clearer than numeric positions.

# 8. Skipping title rows

```{r skip-excel-title-rows, eval=FALSE}
student_data <- readxl::read_excel(
  "data/example_workbook.xlsx",
  sheet = "Students",
  skip = 3
)
```

# 9. Importing a range

```{r import-excel-range, eval=FALSE}
student_data <- readxl::read_excel(
  "data/example_workbook.xlsx",
  sheet = "Students",
  range = "A4:F100"
)
```

Ranges can reduce irrelevant content but may exclude later records if the
workbook grows.

# 10. Missing values in Excel

```{r excel-missing-codes, eval=FALSE}
student_data <- readxl::read_excel(
  "data/example_workbook.xlsx",
  na = c(
    "",
    "NA",
    "N/A",
    "."
  )
)
```

# 11. Column names in Excel

```{r excel-column-names, eval=FALSE}
student_data <- readxl::read_excel(
  "data/example_workbook.xlsx",
  col_names = c(
    "student_id",
    "programme",
    "coursework_score",
    "examination_score",
    "final_score"
  )
)
```

Use only when the imported range does not include a header row.

# 12. Excel column types

```{r excel-column-types, eval=FALSE}
student_data <- readxl::read_excel(
  "data/example_workbook.xlsx",
  col_types = c(
    "text",
    "text",
    "numeric",
    "numeric",
    "numeric"
  )
)
```

Possible types include:

```text
skip
guess
logical
numeric
date
text
list
```

# 13. Mixed-type Excel columns

A column containing numbers and text may be guessed incorrectly.

Example:

```text
100
125
Missing
140
```

The column may become character. Import deliberately or clean after import.

# 14. Excel dates

Excel dates are stored internally as serial numbers but are normally converted
by `readxl` when detected.

Inspect:

```{r inspect-excel-date, eval=FALSE}
class(
  imported_data$visit_date
)

summary(
  imported_data$visit_date
)
```

Always verify date interpretation.

# 15. Writing Excel files

With `writexl`:

```{r write-excel-writexl, eval=FALSE}
writexl::write_xlsx(
  list(
    Students = student_data,
    Summary = summary_table
  ),
  path = "output/student_analysis.xlsx"
)
```

With `openxlsx`:

```{r write-excel-openxlsx, eval=FALSE}
openxlsx::write.xlsx(
  student_data,
  file = "output/student_data.xlsx",
  overwrite = TRUE
)
```

# 16. SPSS files

SPSS files commonly use the `.sav` extension.

```{r import-spss-haven, eval=FALSE}
spss_data <- haven::read_sav(
  "data/survey_data.sav"
)
```

Compressed SPSS files may use `.zsav`.

# 17. Stata files

Stata files commonly use `.dta`.

```{r import-stata-haven, eval=FALSE}
stata_data <- haven::read_dta(
  "data/survey_data.dta"
)
```

# 18. SAS files

SAS datasets commonly use `.sas7bdat`.

```{r import-sas-haven, eval=FALSE}
sas_data <- haven::read_sas(
  "data/survey_data.sas7bdat"
)
```

SAS transport files may use `.xpt`:

```{r import-sas-transport, eval=FALSE}
sas_transport <- haven::read_xpt(
  "data/survey_data.xpt"
)
```

# 19. Haven tibbles

`haven` functions return tibbles.

```{r inspect-haven-object, eval=FALSE}
class(
  spss_data
)

str(
  spss_data
)
```

Convert to base data frame:

```{r convert-haven-data-frame, eval=FALSE}
spss_base <- as.data.frame(
  spss_data
)
```

# 20. Variable labels

```{r inspect-variable-label, eval=FALSE}
attr(
  spss_data$age,
  "label"
)
```

Inspect all labels:

```{r inspect-all-variable-labels, eval=FALSE}
vapply(
  spss_data,
  function(x) {
    label <- attr(
      x,
      "label"
    )

    if (is.null(label)) {
      ""
    } else {
      label
    }
  },
  FUN.VALUE = character(1)
)
```

# 21. Labelled values

A labelled variable stores numeric codes with text labels.

```{r labelled-variable-example}
example_status <- haven::labelled(
  c(
    1,
    2,
    1,
    9
  ),
  labels = c(
    Yes = 1,
    No = 2,
    Missing = 9
  )
)

example_status
```

This chunk requires `haven`.

To keep the document knit-safe when `haven` is unavailable, use:

```{r labelled-variable-safe}
if (requireNamespace(
  "haven",
  quietly = TRUE
)) {
  example_status <- haven::labelled(
    c(
      1,
      2,
      1,
      9
    ),
    labels = c(
      Yes = 1,
      No = 2,
      Missing = 9
    )
  )

  print(
    example_status
  )
}
```

# 22. Converting labelled values to factors

```{r labelled-to-factor, eval=FALSE}
status_factor <- haven::as_factor(
  spss_data$status
)
```

For all labelled variables:

```{r all-labelled-to-factor, eval=FALSE}
factor_data <- haven::as_factor(
  spss_data
)
```

Do not convert labels automatically without considering whether numeric codes
are analytically meaningful.

# 23. Tagged missing values

Haven supports tagged missing values used by Stata and SAS.

```{r tagged-missing-example}
if (requireNamespace(
  "haven",
  quietly = TRUE
)) {
  tagged_values <- c(
    10,
    haven::tagged_na("a"),
    20,
    haven::tagged_na("b")
  )

  print(
    tagged_values
  )

  print(
    haven::is_tagged_na(
      tagged_values
    )
  )
}
```

Tagged missing values may distinguish reasons such as refusal or not
applicable.

# 24. User-defined missing values in SPSS

`read_sav()` may preserve user-defined missing-value metadata.

```{r spss-user-na-example, eval=FALSE}
spss_data <- haven::read_sav(
  "data/survey_data.sav",
  user_na = TRUE
)
```

These values require deliberate treatment.

# 25. Import with `foreign`

Older workflows may use:

```{r foreign-spss-example, eval=FALSE}
foreign::read.spss(
  "data/survey_data.sav",
  to.data.frame = TRUE
)
```

or:

```{r foreign-stata-example, eval=FALSE}
foreign::read.dta(
  "data/survey_data.dta"
)
```

`haven` is generally preferred for modern workflows because it preserves more
metadata and supports newer formats.

# 26. Writing Stata and SPSS files

```{r write-stata-example, eval=FALSE}
haven::write_dta(
  student_data,
  "output/student_data.dta"
)
```

```{r write-spss-example, eval=FALSE}
haven::write_sav(
  student_data,
  "output/student_data.sav"
)
```

# 27. Writing SAS transport files

```{r write-xpt-example, eval=FALSE}
haven::write_xpt(
  student_data,
  "output/student_data.xpt"
)
```

# 28. Metadata inspection

A systematic metadata report:

```{r metadata-report-template}
inspect_metadata <- function(
  data
) {
  data.frame(
    variable = names(data),
    class = vapply(
      data,
      function(x) {
        paste(
          class(x),
          collapse = ", "
        )
      },
      FUN.VALUE = character(1)
    ),
    type = vapply(
      data,
      typeof,
      FUN.VALUE = character(1)
    ),
    missing = vapply(
      data,
      function(x) {
        sum(
          is.na(x)
        )
      },
      FUN.VALUE = integer(1)
    ),
    stringsAsFactors = FALSE
  )
}

inspect_metadata(
  read.csv(
    "data/student_scores.csv",
    stringsAsFactors = FALSE
  )
)
```

# 29. Cross-software type problems

Common issues include:

- numeric identifiers imported as numbers;
- dates imported as numbers or text;
- labelled categories treated as continuous values;
- missing codes preserved as ordinary values;
- long variable names shortened;
- text encoding altered;
- value labels lost during export;
- unsupported software-version features.

# 30. Validation after proprietary import

After import, check:

```{r proprietary-validation-template, eval=FALSE}
dim(imported_data)
names(imported_data)
str(imported_data)
summary(imported_data)
colSums(is.na(imported_data))
```

Check labels:

```{r proprietary-label-check, eval=FALSE}
lapply(
  imported_data,
  attributes
)
```

Check key variables:

```{r proprietary-key-check, eval=FALSE}
!anyDuplicated(
  imported_data$respondent_id
)
```

# 31. Preserving raw imported data

Recommended workflow:

```r
raw_spss_data <- haven::read_sav(
  "data/survey_data.sav"
)

analysis_spss_data <- raw_spss_data
```

Do not overwrite the raw import with cleaned values.

# 32. Package-version documentation

```{r package-version-documentation}
installed_packages <- c(
  readxl = requireNamespace(
    "readxl",
    quietly = TRUE
  ),
  haven = requireNamespace(
    "haven",
    quietly = TRUE
  ),
  writexl = requireNamespace(
    "writexl",
    quietly = TRUE
  )
)

installed_packages
```

When installed:

```{r report-package-versions}
available_packages <- names(
  installed_packages
)[
  installed_packages
]

if (length(
  available_packages
) > 0L) {
  vapply(
    available_packages,
    function(package_name) {
      as.character(
        packageVersion(
          package_name
        )
      )
    },
    FUN.VALUE = character(1)
  )
}
```

# 33. Common mistakes

- assuming the first Excel sheet is correct;
- failing to inspect title rows and merged cells;
- guessing column types incorrectly;
- converting all labelled variables to factors automatically;
- treating tagged missing values as ordinary values;
- losing metadata during conversion or export;
- overwriting the raw imported object;
- using package functions without declaring the package;
- failing to document package versions;
- assuming all software versions are mutually compatible.

# 34. Independent exercises

1. check availability of `readxl`, `haven`, `writexl` and `openxlsx`.
2. write template code to list Excel sheets.
3. write template code to import a named sheet.
4. demonstrate skipping rows and selecting a range.
5. specify Excel missing values and column types.
6. write template code for SPSS, Stata and SAS import.
7. create a labelled variable with `haven`.
8. convert labelled values to factors.
9. demonstrate tagged missing values.
10. create a metadata-inspection function.
11. explain five cross-software type problems.
12. write validation code for an imported proprietary dataset.
13. demonstrate export code for Excel, Stata and SPSS.
14. document package versions.

# 35. Lesson assessment

Create a standalone R Markdown report that:

1. explains workbook and proprietary formats;
2. audits package availability;
3. provides correct Excel import code;
4. provides SPSS, Stata and SAS import code;
5. explains labelled values;
6. demonstrates tagged missing values where possible;
7. creates a metadata report;
8. describes post-import validation;
9. demonstrates export templates; and
10. documents package dependencies.

# 36. Completion checklist

Confirm that you can:

- identify required packages;
- list Excel sheets;
- import named sheets;
- skip rows and select ranges;
- control Excel column types;
- import SPSS, Stata and SAS data;
- inspect labelled variables;
- convert labels deliberately;
- recognise tagged missing values;
- preserve metadata;
- validate proprietary imports;
- preserve raw data;
- export compatible formats; and
- document package versions.

# Lesson conclusion

Proprietary formats may preserve valuable metadata, but they also introduce
software-specific assumptions. Reliable import requires deliberate package
selection, metadata inspection and post-import validation.

\[
\boxed{
\text{identify format}
\rightarrow
\text{load suitable package}
\rightarrow
\text{import}
\rightarrow
\text{inspect metadata}
\rightarrow
\text{convert deliberately}
\rightarrow
\text{validate}
}
\]
