---
title: "Lesson 17: Data Inspection, Validation and Quality Assurance"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Data quality assurance begins before statistical analysis. A dataset may load
successfully and still contain serious problems:

- duplicated identifiers;
- impossible values;
- inconsistent categories;
- missing records;
- contradictory variables;
- unexpected data types;
- invalid dates;
- arithmetic inconsistencies;
- extreme values caused by entry errors;
- and undocumented coding.

This lesson develops a systematic inspection and validation workflow. The aim
is not to repair every problem immediately. The first objective is to identify,
document and classify issues before making changes.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish inspection, validation and cleaning;
2. create a structural data profile;
3. inspect dimensions, names, classes and missingness;
4. validate identifiers;
5. detect duplicates;
6. validate ranges and categories;
7. check cross-variable consistency;
8. detect impossible and suspicious values;
9. create validation flags;
10. summarise data-quality issues;
11. write reusable validation functions;
12. preserve an audit trail;
13. distinguish errors from plausible outliers;
14. create a data-quality report; and
15. decide whether analysis should proceed.

# 1. Inspection, validation and cleaning

**Inspection** asks:

> What does the dataset contain?

**Validation** asks:

> Does the dataset satisfy expected rules?

**Cleaning** asks:

> How should identified problems be corrected or represented?

These are related but distinct stages.

# 2. Import and preserve raw data

```{r import-raw-student-data}
raw_student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

analysis_student_data <-
  raw_student_data
```

The raw object remains unchanged.

# 3. Structural inspection

```{r structural-inspection}
dim(
  raw_student_data
)

names(
  raw_student_data
)

str(
  raw_student_data
)

head(
  raw_student_data
)

tail(
  raw_student_data
)
```

# 4. Column profile

```{r column-profile}
column_profile <- data.frame(
  variable = names(
    raw_student_data
  ),
  class = vapply(
    raw_student_data,
    function(x) {
      paste(
        class(x),
        collapse = ", "
      )
    },
    FUN.VALUE = character(1)
  ),
  type = vapply(
    raw_student_data,
    typeof,
    FUN.VALUE = character(1)
  ),
  length = vapply(
    raw_student_data,
    length,
    FUN.VALUE = integer(1)
  ),
  missing = vapply(
    raw_student_data,
    function(x) {
      sum(
        is.na(x)
      )
    },
    FUN.VALUE = integer(1)
  ),
  unique = vapply(
    raw_student_data,
    function(x) {
      length(
        unique(x)
      )
    },
    FUN.VALUE = integer(1)
  ),
  stringsAsFactors = FALSE
)

column_profile
```

# 5. Missingness profile

```{r missingness-profile}
missing_count <- colSums(
  is.na(
    raw_student_data
  )
)

missing_percentage <-
  missing_count /
  nrow(
    raw_student_data
  ) *
  100

missing_profile <- data.frame(
  variable = names(
    raw_student_data
  ),
  missing_count =
    as.integer(
      missing_count
    ),
  missing_percentage =
    as.numeric(
      missing_percentage
    ),
  stringsAsFactors = FALSE
)

missing_profile
```

# 6. Identifier validation

Check required identifier:

```{r identifier-existence}
"student_id" %in%
  names(
    raw_student_data
  )
```

Check missing identifiers:

```{r missing-identifiers}
sum(
  is.na(
    raw_student_data$student_id
  ) |
    raw_student_data$student_id == ""
)
```

Check duplicates:

```{r duplicate-identifiers}
duplicated_identifier_status <-
  duplicated(
    raw_student_data$student_id
  )

sum(
  duplicated_identifier_status
)

raw_student_data[
  duplicated_identifier_status,
]
```

# 7. All duplicate members

```{r all-duplicate-identifier-members}
all_duplicate_identifier_status <-
  duplicated(
    raw_student_data$student_id
  ) |
  duplicated(
    raw_student_data$student_id,
    fromLast = TRUE
  )

raw_student_data[
  all_duplicate_identifier_status,
]
```

# 8. Range validation

```{r score-range-validation}
coursework_is_valid <- (
  raw_student_data$coursework_score >= 0 &
    raw_student_data$coursework_score <= 40
)

examination_is_valid <- (
  raw_student_data$examination_score >= 0 &
    raw_student_data$examination_score <= 60
)

final_score_is_valid <- (
  raw_student_data$final_score >= 0 &
    raw_student_data$final_score <= 100
)

all(
  coursework_is_valid
)

all(
  examination_is_valid
)

all(
  final_score_is_valid
)
```

# 9. Extract invalid records

```{r extract-invalid-scores}
invalid_score_records <- raw_student_data[
  !coursework_is_valid |
    !examination_is_valid |
    !final_score_is_valid,
]

invalid_score_records
```

# 10. Cross-variable arithmetic validation

```{r arithmetic-consistency}
score_arithmetic_is_valid <-
  raw_student_data$final_score ==
  raw_student_data$coursework_score +
  raw_student_data$examination_score

all(
  score_arithmetic_is_valid
)

raw_student_data[
  !score_arithmetic_is_valid,
]
```

# 11. Category validation

```{r inspect-programme-categories}
sort(
  unique(
    raw_student_data$programme
  )
)
```

Define allowed categories:

```{r allowed-programmes}
allowed_programmes <- c(
  "Statistics",
  "Procurement",
  "Hospitality",
  "Agriculture"
)

programme_is_valid <-
  raw_student_data$programme %in%
  allowed_programmes

all(
  programme_is_valid
)
```

Extract invalid categories:

```{r invalid-programme-records}
raw_student_data[
  !programme_is_valid,
]
```

# 12. Whitespace and case inconsistencies

```{r category-standardisation-example}
messy_programmes <- c(
  "Statistics",
  " statistics ",
  "STATISTICS",
  "Procurement"
)

trimmed_programmes <- trimws(
  messy_programmes
)

standardised_programmes <- tools::toTitleCase(
  tolower(
    trimmed_programmes
  )
)

data.frame(
  raw = messy_programmes,
  trimmed = trimmed_programmes,
  standardised = standardised_programmes,
  stringsAsFactors = FALSE
)
```

# 13. Logical consistency

Suppose a pass-status variable is present:

```{r create-pass-status-for-validation}
analysis_student_data$pass_status <-
  analysis_student_data$final_score >= 50
```

Validate:

```{r validate-pass-status}
calculated_pass_status <-
  analysis_student_data$final_score >= 50

pass_status_is_consistent <-
  analysis_student_data$pass_status ==
  calculated_pass_status

all(
  pass_status_is_consistent
)
```

# 14. Date validation template

```{r date-validation-template}
visit_dates <- as.Date(
  c(
    "2026-01-10",
    "2026-02-15",
    "2027-01-01",
    NA
  )
)

future_date_status <-
  visit_dates >
  Sys.Date()

future_date_status
```

A future date may be invalid depending on the study context.

# 15. Plausibility versus validity

A value can be valid but unusual.

For example, age 95 may be:

- plausible in a general population study;
- impossible in a study limited to adolescents;
- suspicious in a university undergraduate dataset.

Validation rules must reflect the study design.

# 16. Outlier screening with the IQR rule

```{r iqr-outlier-screening}
scores <- raw_student_data$final_score

first_quartile <- quantile(
  scores,
  probs = 0.25,
  na.rm = TRUE
)

third_quartile <- quantile(
  scores,
  probs = 0.75,
  na.rm = TRUE
)

interquartile_range <-
  third_quartile -
  first_quartile

lower_fence <-
  first_quartile -
  1.5 *
  interquartile_range

upper_fence <-
  third_quartile +
  1.5 *
  interquartile_range

outlier_status <-
  scores <
  lower_fence |
  scores >
  upper_fence

raw_student_data[
  outlier_status,
]
```

An outlier flag does not prove that a value is erroneous.

# 17. Z-score screening

```{r z-score-screening}
score_z <- as.numeric(
  scale(
    scores
  )
)

extreme_z_status <-
  abs(
    score_z
  ) > 3

raw_student_data[
  extreme_z_status,
]
```

# 18. Row-level validation flags

```{r row-validation-flags}
validation_data <- raw_student_data

validation_data$invalid_coursework <-
  !coursework_is_valid

validation_data$invalid_examination <-
  !examination_is_valid

validation_data$invalid_final_score <-
  !final_score_is_valid

validation_data$invalid_arithmetic <-
  !score_arithmetic_is_valid

validation_data$invalid_programme <-
  !programme_is_valid

validation_data$duplicate_identifier <-
  all_duplicate_identifier_status

head(
  validation_data
)
```

# 19. Overall issue flag

```{r overall-issue-flag}
issue_columns <- c(
  "invalid_coursework",
  "invalid_examination",
  "invalid_final_score",
  "invalid_arithmetic",
  "invalid_programme",
  "duplicate_identifier"
)

validation_data$has_any_issue <-
  rowSums(
    validation_data[
      issue_columns
    ]
  ) > 0

table(
  validation_data$has_any_issue
)
```

# 20. Issue count per record

```{r issue-count-per-record}
validation_data$number_of_issues <-
  rowSums(
    validation_data[
      issue_columns
    ]
  )

validation_data[
  validation_data$number_of_issues > 0,
]
```

# 21. Issue summary table

```{r issue-summary-table}
issue_summary <- data.frame(
  issue = issue_columns,
  count = vapply(
    validation_data[
      issue_columns
    ],
    sum,
    na.rm = TRUE,
    FUN.VALUE = numeric(1)
  ),
  stringsAsFactors = FALSE
)

issue_summary$percentage <-
  issue_summary$count /
  nrow(
    validation_data
  ) *
  100

issue_summary
```

# 22. Reusable validation function

```{r validate-score-data-function}
validate_score_data <- function(
  data,
  id_variable = "student_id",
  coursework_variable =
    "coursework_score",
  examination_variable =
    "examination_score",
  final_variable =
    "final_score"
) {
  required_variables <- c(
    id_variable,
    coursework_variable,
    examination_variable,
    final_variable
  )

  missing_variables <- setdiff(
    required_variables,
    names(data)
  )

  if (length(
    missing_variables
  ) > 0L) {
    stop(
      paste(
        "Missing variables:",
        paste(
          missing_variables,
          collapse = ", "
        )
      ),
      call. = FALSE
    )
  }

  id_values <-
    data[[
      id_variable
    ]]

  coursework_values <-
    data[[
      coursework_variable
    ]]

  examination_values <-
    data[[
      examination_variable
    ]]

  final_values <-
    data[[
      final_variable
    ]]

  data.frame(
    missing_identifier =
      is.na(
        id_values
      ) |
      id_values == "",
    duplicate_identifier =
      duplicated(
        id_values
      ) |
      duplicated(
        id_values,
        fromLast = TRUE
      ),
    invalid_coursework =
      coursework_values < 0 |
      coursework_values > 40,
    invalid_examination =
      examination_values < 0 |
      examination_values > 60,
    invalid_final =
      final_values < 0 |
      final_values > 100,
    invalid_arithmetic =
      final_values !=
      coursework_values +
      examination_values
  )
}
```

Use:

```{r use-validation-function}
score_validation <- validate_score_data(
  raw_student_data
)

head(
  score_validation
)

colSums(
  score_validation
)
```

# 23. Assertions

```{r assertion-example}
stopifnot(
  is.data.frame(
    raw_student_data
  ),
  nrow(
    raw_student_data
  ) > 0L,
  "student_id" %in%
    names(
      raw_student_data
    )
)
```

For user-facing scripts, customised error messages are often clearer.

# 24. Quality gates

A quality gate decides whether analysis should continue.

```{r quality-gate}
critical_issue_count <- sum(
  validation_data$invalid_arithmetic |
    validation_data$invalid_final_score,
  na.rm = TRUE
)

if (critical_issue_count > 0L) {
  warning(
    paste(
      critical_issue_count,
      "critical score issue(s) detected."
    ),
    call. = FALSE
  )
}
```

A strict workflow may use `stop()` instead of `warning()`.

# 25. Audit trail

Create a correction log:

```{r correction-log}
correction_log <- data.frame(
  record_id = character(0),
  variable = character(0),
  old_value = character(0),
  new_value = character(0),
  reason = character(0),
  corrected_by = character(0),
  correction_date = as.Date(
    character(0)
  ),
  stringsAsFactors = FALSE
)

correction_log
```

Every manual correction should be documented.

# 26. Raw, flagged and cleaned objects

Recommended object sequence:

```text
raw_data
validation_flags
flagged_data
clean_data
analysis_data
```

Do not overwrite the only copy of the raw imported data.

# 27. Integrated clinic quality assessment

```{r import-clinic-validation}
clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)

str(
  clinic_data
)
```

Check identifiers:

```{r clinic-id-validation}
sum(
  is.na(
    clinic_data$patient_id
  ) |
    clinic_data$patient_id == ""
)

anyDuplicated(
  clinic_data$patient_id
)
```

Check ranges:

```{r clinic-range-validation}
age_is_valid <-
  clinic_data$age >= 0 &
  clinic_data$age <= 120

waiting_is_valid <-
  clinic_data$waiting_time >= 0

consultation_is_valid <-
  clinic_data$consultation_duration >= 0

all(
  age_is_valid,
  na.rm = TRUE
)

all(
  waiting_is_valid,
  na.rm = TRUE
)

all(
  consultation_is_valid,
  na.rm = TRUE
)
```

Check categories:

```{r clinic-category-validation}
sort(
  unique(
    clinic_data$sex
  )
)

sort(
  unique(
    clinic_data$visit_type
  )
)
```

# 28. Data-quality report function

```{r quality-report-function}
create_quality_report <- function(
  data
) {
  data.frame(
    variable = names(data),
    class = vapply(
      data,
      function(x) {
        paste(
          class(x),
          collapse = ", "
        )
      },
      FUN.VALUE = character(1)
    ),
    missing = vapply(
      data,
      function(x) {
        sum(
          is.na(x)
        )
      },
      FUN.VALUE = integer(1)
    ),
    missing_percentage = vapply(
      data,
      function(x) {
        mean(
          is.na(x)
        ) * 100
      },
      FUN.VALUE = numeric(1)
    ),
    unique_values = vapply(
      data,
      function(x) {
        length(
          unique(x)
        )
      },
      FUN.VALUE = integer(1)
    ),
    stringsAsFactors = FALSE
  )
}

create_quality_report(
  clinic_data
)
```

# 29. Common mistakes

- cleaning before documenting the original problem;
- treating all outliers as errors;
- validating without a codebook;
- using universal ranges without study context;
- checking only missing values;
- ignoring identifier duplication;
- failing to check cross-variable consistency;
- overwriting raw data;
- applying silent corrections;
- proceeding with critical errors unresolved.

# 30. Independent exercises

1. create a structural profile for a supplied dataset.
2. create a missingness report.
3. validate identifiers.
4. detect all duplicate members.
5. define and test range rules.
6. validate categories.
7. test arithmetic consistency.
8. create row-level issue flags.
9. create issue counts per record.
10. create an issue summary table.
11. write a reusable validation function.
12. create a quality gate.
13. design a correction log.
14. create a complete data-quality report for clinic data.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. preserves raw and analysis copies;
2. profiles structure and missingness;
3. validates identifiers;
4. checks duplicates;
5. checks ranges and categories;
6. checks cross-variable consistency;
7. screens for suspicious outliers;
8. creates row-level flags;
9. summarises issue counts;
10. implements a quality gate;
11. creates a correction log; and
12. states whether analysis should proceed.

# 32. Completion checklist

Confirm that you can:

- distinguish inspection, validation and cleaning;
- profile data structure;
- report missingness;
- validate identifiers;
- detect duplicates;
- validate ranges;
- validate categories;
- check cross-variable consistency;
- screen suspicious values;
- create validation flags;
- summarise issues;
- write reusable validators;
- preserve an audit trail;
- create quality gates; and
- decide whether analysis should proceed.

# Lesson conclusion

Data quality assurance is a structured analytical activity. It requires explicit
rules, transparent flags, documented corrections and a clear decision about
whether the data are fit for analysis.

\[
\boxed{
\text{inspect}
\rightarrow
\text{define rules}
\rightarrow
\text{flag issues}
\rightarrow
\text{summarise}
\rightarrow
\text{document}
\rightarrow
\text{decide}
}
\]
