---
title: "Lesson 18: Missing Values and Incomplete Data"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Missing data are common in surveys, administrative records, experiments and
routine monitoring systems. A missing value may arise because:

- a respondent refused to answer;
- a question was not applicable;
- a measurement was not taken;
- a value was lost during transfer;
- a field was left blank;
- an instrument failed; or
- a value was deliberately suppressed.

Missingness is not simply a software inconvenience. It can affect descriptive
statistics, sample size, uncertainty and the validity of conclusions.

This lesson develops practical methods for detecting, summarising, visualising
and managing incomplete data. It also distinguishes computational handling from
statistical treatment. The use of `na.rm = TRUE` may allow a calculation to run,
but it does not by itself solve a missing-data problem.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish missing values from zero, blanks and special codes;
2. detect missing values in vectors and data frames;
3. count missing and observed values;
4. calculate missingness percentages;
5. identify complete and incomplete records;
6. examine missingness by group;
7. understand propagation of missing values;
8. use `na.rm` appropriately;
9. replace special missing-value codes with `NA`;
10. create missingness indicators;
11. distinguish listwise and pairwise deletion;
12. understand basic missing-data mechanisms;
13. perform simple deterministic replacements cautiously;
14. document missing-data decisions; and
15. create a reproducible missing-data report.

# 1. What counts as missing?

R uses `NA` to represent a missing value.

```{r basic-missing-value}
x <- c(
  10,
  NA,
  20
)

x
```

Missing is not the same as zero:

```{r missing-versus-zero}
is.na(NA)
is.na(0)
```

Missing is also not the same as an empty character string:

```{r missing-versus-empty-string}
is.na("")
"" == ""
```

# 2. Detecting missing values

```{r detect-missing-vector}
scores <- c(
  65,
  NA,
  72,
  81,
  NA
)

is.na(scores)
```

Count missing values:

```{r count-missing-vector}
sum(
  is.na(scores)
)
```

Count observed values:

```{r count-observed-vector}
sum(
  !is.na(scores)
)
```

Check whether any value is missing:

```{r any-missing-vector}
anyNA(scores)
```

# 3. Locating missing positions

```{r locate-missing-positions}
which(
  is.na(scores)
)
```

Extract observed values:

```{r extract-observed-values}
scores[
  !is.na(scores)
]
```

# 4. Missing-value propagation

```{r missing-propagation}
sum(scores)
mean(scores)
min(scores)
max(scores)
```

By default, many functions return `NA` when missing values are present.

# 5. Removing missing values during a calculation

```{r remove-missing-in-summary}
sum(
  scores,
  na.rm = TRUE
)

mean(
  scores,
  na.rm = TRUE
)

min(
  scores,
  na.rm = TRUE
)

max(
  scores,
  na.rm = TRUE
)
```

`na.rm = TRUE` excludes missing values from that operation only.

# 6. Effective sample size

```{r effective-sample-size}
total_positions <- length(scores)

observed_positions <- sum(
  !is.na(scores)
)

missing_positions <- sum(
  is.na(scores)
)

total_positions
observed_positions
missing_positions
```

A summary statistic should often be reported with the observed sample size.

# 7. Missingness percentage

```{r vector-missing-percentage}
missing_percentage <-
  mean(
    is.na(scores)
  ) * 100

missing_percentage
```

# 8. Missing values in data frames

Import rainfall data:

```{r import-rainfall-missing}
rainfall_data <- read.csv(
  "data/rainfall_records.csv",
  stringsAsFactors = FALSE
)

head(rainfall_data)
```

Missing counts by variable:

```{r data-frame-missing-counts}
colSums(
  is.na(
    rainfall_data
  )
)
```

Missing percentages:

```{r data-frame-missing-percentages}
colMeans(
  is.na(
    rainfall_data
  )
) * 100
```

# 9. Missingness report

```{r missingness-report}
create_missingness_report <- function(
  data
) {
  data.frame(
    variable = names(data),
    total = nrow(data),
    observed = vapply(
      data,
      function(x) {
        sum(
          !is.na(x)
        )
      },
      FUN.VALUE = integer(1)
    ),
    missing = vapply(
      data,
      function(x) {
        sum(
          is.na(x)
        )
      },
      FUN.VALUE = integer(1)
    ),
    missing_percentage = vapply(
      data,
      function(x) {
        mean(
          is.na(x)
        ) * 100
      },
      FUN.VALUE = numeric(1)
    ),
    stringsAsFactors = FALSE
  )
}

create_missingness_report(
  rainfall_data
)
```

# 10. Complete and incomplete records

```{r complete-cases-rainfall}
complete_status <- complete.cases(
  rainfall_data
)

complete_status
```

Complete rows:

```{r extract-complete-rainfall}
complete_rainfall_data <- rainfall_data[
  complete_status,
]

complete_rainfall_data
```

Incomplete rows:

```{r extract-incomplete-rainfall}
incomplete_rainfall_data <- rainfall_data[
  !complete_status,
]

incomplete_rainfall_data
```

# 11. Missingness across selected variables

```{r selected-variable-completeness}
selected_complete <- complete.cases(
  rainfall_data[
    c(
      "rainfall_mm"
    )
  ]
)

rainfall_data[
  selected_complete,
]
```

A record may be complete for the variables required by one analysis but
incomplete for another.

# 12. Number of missing values per record

```{r row-missing-count}
rainfall_data$missing_count <-
  rowSums(
    is.na(
      rainfall_data
    )
  )

rainfall_data
```

# 13. Missingness indicator variables

```{r missing-indicator}
rainfall_data$rainfall_missing <-
  is.na(
    rainfall_data$rainfall_mm
  )

table(
  rainfall_data$rainfall_missing
)
```

An indicator may be useful for describing patterns of missingness.

# 14. Special missing-value codes

```{r special-missing-codes}
raw_values <- c(
  10,
  20,
  999,
  -99,
  30
)

raw_values
```

Assume the codebook states:

- `999` means not recorded;
- `-99` means not applicable.

Convert:

```{r replace-special-codes}
clean_values <- raw_values

clean_values[
  clean_values %in%
    c(
      999,
      -99
    )
] <- NA

clean_values
```

Never replace codes without a documented codebook.

# 15. Blank strings and whitespace

```{r blank-string-missing}
raw_text <- c(
  "Statistics",
  "",
  " ",
  NA,
  "Procurement"
)

raw_text
```

Standardise:

```{r standardise-blank-missing}
clean_text <- trimws(
  raw_text
)

clean_text[
  clean_text == ""
] <- NA

clean_text
```

# 16. Character missing codes

```{r character-missing-codes}
responses <- c(
  "Yes",
  "No",
  "Missing",
  "N/A",
  "."
)

responses[
  responses %in%
    c(
      "Missing",
      "N/A",
      "."
    )
] <- NA

responses
```

# 17. Grouped missingness

Import clinic data:

```{r import-clinic-missing}
clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)
```

Missing waiting-time percentage by sex:

```{r grouped-missingness}
missing_by_sex <- tapply(
  is.na(
    clinic_data$waiting_minutes
  ),
  clinic_data$sex,
  mean
) * 100

missing_by_sex
```

Missingness may differ across groups and should be investigated.

# 18. Cross-tabulation of missingness

```{r missingness-cross-tabulation}
table(
  clinic_data$sex,
  is.na(
    clinic_data$waiting_minutes
  ),
  useNA = "ifany"
)
```

# 19. Basic missing-data mechanisms

## 19.1 Missing completely at random

Missingness is unrelated to observed or unobserved values.

## 19.2 Missing at random

Missingness may depend on observed variables but not on the missing value after
conditioning on those observed variables.

## 19.3 Missing not at random

Missingness depends on the missing value itself or an unobserved cause.

These mechanisms are assumptions, not labels that R can determine
automatically.

# 20. Listwise deletion

Listwise deletion removes every record missing any variable required by the
analysis.

```{r listwise-deletion}
analysis_variables <- clinic_data[
  c(
    "age",
    "waiting_minutes",
    "consultation_minutes"
  )
]

complete_analysis_data <-
  analysis_variables[
    complete.cases(
      analysis_variables
    ),
  ]

nrow(
  analysis_variables
)

nrow(
  complete_analysis_data
)
```

# 21. Pairwise deletion

Some functions use available pairs for each calculation.

```{r pairwise-correlation-example}
cor(
  analysis_variables,
  use = "pairwise.complete.obs"
)
```

Pairwise deletion can lead to different effective sample sizes across
statistics.

# 22. Complete-case correlation

```{r complete-case-correlation}
cor(
  analysis_variables,
  use = "complete.obs"
)
```

# 23. Mean replacement

```{r mean-replacement-example}
waiting_values <-
  clinic_data$waiting_minutes

mean_waiting <- mean(
  waiting_values,
  na.rm = TRUE
)

mean_imputed_waiting <-
  waiting_values

mean_imputed_waiting[
  is.na(
    mean_imputed_waiting
  )
] <- mean_waiting

mean_imputed_waiting
```

Mean replacement is simple but can reduce variance and distort relationships.
It should not be treated as a generally preferred method.

# 24. Median replacement

```{r median-replacement-example}
median_waiting <- median(
  waiting_values,
  na.rm = TRUE
)

median_imputed_waiting <-
  waiting_values

median_imputed_waiting[
  is.na(
    median_imputed_waiting
  )
] <- median_waiting

median_imputed_waiting
```

# 25. Group-based replacement

```{r group-based-replacement}
group_medians <- tapply(
  clinic_data$waiting_minutes,
  clinic_data$sex,
  median,
  na.rm = TRUE
)

group_medians
```

Apply cautiously:

```{r apply-group-replacement}
group_imputed_waiting <-
  clinic_data$waiting_minutes

for (i in seq_len(
  nrow(clinic_data)
)) {
  if (is.na(
    group_imputed_waiting[i]
  )) {
    current_group <-
      clinic_data$sex[i]

    group_imputed_waiting[i] <-
      group_medians[
        current_group
      ]
  }
}

group_imputed_waiting
```

# 26. Comparing original and replacement summaries

```{r compare-imputation-summaries}
data.frame(
  dataset = c(
    "Observed only",
    "Mean replacement",
    "Median replacement",
    "Group replacement"
  ),
  mean = c(
    mean(
      waiting_values,
      na.rm = TRUE
    ),
    mean(
      mean_imputed_waiting
    ),
    mean(
      median_imputed_waiting
    ),
    mean(
      group_imputed_waiting
    )
  ),
  standard_deviation = c(
    sd(
      waiting_values,
      na.rm = TRUE
    ),
    sd(
      mean_imputed_waiting
    ),
    sd(
      median_imputed_waiting
    ),
    sd(
      group_imputed_waiting
    )
  ),
  stringsAsFactors = FALSE
)
```

# 27. Missing-data decision log

```{r missing-data-decision-log}
missing_data_log <- data.frame(
  variable = c(
    "rainfall_mm",
    "waiting_minutes"
  ),
  missing_count = c(
    sum(
      is.na(
        rainfall_data$rainfall_mm
      )
    ),
    sum(
      is.na(
        clinic_data$waiting_minutes
      )
    )
  ),
  decision = c(
    "Use observed values for descriptive summary",
    "Retain missing values pending model review"
  ),
  justification = c(
    "Low missingness and descriptive purpose",
    "Cause and pattern require investigation"
  ),
  stringsAsFactors = FALSE
)

missing_data_log
```

# 28. Visualising missingness with base R

```{r missingness-barplot}
missing_counts <- colSums(
  is.na(
    clinic_data
  )
)

barplot(
  missing_counts,
  main = "Missing Values by Variable",
  xlab = "Variable",
  ylab = "Number missing",
  las = 2
)
```

# 29. Optional packages

Useful packages include:

- `naniar`;
- `VIM`;
- `mice`;
- `Amelia`.

Check availability:

```{r check-missing-packages}
vapply(
  c(
    "naniar",
    "VIM",
    "mice",
    "Amelia"
  ),
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
)
```

Multiple imputation is an advanced statistical procedure and requires a later,
more specialised treatment.

# 30. Integrated missing-data audit

```{r integrated-missing-audit}
audit_missing_data <- function(
  data
) {
  report <- create_missingness_report(
    data
  )

  list(
    variable_report = report,
    complete_records = sum(
      complete.cases(data)
    ),
    incomplete_records = sum(
      !complete.cases(data)
    ),
    complete_percentage = mean(
      complete.cases(data)
    ) * 100,
    rows_with_two_or_more_missing =
      sum(
        rowSums(
          is.na(data)
        ) >= 2
      )
  )
}

audit_missing_data(
  clinic_data
)
```

# 31. Common mistakes

- treating zero as missing without evidence;
- treating special codes as valid values;
- using `x == NA`;
- applying `na.rm = TRUE` without reporting effective sample size;
- deleting records before examining missingness patterns;
- replacing missing values silently;
- assuming mean replacement solves missing-data bias;
- confusing listwise and pairwise deletion;
- ignoring group differences in missingness;
- failing to document decisions.

# 32. Independent exercises

1. create vectors containing missing values.
2. count and locate missing positions.
3. calculate observed sample sizes.
4. create a missingness report.
5. identify complete and incomplete rows.
6. calculate row-level missing counts.
7. replace documented special codes.
8. standardise blank strings to `NA`.
9. compare missingness across groups.
10. demonstrate listwise and pairwise deletion.
11. perform and critique simple replacement methods.
12. create a missing-data decision log.
13. visualise missing counts.
14. audit missingness in clinic data.

# 33. Lesson assessment

Create a standalone R Markdown report that:

1. distinguishes `NA`, zero, blanks and special codes;
2. reports variable-level and row-level missingness;
3. identifies complete and incomplete records;
4. compares missingness by group;
5. explains missing-data mechanisms;
6. demonstrates listwise and pairwise deletion;
7. demonstrates two simple replacement methods;
8. compares resulting summaries;
9. creates a missing-data decision log; and
10. states the limitations of the chosen approach.

# 34. Completion checklist

Confirm that you can:

- identify missing values;
- count observed and missing values;
- calculate missing percentages;
- create missingness reports;
- identify complete cases;
- create row missing counts;
- convert special codes;
- standardise blank text;
- compare missingness by group;
- explain missing-data mechanisms;
- distinguish listwise and pairwise deletion;
- perform simple replacement cautiously;
- document missing-data decisions; and
- create a reproducible audit.

# Lesson conclusion

Missing-data management requires more than making functions run. It requires
describing the pattern, considering the likely cause, choosing a defensible
method and documenting the effect on analysis.

\[
\boxed{
\text{detect}
\rightarrow
\text{describe}
\rightarrow
\text{investigate}
\rightarrow
\text{decide}
\rightarrow
\text{document}
}
\]
