---
title: "Lesson 19: Character Data, Text Cleaning and Regular Expressions"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Text data appear in names, addresses, categories, codes, responses, comments
and identifiers. Even apparently simple categories may contain:

- inconsistent capitalisation;
- leading or trailing spaces;
- punctuation differences;
- spelling variants;
- embedded numbers;
- duplicated spaces;
- missing-value codes;
- or multiple pieces of information in one field.

This lesson develops base R tools for character inspection, cleaning,
searching, extraction and replacement. It introduces regular expressions as a
formal language for describing text patterns.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. create and inspect character vectors;
2. measure string length;
3. change case;
4. trim and normalise whitespace;
5. combine and split text;
6. search text with literal and regular-expression patterns;
7. extract and replace matched content;
8. use anchors, character classes and quantifiers;
9. handle punctuation and special characters;
10. standardise categories;
11. validate identifiers;
12. extract numbers from text;
13. work with missing text values;
14. use `stringr` when available; and
15. create reproducible text-cleaning functions.

# 1. Character vectors

```{r create-character-vector}
programmes <- c(
  "Statistics",
  "Procurement",
  "Hospitality",
  "Agriculture"
)

programmes
typeof(programmes)
class(programmes)
```

# 2. String length

```{r string-length}
nchar(
  programmes
)
```

Missing values remain missing:

```{r string-length-missing}
nchar(
  c(
    "R",
    NA,
    ""
  )
)
```

# 3. Changing case

```{r change-case}
tolower(
  programmes
)

toupper(
  programmes
)
```

Title case:

```{r title-case}
tools::toTitleCase(
  tolower(
    c(
      "applied statistics",
      "PROCUREMENT MANAGEMENT"
    )
  )
)
```

# 4. Trimming whitespace

```{r trim-whitespace}
messy_text <- c(
  " Statistics",
  "Procurement ",
  "  Hospitality  "
)

trimws(
  messy_text
)
```

# 5. Repeated internal spaces

```{r collapse-repeated-spaces}
text_with_spaces <- c(
  "Bolgatanga   Technical University",
  "Applied  Statistics"
)

gsub(
  "[[:space:]]+",
  " ",
  text_with_spaces
)
```

# 6. Combining text

```{r paste-example}
paste(
  "Student",
  1:5
)
```

Specify separator:

```{r paste-separator}
paste(
  "Student",
  1:5,
  sep = "_"
)
```

Collapse a vector:

```{r paste-collapse}
paste(
  programmes,
  collapse = ", "
)
```

`paste0()` uses no separator:

```{r paste-zero}
paste0(
  "S",
  sprintf(
    "%03d",
    1:5
  )
)
```

# 7. Formatted identifiers

```{r formatted-identifiers}
student_ids <- sprintf(
  "S%04d",
  1:10
)

student_ids
```

# 8. Splitting text

```{r split-text}
full_names <- c(
  "Ama Mensah",
  "Kojo Asante",
  "Esi Ayisi"
)

strsplit(
  full_names,
  split = " "
)
```

Extract first names:

```{r extract-first-name}
name_parts <- strsplit(
  full_names,
  split = " "
)

first_names <- vapply(
  name_parts,
  `[`,
  FUN.VALUE = character(1),
  1
)

first_names
```

# 9. Substrings

```{r substring-extraction}
substr(
  student_ids,
  start = 1,
  stop = 1
)
```

Extract numeric part:

```{r identifier-number-part}
substr(
  student_ids,
  start = 2,
  stop = 5
)
```

# 10. Literal search with `grepl()`

```{r literal-search}
grepl(
  "Statistics",
  programmes,
  fixed = TRUE
)
```

`grepl()` returns logical values.

# 11. Locate matches with `grep()`

```{r locate-text-matches}
grep(
  "Statistics",
  programmes,
  fixed = TRUE
)
```

Return matched values:

```{r return-matched-values}
grep(
  "Statistics",
  programmes,
  fixed = TRUE,
  value = TRUE
)
```

# 12. Regular expressions

A regular expression describes a text pattern.

```{r basic-regex-search}
codes <- c(
  "S001",
  "S002",
  "P010",
  "X999"
)

grepl(
  "^S",
  codes
)
```

`^` means start of string.

# 13. Anchors

Start anchor:

```{r start-anchor}
grepl(
  "^S",
  codes
)
```

End anchor:

```{r end-anchor}
grepl(
  "9$",
  codes
)
```

Complete pattern:

```{r complete-anchor-pattern}
grepl(
  "^S[0-9]{3}$",
  codes
)
```

# 14. Character classes

Digits:

```{r digit-character-class}
grepl(
  "[0-9]",
  codes
)
```

Letters:

```{r letter-character-class}
grepl(
  "[A-Za-z]",
  codes
)
```

Predefined digit class:

```{r predefined-digit-class}
grepl(
  "\\d",
  codes,
  perl = TRUE
)
```

# 15. Quantifiers

| Quantifier | Meaning |
|---|---|
| `*` | zero or more |
| `+` | one or more |
| `?` | zero or one |
| `{n}` | exactly n |
| `{n,}` | at least n |
| `{n,m}` | between n and m |

Example:

```{r regex-quantifier}
grepl(
  "^[A-Z][0-9]{3}$",
  codes
)
```

# 16. Alternatives

```{r regex-alternatives}
grepl(
  "Statistics|Procurement",
  programmes
)
```

# 17. Dot wildcard

```{r dot-wildcard}
sample_codes <- c(
  "A1",
  "AB",
  "A-"
)

grepl(
  "^A.$",
  sample_codes
)
```

The dot matches any one character.

# 18. Escaping special characters

```{r escape-special-character}
amounts <- c(
  "GH₵100",
  "GH₵250",
  "USD50"
)

grepl(
  "GH₵",
  amounts,
  fixed = TRUE
)
```

Using `fixed = TRUE` is safest when searching literal punctuation.

# 19. Replacement with `sub()`

`sub()` replaces the first match.

```{r sub-replacement}
sub(
  "Statistics",
  "Applied Statistics",
  c(
    "Statistics",
    "Statistics Department"
  )
)
```

# 20. Replacement with `gsub()`

`gsub()` replaces all matches.

```{r gsub-replacement}
gsub(
  "[[:space:]]+",
  " ",
  "Bolgatanga   Technical   University"
)
```

# 21. Removing punctuation

```{r remove-punctuation}
raw_categories <- c(
  "Statistics.",
  "Procurement!",
  "Hospitality,"
)

gsub(
  "[[:punct:]]",
  "",
  raw_categories
)
```

# 22. Extracting numbers from text

```{r extract-numbers}
text_values <- c(
  "Age 25",
  "Score 78",
  "Room 104"
)

numeric_text <- gsub(
  "[^0-9.-]",
  "",
  text_values
)

numeric_values <- as.numeric(
  numeric_text
)

data.frame(
  original = text_values,
  extracted = numeric_text,
  numeric = numeric_values,
  stringsAsFactors = FALSE
)
```

# 23. Standardising categories

```{r standardise-categories}
messy_programmes <- c(
  "statistics",
  " Statistics ",
  "STATISTICS",
  "Procurement",
  "procurement management"
)

standardise_programme <- function(
  x
) {
  cleaned <- trimws(
    tolower(x)
  )

  cleaned <- gsub(
    "[[:space:]]+",
    " ",
    cleaned
  )

  cleaned[
    cleaned %in%
      c(
        "statistics",
        "applied statistics"
      )
  ] <- "Statistics"

  cleaned[
    cleaned %in%
      c(
        "procurement",
        "procurement management"
      )
  ] <- "Procurement"

  cleaned
}

standardise_programme(
  messy_programmes
)
```

# 24. Case-insensitive search

```{r case-insensitive-search}
grepl(
  "statistics",
  messy_programmes,
  ignore.case = TRUE
)
```

# 25. Missing text values

```{r clean-missing-text}
responses <- c(
  "Yes",
  "",
  " ",
  "N/A",
  "No",
  NA
)

clean_responses <- trimws(
  responses
)

clean_responses[
  clean_responses %in%
    c(
      "",
      "N/A"
    )
] <- NA

clean_responses
```

# 26. Identifier validation

Valid form:

```text
S followed by exactly four digits
```

```{r validate-identifiers}
student_ids <- c(
  "S0001",
  "S0012",
  "0013",
  "S14",
  "S12345"
)

valid_identifier <- grepl(
  "^S[0-9]{4}$",
  student_ids
)

data.frame(
  student_id = student_ids,
  valid = valid_identifier,
  stringsAsFactors = FALSE
)
```

# 27. Email validation pattern

```{r simple-email-validation}
emails <- c(
  "ama@example.com",
  "invalid-email",
  "kojo@university.edu.gh"
)

email_pattern <-
  "^[A-Za-z0-9._%+-]+@[A-Za-z0-9.-]+\\.[A-Za-z]{2,}$"

grepl(
  email_pattern,
  emails
)
```

This is a practical pattern, not a complete implementation of every valid
email form.

# 28. Telephone-number cleaning

```{r telephone-cleaning}
telephone_numbers <- c(
  "+233 24 123 4567",
  "024-123-4567",
  "(024) 123 4567"
)

digits_only <- gsub(
  "[^0-9]",
  "",
  telephone_numbers
)

digits_only
```

# 29. Splitting composite variables

```{r composite-variable}
location_values <- c(
  "Bolgatanga|Upper East",
  "Tamale|Northern",
  "Wa|Upper West"
)

location_parts <- strsplit(
  location_values,
  split = "|",
  fixed = TRUE
)

town <- vapply(
  location_parts,
  `[`,
  FUN.VALUE = character(1),
  1
)

region <- vapply(
  location_parts,
  `[`,
  FUN.VALUE = character(1),
  2
)

data.frame(
  original = location_values,
  town = town,
  region = region,
  stringsAsFactors = FALSE
)
```

# 30. Optional `stringr`

Check availability:

```{r check-stringr}
requireNamespace(
  "stringr",
  quietly = TRUE
)
```

Examples:

```{r stringr-examples, eval=FALSE}
stringr::str_trim(
  messy_text
)

stringr::str_to_lower(
  programmes
)

stringr::str_detect(
  codes,
  "^S"
)

stringr::str_replace_all(
  text_with_spaces,
  "\\s+",
  " "
)
```

# 31. Text-cleaning pipeline function

```{r text-cleaning-function}
clean_character_vector <- function(
  x,
  missing_codes = c(
    "",
    "NA",
    "N/A",
    "."
  )
) {
  if (!is.character(x)) {
    stop(
      "`x` must be character.",
      call. = FALSE
    )
  }

  cleaned <- trimws(x)

  cleaned <- gsub(
    "[[:space:]]+",
    " ",
    cleaned
  )

  cleaned[
    cleaned %in%
      missing_codes
  ] <- NA

  cleaned
}

clean_character_vector(
  c(
    " Statistics ",
    "N/A",
    "Procurement   Management"
  )
)
```

# 32. Integrated example: clean programme values

```{r integrated-programme-cleaning}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

raw_programme <- student_data$programme

clean_programme <-
  clean_character_vector(
    raw_programme
  )

sort(
  unique(
    clean_programme
  )
)
```

# 33. Common mistakes

- using regular expressions when literal matching is sufficient;
- forgetting to escape backslashes;
- overlooking leading or trailing spaces;
- changing case without preserving the original value;
- converting identifiers to numeric;
- using an overly broad replacement pattern;
- treating empty strings and `NA` as identical without cleaning;
- validating emails or telephone numbers with unrealistic certainty;
- overwriting raw text before auditing changes;
- failing to review unmatched categories.

# 34. Independent exercises

1. inspect character vectors and string lengths.
2. convert text to lower, upper and title case.
3. trim and normalise whitespace.
4. combine text with `paste()` and `paste0()`.
5. split full names.
6. search with `grep()` and `grepl()`.
7. use anchors and character classes.
8. use quantifiers.
9. replace and remove text.
10. extract numbers from character values.
11. standardise categories.
12. validate student identifiers.
13. clean telephone numbers.
14. split a composite field.
15. write a reusable text-cleaning function.

# 35. Lesson assessment

Create a standalone R Markdown report that:

1. demonstrates character-vector creation;
2. measures and transforms text;
3. trims and normalises whitespace;
4. combines and splits strings;
5. demonstrates literal and regular-expression matching;
6. uses anchors, classes and quantifiers;
7. extracts and replaces text;
8. standardises categories;
9. validates identifiers;
10. handles missing text codes; and
11. creates a reusable cleaning function.

# 36. Completion checklist

Confirm that you can:

- inspect character vectors;
- measure string length;
- change case;
- trim whitespace;
- combine text;
- split text;
- search literally;
- use regular expressions;
- use anchors and quantifiers;
- replace matched text;
- remove punctuation;
- extract numbers;
- standardise categories;
- validate identifiers;
- clean missing text; and
- build reusable text-cleaning functions.

# Lesson conclusion

Text cleaning converts inconsistent human-entered values into analysable and
traceable categories. Reliable cleaning preserves raw values, uses explicit
patterns and documents every transformation.

\[
\boxed{
\text{inspect}
\rightarrow
\text{normalise}
\rightarrow
\text{match}
\rightarrow
\text{replace}
\rightarrow
\text{validate}
}
\]
