---
title: "Lesson 20: Dates, Times and Durations"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Dates and times are not ordinary character strings. They have ordering,
calendar rules, time zones and arithmetic behaviour. A reliable date-time
workflow must distinguish:

- calendar dates;
- clock times;
- date-times;
- durations;
- periods;
- time zones;
- and text representations.

This lesson begins with base R `Date`, `POSIXct` and `difftime` objects, then
introduces `lubridate` as an optional extension.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. create and inspect `Date` objects;
2. convert text to dates;
3. use explicit date formats;
4. extract date components;
5. compare and sort dates;
6. perform date arithmetic;
7. create date sequences;
8. work with `POSIXct` date-times;
9. specify and convert time zones;
10. calculate elapsed time with `difftime`;
11. distinguish durations and calendar periods conceptually;
12. detect invalid or missing dates;
13. validate chronological consistency;
14. use `lubridate` when available; and
15. create reproducible date-time cleaning functions.

# 1. Creating dates

```{r create-date}
study_date <- as.Date(
  "2026-07-14"
)

study_date
class(study_date)
typeof(study_date)
```

A `Date` is stored internally as the number of days since 1 January 1970.

```{r inspect-date-storage}
unclass(
  study_date
)
```

# 2. ISO date format

The recommended unambiguous text format is:

```text
YYYY-MM-DD
```

Example:

```text
2026-07-14
```

This is clearer than:

```text
07/14/2026
14/07/2026
```

because the latter forms depend on regional conventions.

# 3. Parsing custom date formats

Day-month-year:

```{r parse-day-month-year}
date_dmy <- as.Date(
  "14/07/2026",
  format = "%d/%m/%Y"
)

date_dmy
```

Month-day-year:

```{r parse-month-day-year}
date_mdy <- as.Date(
  "07/14/2026",
  format = "%m/%d/%Y"
)

date_mdy
```

Named month:

```{r parse-named-month}
date_named <- as.Date(
  "14 July 2026",
  format = "%d %B %Y"
)

date_named
```

# 4. Common format codes

| Code | Meaning |
|---|---|
| `%Y` | four-digit year |
| `%y` | two-digit year |
| `%m` | two-digit month |
| `%d` | day of month |
| `%B` | full month name |
| `%b` | abbreviated month |
| `%A` | full weekday name |
| `%a` | abbreviated weekday |
| `%H` | hour, 00–23 |
| `%M` | minute |
| `%S` | second |

# 5. Parsing a vector of dates

```{r parse-date-vector}
raw_dates <- c(
  "01/01/2026",
  "15/02/2026",
  "31/03/2026"
)

parsed_dates <- as.Date(
  raw_dates,
  format = "%d/%m/%Y"
)

parsed_dates
```

# 6. Invalid dates

```{r invalid-date}
invalid_date <- as.Date(
  "31/02/2026",
  format = "%d/%m/%Y"
)

invalid_date
```

Invalid dates become `NA`.

# 7. Detecting failed date conversion

```{r detect-date-conversion-failure}
raw_dates_with_error <- c(
  "01/01/2026",
  "31/02/2026",
  "15/03/2026"
)

converted_dates <- as.Date(
  raw_dates_with_error,
  format = "%d/%m/%Y"
)

date_conversion_report <- data.frame(
  raw = raw_dates_with_error,
  converted = converted_dates,
  failed =
    is.na(converted_dates) &
    !is.na(
      raw_dates_with_error
    ),
  stringsAsFactors = FALSE
)

date_conversion_report
```

# 8. Formatting dates

```{r format-dates}
format(
  study_date,
  "%d %B %Y"
)

format(
  study_date,
  "%A, %d %B %Y"
)
```

`format()` returns character output.

# 9. Extracting date components

```{r extract-date-components}
format(
  study_date,
  "%Y"
)

format(
  study_date,
  "%m"
)

format(
  study_date,
  "%d"
)

format(
  study_date,
  "%A"
)
```

Convert numeric components where appropriate:

```{r numeric-date-components}
year_number <- as.integer(
  format(
    study_date,
    "%Y"
  )
)

month_number <- as.integer(
  format(
    study_date,
    "%m"
  )
)

day_number <- as.integer(
  format(
    study_date,
    "%d"
  )
)

year_number
month_number
day_number
```

# 10. Date comparison

```{r compare-dates}
start_date <- as.Date(
  "2026-01-15"
)

end_date <- as.Date(
  "2026-03-20"
)

end_date > start_date
end_date == start_date
```

# 11. Sorting dates

```{r sort-dates}
event_dates <- as.Date(
  c(
    "2026-03-10",
    "2026-01-15",
    "2026-02-20"
  )
)

sort(
  event_dates
)
```

# 12. Date arithmetic

Add days:

```{r add-days}
study_date + 7
```

Subtract days:

```{r subtract-days}
study_date - 30
```

Difference between dates:

```{r date-difference}
end_date - start_date
```

# 13. `difftime`

```{r explicit-difftime}
difftime(
  end_date,
  start_date,
  units = "days"
)
```

Possible units include:

- seconds;
- minutes;
- hours;
- days;
- weeks;
- auto.

# 14. Date sequences

Daily:

```{r daily-date-sequence}
seq(
  from = as.Date(
    "2026-01-01"
  ),
  to = as.Date(
    "2026-01-10"
  ),
  by = "day"
)
```

Weekly:

```{r weekly-date-sequence}
seq(
  from = as.Date(
    "2026-01-01"
  ),
  by = "week",
  length.out = 6
)
```

Monthly:

```{r monthly-date-sequence}
seq(
  from = as.Date(
    "2026-01-01"
  ),
  by = "month",
  length.out = 12
)
```

# 15. Current date and time

```{r current-date-time}
Sys.Date()
Sys.time()
```

# 16. Date-time objects

```{r create-posixct}
appointment_time <- as.POSIXct(
  "2026-07-14 09:30:00",
  format = "%Y-%m-%d %H:%M:%S",
  tz = "Africa/Accra"
)

appointment_time
class(appointment_time)
typeof(appointment_time)
```

# 17. Time zones

```{r inspect-time-zone}
attr(
  appointment_time,
  "tzone"
)
```

Ghana uses:

```text
Africa/Accra
```

List available time zones:

```{r available-time-zones}
head(
  OlsonNames()
)
```

# 18. Converting display time zones

```{r convert-time-zone-display}
format(
  appointment_time,
  tz = "Europe/London",
  usetz = TRUE
)
```

The instant remains the same; the displayed local clock time may change.

# 19. Date-time arithmetic

```{r add-hours}
appointment_time + 2 * 60 * 60
```

Subtract:

```{r subtract-minutes}
appointment_time - 30 * 60
```

# 20. Elapsed time between date-times

```{r elapsed-time}
start_time <- as.POSIXct(
  "2026-07-14 09:15:00",
  tz = "Africa/Accra"
)

end_time <- as.POSIXct(
  "2026-07-14 11:05:00",
  tz = "Africa/Accra"
)

difftime(
  end_time,
  start_time,
  units = "mins"
)
```

# 21. Converting date-time to date

```{r datetime-to-date}
as.Date(
  appointment_time,
  tz = "Africa/Accra"
)
```

# 22. Combining separate date and time text

```{r combine-date-time-text}
date_text <- "2026-07-14"
time_text <- "14:45:00"

combined_text <- paste(
  date_text,
  time_text
)

combined_datetime <- as.POSIXct(
  combined_text,
  format = "%Y-%m-%d %H:%M:%S",
  tz = "Africa/Accra"
)

combined_datetime
```

# 23. Clock-time text

R base does not have a universal standalone time-of-day class. One practical
approach is to attach the time to a reference date.

```{r parse-clock-time}
clock_time <- strptime(
  "14:45:00",
  format = "%H:%M:%S",
  tz = "Africa/Accra"
)

clock_time
```

# 24. Chronological validation

```{r chronological-validation}
registration_date <- as.Date(
  "2026-01-10"
)

completion_date <- as.Date(
  "2026-03-15"
)

completion_date >=
  registration_date
```

Invalid order:

```{r invalid-chronology}
registration_dates <- as.Date(
  c(
    "2026-01-10",
    "2026-03-20"
  )
)

completion_dates <- as.Date(
  c(
    "2026-03-15",
    "2026-02-15"
  )
)

chronology_is_valid <-
  completion_dates >=
  registration_dates

chronology_is_valid
```

# 25. Future-date validation

```{r future-date-check}
dates_to_check <- as.Date(
  c(
    "2025-12-01",
    "2027-01-01",
    NA
  )
)

future_status <-
  dates_to_check >
  Sys.Date()

future_status
```

Interpret future dates according to the data-collection context.

# 26. Age calculation

Approximate age in years:

```{r approximate-age}
birth_date <- as.Date(
  "1990-06-15"
)

reference_date <- as.Date(
  "2026-07-14"
)

approximate_age_years <- as.numeric(
  difftime(
    reference_date,
    birth_date,
    units = "days"
  )
) / 365.2425

approximate_age_years
```

Exact age requires calendar-aware treatment.

# 27. Optional `lubridate`

Check availability:

```{r check-lubridate}
requireNamespace(
  "lubridate",
  quietly = TRUE
)
```

Parsing examples:

```{r lubridate-parsing, eval=FALSE}
lubridate::ymd(
  "2026-07-14"
)

lubridate::dmy(
  "14/07/2026"
)

lubridate::mdy(
  "07/14/2026"
)

lubridate::ymd_hms(
  "2026-07-14 09:30:00",
  tz = "Africa/Accra"
)
```

# 28. Extracting components with `lubridate`

```{r lubridate-components, eval=FALSE}
lubridate::year(
  appointment_time
)

lubridate::month(
  appointment_time,
  label = TRUE
)

lubridate::day(
  appointment_time
)

lubridate::wday(
  appointment_time,
  label = TRUE
)
```

# 29. Durations and periods

A duration represents an exact number of seconds.

A period represents calendar units such as months or years.

```{r lubridate-duration-period, eval=FALSE}
appointment_time +
  lubridate::duration(
    hours = 24
  )

appointment_time +
  lubridate::days(
    1
  )
```

These may differ around daylight-saving changes in some time zones.

# 30. Month arithmetic

Base R:

```{r base-month-sequence}
seq(
  as.Date(
    "2026-01-31"
  ),
  by = "month",
  length.out = 4
)
```

End-of-month arithmetic can be surprising. Calendar-aware functions should be
used deliberately.

# 31. Date cleaning function

```{r date-cleaning-function}
parse_date_safely <- function(
  x,
  format
) {
  if (!is.character(x)) {
    stop(
      "`x` must be character.",
      call. = FALSE
    )
  }

  parsed <- as.Date(
    x,
    format = format
  )

  data.frame(
    raw = x,
    parsed = parsed,
    failed =
      is.na(parsed) &
      !is.na(x) &
      trimws(x) != "",
    stringsAsFactors = FALSE
  )
}

parse_date_safely(
  c(
    "14/07/2026",
    "31/02/2026",
    ""
  ),
  format = "%d/%m/%Y"
)
```

# 32. Date-time validation function

```{r datetime-validation-function}
validate_datetime_order <- function(
  start,
  end
) {
  if (!inherits(
    start,
    "POSIXt"
  ) ||
      !inherits(
        end,
        "POSIXt"
      )) {
    stop(
      "Both inputs must be date-time objects.",
      call. = FALSE
    )
  }

  data.frame(
    start = start,
    end = end,
    valid =
      is.na(start) |
      is.na(end) |
      end >= start
  )
}
```

# 33. Integrated example: monthly reporting dates

```{r integrated-monthly-dates}
reporting_dates <- seq(
  from = as.Date(
    "2026-01-01"
  ),
  by = "month",
  length.out = 12
)

reporting_schedule <- data.frame(
  reporting_date = reporting_dates,
  year = format(
    reporting_dates,
    "%Y"
  ),
  month_number = format(
    reporting_dates,
    "%m"
  ),
  month_name = format(
    reporting_dates,
    "%B"
  ),
  weekday = format(
    reporting_dates,
    "%A"
  ),
  stringsAsFactors = FALSE
)

reporting_schedule
```

# 34. Common mistakes

- storing dates permanently as unvalidated character text;
- confusing day-month-year and month-day-year;
- omitting the parsing format;
- ignoring failed conversion to `NA`;
- treating date-times as ordinary numbers;
- ignoring time zones;
- assuming a day always equals the same calendar behaviour everywhere;
- using 365 days for exact age;
- ignoring end-of-month behaviour;
- comparing dates and character strings directly.

# 35. Independent exercises

1. create and inspect `Date` objects.
2. parse day-month-year and month-day-year text.
3. identify invalid dates.
4. format dates for reporting.
5. extract year, month, day and weekday.
6. compare and sort dates.
7. calculate date differences.
8. create daily, weekly and monthly sequences.
9. create `POSIXct` date-times.
10. specify and convert time zones.
11. calculate elapsed minutes.
12. validate chronological order.
13. detect future dates.
14. write a safe date-parsing function.
15. create a monthly reporting schedule.

# 36. Lesson assessment

Create a standalone R Markdown report that:

1. distinguishes dates, date-times and durations;
2. parses several date formats;
3. identifies failed conversions;
4. formats dates for reporting;
5. extracts components;
6. performs date arithmetic;
7. creates date sequences;
8. creates time-zone-aware date-times;
9. calculates elapsed time;
10. validates chronology;
11. demonstrates optional `lubridate` code; and
12. creates reusable date-cleaning functions.

# 37. Completion checklist

Confirm that you can:

- create `Date` objects;
- parse explicit formats;
- detect invalid dates;
- format dates;
- extract components;
- compare and sort dates;
- perform date arithmetic;
- create sequences;
- create `POSIXct` objects;
- specify time zones;
- calculate elapsed time;
- validate chronology;
- detect future dates;
- explain durations and periods; and
- build reusable date-cleaning functions.

# Lesson conclusion

Dates and times require explicit formats, time zones and validation rules.
Reliable date-time analysis begins by converting raw text into appropriate
classes before calculation.

\[
\boxed{
\text{identify format}
\rightarrow
\text{parse}
\rightarrow
\text{inspect}
\rightarrow
\text{validate}
\rightarrow
\text{calculate}
}
\]
