---
title: "Lesson 21: Factors and Categorical Data"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Categorical variables represent groups, classes, states or ordered levels.
Examples include:

- programme of study;
- sex;
- region;
- employment status;
- procurement method;
- satisfaction category;
- disease severity;
- and educational level.

R represents categorical data using factors. A factor stores integer codes
together with a defined set of labels called levels.

Factors are useful because they:

- distinguish categorical values from ordinary text;
- control category order;
- support statistical modelling;
- define reference groups;
- preserve unobserved but valid categories;
- and provide meaningful summaries.

However, careless factor use can produce incorrect numerical conversion,
unexpected ordering and missing values.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain the purpose of factors;
2. create unordered and ordered factors;
3. inspect levels, labels and internal codes;
4. convert character values to factors;
5. control level order;
6. define and change reference categories;
7. relabel and collapse categories;
8. identify unused levels;
9. drop unused levels;
10. distinguish factors from character vectors;
11. convert factors safely;
12. handle missing factor values;
13. create grouped tables;
14. use factors in statistical modelling; and
15. validate categorical variables.

# 1. Character categories

```{r character-categories}
programme <- c(
  "Statistics",
  "Procurement",
  "Hospitality",
  "Statistics",
  "Agriculture"
)

programme
typeof(programme)
class(programme)
```

The values are character strings.

# 2. Creating a factor

```{r create-factor}
programme_factor <- factor(
  programme
)

programme_factor
class(programme_factor)
typeof(programme_factor)
```

A factor is stored internally as integers with a levels attribute.

# 3. Inspecting factor levels

```{r inspect-factor-levels}
levels(
  programme_factor
)

nlevels(
  programme_factor
)
```

# 4. Internal factor codes

```{r factor-internal-codes}
unclass(
  programme_factor
)
```

The integer codes correspond to the positions of the labels in `levels()`.

# 5. Factor summary

```{r factor-summary}
summary(
  programme_factor
)
```

Equivalent frequency table:

```{r factor-frequency-table}
table(
  programme_factor
)
```

# 6. Specifying level order

By default, factor levels are often arranged alphabetically.

```{r explicit-level-order}
programme_factor <- factor(
  programme,
  levels = c(
    "Statistics",
    "Procurement",
    "Hospitality",
    "Agriculture"
  )
)

levels(
  programme_factor
)
```

# 7. Invalid values become missing

```{r invalid-factor-level}
programme_with_error <- factor(
  c(
    "Statistics",
    "Procurement",
    "Engineering"
  ),
  levels = c(
    "Statistics",
    "Procurement",
    "Hospitality",
    "Agriculture"
  )
)

programme_with_error
```

Because `"Engineering"` was not among the allowed levels, it becomes `NA`.

# 8. Validating categories before conversion

```{r validate-before-factor}
raw_programme <- c(
  "Statistics",
  "Procurement",
  "Engineering"
)

allowed_programmes <- c(
  "Statistics",
  "Procurement",
  "Hospitality",
  "Agriculture"
)

invalid_programmes <- setdiff(
  unique(raw_programme),
  allowed_programmes
)

invalid_programmes
```

Review invalid values before converting.

# 9. Ordered factors

Some categories have a natural order.

```{r ordered-factor}
satisfaction <- factor(
  c(
    "Low",
    "Moderate",
    "High",
    "Moderate",
    "Very high"
  ),
  levels = c(
    "Low",
    "Moderate",
    "High",
    "Very high"
  ),
  ordered = TRUE
)

satisfaction
class(satisfaction)
```

# 10. Comparisons with ordered factors

```{r ordered-factor-comparison}
satisfaction >= "High"
```

This is valid because the factor is ordered.

The same comparison is inappropriate for an unordered factor.

# 11. Relabelling factor levels

```{r relabel-factor-levels}
employment_status <- factor(
  c(
    "FT",
    "PT",
    "UN",
    "FT"
  ),
  levels = c(
    "FT",
    "PT",
    "UN"
  )
)

levels(
  employment_status
) <- c(
  "Full-time",
  "Part-time",
  "Unemployed"
)

employment_status
```

# 12. Recode before creating a factor

```{r recode-character-before-factor}
raw_status <- c(
  "FT",
  "PT",
  "UN",
  "FT"
)

recode_map <- c(
  FT = "Full-time",
  PT = "Part-time",
  UN = "Unemployed"
)

clean_status <- unname(
  recode_map[
    raw_status
  ]
)

clean_status
```

Then convert:

```{r factor-after-recode}
employment_factor <- factor(
  clean_status,
  levels = c(
    "Full-time",
    "Part-time",
    "Unemployed"
  )
)

employment_factor
```

# 13. Collapsing categories

```{r collapse-factor-categories}
programme_group <- as.character(
  programme_factor
)

programme_group[
  programme_group %in%
    c(
      "Statistics",
      "Procurement"
    )
] <- "Quantitative and Business"

programme_group[
  programme_group %in%
    c(
      "Hospitality",
      "Agriculture"
    )
] <- "Applied and Service"

programme_group <- factor(
  programme_group
)

programme_group
```

# 14. Unused levels

```{r unused-factor-levels}
all_regions <- factor(
  c(
    "Upper East",
    "Northern",
    "Upper West"
  ),
  levels = c(
    "Upper East",
    "Northern",
    "Upper West",
    "Savannah"
  )
)

all_regions
levels(all_regions)
```

Subset:

```{r subset-factor-unused-levels}
selected_regions <- all_regions[
  all_regions != "Northern"
]

selected_regions
levels(selected_regions)
```

The unused level remains.

# 15. Dropping unused levels

```{r drop-unused-levels}
selected_regions <- droplevels(
  selected_regions
)

selected_regions
levels(selected_regions)
```

# 16. Reordering factor levels

```{r reorder-factor-levels}
performance <- factor(
  c(
    "Good",
    "Poor",
    "Excellent",
    "Good"
  )
)

performance <- factor(
  performance,
  levels = c(
    "Poor",
    "Good",
    "Excellent"
  ),
  ordered = TRUE
)

performance
```

# 17. Changing the reference level

```{r relevel-factor}
programme_model_factor <- factor(
  programme,
  levels = c(
    "Statistics",
    "Procurement",
    "Hospitality",
    "Agriculture"
  )
)

programme_reference <- relevel(
  programme_model_factor,
  ref = "Statistics"
)

levels(
  programme_reference
)
```

The first level is typically the reference category in regression models.

# 18. Safe factor-to-character conversion

```{r factor-to-character}
as.character(
  programme_factor
)
```

# 19. Dangerous factor-to-numeric conversion

```{r unsafe-factor-numeric}
numeric_factor <- factor(
  c(
    "10",
    "20",
    "30"
  )
)

as.numeric(
  numeric_factor
)
```

This returns internal level codes, not the displayed values.

Correct:

```{r safe-factor-numeric}
as.numeric(
  as.character(
    numeric_factor
  )
)
```

# 20. Missing factor values

```{r missing-factor-values}
response <- factor(
  c(
    "Yes",
    "No",
    NA,
    "Yes"
  ),
  levels = c(
    "No",
    "Yes"
  )
)

response
is.na(response)
summary(response)
```

# 21. Including missing values in tables

```{r factor-table-use-na}
table(
  response,
  useNA = "ifany"
)
```

# 22. Adding an explicit missing category

```{r explicit-missing-category}
response_explicit <- addNA(
  response,
  ifany = TRUE
)

response_explicit
levels(response_explicit)
```

Use an explicit missing category only when it is analytically appropriate.

# 23. Two-way tables

```{r two-way-factor-table}
sex <- factor(
  c(
    "Female",
    "Male",
    "Female",
    "Male",
    "Female"
  )
)

table(
  programme_factor,
  sex
)
```

# 24. Proportions

```{r factor-proportions}
programme_table <- table(
  programme_factor
)

prop.table(
  programme_table
)

prop.table(
  programme_table
) * 100
```

# 25. Row and column percentages

```{r row-column-percentages}
programme_sex_table <- table(
  programme_factor,
  sex
)

prop.table(
  programme_sex_table,
  margin = 1
) * 100

prop.table(
  programme_sex_table,
  margin = 2
) * 100
```

# 26. Factors in imported data

```{r import-student-factor}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_data$programme_factor <- factor(
  student_data$programme
)

str(
  student_data[
    c(
      "programme",
      "programme_factor"
    )
  ]
)
```

# 27. Factor validation function

```{r factor-validation-function}
validate_categories <- function(
  x,
  allowed_levels
) {
  if (!is.character(x) &&
      !is.factor(x)) {
    stop(
      "`x` must be character or factor.",
      call. = FALSE
    )
  }

  raw_values <- as.character(x)

  invalid_values <- setdiff(
    unique(
      raw_values[
        !is.na(raw_values)
      ]
    ),
    allowed_levels
  )

  list(
    valid =
      length(
        invalid_values
      ) == 0L,
    invalid_values =
      invalid_values,
    missing_count =
      sum(
        is.na(raw_values)
      )
  )
}

validate_categories(
  student_data$programme,
  allowed_programmes
)
```

# 28. Frequency-summary function

```{r categorical-summary-function}
summarise_categorical <- function(
  x,
  include_missing = TRUE
) {
  if (!is.factor(x)) {
    x <- factor(x)
  }

  counts <- table(
    x,
    useNA = if (
      include_missing
    ) {
      "ifany"
    } else {
      "no"
    }
  )

  data.frame(
    category = names(counts),
    frequency =
      as.integer(
        counts
      ),
    percentage =
      as.numeric(
        counts /
          sum(counts) *
          100
      ),
    stringsAsFactors = FALSE
  )
}

summarise_categorical(
  student_data$programme
)
```

# 29. Model matrix demonstration

```{r model-matrix-factor}
model.matrix(
  ~ programme_factor
)
```

The factor is represented using indicator variables relative to a reference
level.

# 30. Common mistakes

- assuming alphabetical order is meaningful;
- converting invalid categories directly to factors;
- changing level labels without checking correspondence;
- using `as.numeric()` directly on a factor;
- failing to drop unused levels;
- using ordered factors for nominal categories;
- ignoring reference-category choice;
- treating missingness as a substantive category without justification;
- collapsing categories without documenting the rule;
- using factors before cleaning character inconsistencies.

# 31. Independent exercises

1. create an unordered factor.
2. inspect levels and internal codes.
3. specify a non-alphabetical level order.
4. validate raw categories before conversion.
5. create an ordered factor.
6. relabel factor levels.
7. collapse several categories.
8. demonstrate unused levels and `droplevels()`.
9. change a reference category.
10. convert a numeric-looking factor safely.
11. handle missing factor values.
12. create one-way and two-way tables.
13. calculate row and column percentages.
14. create a factor-validation function.
15. create a categorical-summary function.

# 32. Lesson assessment

Create a standalone R Markdown report that:

1. explains factor structure;
2. creates nominal and ordered factors;
3. controls level order;
4. validates categories;
5. relabels and collapses levels;
6. removes unused levels;
7. changes a reference category;
8. demonstrates safe numeric conversion;
9. handles missing categories;
10. creates frequency and percentage tables;
11. demonstrates model-matrix coding; and
12. applies factors to supplied data.

# 33. Completion checklist

Confirm that you can:

- create factors;
- inspect levels and codes;
- define allowed levels;
- create ordered factors;
- relabel categories;
- collapse categories;
- detect unused levels;
- drop unused levels;
- change reference levels;
- convert factors safely;
- handle missing factor values;
- create frequency tables;
- calculate proportions;
- validate categorical data; and
- understand factor coding in models.

# Lesson conclusion

Factors give categorical data an explicit analytical structure. Reliable use
requires cleaned values, deliberate level ordering and careful conversion.

\[
\boxed{
\text{clean text}
\rightarrow
\text{validate categories}
\rightarrow
\text{define levels}
\rightarrow
\text{summarise}
\rightarrow
\text{model}
}
\]
