---
title: "Lesson 22: Sorting, Filtering, Recoding and Creating Variables"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Data transformation converts raw variables into forms required for analysis.
Common tasks include:

- sorting records;
- filtering observations;
- selecting variables;
- creating totals and percentages;
- recoding categories;
- grouping numerical values;
- creating indicators;
- standardising units;
- and validating derived variables.

This lesson uses base R so that the underlying logic remains visible. Optional
`dplyr` equivalents are introduced near the end.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. sort vectors and data frames;
2. order records by one or more variables;
3. filter rows using logical conditions;
4. select and exclude columns;
5. create new variables;
6. recode character and numeric values;
7. create binary indicators;
8. group continuous variables into categories;
9. use `ifelse()`, `cut()` and named mappings;
10. transform units and scales;
11. create centred and standardised variables;
12. validate derived variables;
13. preserve raw and transformed versions; and
14. document transformation rules.

# 1. Import data

```{r import-student-transform}
raw_student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_data <- raw_student_data
```

# 2. Sorting vectors

```{r sort-vector}
scores <- student_data$final_score

sort(scores)
sort(
  scores,
  decreasing = TRUE
)
```

# 3. Ordering indices

```{r order-indices-transform}
ascending_index <- order(
  student_data$final_score
)

descending_index <- order(
  student_data$final_score,
  decreasing = TRUE
)

ascending_index
descending_index
```

# 4. Sorting a data frame

```{r sort-data-frame}
students_high_to_low <- student_data[
  descending_index,
]

students_high_to_low
```

# 5. Sorting by several variables

```{r multi-sort-data-frame}
programme_score_order <- order(
  student_data$programme,
  -student_data$final_score
)

students_by_programme <- student_data[
  programme_score_order,
]

students_by_programme
```

# 6. Filtering rows

Passing students:

```{r filter-passing-students}
passing_students <- student_data[
  student_data$final_score >= 50,
]

passing_students
```

Distinction students:

```{r filter-distinction-students}
distinction_students <- student_data[
  student_data$final_score >= 80,
]

distinction_students
```

# 7. Filtering with several conditions

```{r filter-several-conditions}
selected_students <- student_data[
  student_data$final_score >= 70 &
    student_data$programme %in%
    c(
      "Statistics",
      "Procurement"
    ),
]

selected_students
```

# 8. Filtering missing and complete values

```{r filter-complete-values}
complete_score_records <- student_data[
  !is.na(
    student_data$final_score
  ),
]

complete_score_records
```

# 9. Selecting columns

```{r select-transform-columns}
selected_columns <- student_data[
  c(
    "student_id",
    "programme",
    "final_score"
  )
]

selected_columns
```

# 10. Excluding columns

```{r exclude-transform-columns}
without_components <- student_data[
  ,
  !names(student_data) %in%
    c(
      "coursework_score",
      "examination_score"
    )
]

without_components
```

# 11. Creating arithmetic variables

```{r create-total-variable}
student_data$calculated_final_score <-
  student_data$coursework_score +
  student_data$examination_score

head(
  student_data[
    c(
      "coursework_score",
      "examination_score",
      "final_score",
      "calculated_final_score"
    )
  ]
)
```

# 12. Validating derived arithmetic

```{r validate-derived-score}
student_data$final_score_matches <-
  student_data$final_score ==
  student_data$calculated_final_score

table(
  student_data$final_score_matches,
  useNA = "ifany"
)
```

# 13. Creating percentage variables

```{r create-percentage-variable}
student_data$coursework_percentage <-
  student_data$coursework_score /
  40 *
  100

student_data$examination_percentage <-
  student_data$examination_score /
  60 *
  100

head(
  student_data[
    c(
      "coursework_percentage",
      "examination_percentage"
    )
  ]
)
```

# 14. Creating binary indicators

```{r create-binary-indicator}
student_data$passed <-
  student_data$final_score >= 50

student_data$distinction <-
  student_data$final_score >= 80

head(
  student_data[
    c(
      "final_score",
      "passed",
      "distinction"
    )
  ]
)
```

# 15. Character recoding with named mappings

```{r character-recode-map}
programme_code <- c(
  Statistics = "STAT",
  Procurement = "PROC",
  Hospitality = "HOSP",
  Agriculture = "AGRIC"
)

student_data$programme_code <- unname(
  programme_code[
    student_data$programme
  ]
)

head(
  student_data[
    c(
      "programme",
      "programme_code"
    )
  ]
)
```

# 16. Detecting failed recodes

```{r detect-failed-recodes}
failed_recode <-
  is.na(
    student_data$programme_code
  ) &
  !is.na(
    student_data$programme
  )

student_data[
  failed_recode,
  c(
    "programme",
    "programme_code"
  )
]
```

# 17. Recoding with `ifelse()`

```{r ifelse-recode}
student_data$result_label <- ifelse(
  student_data$final_score >= 50,
  "Pass",
  "Fail"
)

head(
  student_data[
    c(
      "final_score",
      "result_label"
    )
  ]
)
```

# 18. Multi-category recoding with nested `ifelse()`

```{r nested-ifelse-recode}
student_data$performance_band <- ifelse(
  student_data$final_score >= 80,
  "Excellent",
  ifelse(
    student_data$final_score >= 70,
    "Very good",
    ifelse(
      student_data$final_score >= 60,
      "Good",
      ifelse(
        student_data$final_score >= 50,
        "Satisfactory",
        "Unsatisfactory"
      )
    )
  )
)

table(
  student_data$performance_band
)
```

# 19. Grouping with `cut()`

```{r cut-performance-band}
student_data$grade <- cut(
  student_data$final_score,
  breaks = c(
    -Inf,
    49,
    59,
    69,
    79,
    Inf
  ),
  labels = c(
    "F",
    "D",
    "C",
    "B",
    "A"
  ),
  right = TRUE
)

table(
  student_data$grade,
  useNA = "ifany"
)
```

# 20. Interval boundaries

```{r inspect-cut-boundaries}
boundary_scores <- c(
  49,
  50,
  59,
  60,
  69,
  70,
  79,
  80
)

cut(
  boundary_scores,
  breaks = c(
    -Inf,
    49,
    59,
    69,
    79,
    Inf
  ),
  labels = c(
    "F",
    "D",
    "C",
    "B",
    "A"
  ),
  right = TRUE
)
```

Boundary testing is essential.

# 21. Creating quantile groups

```{r quantile-groups}
quartile_breaks <- quantile(
  student_data$final_score,
  probs = seq(
    0,
    1,
    0.25
  ),
  na.rm = TRUE
)

quartile_breaks
```

Create groups only if break values are unique:

```{r create-quantile-groups}
unique_breaks <- unique(
  quartile_breaks
)

if (length(
  unique_breaks
) >= 2L) {
  student_data$score_quartile <- cut(
    student_data$final_score,
    breaks = unique_breaks,
    include.lowest = TRUE
  )
}

head(
  student_data
)
```

# 22. Unit conversion

```{r unit-conversion}
distance_kilometres <- c(
  2.5,
  5,
  10
)

distance_metres <-
  distance_kilometres *
  1000

distance_metres
```

# 23. Currency conversion example

```{r currency-conversion-example}
revenue_ghs <- c(
  1000,
  2500,
  4200
)

exchange_rate <- 15.5

revenue_usd <-
  revenue_ghs /
  exchange_rate

revenue_usd
```

Exchange rates change over time and should be documented with date and source.

# 24. Log transformation

```{r log-transformation}
positive_values <- c(
  10,
  100,
  1000
)

log_values <- log(
  positive_values
)

log10_values <- log10(
  positive_values
)

log_values
log10_values
```

Do not apply logarithms to zero or negative values without a defensible rule.

# 25. Centring

```{r centre-variable}
mean_final_score <- mean(
  student_data$final_score,
  na.rm = TRUE
)

student_data$final_score_centred <-
  student_data$final_score -
  mean_final_score

mean(
  student_data$final_score_centred,
  na.rm = TRUE
)
```

# 26. Standardising

```{r standardise-variable}
student_data$final_score_z <- as.numeric(
  scale(
    student_data$final_score
  )
)

mean(
  student_data$final_score_z,
  na.rm = TRUE
)

sd(
  student_data$final_score_z,
  na.rm = TRUE
)
```

# 27. Ranking

```{r rank-variable}
student_data$score_rank <- rank(
  -student_data$final_score,
  ties.method = "min"
)

student_data[
  order(
    student_data$score_rank
  ),
  c(
    "student_id",
    "final_score",
    "score_rank"
  )
]
```

# 28. Within-group ranks

```{r within-group-rank}
student_data$programme_rank <- ave(
  -student_data$final_score,
  student_data$programme,
  FUN = function(x) {
    rank(
      x,
      ties.method = "min"
    )
  }
)

student_data[
  order(
    student_data$programme,
    student_data$programme_rank
  ),
  c(
    "programme",
    "student_id",
    "final_score",
    "programme_rank"
  )
]
```

# 29. Row-wise totals

```{r row-wise-total}
score_components <- student_data[
  c(
    "coursework_score",
    "examination_score"
  )
]

student_data$row_total <- rowSums(
  score_components,
  na.rm = FALSE
)

head(
  student_data[
    c(
      "row_total",
      "final_score"
    )
  ]
)
```

# 30. Row-wise missing count

```{r row-wise-missing-count}
student_data$component_missing_count <-
  rowSums(
    is.na(
      score_components
    )
  )

table(
  student_data$component_missing_count
)
```

# 31. Preserving raw variables

```{r preserve-raw-variable}
student_data$programme_raw <-
  student_data$programme

student_data$programme_clean <-
  trimws(
    tools::toTitleCase(
      tolower(
        student_data$programme
      )
    )
  )
```

Do not overwrite the original variable until transformations have been
verified.

# 32. Transformation log

```{r transformation-log}
transformation_log <- data.frame(
  new_variable = c(
    "calculated_final_score",
    "passed",
    "grade",
    "final_score_z"
  ),
  source_variables = c(
    "coursework_score + examination_score",
    "final_score",
    "final_score",
    "final_score"
  ),
  rule = c(
    "Arithmetic sum",
    "final_score >= 50",
    "Cut points: 49, 59, 69, 79",
    "Mean-centred and divided by standard deviation"
  ),
  stringsAsFactors = FALSE
)

transformation_log
```

# 33. Optional `dplyr`

```{r check-dplyr-transform}
requireNamespace(
  "dplyr",
  quietly = TRUE
)
```

Template:

```{r dplyr-transform-template, eval=FALSE}
student_transformed <- student_data |>
  dplyr::filter(
    final_score >= 50
  ) |>
  dplyr::mutate(
    passed =
      final_score >= 50,
    grade = dplyr::case_when(
      final_score >= 80 ~ "A",
      final_score >= 70 ~ "B",
      final_score >= 60 ~ "C",
      final_score >= 50 ~ "D",
      TRUE ~ "F"
    )
  ) |>
  dplyr::arrange(
    dplyr::desc(
      final_score
    )
  )
```

# 34. Validation after transformation

```{r validate-transformations}
all(
  student_data$calculated_final_score ==
    student_data$final_score
)

all(
  student_data$passed ==
    (
      student_data$final_score >= 50
    )
)

all(
  !is.na(
    student_data$programme_code
  )
)
```

# 35. Common mistakes

- sorting one vector without reordering the whole data frame;
- filtering before preserving raw data;
- using nested `ifelse()` without testing boundaries;
- recoding unmatched values silently to `NA`;
- changing units without documenting the conversion;
- applying logarithms to invalid values;
- overwriting original variables;
- calculating standard scores without checking missing values;
- using row sums with inappropriate `na.rm`;
- failing to validate derived variables.

# 36. Independent exercises

1. sort a vector and a data frame.
2. order by two variables.
3. filter using one and multiple conditions.
4. select and exclude columns.
5. create arithmetic variables.
6. create percentage variables.
7. create binary indicators.
8. recode categories with a named mapping.
9. group scores with `cut()`.
10. test interval boundaries.
11. create centred and standardised variables.
12. rank observations.
13. create within-group ranks.
14. build a transformation log.
15. validate all derived variables.

# 37. Lesson assessment

Create a standalone R Markdown report that:

1. preserves raw data;
2. sorts and filters records;
3. selects variables;
4. creates arithmetic and percentage variables;
5. creates indicators;
6. recodes categories;
7. groups a continuous variable;
8. performs unit and scale transformations;
9. creates ranks;
10. logs every transformation; and
11. validates the final dataset.

# 38. Completion checklist

Confirm that you can:

- sort vectors;
- order data frames;
- filter rows;
- select columns;
- exclude columns;
- create derived variables;
- create indicators;
- recode categories;
- use `cut()`;
- test boundaries;
- convert units;
- centre and standardise;
- rank observations;
- preserve raw variables;
- document transformations; and
- validate transformed data.

# Lesson conclusion

Data transformation should be transparent, reversible and validated. Every
derived variable must have a documented source and rule.

\[
\boxed{
\text{preserve raw data}
\rightarrow
\text{transform}
\rightarrow
\text{label}
\rightarrow
\text{validate}
\rightarrow
\text{document}
}
\]
