---
title: "Lesson 23: Combining, Joining and Reshaping Data"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Many analyses require information from more than one table. Data may be split
by:

- year;
- institution;
- department;
- questionnaire section;
- transaction type;
- respondent and household;
- or measurement occasion.

Combining data correctly requires understanding the structure of each table and
the keys that identify matching records.

This lesson develops:

- row binding;
- column binding;
- base R joins with `merge()`;
- one-to-one, one-to-many and many-to-many relationships;
- duplicate-key checks;
- unmatched-record diagnostics;
- wide-to-long and long-to-wide reshaping;
- and optional `dplyr` and `tidyr` equivalents.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish row binding, column binding and relational joining;
2. combine compatible rows with `rbind()`;
3. combine aligned columns with `cbind()`;
4. join tables using `merge()`;
5. perform inner, left, right and full joins;
6. validate key variables;
7. identify unmatched records;
8. detect one-to-many and many-to-many relationships;
9. avoid accidental row multiplication;
10. reshape data from wide to long;
11. reshape data from long to wide;
12. preserve identifiers during reshaping;
13. use optional `dplyr` and `tidyr` joins; and
14. audit combined datasets.

# 1. Row binding

Create two compatible tables:

```{r create-row-bind-data}
students_2025 <- data.frame(
  student_id = c(
    "S001",
    "S002"
  ),
  programme = c(
    "Statistics",
    "Procurement"
  ),
  final_score = c(
    72,
    68
  ),
  year = 2025L,
  stringsAsFactors = FALSE
)

students_2026 <- data.frame(
  student_id = c(
    "S003",
    "S004"
  ),
  programme = c(
    "Hospitality",
    "Agriculture"
  ),
  final_score = c(
    81,
    75
  ),
  year = 2026L,
  stringsAsFactors = FALSE
)
```

Combine:

```{r row-bind-data}
all_students <- rbind(
  students_2025,
  students_2026
)

all_students
```

# 2. Requirements for `rbind()`

Tables should have:

- compatible column names;
- compatible column types;
- compatible meanings;
- and equivalent units.

Inspect:

```{r inspect-row-bind-structure}
str(
  students_2025
)

str(
  students_2026
)
```

# 3. Incompatible columns

```{r incompatible-rbind-example, error=TRUE}
incomplete_students <- data.frame(
  student_id = "S005",
  final_score = 77,
  stringsAsFactors = FALSE
)

rbind(
  students_2025,
  incomplete_students
)
```

Align variables before binding.

# 4. Column binding

```{r create-column-bind-data}
student_identity <- data.frame(
  student_id = c(
    "S001",
    "S002",
    "S003"
  ),
  programme = c(
    "Statistics",
    "Procurement",
    "Hospitality"
  ),
  stringsAsFactors = FALSE
)

student_results <- data.frame(
  coursework = c(
    30,
    28,
    35
  ),
  examination = c(
    48,
    42,
    51
  )
)

combined_columns <- cbind(
  student_identity,
  student_results
)

combined_columns
```

# 5. Risk of `cbind()`

`cbind()` assumes row alignment by position.

```{r misaligned-cbind-risk}
shuffled_results <- student_results[
  c(
    3,
    1,
    2
  ),
]

cbind(
  student_identity,
  shuffled_results
)
```

The result is structurally valid but analytically wrong.

Use key-based joins when rows must be matched by identifiers.

# 6. Relational keys

A key is a variable or combination of variables that identifies records.

```{r create-join-tables}
student_master <- data.frame(
  student_id = c(
    "S001",
    "S002",
    "S003",
    "S004"
  ),
  programme = c(
    "Statistics",
    "Procurement",
    "Hospitality",
    "Agriculture"
  ),
  stringsAsFactors = FALSE
)

score_table <- data.frame(
  student_id = c(
    "S001",
    "S002",
    "S003",
    "S005"
  ),
  final_score = c(
    72,
    68,
    81,
    77
  ),
  stringsAsFactors = FALSE
)
```

# 7. Key validation

```{r validate-join-keys}
anyDuplicated(
  student_master$student_id
)

anyDuplicated(
  score_table$student_id
)
```

Both should be zero for a one-to-one join.

# 8. Inner join with `merge()`

```{r inner-merge}
inner_joined <- merge(
  student_master,
  score_table,
  by = "student_id"
)

inner_joined
```

Only matching identifiers are retained.

# 9. Left join

```{r left-merge}
left_joined <- merge(
  student_master,
  score_table,
  by = "student_id",
  all.x = TRUE
)

left_joined
```

All rows from the left table are retained.

# 10. Right join

```{r right-merge}
right_joined <- merge(
  student_master,
  score_table,
  by = "student_id",
  all.y = TRUE
)

right_joined
```

# 11. Full join

```{r full-merge}
full_joined <- merge(
  student_master,
  score_table,
  by = "student_id",
  all = TRUE
)

full_joined
```

# 12. Unmatched records

Identifiers in master but not scores:

```{r unmatched-left-keys}
setdiff(
  student_master$student_id,
  score_table$student_id
)
```

Identifiers in scores but not master:

```{r unmatched-right-keys}
setdiff(
  score_table$student_id,
  student_master$student_id
)
```

# 13. Joining on differently named keys

```{r differently-named-keys}
identity_data <- data.frame(
  student_code = c(
    "S001",
    "S002"
  ),
  programme = c(
    "Statistics",
    "Procurement"
  ),
  stringsAsFactors = FALSE
)

result_data <- data.frame(
  student_id = c(
    "S001",
    "S002"
  ),
  final_score = c(
    72,
    68
  ),
  stringsAsFactors = FALSE
)

merge(
  identity_data,
  result_data,
  by.x = "student_code",
  by.y = "student_id"
)
```

# 14. Composite keys

```{r composite-key-tables}
semester_records <- data.frame(
  student_id = c(
    "S001",
    "S001",
    "S002",
    "S002"
  ),
  semester = c(
    1,
    2,
    1,
    2
  ),
  final_score = c(
    70,
    75,
    68,
    72
  )
)

semester_status <- data.frame(
  student_id = c(
    "S001",
    "S001",
    "S002",
    "S002"
  ),
  semester = c(
    1,
    2,
    1,
    2
  ),
  status = c(
    "Completed",
    "Completed",
    "Completed",
    "Completed"
  ),
  stringsAsFactors = FALSE
)

merge(
  semester_records,
  semester_status,
  by = c(
    "student_id",
    "semester"
  )
)
```

# 15. One-to-many relationships

```{r one-to-many-example}
student_households <- data.frame(
  household_id = c(
    "H001",
    "H002"
  ),
  district = c(
    "Bolgatanga",
    "Bawku West"
  ),
  stringsAsFactors = FALSE
)

household_members <- data.frame(
  household_id = c(
    "H001",
    "H001",
    "H002"
  ),
  member_name = c(
    "Ama",
    "Kojo",
    "Esi"
  ),
  stringsAsFactors = FALSE
)

merge(
  student_households,
  household_members,
  by = "household_id"
)
```

One household appears in several member rows.

# 16. Many-to-many risk

```{r many-to-many-example}
table_a <- data.frame(
  key = c(
    "A",
    "A"
  ),
  value_a = c(
    1,
    2
  )
)

table_b <- data.frame(
  key = c(
    "A",
    "A",
    "A"
  ),
  value_b = c(
    10,
    20,
    30
  )
)

merge(
  table_a,
  table_b,
  by = "key"
)
```

Two rows multiplied by three rows produce six combinations.

# 17. Duplicate-key audit

```{r duplicate-key-audit}
audit_key <- function(
  data,
  key
) {
  key_values <- data[[
    key
  ]]

  data.frame(
    total_rows = nrow(data),
    missing_keys = sum(
      is.na(key_values) |
        key_values == ""
    ),
    duplicated_rows = sum(
      duplicated(key_values)
    ),
    unique_keys = length(
      unique(key_values)
    )
  )
}

audit_key(
  student_master,
  "student_id"
)
```

# 18. Join audit

```{r join-audit-function}
audit_join <- function(
  left_data,
  right_data,
  key
) {
  left_keys <-
    left_data[[
      key
    ]]

  right_keys <-
    right_data[[
      key
    ]]

  list(
    left_rows =
      nrow(left_data),
    right_rows =
      nrow(right_data),
    left_duplicate_keys =
      sum(
        duplicated(left_keys)
      ),
    right_duplicate_keys =
      sum(
        duplicated(right_keys)
      ),
    unmatched_left =
      setdiff(
        unique(left_keys),
        unique(right_keys)
      ),
    unmatched_right =
      setdiff(
        unique(right_keys),
        unique(left_keys)
      )
  )
}

audit_join(
  student_master,
  score_table,
  "student_id"
)
```

# 19. Suffixes for duplicate variable names

```{r merge-suffixes}
left_data <- data.frame(
  student_id = c(
    "S001",
    "S002"
  ),
  status = c(
    "Registered",
    "Registered"
  )
)

right_data <- data.frame(
  student_id = c(
    "S001",
    "S002"
  ),
  status = c(
    "Passed",
    "Failed"
  )
)

merge(
  left_data,
  right_data,
  by = "student_id",
  suffixes = c(
    "_registration",
    "_result"
  )
)
```

# 20. Wide data

```{r create-wide-data}
wide_scores <- data.frame(
  student_id = c(
    "S001",
    "S002",
    "S003"
  ),
  semester_1 = c(
    65,
    72,
    81
  ),
  semester_2 = c(
    68,
    75,
    84
  ),
  stringsAsFactors = FALSE
)

wide_scores
```

Each semester is stored in a separate column.

# 21. Long data

```{r create-long-data}
long_scores <- data.frame(
  student_id = rep(
    wide_scores$student_id,
    each = 2
  ),
  semester = rep(
    c(
      "semester_1",
      "semester_2"
    ),
    times = nrow(
      wide_scores
    )
  ),
  score = as.vector(
    t(
      wide_scores[
        c(
          "semester_1",
          "semester_2"
        )
      ]
    )
  ),
  stringsAsFactors = FALSE
)

long_scores
```

# 22. Base R wide-to-long with `reshape()`

```{r base-wide-to-long}
wide_for_reshape <- wide_scores

long_with_reshape <- reshape(
  wide_for_reshape,
  varying = c(
    "semester_1",
    "semester_2"
  ),
  v.names = "score",
  timevar = "semester",
  times = c(
    1,
    2
  ),
  direction = "long"
)

long_with_reshape
```

# 23. Base R long-to-wide

```{r base-long-to-wide}
wide_with_reshape <- reshape(
  long_with_reshape,
  idvar = "student_id",
  timevar = "semester",
  direction = "wide"
)

wide_with_reshape
```

# 24. Preserving identifiers during reshaping

Every repeated row in long format should retain the unit identifier.

```{r validate-long-identifiers}
table(
  long_scores$student_id
)
```

Each student appears twice because there are two semesters.

# 25. Duplicate long-format cells

A long dataset should have one row per identifier-time combination when it is
intended to reshape to wide.

```{r duplicate-long-cell-check}
long_key <- paste(
  long_scores$student_id,
  long_scores$semester,
  sep = "::"
)

anyDuplicated(
  long_key
)
```

# 26. Optional `dplyr` joins

```{r check-dplyr-join}
requireNamespace(
  "dplyr",
  quietly = TRUE
)
```

Templates:

```{r dplyr-join-templates, eval=FALSE}
dplyr::inner_join(
  student_master,
  score_table,
  by = "student_id"
)

dplyr::left_join(
  student_master,
  score_table,
  by = "student_id"
)

dplyr::right_join(
  student_master,
  score_table,
  by = "student_id"
)

dplyr::full_join(
  student_master,
  score_table,
  by = "student_id"
)

dplyr::anti_join(
  student_master,
  score_table,
  by = "student_id"
)
```

# 27. Optional `tidyr` reshaping

```{r check-tidyr}
requireNamespace(
  "tidyr",
  quietly = TRUE
)
```

Wide to long:

```{r pivot-longer-template, eval=FALSE}
long_scores <- tidyr::pivot_longer(
  wide_scores,
  cols = c(
    semester_1,
    semester_2
  ),
  names_to = "semester",
  values_to = "score"
)
```

Long to wide:

```{r pivot-wider-template, eval=FALSE}
wide_scores <- tidyr::pivot_wider(
  long_scores,
  names_from = semester,
  values_from = score
)
```

# 28. Binding many files

```{r bind-many-files-template}
csv_files <- list.files(
  path = "data",
  pattern = "\\.csv$",
  full.names = TRUE
)

file_data <- lapply(
  csv_files,
  read.csv,
  stringsAsFactors = FALSE
)

names(file_data) <-
  basename(csv_files)

lapply(
  file_data,
  dim
)
```

Do not row-bind unrelated datasets merely because they are all CSV files.

# 29. Source-file tracking

When binding similar files, create a source variable.

```{r source-file-tracking}
year_2025 <- students_2025
year_2026 <- students_2026

year_2025$source_file <-
  "students_2025.csv"

year_2026$source_file <-
  "students_2026.csv"

combined_with_source <- rbind(
  year_2025,
  year_2026
)

combined_with_source
```

# 30. Post-combination validation

```{r post-combination-validation}
nrow(
  all_students
)

names(
  all_students
)

str(
  all_students
)

anyDuplicated(
  paste(
    all_students$student_id,
    all_students$year,
    sep = "::"
  )
)
```

# 31. Join-result row-count check

```{r join-row-count-check}
nrow(
  student_master
)

nrow(
  left_joined
)
```

For a one-to-one left join, row counts should usually match. Extra rows may
indicate duplicated keys in the right table.

# 32. Combined-data audit log

```{r combination-log}
combination_log <- data.frame(
  operation = c(
    "Row bind",
    "Left join",
    "Wide-to-long"
  ),
  input_objects = c(
    "students_2025; students_2026",
    "student_master; score_table",
    "wide_scores"
  ),
  key = c(
    "student_id + year",
    "student_id",
    "student_id + semester"
  ),
  expected_rows = c(
    nrow(students_2025) +
      nrow(students_2026),
    nrow(student_master),
    nrow(wide_scores) * 2
  ),
  stringsAsFactors = FALSE
)

combination_log
```

# 33. Common mistakes

- using `cbind()` for records that are not aligned;
- row-binding incompatible variables;
- joining without checking key uniqueness;
- ignoring unmatched records;
- accepting accidental many-to-many joins;
- forgetting suffixes for duplicate variable names;
- reshaping without preserving identifiers;
- reshaping long data containing duplicate identifier-time cells;
- binding unrelated files;
- failing to record source files and row counts.

# 34. Independent exercises

1. row-bind two compatible datasets.
2. demonstrate an incompatible row bind.
3. column-bind aligned tables.
4. explain the risk of positional alignment.
5. perform inner, left, right and full joins.
6. identify unmatched keys.
7. join on differently named variables.
8. join on a composite key.
9. demonstrate one-to-many and many-to-many relationships.
10. audit key uniqueness.
11. reshape wide data to long.
12. reshape long data to wide.
13. validate identifier-time uniqueness.
14. add source-file tracking.
15. create a post-combination audit.

# 35. Lesson assessment

Create a standalone R Markdown report that:

1. demonstrates row and column binding;
2. validates compatibility;
3. performs four join types;
4. audits unmatched and duplicate keys;
5. explains relationship types;
6. detects accidental row multiplication;
7. reshapes wide to long;
8. reshapes long to wide;
9. preserves identifiers;
10. tracks source files;
11. validates row counts; and
12. documents all combination operations.

# 36. Completion checklist

Confirm that you can:

- row-bind tables;
- column-bind aligned data;
- explain relational keys;
- perform inner joins;
- perform left, right and full joins;
- identify unmatched records;
- join differently named keys;
- use composite keys;
- distinguish relationship types;
- detect duplicate keys;
- reshape wide to long;
- reshape long to wide;
- validate identifier-time cells;
- track source files; and
- audit combined data.

# Lesson conclusion

Combining data requires more than matching column names. The analyst must
understand keys, relationships, unmatched records and expected row counts.

\[
\boxed{
\text{inspect tables}
\rightarrow
\text{validate keys}
\rightarrow
\text{combine}
\rightarrow
\text{audit rows}
\rightarrow
\text{verify meaning}
}
\]
