---
title: "Lesson 24: Descriptive Statistics for Numerical Data"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Descriptive statistics summarise the main features of numerical data before
formal modelling. They help the analyst understand:

- centre;
- spread;
- range;
- shape;
- unusual values;
- missingness;
- and differences across groups.

A reliable descriptive analysis should not consist of a single mean. It should
combine several summaries and explain what each contributes.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish descriptive and inferential statistics;
2. calculate counts and effective sample sizes;
3. calculate means, medians and modes;
4. calculate ranges, quartiles and interquartile ranges;
5. calculate variance and standard deviation;
6. calculate coefficients of variation;
7. calculate quantiles and percentiles;
8. understand resistant and non-resistant statistics;
9. summarise missing values;
10. identify possible outliers;
11. compare summaries across groups;
12. write reusable summary functions;
13. produce publication-ready descriptive tables; and
14. interpret descriptive statistics in context.

# 1. Importing the working data

```{r import-student-descriptive}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

head(student_data)
str(student_data)
```

# 2. The analytical variable

```{r select-final-score}
final_scores <- student_data$final_score

final_scores
```

# 3. Sample size and observed count

```{r descriptive-counts}
total_count <- length(final_scores)

observed_count <- sum(
  !is.na(final_scores)
)

missing_count <- sum(
  is.na(final_scores)
)

total_count
observed_count
missing_count
```

# 4. Arithmetic mean

```{r arithmetic-mean}
mean(
  final_scores,
  na.rm = TRUE
)
```

The arithmetic mean is:

\[
\bar{x}
=
\frac{1}{n}
\sum_{i=1}^{n}x_i.
\]

It uses every observed value and is sensitive to extreme observations.

# 5. Median

```{r descriptive-median}
median(
  final_scores,
  na.rm = TRUE
)
```

The median is the middle ordered value and is more resistant to extreme values.

# 6. Mode

Base R does not provide a direct statistical mode function.

```{r statistical-mode-function}
statistical_mode <- function(
  x,
  remove_missing = TRUE
) {
  if (remove_missing) {
    x <- x[
      !is.na(x)
    ]
  }

  frequencies <- table(x)

  if (length(frequencies) == 0L) {
    return(NA)
  }

  maximum_frequency <- max(frequencies)

  modes <- names(
    frequencies
  )[
    frequencies ==
      maximum_frequency
  ]

  list(
    mode = modes,
    frequency =
      as.integer(
        maximum_frequency
      )
  )
}

statistical_mode(
  final_scores
)
```

A distribution may have more than one mode.

# 7. Minimum and maximum

```{r minimum-maximum}
minimum_score <- min(
  final_scores,
  na.rm = TRUE
)

maximum_score <- max(
  final_scores,
  na.rm = TRUE
)

minimum_score
maximum_score
```

# 8. Range

```{r descriptive-range}
range(
  final_scores,
  na.rm = TRUE
)

score_range <-
  maximum_score -
  minimum_score

score_range
```

# 9. Quartiles

```{r descriptive-quartiles}
quantile(
  final_scores,
  probs = c(
    0.25,
    0.50,
    0.75
  ),
  na.rm = TRUE
)
```

# 10. Interquartile range

```{r descriptive-iqr}
IQR(
  final_scores,
  na.rm = TRUE
)
```

The interquartile range is:

\[
IQR = Q_3 - Q_1.
\]

# 11. Variance

```{r descriptive-variance}
var(
  final_scores,
  na.rm = TRUE
)
```

Sample variance is:

\[
s^2
=
\frac{
\sum_{i=1}^{n}
(x_i-\bar{x})^2
}{
n-1
}.
\]

# 12. Standard deviation

```{r descriptive-standard-deviation}
sd(
  final_scores,
  na.rm = TRUE
)
```

The standard deviation is the square root of the variance.

# 13. Coefficient of variation

```{r coefficient-of-variation}
mean_score <- mean(
  final_scores,
  na.rm = TRUE
)

standard_deviation <- sd(
  final_scores,
  na.rm = TRUE
)

coefficient_of_variation <-
  standard_deviation /
  mean_score *
  100

coefficient_of_variation
```

The coefficient of variation is meaningful mainly for ratio-scale variables
with a meaningful zero.

# 14. Quantiles and percentiles

```{r descriptive-percentiles}
quantile(
  final_scores,
  probs = c(
    0.05,
    0.10,
    0.25,
    0.50,
    0.75,
    0.90,
    0.95
  ),
  na.rm = TRUE
)
```

# 15. Five-number summary

```{r five-number-summary}
fivenum(
  final_scores,
  na.rm = TRUE
)
```

# 16. General summary

```{r general-summary}
summary(
  final_scores
)
```

# 17. Mean versus median under an extreme value

```{r extreme-value-example}
ordinary_values <- c(
  10,
  12,
  13,
  14,
  15
)

values_with_extreme <- c(
  ordinary_values,
  100
)

data.frame(
  dataset = c(
    "Ordinary",
    "With extreme value"
  ),
  mean = c(
    mean(ordinary_values),
    mean(values_with_extreme)
  ),
  median = c(
    median(ordinary_values),
    median(values_with_extreme)
  ),
  stringsAsFactors = FALSE
)
```

# 18. Resistant statistics

More resistant:

- median;
- quartiles;
- interquartile range.

Less resistant:

- mean;
- variance;
- standard deviation;
- range.

# 19. Z-scores

```{r descriptive-z-scores}
score_z <- as.numeric(
  scale(
    final_scores
  )
)

head(score_z)
mean(
  score_z,
  na.rm = TRUE
)
sd(
  score_z,
  na.rm = TRUE
)
```

# 20. IQR outlier screening

```{r descriptive-outlier-screen}
q1 <- quantile(
  final_scores,
  0.25,
  na.rm = TRUE
)

q3 <- quantile(
  final_scores,
  0.75,
  na.rm = TRUE
)

iqr_value <- q3 - q1

lower_fence <- q1 - 1.5 * iqr_value
upper_fence <- q3 + 1.5 * iqr_value

outlier_status <-
  final_scores < lower_fence |
  final_scores > upper_fence

student_data[
  outlier_status,
]
```

An outlier flag is not proof of error.

# 21. Skewness by moment calculation

```{r descriptive-skewness-function}
calculate_skewness <- function(
  x
) {
  x <- x[
    !is.na(x)
  ]

  n <- length(x)
  centre <- mean(x)
  spread <- sd(x)

  if (n < 3L ||
      spread == 0) {
    return(NA_real_)
  }

  mean(
    (
      (x - centre) /
        spread
    ) ^ 3
  )
}

calculate_skewness(
  final_scores
)
```

# 22. Grouped descriptive statistics with `tapply()`

```{r grouped-means}
tapply(
  student_data$final_score,
  student_data$programme,
  mean,
  na.rm = TRUE
)
```

Grouped standard deviations:

```{r grouped-standard-deviations}
tapply(
  student_data$final_score,
  student_data$programme,
  sd,
  na.rm = TRUE
)
```

# 23. Grouped summaries with `aggregate()`

```{r aggregate-grouped-summary}
aggregate(
  final_score ~ programme,
  data = student_data,
  FUN = mean
)
```

# 24. A complete numerical summary function

```{r complete-numerical-summary}
summarise_numeric <- function(
  x,
  variable_name = "x"
) {
  if (!is.numeric(x)) {
    stop(
      "`x` must be numeric.",
      call. = FALSE
    )
  }

  observed <- x[
    !is.na(x)
  ]

  if (length(observed) == 0L) {
    stop(
      "No observed values are available.",
      call. = FALSE
    )
  }

  q <- quantile(
    observed,
    probs = c(
      0.25,
      0.50,
      0.75
    )
  )

  data.frame(
    variable = variable_name,
    total = length(x),
    observed = length(observed),
    missing = sum(
      is.na(x)
    ),
    mean = mean(observed),
    median = median(observed),
    standard_deviation = sd(observed),
    variance = var(observed),
    minimum = min(observed),
    first_quartile = q[1],
    third_quartile = q[3],
    maximum = max(observed),
    interquartile_range = IQR(observed),
    stringsAsFactors = FALSE,
    row.names = NULL
  )
}

summarise_numeric(
  final_scores,
  "final_score"
)
```

# 25. Summarising several numerical variables

```{r summarise-several-numeric}
numeric_variables <- names(
  student_data
)[
  vapply(
    student_data,
    is.numeric,
    FUN.VALUE = logical(1)
  )
]

numeric_summaries <- lapply(
  numeric_variables,
  function(variable_name) {
    summarise_numeric(
      student_data[[
        variable_name
      ]],
      variable_name
    )
  }
)

numeric_summary_table <- do.call(
  rbind,
  numeric_summaries
)

numeric_summary_table
```

# 26. Grouped summary function

```{r grouped-summary-function}
summarise_by_group <- function(
  data,
  numerical_variable,
  group_variable
) {
  if (!all(
    c(
      numerical_variable,
      group_variable
    ) %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  groups <- unique(
    data[[
      group_variable
    ]]
  )

  results <- lapply(
    groups,
    function(group_value) {
      selected <- data[
        data[[
          group_variable
        ]] == group_value &
          !is.na(
            data[[
              group_variable
            ]]
          ),
        numerical_variable
      ]

      summary_row <- summarise_numeric(
        selected,
        numerical_variable
      )

      summary_row[
        group_variable
      ] <- group_value

      summary_row
    }
  )

  result <- do.call(
    rbind,
    results
  )

  result[
    c(
      group_variable,
      setdiff(
        names(result),
        group_variable
      )
    )
  ]
}

summarise_by_group(
  student_data,
  "final_score",
  "programme"
)
```

# 27. Reporting precision

```{r reporting-precision}
summary_table <- summarise_numeric(
  final_scores,
  "final_score"
)

numeric_columns <- vapply(
  summary_table,
  is.numeric,
  FUN.VALUE = logical(1)
)

summary_table[
  numeric_columns
] <- lapply(
  summary_table[
    numeric_columns
  ],
  round,
  digits = 2
)

summary_table
```

# 28. Interpreting descriptive statistics

A good interpretation should state:

- the number of observed values;
- the main measure of centre;
- the spread;
- the range or quartiles;
- missingness;
- and any important asymmetry or unusual values.

# 29. Common mistakes

- reporting the mean without the sample size;
- using `na.rm = TRUE` without reporting missingness;
- presenting standard deviation without the mean;
- interpreting an outlier flag as an error;
- using coefficient of variation for unsuitable scales;
- comparing group means without group sizes;
- rounding too early;
- confusing variance and standard deviation;
- using mode for continuous variables without purpose;
- relying on one statistic only.

# 30. Independent exercises

1. calculate observed and missing counts.
2. calculate mean, median and mode.
3. calculate minimum, maximum and range.
4. calculate quartiles and IQR.
5. calculate variance and standard deviation.
6. calculate coefficient of variation.
7. calculate selected percentiles.
8. compare mean and median under an extreme value.
9. identify possible outliers.
10. calculate grouped means and standard deviations.
11. create a complete numerical summary function.
12. summarise every numerical variable.
13. create grouped descriptive tables.
14. write a contextual interpretation.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. imports supplied data;
2. describes sample size and missingness;
3. reports measures of centre;
4. reports measures of spread;
5. reports quartiles and percentiles;
6. screens for possible outliers;
7. compares programme groups;
8. creates reusable summary functions;
9. produces a rounded final table; and
10. interprets the results in complete sentences.

# 32. Completion checklist

Confirm that you can:

- calculate observed sample size;
- calculate mean, median and mode;
- calculate range and quartiles;
- calculate variance and standard deviation;
- calculate coefficient of variation;
- calculate percentiles;
- identify resistant statistics;
- calculate z-scores;
- screen possible outliers;
- compare groups;
- create reusable summary functions; and
- interpret descriptive statistics.

# Lesson conclusion

Descriptive analysis should combine sample size, centre, spread, distribution
shape and missingness. The purpose is to understand the data before making
formal statistical claims.

\[
\boxed{
\text{inspect}
\rightarrow
\text{count}
\rightarrow
\text{summarise centre}
\rightarrow
\text{summarise spread}
\rightarrow
\text{interpret}
}
\]
