---
title: "Lesson 25: Frequency Tables, Cross-Tabulation and Grouped Summaries"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Categorical analysis begins with frequencies and percentages. Cross-tabulation
extends this by examining how two categorical variables are distributed
together.

Grouped summaries also combine categorical and numerical information, such as:

- mean score by programme;
- waiting time by sex;
- sales by month;
- rainfall by location;
- or pass rate by department.

This lesson develops one-way tables, two-way tables, margins, proportions,
grouped numerical summaries and reusable reporting functions.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. create one-way frequency tables;
2. include or exclude missing values;
3. calculate percentages;
4. create two-way cross-tabulations;
5. calculate row, column and overall percentages;
6. add marginal totals;
7. create grouped numerical summaries;
8. calculate group sample sizes;
9. calculate grouped proportions;
10. create multi-function summaries;
11. reshape table outputs into data frames;
12. create reusable tabulation functions;
13. validate percentages and totals; and
14. interpret grouped tables.

# 1. Importing data

```{r import-student-frequency}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

head(student_data)
```

# 2. One-way frequency table

```{r one-way-frequency}
programme_frequency <- table(
  student_data$programme
)

programme_frequency
```

# 3. Including missing values

```{r frequency-use-na}
table(
  student_data$programme,
  useNA = "ifany"
)
```

# 4. Percentage table

```{r percentage-frequency}
programme_percentage <- prop.table(
  programme_frequency
) * 100

programme_percentage
```

# 5. Frequency and percentage data frame

```{r frequency-percentage-data-frame}
programme_summary <- data.frame(
  programme = names(
    programme_frequency
  ),
  frequency =
    as.integer(
      programme_frequency
    ),
  percentage =
    as.numeric(
      programme_percentage
    ),
  stringsAsFactors = FALSE
)

programme_summary
```

# 6. Cumulative frequency

Cumulative frequency is meaningful when categories have a logical order.

```{r cumulative-frequency}
grade <- cut(
  student_data$final_score,
  breaks = c(
    -Inf,
    49,
    59,
    69,
    79,
    Inf
  ),
  labels = c(
    "F",
    "D",
    "C",
    "B",
    "A"
  ),
  ordered_result = TRUE
)

grade_frequency <- table(
  grade
)

cumulative_frequency <- cumsum(
  grade_frequency
)

cumulative_percentage <- cumsum(
  prop.table(
    grade_frequency
  ) * 100
)

data.frame(
  grade = names(
    grade_frequency
  ),
  frequency =
    as.integer(
      grade_frequency
    ),
  cumulative_frequency =
    as.integer(
      cumulative_frequency
    ),
  cumulative_percentage =
    as.numeric(
      cumulative_percentage
    ),
  stringsAsFactors = FALSE
)
```

# 7. Creating a second category

```{r create-result-category}
student_data$result <- ifelse(
  student_data$final_score >= 50,
  "Pass",
  "Fail"
)

table(
  student_data$result
)
```

# 8. Two-way cross-tabulation

```{r two-way-cross-tabulation}
programme_result_table <- table(
  student_data$programme,
  student_data$result
)

programme_result_table
```

# 9. Marginal totals

```{r add-margins}
addmargins(
  programme_result_table
)
```

# 10. Overall proportions

```{r overall-cross-proportions}
prop.table(
  programme_result_table
) * 100
```

# 11. Row percentages

```{r row-cross-percentages}
row_percentages <- prop.table(
  programme_result_table,
  margin = 1
) * 100

row_percentages
```

# 12. Column percentages

```{r column-cross-percentages}
column_percentages <- prop.table(
  programme_result_table,
  margin = 2
) * 100

column_percentages
```

# 13. Verifying percentage totals

```{r verify-row-percentages}
rowSums(
  row_percentages
)

colSums(
  column_percentages
)
```

Minor deviations from 100 may appear after rounding.

# 14. `xtabs()`

```{r xtabs-example}
xtabs(
  ~ programme + result,
  data = student_data
)
```

# 15. Weighted frequencies

```{r weighted-frequency-example}
weighted_data <- data.frame(
  programme = c(
    "Statistics",
    "Procurement",
    "Hospitality"
  ),
  frequency = c(
    35,
    42,
    28
  ),
  stringsAsFactors = FALSE
)

xtabs(
  frequency ~ programme,
  data = weighted_data
)
```

# 16. Group sizes

```{r grouped-sample-size}
programme_n <- table(
  student_data$programme
)

programme_n
```

# 17. Grouped means

```{r grouped-frequency-means}
programme_mean <- tapply(
  student_data$final_score,
  student_data$programme,
  mean,
  na.rm = TRUE
)

programme_mean
```

# 18. Grouped medians

```{r grouped-frequency-medians}
programme_median <- tapply(
  student_data$final_score,
  student_data$programme,
  median,
  na.rm = TRUE
)

programme_median
```

# 19. Grouped standard deviations

```{r grouped-frequency-sd}
programme_sd <- tapply(
  student_data$final_score,
  student_data$programme,
  sd,
  na.rm = TRUE
)

programme_sd
```

# 20. Combined grouped summary

```{r combined-grouped-summary}
programme_descriptive <- data.frame(
  programme = names(
    programme_n
  ),
  n =
    as.integer(
      programme_n
    ),
  mean =
    as.numeric(
      programme_mean[
        names(
          programme_n
        )
      ]
    ),
  median =
    as.numeric(
      programme_median[
        names(
          programme_n
        )
      ]
    ),
  standard_deviation =
    as.numeric(
      programme_sd[
        names(
          programme_n
        )
      ]
    ),
  stringsAsFactors = FALSE
)

programme_descriptive
```

# 21. Grouped pass percentages

```{r grouped-pass-percentage}
pass_percentage <- tapply(
  student_data$final_score >= 50,
  student_data$programme,
  mean,
  na.rm = TRUE
) * 100

pass_percentage
```

# 22. Grouped distinction percentages

```{r grouped-distinction-percentage}
distinction_percentage <- tapply(
  student_data$final_score >= 80,
  student_data$programme,
  mean,
  na.rm = TRUE
) * 100

distinction_percentage
```

# 23. `aggregate()` with custom function

```{r aggregate-custom-function}
aggregate(
  final_score ~ programme,
  data = student_data,
  FUN = function(x) {
    c(
      n = sum(
        !is.na(x)
      ),
      mean = mean(
        x,
        na.rm = TRUE
      ),
      median = median(
        x,
        na.rm = TRUE
      ),
      sd = sd(
        x,
        na.rm = TRUE
      )
    )
  }
)
```

# 24. Splitting and applying

```{r split-apply-example}
scores_by_programme <- split(
  student_data$final_score,
  student_data$programme
)

lapply(
  scores_by_programme,
  summary
)
```

# 25. Reusable one-way table function

```{r one-way-table-function}
create_one_way_table <- function(
  x,
  variable_name = "category",
  include_missing = TRUE
) {
  counts <- table(
    x,
    useNA = if (
      include_missing
    ) {
      "ifany"
    } else {
      "no"
    }
  )

  result <- data.frame(
    category = names(counts),
    frequency =
      as.integer(counts),
    percentage =
      as.numeric(
        counts /
          sum(counts) *
          100
      ),
    stringsAsFactors = FALSE
  )

  names(result)[1] <-
    variable_name

  result
}

create_one_way_table(
  student_data$programme,
  "programme"
)
```

# 26. Reusable cross-tabulation function

```{r cross-tabulation-function}
create_cross_tabulation <- function(
  row_variable,
  column_variable,
  percentage = c(
    "count",
    "row",
    "column",
    "overall"
  )
) {
  percentage <- match.arg(
    percentage
  )

  tab <- table(
    row_variable,
    column_variable,
    useNA = "ifany"
  )

  switch(
    percentage,
    count = tab,
    row = prop.table(
      tab,
      margin = 1
    ) * 100,
    column = prop.table(
      tab,
      margin = 2
    ) * 100,
    overall = prop.table(
      tab
    ) * 100
  )
}

create_cross_tabulation(
  student_data$programme,
  student_data$result,
  percentage = "row"
)
```

# 27. Grouped summary function

```{r tabulation-grouped-summary-function}
create_grouped_summary <- function(
  data,
  group_variable,
  numeric_variable
) {
  groups <- split(
    data[[
      numeric_variable
    ]],
    data[[
      group_variable
    ]]
  )

  results <- lapply(
    names(groups),
    function(group_name) {
      x <- groups[[
        group_name
      ]]

      data.frame(
        group = group_name,
        n = sum(
          !is.na(x)
        ),
        missing = sum(
          is.na(x)
        ),
        mean = mean(
          x,
          na.rm = TRUE
        ),
        median = median(
          x,
          na.rm = TRUE
        ),
        standard_deviation = sd(
          x,
          na.rm = TRUE
        ),
        stringsAsFactors = FALSE
      )
    }
  )

  result <- do.call(
    rbind,
    results
  )

  names(result)[1] <-
    group_variable

  result
}

create_grouped_summary(
  student_data,
  "programme",
  "final_score"
)
```

# 28. Converting a table to a data frame

```{r table-to-data-frame}
programme_result_data <- as.data.frame(
  programme_result_table,
  stringsAsFactors = FALSE
)

programme_result_data
```

# 29. Clinic grouped summaries

```{r clinic-grouped-summary}
clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)

create_grouped_summary(
  clinic_data,
  "sex",
  "waiting_minutes"
)
```

# 30. Interpretation principles

A cross-tabulation interpretation should state:

- which percentage base is used;
- the relevant denominator;
- the largest or smallest category;
- the magnitude of differences;
- and whether missing values were included.

# 31. Common mistakes

- reporting percentages without frequencies;
- failing to state row or column percentages;
- using cumulative percentages for unordered categories;
- excluding missing values silently;
- comparing group means without group sizes;
- rounding percentages before verification;
- treating descriptive differences as statistically significant;
- using the wrong denominator;
- confusing a table margin with a percentage margin;
- failing to check that percentages total 100.

# 32. Independent exercises

1. create a one-way frequency table.
2. include missing categories.
3. calculate frequency percentages.
4. create cumulative frequencies for ordered grades.
5. create a two-way table.
6. add margins.
7. calculate row, column and overall percentages.
8. verify percentage totals.
9. calculate grouped sample sizes.
10. calculate grouped means, medians and standard deviations.
11. calculate pass and distinction percentages.
12. create reusable one-way and cross-tabulation functions.
13. create grouped summaries for clinic data.
14. write a clear interpretation.

# 33. Lesson assessment

Create a standalone R Markdown report that:

1. produces one-way frequency tables;
2. reports frequencies and percentages;
3. handles missing categories explicitly;
4. creates an ordered cumulative table;
5. creates a two-way cross-tabulation;
6. reports row and column percentages;
7. validates totals;
8. calculates grouped numerical summaries;
9. creates reusable tabulation functions; and
10. interprets all tables accurately.

# 34. Completion checklist

Confirm that you can:

- create frequency tables;
- include missing values;
- calculate percentages;
- create cumulative frequencies;
- create cross-tabulations;
- add margins;
- calculate row percentages;
- calculate column percentages;
- verify totals;
- calculate grouped numerical summaries;
- calculate grouped proportions;
- convert tables to data frames; and
- interpret grouped tables.

# Lesson conclusion

Frequency tables and grouped summaries organise categorical and numerical
information into interpretable patterns. Every percentage must have a clear
denominator.

\[
\boxed{
\text{count}
\rightarrow
\text{calculate denominator}
\rightarrow
\text{convert to percentage}
\rightarrow
\text{verify}
\rightarrow
\text{interpret}
}
\]
