---
title: "Lesson 26: Data Visualisation with Base R"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Data visualisation helps the analyst inspect distributions, compare groups,
identify unusual values and communicate findings.

A graph should be chosen according to the variable type and analytical purpose.
The goal is not decoration. The goal is accurate and efficient communication.

This lesson introduces base R graphics for:

- bar charts;
- pie charts;
- histograms;
- box plots;
- scatter plots;
- line graphs;
- dot charts;
- grouped displays;
- and multi-panel layouts.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. choose a graph for a variable type;
2. create bar charts;
3. create histograms;
4. create box plots;
5. create scatter plots;
6. create line graphs;
7. create dot charts;
8. add titles, labels and legends;
9. control axes and plotting symbols;
10. compare groups visually;
11. create multi-panel displays;
12. save graphs to files;
13. avoid misleading graphics; and
14. interpret graphs in context.

# 1. Importing data

```{r import-visualisation-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)
```

# 2. Choosing a graph

| Purpose | Common graph |
|---|---|
| category frequencies | bar chart |
| numerical distribution | histogram |
| numerical distribution and outliers | box plot |
| relationship between two numerical variables | scatter plot |
| trend over ordered time | line graph |
| compact category comparison | dot chart |

# 3. Bar chart

```{r basic-bar-chart}
programme_counts <- table(
  student_data$programme
)

barplot(
  programme_counts,
  main = "Students by Programme",
  xlab = "Programme",
  ylab = "Number of students",
  las = 2
)
```

# 4. Horizontal bar chart

```{r horizontal-bar-chart}
barplot(
  programme_counts,
  horiz = TRUE,
  main = "Students by Programme",
  xlab = "Number of students",
  ylab = "Programme",
  las = 1
)
```

# 5. Ordered bar chart

```{r ordered-bar-chart}
ordered_programme_counts <- sort(
  programme_counts,
  decreasing = TRUE
)

barplot(
  ordered_programme_counts,
  main = "Students by Programme",
  xlab = "Programme",
  ylab = "Number of students",
  las = 2
)
```

# 6. Grouped bar chart

```{r grouped-bar-chart}
student_data$result <- ifelse(
  student_data$final_score >= 50,
  "Pass",
  "Fail"
)

programme_result_table <- table(
  student_data$result,
  student_data$programme
)

barplot(
  programme_result_table,
  beside = TRUE,
  main = "Results by Programme",
  xlab = "Programme",
  ylab = "Number of students",
  legend.text = rownames(
    programme_result_table
  ),
  args.legend = list(
    x = "topright"
  ),
  las = 2
)
```

# 7. Stacked bar chart

```{r stacked-bar-chart}
barplot(
  programme_result_table,
  beside = FALSE,
  main = "Results by Programme",
  xlab = "Programme",
  ylab = "Number of students",
  legend.text = rownames(
    programme_result_table
  ),
  args.legend = list(
    x = "topright"
  ),
  las = 2
)
```

# 8. Percentage bar chart

```{r percentage-bar-chart}
programme_result_percentage <- prop.table(
  programme_result_table,
  margin = 2
) * 100

barplot(
  programme_result_percentage,
  beside = FALSE,
  main = "Percentage Results by Programme",
  xlab = "Programme",
  ylab = "Percentage",
  legend.text = rownames(
    programme_result_percentage
  ),
  args.legend = list(
    x = "topright"
  ),
  las = 2
)
```

# 9. Pie chart

```{r basic-pie-chart}
pie(
  programme_counts,
  main = "Programme Distribution"
)
```

Pie charts are less effective when there are many categories or similar
proportions.

# 10. Histogram

```{r basic-histogram}
hist(
  student_data$final_score,
  main = "Distribution of Final Scores",
  xlab = "Final score",
  ylab = "Frequency"
)
```

# 11. Histogram breaks

```{r histogram-breaks}
hist(
  student_data$final_score,
  breaks = 10,
  main = "Distribution of Final Scores",
  xlab = "Final score",
  ylab = "Frequency"
)
```

The choice of breaks affects the displayed shape.

# 12. Density-scaled histogram

```{r density-histogram}
hist(
  student_data$final_score,
  probability = TRUE,
  main = "Density of Final Scores",
  xlab = "Final score",
  ylab = "Density"
)

lines(
  density(
    student_data$final_score,
    na.rm = TRUE
  )
)
```

# 13. Box plot

```{r basic-boxplot}
boxplot(
  student_data$final_score,
  main = "Box Plot of Final Scores",
  ylab = "Final score"
)
```

# 14. Grouped box plot

```{r grouped-boxplot}
boxplot(
  final_score ~ programme,
  data = student_data,
  main = "Final Scores by Programme",
  xlab = "Programme",
  ylab = "Final score",
  las = 2
)
```

# 15. Extracting box-plot statistics

```{r boxplot-statistics}
boxplot.stats(
  student_data$final_score
)
```

# 16. Scatter plot

```{r basic-scatterplot}
plot(
  student_data$coursework_score,
  student_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)
```

# 17. Adding a fitted line

```{r scatterplot-fitted-line}
plot(
  student_data$coursework_score,
  student_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)

fitted_model <- lm(
  examination_score ~ coursework_score,
  data = student_data
)

abline(
  fitted_model,
  lwd = 2
)
```

# 18. Adding reference lines

```{r scatterplot-reference-lines}
plot(
  student_data$coursework_score,
  student_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)

abline(
  v = mean(
    student_data$coursework_score
  ),
  lty = 2
)

abline(
  h = mean(
    student_data$examination_score
  ),
  lty = 2
)
```

# 19. Line graph

```{r line-graph-example}
monthly_values <- c(
  12,
  15,
  14,
  18,
  20,
  22
)

month_number <- 1:6

plot(
  month_number,
  monthly_values,
  type = "o",
  main = "Monthly Trend",
  xlab = "Month",
  ylab = "Value",
  xaxt = "n"
)

axis(
  1,
  at = month_number,
  labels = month.abb[
    month_number
  ]
)
```

# 20. Dot chart

```{r dot-chart-example}
programme_means <- tapply(
  student_data$final_score,
  student_data$programme,
  mean,
  na.rm = TRUE
)

dotchart(
  programme_means,
  main = "Mean Final Score by Programme",
  xlab = "Mean final score"
)
```

# 21. Stem-and-leaf display

```{r stem-and-leaf}
stem(
  student_data$final_score
)
```

# 22. Empirical cumulative distribution

```{r empirical-cdf}
plot(
  ecdf(
    student_data$final_score
  ),
  main = "Empirical Cumulative Distribution of Scores",
  xlab = "Final score",
  ylab = "Cumulative proportion"
)
```

# 23. Plotting missingness counts

```{r visualise-missingness}
missing_counts <- colSums(
  is.na(
    clinic_data
  )
)

barplot(
  missing_counts,
  main = "Missing Values by Clinic Variable",
  xlab = "Variable",
  ylab = "Number missing",
  las = 2
)
```

# 24. Multi-panel graphics

```{r multi-panel-graphics}
old_parameters <- par(
  mfrow = c(
    2,
    2
  )
)

hist(
  student_data$final_score,
  main = "Histogram",
  xlab = "Final score"
)

boxplot(
  student_data$final_score,
  main = "Box Plot",
  ylab = "Final score"
)

plot(
  student_data$coursework_score,
  student_data$examination_score,
  main = "Scatter Plot",
  xlab = "Coursework",
  ylab = "Examination",
  pch = 19
)

barplot(
  programme_counts,
  main = "Programme Counts",
  las = 2
)

par(
  old_parameters
)
```

# 25. Axis limits

```{r plot-axis-limits}
plot(
  student_data$coursework_score,
  student_data$examination_score,
  xlim = c(
    0,
    40
  ),
  ylim = c(
    0,
    60
  ),
  main = "Assessment Components",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)
```

Axis limits should not be used to conceal relevant data.

# 26. Text annotations

```{r plot-text-annotations}
plot(
  student_data$coursework_score,
  student_data$examination_score,
  main = "Assessment Components",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)

highest_index <- which.max(
  student_data$final_score
)

text(
  student_data$coursework_score[
    highest_index
  ],
  student_data$examination_score[
    highest_index
  ],
  labels =
    student_data$student_id[
      highest_index
    ],
  pos = 3
)
```

# 27. Saving a PNG graph

```{r save-png-graph}
output_directory <- "lesson_26_output"

if (!dir.exists(
  output_directory
)) {
  dir.create(
    output_directory
  )
}

png_path <- file.path(
  output_directory,
  "final_score_histogram.png"
)

png(
  filename = png_path,
  width = 1200,
  height = 800,
  res = 150
)

hist(
  student_data$final_score,
  main = "Distribution of Final Scores",
  xlab = "Final score",
  ylab = "Frequency"
)

dev.off()

file.exists(
  png_path
)
```

# 28. Saving a PDF graph

```{r save-pdf-graph}
pdf_path <- file.path(
  output_directory,
  "final_score_boxplot.pdf"
)

pdf(
  file = pdf_path,
  width = 7,
  height = 5
)

boxplot(
  final_score ~ programme,
  data = student_data,
  main = "Final Scores by Programme",
  xlab = "Programme",
  ylab = "Final score",
  las = 2
)

dev.off()

file.exists(
  pdf_path
)
```

Cleanup:

```{r cleanup-visualisation-output}
unlink(
  output_directory,
  recursive = TRUE,
  force = TRUE
)
```

# 29. Reusable histogram function

```{r histogram-function}
plot_histogram <- function(
  x,
  title,
  x_label,
  breaks = "Sturges"
) {
  if (!is.numeric(x)) {
    stop(
      "`x` must be numeric.",
      call. = FALSE
    )
  }

  hist(
    x,
    breaks = breaks,
    main = title,
    xlab = x_label,
    ylab = "Frequency"
  )

  invisible(NULL)
}

plot_histogram(
  student_data$final_score,
  title = "Distribution of Final Scores",
  x_label = "Final score"
)
```

# 30. Reusable grouped box-plot function

```{r grouped-boxplot-function}
plot_grouped_boxplot <- function(
  data,
  numeric_variable,
  group_variable,
  title = NULL
) {
  if (!all(
    c(
      numeric_variable,
      group_variable
    ) %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  formula_object <- reformulate(
    group_variable,
    response =
      numeric_variable
  )

  boxplot(
    formula_object,
    data = data,
    main = title,
    xlab = group_variable,
    ylab = numeric_variable,
    las = 2
  )

  invisible(NULL)
}

plot_grouped_boxplot(
  student_data,
  "final_score",
  "programme",
  "Final Scores by Programme"
)
```

# 31. Misleading graph practices

Avoid:

- truncated axes that exaggerate differences;
- unequal category spacing;
- three-dimensional effects;
- excessive decoration;
- too many colours or symbols;
- unreadable labels;
- inconsistent scales across panels;
- graphs without denominators;
- hiding missing values;
- using line graphs for unordered categories.

# 32. Graph interpretation

A useful interpretation identifies:

- the variable or relationship shown;
- the dominant pattern;
- variation;
- unusual observations;
- group differences;
- and limitations of the graph.

# 33. Common mistakes

- choosing a graph unsuitable for the variable type;
- using a pie chart with too many categories;
- treating histogram bars as categories;
- interpreting box-plot outliers as errors;
- drawing a line through unordered observations;
- adding a fitted line without interpretation;
- failing to label axes;
- saving low-resolution images;
- forgetting to close graphics devices;
- creating decorative rather than analytical graphs.

# 34. Independent exercises

1. create vertical and horizontal bar charts.
2. create ordered frequency charts.
3. create grouped and stacked bar charts.
4. create a histogram with different breaks.
5. create a density-scaled histogram.
6. create a box plot and grouped box plot.
7. create a scatter plot with a fitted line.
8. create a line graph.
9. create a dot chart.
10. create an empirical cumulative distribution.
11. create a multi-panel display.
12. annotate an unusual point.
13. save PNG and PDF graphs.
14. write reusable plotting functions.
15. critique misleading graph practices.

# 35. Lesson assessment

Create a standalone R Markdown report that:

1. chooses graphs based on variable type;
2. produces bar charts;
3. produces a histogram;
4. produces a grouped box plot;
5. produces a scatter plot with a fitted line;
6. produces a line graph;
7. creates a multi-panel display;
8. saves at least two graph formats;
9. uses reusable plotting functions; and
10. interprets every graph.

# 36. Completion checklist

Confirm that you can:

- choose an appropriate graph;
- create bar charts;
- create histograms;
- create box plots;
- create scatter plots;
- create line graphs;
- create dot charts;
- create grouped displays;
- add labels and legends;
- annotate points;
- create multi-panel graphics;
- save graphics;
- avoid misleading designs; and
- interpret visual patterns.

# Lesson conclusion

A graph is an analytical argument presented visually. It must match the data,
show the relevant pattern and avoid distortions.

\[
\boxed{
\text{define purpose}
\rightarrow
\text{identify variable types}
\rightarrow
\text{choose graph}
\rightarrow
\text{label clearly}
\rightarrow
\text{interpret}
}
\]
