---
title: "Lesson 27: Data Visualisation with ggplot2"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

`ggplot2` is a widely used R package for building graphics from a consistent
grammar. A plot is constructed by combining:

- data;
- aesthetic mappings;
- geometric objects;
- statistical transformations;
- scales;
- coordinate systems;
- facets;
- and themes.

This approach separates what the graph shows from how it is displayed.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain the grammar of graphics;
2. create a basic `ggplot` object;
3. map variables to aesthetics;
4. distinguish mapped and fixed aesthetics;
5. create bar charts, histograms, box plots and scatter plots;
6. add fitted lines;
7. create grouped plots;
8. use facets;
9. control scales and labels;
10. apply themes;
11. annotate graphs;
12. save plots with `ggsave()`;
13. create reusable plotting functions; and
14. avoid misleading designs.

# 1. Package availability

```{r check-ggplot2}
ggplot2_available <- requireNamespace(
  "ggplot2",
  quietly = TRUE
)

ggplot2_available
```

Install once when needed:

```{r install-ggplot2, eval=FALSE}
install.packages(
  "ggplot2"
)
```

# 2. Importing data

```{r import-ggplot-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)
```

# 3. Grammar of graphics

A typical plot has the form:

```r
ggplot2::ggplot(
  data,
  ggplot2::aes(
    x = variable_x,
    y = variable_y
  )
) +
  ggplot2::geom_point()
```

# 4. Basic scatter plot

```{r ggplot-basic-scatter, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score
  )
) +
  ggplot2::geom_point()
```

# 5. Mapped versus fixed aesthetics

Mapped:

```{r ggplot-mapped-aesthetic, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score,
    shape = programme
  )
) +
  ggplot2::geom_point()
```

Fixed:

```{r ggplot-fixed-aesthetic, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score
  )
) +
  ggplot2::geom_point(
    size = 3
  )
```

# 6. Bar chart of counts

```{r ggplot-bar-counts, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = programme
  )
) +
  ggplot2::geom_bar() +
  ggplot2::labs(
    title = "Students by Programme",
    x = "Programme",
    y = "Number of students"
  )
```

# 7. Pre-computed bar chart

```{r ggplot-precomputed-bar, eval=FALSE}
programme_summary <- as.data.frame(
  table(
    student_data$programme
  ),
  stringsAsFactors = FALSE
)

names(programme_summary) <- c(
  "programme",
  "frequency"
)

ggplot2::ggplot(
  programme_summary,
  ggplot2::aes(
    x = programme,
    y = frequency
  )
) +
  ggplot2::geom_col()
```

# 8. Horizontal bars

```{r ggplot-horizontal-bars, eval=FALSE}
ggplot2::ggplot(
  programme_summary,
  ggplot2::aes(
    x = programme,
    y = frequency
  )
) +
  ggplot2::geom_col() +
  ggplot2::coord_flip()
```

# 9. Histogram

```{r ggplot-histogram, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = final_score
  )
) +
  ggplot2::geom_histogram(
    bins = 10
  ) +
  ggplot2::labs(
    title = "Distribution of Final Scores",
    x = "Final score",
    y = "Frequency"
  )
```

# 10. Histogram with bin width

```{r ggplot-binwidth, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = final_score
  )
) +
  ggplot2::geom_histogram(
    binwidth = 5,
    boundary = 0
  )
```

# 11. Density plot

```{r ggplot-density, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = final_score
  )
) +
  ggplot2::geom_density()
```

# 12. Box plot

```{r ggplot-boxplot, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    y = final_score
  )
) +
  ggplot2::geom_boxplot()
```

# 13. Grouped box plot

```{r ggplot-grouped-boxplot, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = programme,
    y = final_score
  )
) +
  ggplot2::geom_boxplot() +
  ggplot2::labs(
    title = "Final Scores by Programme",
    x = "Programme",
    y = "Final score"
  )
```

# 14. Scatter plot with fitted line

```{r ggplot-fitted-line, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score
  )
) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(
    method = "lm",
    se = TRUE
  )
```

# 15. Group mapping

```{r ggplot-group-mapping, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score,
    shape = programme
  )
) +
  ggplot2::geom_point(
    size = 3
  )
```

# 16. Position adjustments

```{r ggplot-position-dodge, eval=FALSE}
student_data$result <- ifelse(
  student_data$final_score >= 50,
  "Pass",
  "Fail"
)

ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = programme,
    fill = result
  )
) +
  ggplot2::geom_bar(
    position = "dodge"
  )
```

Stacked proportions:

```{r ggplot-position-fill, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = programme,
    fill = result
  )
) +
  ggplot2::geom_bar(
    position = "fill"
  ) +
  ggplot2::labs(
    y = "Proportion"
  )
```

# 17. Faceting

```{r ggplot-facet-wrap, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = final_score
  )
) +
  ggplot2::geom_histogram(
    bins = 8
  ) +
  ggplot2::facet_wrap(
    ~ programme
  )
```

# 18. Facet grid

```{r ggplot-facet-grid, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score
  )
) +
  ggplot2::geom_point() +
  ggplot2::facet_grid(
    rows = ggplot2::vars(
      programme
    )
  )
```

# 19. Labels

```{r ggplot-labels, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score
  )
) +
  ggplot2::geom_point() +
  ggplot2::labs(
    title = "Relationship between Assessment Components",
    subtitle = "Student assessment records",
    x = "Coursework score",
    y = "Examination score",
    caption = "Source: Synthetic teaching dataset"
  )
```

# 20. Themes

```{r ggplot-themes, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = final_score
  )
) +
  ggplot2::geom_histogram(
    bins = 10
  ) +
  ggplot2::theme_minimal()
```

Other built-in themes include:

- `theme_classic()`;
- `theme_bw()`;
- `theme_light()`;
- `theme_void()`.

# 21. Rotating axis text

```{r ggplot-rotate-axis, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = programme
  )
) +
  ggplot2::geom_bar() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 45,
      hjust = 1
    )
  )
```

# 22. Continuous scales

```{r ggplot-continuous-scale, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score
  )
) +
  ggplot2::geom_point() +
  ggplot2::scale_x_continuous(
    limits = c(
      0,
      40
    ),
    breaks = seq(
      0,
      40,
      10
    )
  ) +
  ggplot2::scale_y_continuous(
    limits = c(
      0,
      60
    ),
    breaks = seq(
      0,
      60,
      10
    )
  )
```

# 23. Discrete scales

```{r ggplot-discrete-scale, eval=FALSE}
ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = programme
  )
) +
  ggplot2::geom_bar() +
  ggplot2::scale_x_discrete(
    limits = c(
      "Statistics",
      "Procurement",
      "Hospitality",
      "Agriculture"
    )
  )
```

# 24. Annotation

```{r ggplot-annotation, eval=FALSE}
highest_index <- which.max(
  student_data$final_score
)

highest_student <- student_data[
  highest_index,
]

ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = coursework_score,
    y = examination_score
  )
) +
  ggplot2::geom_point() +
  ggplot2::annotate(
    "text",
    x = highest_student$coursework_score,
    y = highest_student$examination_score,
    label = highest_student$student_id,
    vjust = -1
  )
```

# 25. Saving a plot object

```{r ggplot-save-object, eval=FALSE}
score_plot <- ggplot2::ggplot(
  student_data,
  ggplot2::aes(
    x = final_score
  )
) +
  ggplot2::geom_histogram(
    bins = 10
  ) +
  ggplot2::labs(
    title = "Distribution of Final Scores",
    x = "Final score",
    y = "Frequency"
  )

score_plot
```

# 26. Saving with `ggsave()`

```{r ggsave-example, eval=FALSE}
ggplot2::ggsave(
  filename = "final_score_histogram.png",
  plot = score_plot,
  width = 7,
  height = 5,
  dpi = 300
)
```

# 27. Reusable plotting function

```{r ggplot-function, eval=FALSE}
create_scatter_plot <- function(
  data,
  x_variable,
  y_variable,
  title = NULL
) {
  if (!all(
    c(
      x_variable,
      y_variable
    ) %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  ggplot2::ggplot(
    data,
    ggplot2::aes(
      x = .data[[
        x_variable
      ]],
      y = .data[[
        y_variable
      ]]
    )
  ) +
    ggplot2::geom_point() +
    ggplot2::geom_smooth(
      method = "lm"
    ) +
    ggplot2::labs(
      title = title,
      x = x_variable,
      y = y_variable
    )
}
```

# 28. Safe execution when `ggplot2` is absent

```{r safe-ggplot-execution}
if (ggplot2_available) {
  basic_plot <- ggplot2::ggplot(
    student_data,
    ggplot2::aes(
      x = final_score
    )
  ) +
    ggplot2::geom_histogram(
      bins = 10
    )

  print(basic_plot)
}
```

# 29. Common mistakes

- placing fixed values inside `aes()`;
- placing mapped variables outside `aes()`;
- using `geom_bar()` when values are already summarised;
- ignoring the denominator in stacked proportion charts;
- adding too many aesthetics;
- using facets with too few observations;
- truncating scales misleadingly;
- failing to label axes;
- saving without specifying size or resolution;
- treating theme changes as substantive analysis.

# 30. Independent exercises

1. check and install `ggplot2`.
2. create a basic scatter plot.
3. demonstrate mapped and fixed aesthetics.
4. create count and pre-computed bar charts.
5. create histograms and density plots.
6. create grouped box plots.
7. add fitted lines.
8. create grouped and proportional bar charts.
9. use faceting.
10. add labels and themes.
11. adjust scales.
12. annotate a point.
13. save a plot.
14. create a reusable plotting function.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. explains the grammar of graphics;
2. creates at least six `ggplot2` graphs;
3. distinguishes mapped and fixed aesthetics;
4. uses grouping and faceting;
5. controls labels and scales;
6. applies a suitable theme;
7. annotates an important point;
8. saves a high-resolution graph;
9. uses a reusable plotting function; and
10. interprets each graph.

# 32. Completion checklist

Confirm that you can:

- create `ggplot` objects;
- map aesthetics;
- fix aesthetic values;
- create bar charts;
- create histograms;
- create box plots;
- create scatter plots;
- add fitted lines;
- use groups and facets;
- control scales and labels;
- apply themes;
- annotate graphs;
- save plots; and
- create reusable graph functions.

# Lesson conclusion

`ggplot2` builds graphics by combining meaningful layers. A successful plot
begins with a clear analytical purpose and ends with an accurate visual
interpretation.

\[
\boxed{
\text{data}
+
\text{aesthetics}
+
\text{geometry}
+
\text{scales}
+
\text{theme}
}
\]
