---
title: "Lesson 28: Distribution Assessment and Exploratory Data Analysis"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Exploratory data analysis is the systematic process of understanding data
before formal modelling. It combines numerical summaries, visual displays and
quality checks.

Distribution assessment asks:

- where the data are centred;
- how widely values vary;
- whether the distribution is symmetric;
- whether there are multiple peaks;
- whether extreme observations exist;
- whether transformations may be useful;
- and whether patterns differ across groups.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain exploratory data analysis;
2. inspect numerical distributions;
3. compare mean and median;
4. assess skewness visually and numerically;
5. assess possible outliers;
6. use histograms, density plots and box plots;
7. use quantile-quantile plots;
8. perform normality tests cautiously;
9. compare distributions across groups;
10. assess transformations;
11. identify data-quality concerns;
12. create an EDA function;
13. document exploratory findings; and
14. distinguish exploration from confirmation.

# 1. Importing data

```{r import-eda-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

final_scores <- student_data$final_score
```

# 2. Initial summary

```{r eda-initial-summary}
summary(
  final_scores
)

mean(
  final_scores,
  na.rm = TRUE
)

median(
  final_scores,
  na.rm = TRUE
)

sd(
  final_scores,
  na.rm = TRUE
)

IQR(
  final_scores,
  na.rm = TRUE
)
```

# 3. Shape from mean and median

```{r eda-mean-median}
mean_score <- mean(
  final_scores,
  na.rm = TRUE
)

median_score <- median(
  final_scores,
  na.rm = TRUE
)

mean_score
median_score
mean_score - median_score
```

The difference offers only a preliminary indication of asymmetry.

# 4. Histogram

```{r eda-histogram}
hist(
  final_scores,
  main = "Distribution of Final Scores",
  xlab = "Final score",
  ylab = "Frequency"
)
```

# 5. Alternative histogram breaks

```{r eda-histogram-breaks}
hist(
  final_scores,
  breaks = 5,
  main = "Final Scores: Five Bins",
  xlab = "Final score"
)

hist(
  final_scores,
  breaks = 12,
  main = "Final Scores: Twelve Bins",
  xlab = "Final score"
)
```

# 6. Density plot

```{r eda-density-plot}
plot(
  density(
    final_scores,
    na.rm = TRUE
  ),
  main = "Density of Final Scores",
  xlab = "Final score",
  ylab = "Density"
)
```

# 7. Box plot

```{r eda-boxplot}
boxplot(
  final_scores,
  main = "Box Plot of Final Scores",
  ylab = "Final score"
)
```

# 8. Box-plot statistics

```{r eda-boxplot-stats}
boxplot.stats(
  final_scores
)
```

# 9. IQR outlier rule

```{r eda-iqr-outliers}
q1 <- quantile(
  final_scores,
  0.25,
  na.rm = TRUE
)

q3 <- quantile(
  final_scores,
  0.75,
  na.rm = TRUE
)

iqr_value <- q3 - q1

lower_fence <- q1 - 1.5 * iqr_value
upper_fence <- q3 + 1.5 * iqr_value

iqr_outlier <- final_scores < lower_fence |
  final_scores > upper_fence

student_data[
  iqr_outlier,
]
```

# 10. Z-score screening

```{r eda-z-screen}
z_scores <- as.numeric(
  scale(
    final_scores
  )
)

extreme_z <- abs(
  z_scores
) > 3

student_data[
  extreme_z,
]
```

# 11. Skewness function

```{r eda-skewness-function}
calculate_skewness <- function(
  x
) {
  observed <- x[
    !is.na(x)
  ]

  n <- length(observed)

  if (n < 3L) {
    return(NA_real_)
  }

  centre <- mean(observed)
  spread <- sd(observed)

  if (spread == 0) {
    return(0)
  }

  mean(
    (
      (observed - centre) /
        spread
    ) ^ 3
  )
}

calculate_skewness(
  final_scores
)
```

# 12. Kurtosis function

```{r eda-kurtosis-function}
calculate_excess_kurtosis <- function(
  x
) {
  observed <- x[
    !is.na(x)
  ]

  n <- length(observed)

  if (n < 4L) {
    return(NA_real_)
  }

  centre <- mean(observed)
  spread <- sd(observed)

  if (spread == 0) {
    return(NA_real_)
  }

  mean(
    (
      (observed - centre) /
        spread
    ) ^ 4
  ) - 3
}

calculate_excess_kurtosis(
  final_scores
)
```

# 13. Quantile-quantile plot

```{r eda-qq-plot}
qqnorm(
  final_scores,
  main = "Normal Q-Q Plot of Final Scores"
)

qqline(
  final_scores
)
```

Points close to the line are broadly consistent with normality.

# 14. Shapiro-Wilk test

```{r eda-shapiro-wilk}
shapiro.test(
  final_scores
)
```

Interpretation:

- null hypothesis: the data are normally distributed;
- small p-values indicate evidence against normality;
- large p-values do not prove normality.

# 15. Normality tests and sample size

Normality tests may:

- have low power in small samples;
- detect trivial deviations in very large samples;
- be affected by ties and rounding.

Use tests together with plots and contextual judgement.

# 16. Grouped box plots

```{r eda-grouped-boxplots}
boxplot(
  final_score ~ programme,
  data = student_data,
  main = "Final Scores by Programme",
  xlab = "Programme",
  ylab = "Final score",
  las = 2
)
```

# 17. Grouped histograms

```{r eda-grouped-histograms}
programme_names <- unique(
  student_data$programme
)

old_parameters <- par(
  mfrow = c(
    2,
    2
  )
)

for (programme_name in programme_names) {
  selected_scores <- student_data$final_score[
    student_data$programme ==
      programme_name
  ]

  hist(
    selected_scores,
    main = programme_name,
    xlab = "Final score"
  )
}

par(
  old_parameters
)
```

# 18. Group summaries

```{r eda-group-summary}
aggregate(
  final_score ~ programme,
  data = student_data,
  FUN = function(x) {
    c(
      n = length(x),
      mean = mean(x),
      median = median(x),
      sd = sd(x),
      iqr = IQR(x),
      skewness =
        calculate_skewness(x)
    )
  }
)
```

# 19. Transformation assessment

Create positive skewed data:

```{r eda-skewed-data}
skewed_values <- rexp(
  n = 200,
  rate = 0.10
)

summary(
  skewed_values
)

calculate_skewness(
  skewed_values
)
```

# 20. Log transformation

```{r eda-log-transform}
log_values <- log(
  skewed_values
)

calculate_skewness(
  log_values
)
```

Compare histograms:

```{r eda-transform-comparison}
old_parameters <- par(
  mfrow = c(
    1,
    2
  )
)

hist(
  skewed_values,
  main = "Original Values",
  xlab = "Value"
)

hist(
  log_values,
  main = "Log-Transformed Values",
  xlab = "Log value"
)

par(
  old_parameters
)
```

# 21. Square-root transformation

```{r eda-square-root-transform}
square_root_values <- sqrt(
  skewed_values
)

calculate_skewness(
  square_root_values
)
```

# 22. Transformation cautions

A transformation should:

- match the measurement scale;
- preserve substantive meaning where possible;
- be applied before modelling for a clear reason;
- not be used merely to force a desired result;
- be documented and interpreted on the correct scale.

# 23. Missingness in EDA

```{r eda-missingness}
missing_summary <- data.frame(
  variable = names(
    student_data
  ),
  missing = vapply(
    student_data,
    function(x) {
      sum(
        is.na(x)
      )
    },
    FUN.VALUE = integer(1)
  ),
  missing_percentage = vapply(
    student_data,
    function(x) {
      mean(
        is.na(x)
      ) * 100
    },
    FUN.VALUE = numeric(1)
  ),
  stringsAsFactors = FALSE
)

missing_summary
```

# 24. Duplicate identifiers

```{r eda-duplicate-identifiers}
duplicated_student_ids <- duplicated(
  student_data$student_id
) |
  duplicated(
    student_data$student_id,
    fromLast = TRUE
  )

student_data[
  duplicated_student_ids,
]
```

# 25. Range and consistency checks

```{r eda-range-checks}
range_checks <- list(
  coursework_valid = all(
    student_data$coursework_score >= 0 &
      student_data$coursework_score <= 40
  ),
  examination_valid = all(
    student_data$examination_score >= 0 &
      student_data$examination_score <= 60
  ),
  final_valid = all(
    student_data$final_score >= 0 &
      student_data$final_score <= 100
  ),
  arithmetic_valid = all(
    student_data$final_score ==
      student_data$coursework_score +
      student_data$examination_score
  )
)

range_checks
```

# 26. Reusable EDA summary function

```{r eda-summary-function}
create_eda_summary <- function(
  x,
  variable_name = "x"
) {
  if (!is.numeric(x)) {
    stop(
      "`x` must be numeric.",
      call. = FALSE
    )
  }

  observed <- x[
    !is.na(x)
  ]

  q <- quantile(
    observed,
    c(
      0.25,
      0.50,
      0.75
    )
  )

  data.frame(
    variable = variable_name,
    n = length(observed),
    missing = sum(
      is.na(x)
    ),
    mean = mean(observed),
    median = median(observed),
    standard_deviation = sd(observed),
    minimum = min(observed),
    first_quartile = q[1],
    third_quartile = q[3],
    maximum = max(observed),
    skewness =
      calculate_skewness(
        observed
      ),
    excess_kurtosis =
      calculate_excess_kurtosis(
        observed
      ),
    stringsAsFactors = FALSE,
    row.names = NULL
  )
}

create_eda_summary(
  final_scores,
  "final_score"
)
```

# 27. EDA report for several variables

```{r eda-several-variables}
numeric_variables <- names(
  student_data
)[
  vapply(
    student_data,
    is.numeric,
    FUN.VALUE = logical(1)
  )
]

eda_tables <- lapply(
  numeric_variables,
  function(variable_name) {
    create_eda_summary(
      student_data[[
        variable_name
      ]],
      variable_name
    )
  }
)

do.call(
  rbind,
  eda_tables
)
```

# 28. Exploration versus confirmation

Exploratory analysis:

- discovers patterns;
- generates questions;
- checks assumptions;
- identifies anomalies.

Confirmatory analysis:

- evaluates pre-specified hypotheses;
- uses inferential procedures;
- controls error rates;
- requires planned interpretation.

The same dataset can support both stages, but the distinction should be clear.

# 29. Common mistakes

- relying only on a normality test;
- declaring data normal because p > 0.05;
- treating all outliers as errors;
- ignoring small group sizes;
- applying transformations without justification;
- using too many histogram bins;
- interpreting exploratory patterns as confirmed effects;
- overlooking missingness and duplication;
- checking distribution only for the pooled sample;
- failing to document exploratory decisions.

# 30. Independent exercises

1. create a complete numerical summary.
2. compare mean and median.
3. create histograms with different breaks.
4. create density and box plots.
5. screen outliers with IQR and z-scores.
6. calculate skewness and kurtosis.
7. create a Q-Q plot.
8. conduct and interpret Shapiro-Wilk.
9. compare distributions across groups.
10. assess log and square-root transformations.
11. audit missingness and duplicate identifiers.
12. create an EDA function.
13. distinguish exploratory and confirmatory findings.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. profiles all numerical variables;
2. uses several distribution plots;
3. assesses skewness and outliers;
4. creates Q-Q plots;
5. uses normality tests cautiously;
6. compares groups;
7. evaluates a transformation;
8. includes data-quality checks;
9. creates reusable EDA functions; and
10. distinguishes exploration from confirmation.

# 32. Completion checklist

Confirm that you can:

- conduct EDA;
- assess centre and spread;
- compare mean and median;
- use histograms and densities;
- use box plots;
- screen outliers;
- calculate skewness and kurtosis;
- create Q-Q plots;
- interpret normality tests;
- compare group distributions;
- assess transformations;
- audit data quality; and
- create reusable EDA summaries.

# Lesson conclusion

Exploratory analysis combines numerical, visual and structural evidence. Its
purpose is to understand the data before formal conclusions are drawn.

\[
\boxed{
\text{summarise}
\rightarrow
\text{visualise}
\rightarrow
\text{check quality}
\rightarrow
\text{compare groups}
\rightarrow
\text{document findings}
}
\]
