---
title: "Lesson 29: Correlation, Covariance and Visual Association"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Correlation and covariance describe how numerical variables vary together.
They are useful for exploring association, screening predictors and checking
redundancy.

Correlation does not establish causation. A strong association may arise from:

- direct influence;
- reverse influence;
- a common cause;
- selection;
- measurement structure;
- or coincidence.

This lesson develops Pearson, Spearman and Kendall correlations, covariance,
correlation matrices, significance tests and visual assessment.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain covariance and correlation;
2. calculate covariance;
3. calculate Pearson correlation;
4. calculate Spearman correlation;
5. calculate Kendall correlation;
6. distinguish linear and monotonic relationships;
7. handle missing values appropriately;
8. create correlation matrices;
9. test individual correlations;
10. calculate confidence intervals;
11. inspect scatter plots;
12. detect non-linear relationships and outliers;
13. create grouped correlations;
14. avoid causal interpretation; and
15. produce a correlation report.

# 1. Importing data

```{r import-correlation-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

clinic_data <- read.csv(
  "data/clinic_visits.csv",
  stringsAsFactors = FALSE
)
```

# 2. Covariance

```{r covariance-example}
cov(
  student_data$coursework_score,
  student_data$examination_score,
  use = "complete.obs"
)
```

Sample covariance is:

\[
s_{xy}
=
\frac{
\sum_{i=1}^{n}
(x_i-\bar{x})
(y_i-\bar{y})
}{
n-1
}.
\]

# 3. Interpreting covariance

- positive covariance: variables tend to increase together;
- negative covariance: one tends to increase as the other decreases;
- near-zero covariance: little linear co-movement.

Magnitude depends on measurement units.

# 4. Pearson correlation

```{r pearson-correlation}
cor(
  student_data$coursework_score,
  student_data$examination_score,
  method = "pearson",
  use = "complete.obs"
)
```

Pearson correlation standardises covariance:

\[
r
=
\frac{s_{xy}}{s_xs_y}.
\]

# 5. Range of correlation

Correlation lies between -1 and 1.

- `1`: perfect positive linear association;
- `-1`: perfect negative linear association;
- `0`: no linear association.

# 6. Scatter plot

```{r correlation-scatterplot}
plot(
  student_data$coursework_score,
  student_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)
```

# 7. Fitted line

```{r correlation-fitted-line}
plot(
  student_data$coursework_score,
  student_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)

abline(
  lm(
    examination_score ~ coursework_score,
    data = student_data
  ),
  lwd = 2
)
```

# 8. Spearman correlation

```{r spearman-correlation}
cor(
  student_data$coursework_score,
  student_data$examination_score,
  method = "spearman",
  use = "complete.obs"
)
```

Spearman correlation is based on ranks and assesses monotonic association.

# 9. Kendall correlation

```{r kendall-correlation}
cor(
  student_data$coursework_score,
  student_data$examination_score,
  method = "kendall",
  use = "complete.obs"
)
```

Kendall's tau is based on concordant and discordant pairs.

# 10. Comparing methods

```{r compare-correlation-methods}
correlation_comparison <- data.frame(
  method = c(
    "Pearson",
    "Spearman",
    "Kendall"
  ),
  coefficient = c(
    cor(
      student_data$coursework_score,
      student_data$examination_score,
      method = "pearson"
    ),
    cor(
      student_data$coursework_score,
      student_data$examination_score,
      method = "spearman"
    ),
    cor(
      student_data$coursework_score,
      student_data$examination_score,
      method = "kendall"
    )
  ),
  stringsAsFactors = FALSE
)

correlation_comparison
```

# 11. Linear versus monotonic association

```{r monotonic-example}
x <- 1:20
y <- x ^ 2

cor(
  x,
  y,
  method = "pearson"
)

cor(
  x,
  y,
  method = "spearman"
)
```

The relationship is perfectly monotonic but not perfectly linear.

# 12. Zero correlation can hide structure

```{r nonlinear-zero-example}
x_curve <- seq(
  -3,
  3,
  length.out = 100
)

y_curve <- x_curve ^ 2

cor(
  x_curve,
  y_curve
)

plot(
  x_curve,
  y_curve,
  main = "Non-linear Association",
  xlab = "x",
  ylab = "x squared",
  pch = 19
)
```

# 13. Outlier influence

```{r correlation-outlier-influence}
x_regular <- 1:10
y_regular <- 2 * x_regular

pearson_regular <- cor(
  x_regular,
  y_regular
)

x_with_outlier <- c(
  x_regular,
  20
)

y_with_outlier <- c(
  y_regular,
  -20
)

pearson_with_outlier <- cor(
  x_with_outlier,
  y_with_outlier
)

spearman_with_outlier <- cor(
  x_with_outlier,
  y_with_outlier,
  method = "spearman"
)

data.frame(
  statistic = c(
    "Pearson without outlier",
    "Pearson with outlier",
    "Spearman with outlier"
  ),
  value = c(
    pearson_regular,
    pearson_with_outlier,
    spearman_with_outlier
  ),
  stringsAsFactors = FALSE
)
```

# 14. Correlation matrix

```{r correlation-matrix}
numeric_student_data <- student_data[
  vapply(
    student_data,
    is.numeric,
    FUN.VALUE = logical(1)
  )
]

student_correlation_matrix <- cor(
  numeric_student_data,
  use = "pairwise.complete.obs",
  method = "pearson"
)

student_correlation_matrix
```

# 15. Covariance matrix

```{r covariance-matrix}
cov(
  numeric_student_data,
  use = "pairwise.complete.obs"
)
```

# 16. Rounding a matrix

```{r round-correlation-matrix}
round(
  student_correlation_matrix,
  digits = 3
)
```

# 17. Missing-value options

Common `use` choices:

- `"everything"`;
- `"all.obs"`;
- `"complete.obs"`;
- `"pairwise.complete.obs"`.

Pairwise deletion can produce matrices based on different sample sizes.

# 18. Correlation test

```{r pearson-correlation-test}
pearson_test <- cor.test(
  student_data$coursework_score,
  student_data$examination_score,
  method = "pearson"
)

pearson_test
```

# 19. Extracting test results

```{r extract-correlation-test}
data.frame(
  estimate =
    unname(
      pearson_test$estimate
    ),
  statistic =
    unname(
      pearson_test$statistic
    ),
  degrees_of_freedom =
    unname(
      pearson_test$parameter
    ),
  p_value =
    pearson_test$p.value,
  confidence_lower =
    pearson_test$conf.int[1],
  confidence_upper =
    pearson_test$conf.int[2]
)
```

# 20. Spearman test

```{r spearman-test}
cor.test(
  student_data$coursework_score,
  student_data$examination_score,
  method = "spearman",
  exact = FALSE
)
```

# 21. Kendall test

```{r kendall-test}
cor.test(
  student_data$coursework_score,
  student_data$examination_score,
  method = "kendall",
  exact = FALSE
)
```

# 22. Grouped correlations

```{r grouped-correlations}
programme_groups <- split(
  student_data,
  student_data$programme
)

grouped_correlation <- lapply(
  programme_groups,
  function(group_data) {
    cor(
      group_data$coursework_score,
      group_data$examination_score,
      use = "complete.obs"
    )
  }
)

grouped_correlation
```

Small group sizes make grouped correlations unstable.

# 23. Correlation by group function

```{r grouped-correlation-function}
calculate_grouped_correlation <- function(
  data,
  x_variable,
  y_variable,
  group_variable,
  method = "pearson"
) {
  required <- c(
    x_variable,
    y_variable,
    group_variable
  )

  if (!all(
    required %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  groups <- split(
    data,
    data[[
      group_variable
    ]]
  )

  results <- lapply(
    names(groups),
    function(group_name) {
      group_data <- groups[[
        group_name
      ]]

      complete_status <- complete.cases(
        group_data[
          c(
            x_variable,
            y_variable
          )
        ]
      )

      n_complete <- sum(
        complete_status
      )

      estimate <- if (
        n_complete >= 3L
      ) {
        cor(
          group_data[[
            x_variable
          ]][
            complete_status
          ],
          group_data[[
            y_variable
          ]][
            complete_status
          ],
          method = method
        )
      } else {
        NA_real_
      }

      data.frame(
        group = group_name,
        n = n_complete,
        correlation = estimate,
        stringsAsFactors = FALSE
      )
    }
  )

  result <- do.call(
    rbind,
    results
  )

  names(result)[1] <-
    group_variable

  result
}

calculate_grouped_correlation(
  student_data,
  "coursework_score",
  "examination_score",
  "programme"
)
```

# 24. Pairwise correlation report function

```{r correlation-report-function}
create_correlation_report <- function(
  x,
  y,
  x_name = "x",
  y_name = "y",
  method = c(
    "pearson",
    "spearman",
    "kendall"
  )
) {
  method <- match.arg(
    method
  )

  complete_status <- complete.cases(
    x,
    y
  )

  x_complete <- x[
    complete_status
  ]

  y_complete <- y[
    complete_status
  ]

  test_result <- cor.test(
    x_complete,
    y_complete,
    method = method,
    exact = FALSE
  )

  data.frame(
    variable_x = x_name,
    variable_y = y_name,
    method = method,
    n = length(
      x_complete
    ),
    estimate =
      unname(
        test_result$estimate
      ),
    p_value =
      test_result$p.value,
    stringsAsFactors = FALSE
  )
}

create_correlation_report(
  student_data$coursework_score,
  student_data$examination_score,
  "coursework_score",
  "examination_score",
  method = "pearson"
)
```

# 25. Correlation matrix function

```{r safe-correlation-matrix-function}
create_correlation_matrix <- function(
  data,
  method = c(
    "pearson",
    "spearman",
    "kendall"
  ),
  digits = 3
) {
  method <- match.arg(
    method
  )

  numeric_data <- data[
    vapply(
      data,
      is.numeric,
      FUN.VALUE = logical(1)
    )
  ]

  if (ncol(
    numeric_data
  ) < 2L) {
    stop(
      "At least two numeric variables are required.",
      call. = FALSE
    )
  }

  round(
    cor(
      numeric_data,
      use =
        "pairwise.complete.obs",
      method = method
    ),
    digits = digits
  )
}

create_correlation_matrix(
  student_data
)
```

# 26. Visual matrix with `pairs()`

```{r pairs-plot}
pairs(
  numeric_student_data,
  main = "Scatter-Plot Matrix"
)
```

# 27. Correlation is not causation

A correlation does not identify:

- temporal order;
- causal direction;
- confounding variables;
- selection effects;
- measurement bias;
- or intervention effects.

# 28. Practical interpretation

A complete report should state:

- variables;
- method;
- sample size;
- coefficient;
- direction;
- magnitude;
- uncertainty or p-value;
- visual pattern;
- and limitations.

# 29. Common mistakes

- interpreting correlation causally;
- using Pearson correlation for a strongly non-linear relationship;
- ignoring outliers;
- reporting coefficients without sample size;
- using pairwise deletion without noting varying sample sizes;
- treating p-values as effect sizes;
- correlating categorical codes as if continuous;
- reporting a large matrix without purpose;
- ignoring group structure;
- using arbitrary labels such as weak or strong without context.

# 30. Independent exercises

1. calculate covariance.
2. calculate Pearson correlation.
3. calculate Spearman correlation.
4. calculate Kendall correlation.
5. compare methods.
6. demonstrate monotonic versus linear association.
7. demonstrate non-linear zero correlation.
8. assess outlier influence.
9. create correlation and covariance matrices.
10. perform correlation tests.
11. extract confidence intervals.
12. calculate grouped correlations.
13. create reusable report functions.
14. write a non-causal interpretation.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. explains covariance and correlation;
2. calculates three correlation methods;
3. includes scatter plots;
4. demonstrates a non-linear relationship;
5. demonstrates outlier influence;
6. creates a correlation matrix;
7. performs significance tests;
8. calculates grouped correlations;
9. uses reusable functions; and
10. provides cautious interpretations.

# 32. Completion checklist

Confirm that you can:

- calculate covariance;
- calculate Pearson correlation;
- calculate Spearman correlation;
- calculate Kendall correlation;
- distinguish linear and monotonic association;
- inspect scatter plots;
- detect non-linear patterns;
- assess outlier influence;
- create correlation matrices;
- test correlations;
- extract confidence intervals;
- calculate grouped correlations;
- create report functions; and
- avoid causal claims.

# Lesson conclusion

Correlation summarises association, not causation. Reliable interpretation
requires a suitable method, visual inspection, adequate sample size and clear
acknowledgement of limitations.

\[
\boxed{
\text{visualise}
\rightarrow
\text{choose method}
\rightarrow
\text{calculate}
\rightarrow
\text{test}
\rightarrow
\text{interpret cautiously}
}
\]
