---
title: "Lesson 30: Sampling Distributions and Standard Errors"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Statistical inference uses sample information to learn about a population.
Because different random samples produce different results, every sample
statistic has sampling variability.

A sampling distribution describes how a statistic would vary over repeated
samples drawn under the same process. Standard errors quantify the expected
sampling variation.

This lesson develops the logic of repeated sampling through simulation and
connects sample size, variability and precision.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish populations, samples, parameters and statistics;
2. explain sampling variability;
3. define a sampling distribution;
4. simulate repeated random samples;
5. calculate the standard error of a mean;
6. explain the central limit theorem;
7. compare sampling distributions across sample sizes;
8. calculate the standard error of a proportion;
9. distinguish standard deviation from standard error;
10. understand bias and precision;
11. assess Monte Carlo error;
12. create reusable sampling-distribution functions; and
13. interpret simulation results.

# 1. Population and sample

A **population** is the complete set of units of interest.

A **sample** is a subset observed from that population.

A **parameter** summarises the population.

A **statistic** summarises a sample.

Examples:

- population mean: \(\mu\);
- sample mean: \(\bar{x}\);
- population proportion: \(p\);
- sample proportion: \(\hat{p}\).

# 2. Creating a teaching population

```{r create-teaching-population}
population <- rnorm(
  n = 100000,
  mean = 70,
  sd = 12
)

mean(population)
sd(population)
```

# 3. Drawing one random sample

```{r draw-one-sample}
one_sample <- sample(
  population,
  size = 30,
  replace = FALSE
)

mean(one_sample)
sd(one_sample)
```

The sample mean is unlikely to equal the population mean exactly.

# 4. Repeated sample means

```{r repeated-sample-means}
number_of_samples <- 1000
sample_size <- 30

sample_means <- replicate(
  number_of_samples,
  mean(
    sample(
      population,
      size = sample_size,
      replace = TRUE
    )
  )
)

head(sample_means)
```

# 5. Sampling distribution

```{r sampling-distribution-summary}
mean(sample_means)
sd(sample_means)
summary(sample_means)
```

The mean of the sampling distribution should be close to the population mean.

# 6. Standard error of a mean

The theoretical standard error is:

\[
SE(\bar{x})
=
\frac{\sigma}{\sqrt{n}}.
\]

```{r theoretical-standard-error}
population_sd <- sd(population)

theoretical_standard_error <-
  population_sd /
  sqrt(sample_size)

simulated_standard_error <-
  sd(sample_means)

theoretical_standard_error
simulated_standard_error
```

# 7. Visualising the sampling distribution

```{r sampling-distribution-histogram}
hist(
  sample_means,
  breaks = 30,
  main = "Sampling Distribution of the Mean",
  xlab = "Sample mean",
  ylab = "Frequency"
)

abline(
  v = mean(population),
  lwd = 2
)
```

# 8. Effect of sample size

```{r compare-sample-sizes}
sample_sizes <- c(
  5,
  30,
  100
)

sampling_results <- lapply(
  sample_sizes,
  function(current_size) {
    replicate(
      2000,
      mean(
        sample(
          population,
          size = current_size,
          replace = TRUE
        )
      )
    )
  }
)

names(sampling_results) <- paste0(
  "n_",
  sample_sizes
)

vapply(
  sampling_results,
  sd,
  FUN.VALUE = numeric(1)
)
```

Larger samples produce smaller standard errors.

# 9. Multi-panel comparison

```{r sample-size-histograms}
old_parameters <- par(
  mfrow = c(
    1,
    3
  )
)

for (i in seq_along(
  sampling_results
)) {
  hist(
    sampling_results[[i]],
    breaks = 25,
    main = names(
      sampling_results
    )[i],
    xlab = "Sample mean"
  )
}

par(
  old_parameters
)
```

# 10. Central limit theorem

The central limit theorem states that, under broad conditions, the sampling
distribution of the mean becomes approximately normal as sample size increases,
even when the population distribution is not normal.

# 11. Skewed population demonstration

```{r skewed-population}
skewed_population <- rexp(
  n = 100000,
  rate = 0.10
)

hist(
  skewed_population,
  breaks = 40,
  main = "Skewed Population",
  xlab = "Value"
)
```

# 12. Sample means from a skewed population

```{r skewed-population-sample-means}
skewed_means_small <- replicate(
  2000,
  mean(
    sample(
      skewed_population,
      size = 5,
      replace = TRUE
    )
  )
)

skewed_means_large <- replicate(
  2000,
  mean(
    sample(
      skewed_population,
      size = 50,
      replace = TRUE
    )
  )
)
```

```{r skewed-sampling-comparison}
old_parameters <- par(
  mfrow = c(
    1,
    2
  )
)

hist(
  skewed_means_small,
  breaks = 30,
  main = "Sample Means: n = 5",
  xlab = "Sample mean"
)

hist(
  skewed_means_large,
  breaks = 30,
  main = "Sample Means: n = 50",
  xlab = "Sample mean"
)

par(
  old_parameters
)
```

# 13. Standard deviation versus standard error

**Standard deviation** describes variation among observed values.

**Standard error** describes variation among repeated sample statistics.

```{r sd-versus-se}
sample_standard_deviation <- sd(
  one_sample
)

estimated_standard_error <-
  sample_standard_deviation /
  sqrt(
    length(one_sample)
  )

sample_standard_deviation
estimated_standard_error
```

# 14. Sampling distribution of a proportion

Create a binary population:

```{r binary-population}
binary_population <- rbinom(
  n = 100000,
  size = 1,
  prob = 0.60
)

mean(binary_population)
```

Repeated sample proportions:

```{r repeated-sample-proportions}
sample_proportions <- replicate(
  2000,
  mean(
    sample(
      binary_population,
      size = 100,
      replace = TRUE
    )
  )
)

mean(sample_proportions)
sd(sample_proportions)
```

# 15. Standard error of a proportion

\[
SE(\hat{p})
=
\sqrt{
\frac{
p(1-p)
}{
n
}
}.
\]

```{r theoretical-proportion-se}
population_proportion <- mean(
  binary_population
)

theoretical_proportion_se <- sqrt(
  population_proportion *
    (
      1 -
      population_proportion
    ) /
    100
)

simulated_proportion_se <- sd(
  sample_proportions
)

theoretical_proportion_se
simulated_proportion_se
```

# 16. Bias

An estimator is unbiased if its expected value equals the population parameter.

```{r estimate-bias}
estimated_bias <- mean(
  sample_means
) - mean(
  population
)

estimated_bias
```

# 17. Precision

Precision refers to how tightly the sampling distribution is concentrated.

A smaller standard error indicates greater precision.

# 18. Monte Carlo error

Simulation estimates vary because only a finite number of replications are
used.

```{r monte-carlo-comparison}
replication_counts <- c(
  100,
  1000,
  10000
)

monte_carlo_means <- vapply(
  replication_counts,
  function(repetitions) {
    simulated_means <- replicate(
      repetitions,
      mean(
        sample(
          population,
          size = 30,
          replace = TRUE
        )
      )
    )

    mean(simulated_means)
  },
  FUN.VALUE = numeric(1)
)

data.frame(
  replications =
    replication_counts,
  estimated_sampling_mean =
    monte_carlo_means
)
```

More replications generally reduce Monte Carlo uncertainty.

# 19. Bootstrap-style resampling

```{r bootstrap-sample-means}
observed_sample <- sample(
  population,
  size = 40,
  replace = FALSE
)

bootstrap_means <- replicate(
  2000,
  mean(
    sample(
      observed_sample,
      size = length(
        observed_sample
      ),
      replace = TRUE
    )
  )
)

sd(
  bootstrap_means
)
```

This estimates the sampling variability from the observed sample.

# 20. Reusable simulation function

```{r sampling-function}
simulate_sample_means <- function(
  population,
  sample_size,
  replications = 1000
) {
  if (!is.numeric(population)) {
    stop(
      "`population` must be numeric.",
      call. = FALSE
    )
  }

  if (sample_size < 1 ||
      sample_size !=
      as.integer(sample_size)) {
    stop(
      "`sample_size` must be a positive whole number.",
      call. = FALSE
    )
  }

  if (replications < 1 ||
      replications !=
      as.integer(replications)) {
    stop(
      "`replications` must be a positive whole number.",
      call. = FALSE
    )
  }

  replicate(
    replications,
    mean(
      sample(
        population,
        size = sample_size,
        replace = TRUE
      )
    )
  )
}

simulated_means <- simulate_sample_means(
  population,
  sample_size = 50,
  replications = 2000
)

summary(simulated_means)
```

# 21. Sampling-distribution report

```{r sampling-report-function}
create_sampling_report <- function(
  population,
  sample_size,
  replications = 2000
) {
  simulated_statistics <- simulate_sample_means(
    population,
    sample_size,
    replications
  )

  data.frame(
    population_mean =
      mean(population),
    population_sd =
      sd(population),
    sample_size =
      sample_size,
    replications =
      replications,
    simulated_mean =
      mean(
        simulated_statistics
      ),
    simulated_standard_error =
      sd(
        simulated_statistics
      ),
    theoretical_standard_error =
      sd(population) /
      sqrt(sample_size),
    estimated_bias =
      mean(
        simulated_statistics
      ) -
      mean(population)
  )
}

create_sampling_report(
  population,
  sample_size = 30
)
```

# 22. Common mistakes

- confusing the sample distribution with the sampling distribution;
- confusing standard deviation with standard error;
- assuming larger samples reduce population variability;
- treating one simulated result as proof;
- failing to set a random seed;
- using too few simulation replications;
- assuming the central limit theorem applies without conditions;
- ignoring dependence between observations;
- using the wrong standard-error formula;
- interpreting precision as accuracy.

# 23. Independent exercises

1. distinguish populations, samples, parameters and statistics.
2. draw one random sample.
3. simulate repeated sample means.
4. estimate the sampling mean and standard error.
5. compare theoretical and simulated standard errors.
6. compare several sample sizes.
7. demonstrate the central limit theorem with skewed data.
8. compare standard deviation and standard error.
9. simulate sample proportions.
10. estimate bias.
11. compare Monte Carlo replication counts.
12. create reusable simulation functions.
13. write a sampling-distribution interpretation.

# 24. Lesson assessment

Create a standalone R Markdown report that:

1. defines the inferential concepts;
2. creates a population;
3. draws repeated samples;
4. compares simulated and theoretical standard errors;
5. examines three sample sizes;
6. demonstrates the central limit theorem;
7. simulates proportions;
8. evaluates bias and precision;
9. creates reusable simulation functions; and
10. interprets all results.

# 25. Completion checklist

Confirm that you can:

- distinguish parameters and statistics;
- explain sampling variability;
- simulate sampling distributions;
- calculate standard errors;
- explain the central limit theorem;
- assess sample-size effects;
- simulate proportions;
- distinguish standard deviation and standard error;
- assess bias and precision;
- understand Monte Carlo error; and
- create reusable simulation functions.

# Lesson conclusion

Sampling distributions connect observed samples to population inference.
Standard errors quantify uncertainty and become smaller as effective sample size
increases.

\[
\boxed{
\text{population}
\rightarrow
\text{sample repeatedly}
\rightarrow
\text{sampling distribution}
\rightarrow
\text{standard error}
\rightarrow
\text{inference}
}
\]
