---
title: "Lesson 31: Confidence Intervals and Estimation"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Estimation uses sample data to approximate unknown population parameters.

A point estimate gives one value.

An interval estimate gives a range of plausible values, together with a stated
confidence level.

Confidence intervals combine:

- an estimate;
- a standard error;
- a critical value;
- and a margin of error.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish point and interval estimation;
2. explain confidence level and margin of error;
3. construct a confidence interval for a mean;
4. use z and t critical values appropriately;
5. construct a confidence interval for a proportion;
6. calculate confidence intervals with built-in functions;
7. explain the repeated-sampling interpretation;
8. examine the effect of sample size and confidence level;
9. simulate confidence-interval coverage;
10. create bootstrap percentile intervals;
11. report estimates and intervals clearly;
12. create reusable confidence-interval functions; and
13. avoid common interpretation errors.

# 1. Importing data

```{r import-confidence-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

final_scores <- student_data$final_score
```

# 2. Point estimate of a mean

```{r point-estimate-mean}
sample_mean <- mean(
  final_scores,
  na.rm = TRUE
)

sample_mean
```

# 3. Standard error from sample data

```{r sample-standard-error}
observed_scores <- final_scores[
  !is.na(final_scores)
]

sample_size <- length(
  observed_scores
)

sample_sd <- sd(
  observed_scores
)

standard_error <-
  sample_sd /
  sqrt(sample_size)

standard_error
```

# 4. Confidence interval with the t distribution

For an unknown population standard deviation:

\[
\bar{x}
\pm
t_{\alpha/2,n-1}
\frac{s}{\sqrt{n}}.
\]

```{r manual-mean-confidence-interval}
confidence_level <- 0.95

alpha <- 1 -
  confidence_level

critical_t <- qt(
  1 -
    alpha / 2,
  df =
    sample_size - 1
)

margin_of_error <-
  critical_t *
  standard_error

confidence_interval <- c(
  lower =
    sample_mean -
    margin_of_error,
  upper =
    sample_mean +
    margin_of_error
)

confidence_interval
```

# 5. Using `t.test()`

```{r t-test-confidence-interval}
mean_interval_test <- t.test(
  observed_scores,
  conf.level = 0.95
)

mean_interval_test
```

Extract:

```{r extract-mean-interval}
data.frame(
  estimate =
    unname(
      mean_interval_test$estimate
    ),
  lower =
    mean_interval_test$conf.int[1],
  upper =
    mean_interval_test$conf.int[2],
  confidence_level =
    attr(
      mean_interval_test$conf.int,
      "conf.level"
    )
)
```

# 6. Confidence-level comparison

```{r compare-confidence-levels}
confidence_levels <- c(
  0.90,
  0.95,
  0.99
)

intervals_by_level <- lapply(
  confidence_levels,
  function(level) {
    test_result <- t.test(
      observed_scores,
      conf.level = level
    )

    data.frame(
      confidence_level = level,
      lower =
        test_result$conf.int[1],
      upper =
        test_result$conf.int[2],
      width =
        diff(
          test_result$conf.int
        )
    )
  }
)

do.call(
  rbind,
  intervals_by_level
)
```

Higher confidence produces a wider interval.

# 7. Effect of sample size

```{r confidence-sample-size-effect}
population <- rnorm(
  100000,
  mean = 70,
  sd = 12
)

sample_sizes <- c(
  20,
  50,
  200
)

interval_widths <- lapply(
  sample_sizes,
  function(current_size) {
    selected_sample <- sample(
      population,
      size = current_size
    )

    test_result <- t.test(
      selected_sample
    )

    data.frame(
      sample_size =
        current_size,
      estimate =
        mean(
          selected_sample
        ),
      lower =
        test_result$conf.int[1],
      upper =
        test_result$conf.int[2],
      width =
        diff(
          test_result$conf.int
        )
    )
  }
)

do.call(
  rbind,
  interval_widths
)
```

# 8. Confidence interval for a proportion

```{r create-pass-proportion}
pass_status <-
  student_data$final_score >= 50

number_passing <- sum(
  pass_status,
  na.rm = TRUE
)

number_observed <- sum(
  !is.na(
    pass_status
  )
)

sample_proportion <-
  number_passing /
  number_observed

sample_proportion
```

# 9. Approximate Wald interval

\[
\hat{p}
\pm
z_{\alpha/2}
\sqrt{
\frac{
\hat{p}(1-\hat{p})
}{
n
}
}.
\]

```{r wald-proportion-interval}
critical_z <- qnorm(
  0.975
)

proportion_se <- sqrt(
  sample_proportion *
    (
      1 -
      sample_proportion
    ) /
    number_observed
)

wald_interval <- c(
  lower =
    sample_proportion -
    critical_z *
    proportion_se,
  upper =
    sample_proportion +
    critical_z *
    proportion_se
)

wald_interval
```

The Wald interval can perform poorly with small samples or proportions near 0
or 1.

# 10. `prop.test()`

```{r proportion-test-interval}
proportion_test <- prop.test(
  x = number_passing,
  n = number_observed,
  conf.level = 0.95,
  correct = FALSE
)

proportion_test
```

# 11. `binom.test()`

```{r exact-binomial-interval}
exact_proportion_test <- binom.test(
  x = number_passing,
  n = number_observed,
  conf.level = 0.95
)

exact_proportion_test
```

# 12. Confidence interval for a difference in means

```{r two-group-confidence-interval}
selected_programmes <- student_data[
  student_data$programme %in%
    c(
      "Statistics",
      "Procurement"
    ),
]

two_group_test <- t.test(
  final_score ~ programme,
  data =
    selected_programmes,
  var.equal = FALSE,
  conf.level = 0.95
)

two_group_test
```

The interval concerns the difference between group means.

# 13. Paired confidence interval

```{r paired-confidence-example}
before_scores <- c(
  60,
  65,
  70,
  72,
  68,
  75
)

after_scores <- c(
  64,
  67,
  74,
  73,
  72,
  80
)

paired_interval <- t.test(
  after_scores,
  before_scores,
  paired = TRUE,
  conf.level = 0.95
)

paired_interval
```

# 14. Repeated-sampling interpretation

A 95 per cent confidence procedure is designed so that, over repeated sampling,
approximately 95 per cent of intervals constructed by that procedure contain
the true parameter.

It does not mean that there is a 95 per cent probability that a fixed parameter
lies in one already-calculated interval.

# 15. Simulating confidence-interval coverage

```{r confidence-coverage-simulation}
population_mean <- 70
population_sd <- 12
sample_size <- 30
replications <- 2000

coverage_status <- replicate(
  replications,
  {
    current_sample <- rnorm(
      sample_size,
      mean = population_mean,
      sd = population_sd
    )

    interval <- t.test(
      current_sample,
      conf.level = 0.95
    )$conf.int

    interval[1] <=
      population_mean &&
      population_mean <=
      interval[2]
  }
)

mean(
  coverage_status
)
```

# 16. Visualising interval coverage

```{r confidence-coverage-visual}
number_to_plot <- 50

interval_matrix <- replicate(
  number_to_plot,
  {
    current_sample <- rnorm(
      sample_size,
      mean = population_mean,
      sd = population_sd
    )

    test_result <- t.test(
      current_sample
    )

    c(
      estimate = mean(
        current_sample
      ),
      lower =
        test_result$conf.int[1],
      upper =
        test_result$conf.int[2]
    )
  }
)

interval_matrix <- t(
  interval_matrix
)

plot(
  interval_matrix[
    ,
    "estimate"
  ],
  seq_len(
    number_to_plot
  ),
  xlim = range(
    interval_matrix[
      ,
      c(
        "lower",
        "upper"
      )
    ]
  ),
  pch = 19,
  xlab = "Mean and 95% confidence interval",
  ylab = "Simulation",
  main = "Repeated Confidence Intervals"
)

segments(
  x0 =
    interval_matrix[
      ,
      "lower"
    ],
  y0 =
    seq_len(
      number_to_plot
    ),
  x1 =
    interval_matrix[
      ,
      "upper"
    ],
  y1 =
    seq_len(
      number_to_plot
    )
)

abline(
  v =
    population_mean,
  lwd = 2
)
```

# 17. Bootstrap percentile interval

```{r bootstrap-percentile-interval}
set.seed(20260714)

bootstrap_means <- replicate(
  5000,
  mean(
    sample(
      observed_scores,
      size = length(
        observed_scores
      ),
      replace = TRUE
    )
  )
)

bootstrap_interval <- quantile(
  bootstrap_means,
  probs = c(
    0.025,
    0.975
  )
)

bootstrap_interval
```

# 18. Comparing t and bootstrap intervals

```{r compare-interval-methods}
data.frame(
  method = c(
    "t interval",
    "Bootstrap percentile"
  ),
  lower = c(
    mean_interval_test$conf.int[1],
    bootstrap_interval[1]
  ),
  upper = c(
    mean_interval_test$conf.int[2],
    bootstrap_interval[2]
  ),
  stringsAsFactors = FALSE
)
```

# 19. Reusable mean interval function

```{r mean-confidence-function}
mean_confidence_interval <- function(
  x,
  confidence_level = 0.95
) {
  if (!is.numeric(x)) {
    stop(
      "`x` must be numeric.",
      call. = FALSE
    )
  }

  observed <- x[
    !is.na(x)
  ]

  if (length(observed) < 2L) {
    stop(
      "At least two observed values are required.",
      call. = FALSE
    )
  }

  result <- t.test(
    observed,
    conf.level =
      confidence_level
  )

  data.frame(
    n = length(observed),
    estimate =
      mean(observed),
    standard_error =
      sd(observed) /
      sqrt(
        length(observed)
      ),
    lower =
      result$conf.int[1],
    upper =
      result$conf.int[2],
    confidence_level =
      confidence_level
  )
}

mean_confidence_interval(
  final_scores
)
```

# 20. Reusable proportion interval function

```{r proportion-confidence-function}
proportion_confidence_interval <- function(
  successes,
  trials,
  confidence_level = 0.95,
  method = c(
    "exact",
    "score"
  )
) {
  method <- match.arg(
    method
  )

  if (successes < 0 ||
      trials <= 0 ||
      successes > trials) {
    stop(
      "Counts are invalid.",
      call. = FALSE
    )
  }

  test_result <- if (
    method == "exact"
  ) {
    binom.test(
      successes,
      trials,
      conf.level =
        confidence_level
    )
  } else {
    prop.test(
      successes,
      trials,
      conf.level =
        confidence_level,
      correct = FALSE
    )
  }

  data.frame(
    successes =
      successes,
    trials =
      trials,
    estimate =
      successes /
      trials,
    lower =
      test_result$conf.int[1],
    upper =
      test_result$conf.int[2],
    confidence_level =
      confidence_level,
    method =
      method
  )
}

proportion_confidence_interval(
  number_passing,
  number_observed
)
```

# 21. Reporting intervals

A useful report states:

- parameter estimated;
- point estimate;
- confidence level;
- lower and upper limits;
- sample size;
- method;
- and substantive interpretation.

# 22. Common mistakes

- interpreting confidence as probability for a fixed parameter;
- reporting an interval without its confidence level;
- ignoring the method used;
- assuming narrow intervals are automatically unbiased;
- confusing standard deviation and standard error;
- using a Wald proportion interval with unsuitable data;
- treating overlap of confidence intervals as a formal test;
- rounding before calculation;
- hiding the sample size;
- reporting precision without considering design effects.

# 23. Independent exercises

1. calculate a point estimate.
2. calculate a standard error.
3. construct a manual t interval.
4. verify it with `t.test()`.
5. compare 90%, 95% and 99% intervals.
6. assess the effect of sample size.
7. construct proportion intervals.
8. compare `prop.test()` and `binom.test()`.
9. construct a difference-in-means interval.
10. construct a paired interval.
11. simulate confidence-interval coverage.
12. create a bootstrap percentile interval.
13. create reusable interval functions.
14. write an accurate interpretation.

# 24. Lesson assessment

Create a standalone R Markdown report that:

1. distinguishes point and interval estimates;
2. constructs manual and built-in mean intervals;
3. compares confidence levels;
4. examines sample-size effects;
5. constructs proportion intervals;
6. constructs independent and paired mean-difference intervals;
7. simulates coverage;
8. creates a bootstrap interval;
9. uses reusable functions; and
10. reports all estimates accurately.

# 25. Completion checklist

Confirm that you can:

- distinguish point and interval estimates;
- calculate standard errors;
- construct t intervals;
- construct proportion intervals;
- use built-in interval functions;
- compare confidence levels;
- assess sample-size effects;
- explain repeated-sampling coverage;
- create bootstrap intervals;
- report margins of error; and
- create reusable interval functions.

# Lesson conclusion

Confidence intervals combine an estimate with its uncertainty. Their correct
interpretation concerns the long-run performance of the interval procedure.

\[
\boxed{
\text{estimate}
\pm
\text{critical value}
\times
\text{standard error}
}
\]
