---
title: "Lesson 32: Hypothesis Testing and t-Tests"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Hypothesis testing evaluates how compatible observed data are with a specified
null hypothesis.

A test combines:

- a null hypothesis;
- an alternative hypothesis;
- a test statistic;
- a reference distribution;
- a p-value;
- a significance level;
- and a decision.

This lesson develops one-sample, independent-samples and paired t-tests, while
emphasising assumptions, effect sizes and interpretation.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. formulate null and alternative hypotheses;
2. distinguish one-sided and two-sided tests;
3. explain test statistics and p-values;
4. distinguish statistical and substantive significance;
5. conduct one-sample t-tests;
6. conduct independent-samples t-tests;
7. distinguish Welch and pooled-variance tests;
8. conduct paired t-tests;
9. calculate confidence intervals;
10. calculate effect sizes;
11. assess assumptions;
12. understand Type I and Type II errors;
13. conduct power calculations conceptually;
14. create reusable t-test reporting functions; and
15. report results clearly.

# 1. Hypotheses

Example one-sample hypotheses:

\[
H_0:\mu=70
\]

\[
H_1:\mu\neq70
\]

The null hypothesis states the reference claim.

The alternative states the departure being investigated.

# 2. Significance level

The significance level, commonly \(\alpha=0.05\), is the tolerated long-run
Type I error rate under the null hypothesis.

# 3. P-value

A p-value is the probability, assuming the null hypothesis and model conditions
are true, of obtaining a result at least as incompatible with the null as the
observed result.

It is not:

- the probability that the null hypothesis is true;
- the probability that the result occurred by chance;
- or the size of the effect.

# 4. Importing data

```{r import-t-test-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

final_scores <- student_data$final_score
```

# 5. One-sample t-test

Test whether the mean final score differs from 70.

```{r one-sample-t-test}
one_sample_test <- t.test(
  final_scores,
  mu = 70,
  alternative = "two.sided",
  conf.level = 0.95
)

one_sample_test
```

# 6. Extracting one-sample results

```{r extract-one-sample-results}
data.frame(
  estimate =
    unname(
      one_sample_test$estimate
    ),
  null_value =
    unname(
      one_sample_test$null.value
    ),
  statistic =
    unname(
      one_sample_test$statistic
    ),
  degrees_of_freedom =
    unname(
      one_sample_test$parameter
    ),
  p_value =
    one_sample_test$p.value,
  confidence_lower =
    one_sample_test$conf.int[1],
  confidence_upper =
    one_sample_test$conf.int[2]
)
```

# 7. Manual one-sample test statistic

\[
t
=
\frac{
\bar{x}-\mu_0
}{
s/\sqrt{n}
}.
\]

```{r manual-one-sample-t}
observed_scores <- final_scores[
  !is.na(final_scores)
]

sample_mean <- mean(
  observed_scores
)

sample_sd <- sd(
  observed_scores
)

sample_size <- length(
  observed_scores
)

null_mean <- 70

manual_t <- (
  sample_mean -
    null_mean
) / (
  sample_sd /
    sqrt(sample_size)
)

manual_t
```

# 8. One-sided tests

```{r one-sided-t-tests}
greater_test <- t.test(
  observed_scores,
  mu = 70,
  alternative = "greater"
)

less_test <- t.test(
  observed_scores,
  mu = 70,
  alternative = "less"
)

greater_test
less_test
```

The direction must be specified before looking at the result.

# 9. Independent-samples t-test

Select two programmes:

```{r select-two-programmes}
two_programmes <- student_data[
  student_data$programme %in%
    c(
      "Statistics",
      "Procurement"
    ),
]

table(
  two_programmes$programme
)
```

Welch test:

```{r welch-two-sample-test}
welch_test <- t.test(
  final_score ~ programme,
  data =
    two_programmes,
  var.equal = FALSE
)

welch_test
```

# 10. Pooled-variance t-test

```{r pooled-two-sample-test}
pooled_test <- t.test(
  final_score ~ programme,
  data =
    two_programmes,
  var.equal = TRUE
)

pooled_test
```

Welch's test is generally preferred unless equal variances are well justified.

# 11. Group summaries

```{r t-test-group-summaries}
aggregate(
  final_score ~ programme,
  data =
    two_programmes,
  FUN = function(x) {
    c(
      n = length(x),
      mean = mean(x),
      sd = sd(x)
    )
  }
)
```

# 12. Variance comparison

```{r variance-ratio-test}
var.test(
  final_score ~ programme,
  data =
    two_programmes
)
```

This test is sensitive to non-normality and should not be used mechanically to
choose the t-test version.

# 13. Paired t-test

```{r paired-test-data}
before_scores <- c(
  60,
  65,
  70,
  72,
  68,
  75,
  66,
  71
)

after_scores <- c(
  64,
  67,
  74,
  73,
  72,
  80,
  70,
  74
)

paired_test <- t.test(
  after_scores,
  before_scores,
  paired = TRUE
)

paired_test
```

# 14. Analysing paired differences

```{r paired-differences}
differences <-
  after_scores -
  before_scores

summary(differences)
mean(differences)
sd(differences)

t.test(
  differences,
  mu = 0
)
```

A paired t-test is a one-sample t-test on the within-pair differences.

# 15. Assumption checks

The one-sample and paired t-tests assume the observations or differences are
approximately normal when sample sizes are small.

The independent t-test also assumes:

- independent groups;
- numerical outcome;
- no severe outlier domination;
- and, for the pooled test, equal population variances.

# 16. Q-Q plot for one-sample test

```{r one-sample-qqplot}
qqnorm(
  observed_scores,
  main = "Q-Q Plot of Final Scores"
)

qqline(
  observed_scores
)
```

# 17. Q-Q plot of paired differences

```{r paired-difference-qqplot}
qqnorm(
  differences,
  main = "Q-Q Plot of Paired Differences"
)

qqline(
  differences
)
```

# 18. Shapiro-Wilk checks

```{r t-test-shapiro}
shapiro.test(
  observed_scores
)

shapiro.test(
  differences
)
```

Normality tests should be interpreted with plots and sample size.

# 19. Effect size for one-sample test

Cohen's \(d\):

\[
d
=
\frac{
\bar{x}-\mu_0
}{
s
}.
\]

```{r one-sample-cohens-d}
one_sample_d <- (
  sample_mean -
    null_mean
) / sample_sd

one_sample_d
```

# 20. Effect size for independent groups

```{r independent-cohens-d}
group_split <- split(
  two_programmes$final_score,
  two_programmes$programme
)

group_names <- names(
  group_split
)

group_1 <- group_split[[
  group_names[1]
]]

group_2 <- group_split[[
  group_names[2]
]]

n_1 <- length(group_1)
n_2 <- length(group_2)

pooled_sd <- sqrt(
  (
    (
      n_1 - 1
    ) *
    var(group_1) +
    (
      n_2 - 1
    ) *
    var(group_2)
  ) /
  (
    n_1 +
    n_2 -
    2
  )
)

independent_d <- (
  mean(group_1) -
    mean(group_2)
) / pooled_sd

independent_d
```

# 21. Effect size for paired data

```{r paired-cohens-d}
paired_d <- mean(
  differences
) / sd(
  differences
)

paired_d
```

# 22. Type I and Type II errors

**Type I error:** rejecting a true null hypothesis.

**Type II error:** failing to reject a false null hypothesis.

Power is:

\[
1-\beta.
\]

It is the probability of rejecting the null when a specified alternative is
true.

# 23. Power calculation

```{r power-one-sample-example}
power.t.test(
  n = 30,
  delta = 5,
  sd = 12,
  sig.level = 0.05,
  type = "one.sample",
  alternative = "two.sided"
)
```

Required sample size:

```{r power-required-sample-size}
power.t.test(
  power = 0.80,
  delta = 5,
  sd = 12,
  sig.level = 0.05,
  type = "one.sample",
  alternative = "two.sided"
)
```

# 24. Statistical versus practical significance

A small p-value does not guarantee an important effect.

A useful interpretation considers:

- effect size;
- confidence interval;
- sample size;
- measurement scale;
- and substantive context.

# 25. Decision language

Preferred language:

- reject the null hypothesis;
- fail to reject the null hypothesis;
- evidence supports a difference;
- evidence is insufficient to support a difference.

Avoid saying the null hypothesis has been proved true.

# 26. Reusable one-sample reporting function

```{r one-sample-report-function}
report_one_sample_t <- function(
  x,
  null_mean,
  confidence_level = 0.95,
  alternative = c(
    "two.sided",
    "less",
    "greater"
  )
) {
  alternative <- match.arg(
    alternative
  )

  observed <- x[
    !is.na(x)
  ]

  result <- t.test(
    observed,
    mu = null_mean,
    alternative =
      alternative,
    conf.level =
      confidence_level
  )

  data.frame(
    n = length(observed),
    mean = mean(observed),
    standard_deviation =
      sd(observed),
    null_mean =
      null_mean,
    mean_difference =
      mean(observed) -
      null_mean,
    t =
      unname(
        result$statistic
      ),
    degrees_of_freedom =
      unname(
        result$parameter
      ),
    p_value =
      result$p.value,
    confidence_lower =
      result$conf.int[1],
    confidence_upper =
      result$conf.int[2],
    cohens_d =
      (
        mean(observed) -
        null_mean
      ) /
      sd(observed)
  )
}

report_one_sample_t(
  final_scores,
  null_mean = 70
)
```

# 27. Reusable independent t-test report

```{r independent-t-report-function}
report_independent_t <- function(
  data,
  outcome,
  group,
  confidence_level = 0.95
) {
  required <- c(
    outcome,
    group
  )

  if (!all(
    required %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  group_values <- unique(
    data[[
      group
    ]][
      !is.na(
        data[[
          group
        ]]
      )
    ]
  )

  if (length(group_values) != 2L) {
    stop(
      "Exactly two groups are required.",
      call. = FALSE
    )
  }

  formula_object <- reformulate(
    group,
    response =
      outcome
  )

  result <- t.test(
    formula_object,
    data = data,
    var.equal = FALSE,
    conf.level =
      confidence_level
  )

  data.frame(
    group_1 =
      group_values[1],
    group_2 =
      group_values[2],
    estimate_1 =
      unname(
        result$estimate[1]
      ),
    estimate_2 =
      unname(
        result$estimate[2]
      ),
    mean_difference =
      unname(
        result$estimate[1] -
        result$estimate[2]
      ),
    t =
      unname(
        result$statistic
      ),
    degrees_of_freedom =
      unname(
        result$parameter
      ),
    p_value =
      result$p.value,
    confidence_lower =
      result$conf.int[1],
    confidence_upper =
      result$conf.int[2]
  )
}

report_independent_t(
  two_programmes,
  outcome = "final_score",
  group = "programme"
)
```

# 28. Reusable paired t-test report

```{r paired-t-report-function}
report_paired_t <- function(
  before,
  after,
  confidence_level = 0.95
) {
  complete_status <- complete.cases(
    before,
    after
  )

  before_complete <- before[
    complete_status
  ]

  after_complete <- after[
    complete_status
  ]

  differences <-
    after_complete -
    before_complete

  result <- t.test(
    after_complete,
    before_complete,
    paired = TRUE,
    conf.level =
      confidence_level
  )

  data.frame(
    n = length(
      differences
    ),
    before_mean =
      mean(
        before_complete
      ),
    after_mean =
      mean(
        after_complete
      ),
    mean_difference =
      mean(
        differences
      ),
    t =
      unname(
        result$statistic
      ),
    degrees_of_freedom =
      unname(
        result$parameter
      ),
    p_value =
      result$p.value,
    confidence_lower =
      result$conf.int[1],
    confidence_upper =
      result$conf.int[2],
    cohens_d =
      mean(
        differences
      ) /
      sd(
        differences
      )
  )
}

report_paired_t(
  before_scores,
  after_scores
)
```

# 29. Reporting template

A complete result may be written as:

> The mean score was compared with the reference value using a one-sample
> t-test. The report includes the sample mean, standard deviation, mean
> difference, test statistic, degrees of freedom, p-value, confidence interval
> and effect size.

# 30. Common mistakes

- choosing one-sided hypotheses after seeing the result;
- interpreting p-values as effect sizes;
- saying the null hypothesis is accepted or proved;
- using an independent test for paired data;
- using a paired test for independent groups;
- selecting pooled variance mechanically;
- ignoring outliers and data quality;
- reporting p-values without confidence intervals;
- failing to report group means and standard deviations;
- ignoring practical significance.

# 31. Independent exercises

1. formulate null and alternative hypotheses.
2. conduct a one-sample t-test.
3. calculate the test statistic manually.
4. compare one-sided and two-sided tests.
5. conduct Welch and pooled independent tests.
6. create group summaries.
7. conduct a paired t-test.
8. analyse paired differences.
9. inspect assumptions.
10. calculate Cohen's d.
11. explain Type I and Type II errors.
12. conduct a power calculation.
13. create reusable t-test functions.
14. write a complete interpretation.

# 32. Lesson assessment

Create a standalone R Markdown report that:

1. states hypotheses;
2. conducts a one-sample test;
3. conducts an independent-samples test;
4. conducts a paired test;
5. checks assumptions;
6. reports confidence intervals;
7. calculates effect sizes;
8. explains errors and power;
9. uses reusable reporting functions; and
10. distinguishes statistical and practical significance.

# 33. Completion checklist

Confirm that you can:

- formulate hypotheses;
- explain p-values;
- conduct one-sample t-tests;
- conduct Welch tests;
- conduct pooled tests;
- conduct paired tests;
- assess assumptions;
- calculate confidence intervals;
- calculate effect sizes;
- explain Type I and Type II errors;
- conduct basic power calculations;
- write reusable test functions; and
- report results accurately.

# Lesson conclusion

Hypothesis testing evaluates evidence against a specified null model. Good
reporting combines p-values with estimates, confidence intervals, effect sizes
and substantive interpretation.

\[
\boxed{
\text{question}
\rightarrow
\text{hypotheses}
\rightarrow
\text{test statistic}
\rightarrow
\text{p-value}
\rightarrow
\text{effect size and interval}
\rightarrow
\text{interpretation}
}
\]
