---
title: "Lesson 33: Analysis of Variance and Post Hoc Comparisons"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Analysis of variance compares the means of three or more groups through a
single inferential framework.

Rather than conducting many separate t-tests, ANOVA partitions total variation
into:

- variation between groups; and
- variation within groups.

The F statistic compares these two sources of variation.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain the logic of one-way ANOVA;
2. state ANOVA hypotheses;
3. calculate group summaries;
4. fit an ANOVA model with `aov()` and `lm()`;
5. interpret the ANOVA table;
6. assess assumptions;
7. calculate effect sizes;
8. conduct Tukey post hoc comparisons;
9. perform planned contrasts;
10. distinguish omnibus and pairwise tests;
11. conduct Welch's ANOVA;
12. create diagnostic plots;
13. build reusable ANOVA reporting functions; and
14. report results accurately.

# 1. Importing data

```{r import-anova-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_data$programme <- factor(
  student_data$programme
)
```

# 2. Research question

Do mean final scores differ across programmes?

The hypotheses are:

\[
H_0:
\mu_1 =
\mu_2 =
\mu_3 =
\cdots
\]

\[
H_1:
\text{At least one group mean differs.}
\]

# 3. Group summaries

```{r anova-group-summary}
aggregate(
  final_score ~ programme,
  data = student_data,
  FUN = function(x) {
    c(
      n = length(x),
      mean = mean(x),
      sd = sd(x)
    )
  }
)
```

# 4. Visual comparison

```{r anova-boxplot}
boxplot(
  final_score ~ programme,
  data = student_data,
  main = "Final Scores by Programme",
  xlab = "Programme",
  ylab = "Final score",
  las = 2
)
```

# 5. Fitting one-way ANOVA

```{r fit-one-way-anova}
anova_model <- aov(
  final_score ~ programme,
  data = student_data
)

summary(
  anova_model
)
```

# 6. ANOVA through linear modelling

```{r anova-lm-equivalent}
linear_model <- lm(
  final_score ~ programme,
  data = student_data
)

anova(
  linear_model
)
```

# 7. ANOVA decomposition

Total variation is divided into:

\[
SS_{\text{Total}}
=
SS_{\text{Between}}
+
SS_{\text{Within}}.
\]

The F statistic is:

\[
F
=
\frac{
MS_{\text{Between}}
}{
MS_{\text{Within}}
}.
\]

# 8. Extracting the ANOVA table

```{r extract-anova-table}
anova_table <- summary(
  anova_model
)[[1]]

anova_table
```

# 9. Manual F-ratio extraction

```{r extract-anova-components}
between_mean_square <-
  anova_table[
    "programme",
    "Mean Sq"
  ]

within_mean_square <-
  anova_table[
    "Residuals",
    "Mean Sq"
  ]

manual_f <-
  between_mean_square /
  within_mean_square

manual_f
```

# 10. Assumption: independent observations

Independence is primarily a design assumption. It cannot be verified by a
single statistical test.

Potential violations include:

- repeated measures treated as independent;
- clustered observations;
- multiple records from the same respondent;
- paired data analysed as separate groups.

# 11. Assumption: normality of residuals

```{r anova-residuals}
anova_residuals <- residuals(
  anova_model
)

hist(
  anova_residuals,
  main = "Distribution of ANOVA Residuals",
  xlab = "Residual"
)

qqnorm(
  anova_residuals,
  main = "Q-Q Plot of ANOVA Residuals"
)

qqline(
  anova_residuals
)
```

# 12. Shapiro-Wilk test of residuals

```{r anova-shapiro}
shapiro.test(
  anova_residuals
)
```

Use this together with the Q-Q plot and sample size.

# 13. Assumption: homogeneity of variances

```{r bartlett-test}
bartlett.test(
  final_score ~ programme,
  data = student_data
)
```

Bartlett's test is sensitive to non-normality.

# 14. Fligner-Killeen test

```{r fligner-test}
fligner.test(
  final_score ~ programme,
  data = student_data
)
```

This is more robust to non-normality.

# 15. Diagnostic plots

```{r anova-diagnostic-plots}
old_parameters <- par(
  mfrow = c(
    2,
    2
  )
)

plot(
  anova_model
)

par(
  old_parameters
)
```

# 16. Eta squared

\[
\eta^2
=
\frac{
SS_{\text{Between}}
}{
SS_{\text{Total}}
}.
\]

```{r eta-squared}
between_sum_squares <-
  anova_table[
    "programme",
    "Sum Sq"
  ]

within_sum_squares <-
  anova_table[
    "Residuals",
    "Sum Sq"
  ]

total_sum_squares <-
  between_sum_squares +
  within_sum_squares

eta_squared <-
  between_sum_squares /
  total_sum_squares

eta_squared
```

# 17. Omega squared

\[
\omega^2
=
\frac{
SS_{\text{Between}}
-
df_{\text{Between}}
MS_{\text{Within}}
}{
SS_{\text{Total}}
+
MS_{\text{Within}}
}.
\]

```{r omega-squared}
between_df <-
  anova_table[
    "programme",
    "Df"
  ]

omega_squared <- (
  between_sum_squares -
    between_df *
    within_mean_square
) / (
  total_sum_squares +
    within_mean_square
)

omega_squared
```

# 18. Tukey post hoc comparisons

```{r tukey-post-hoc}
tukey_results <- TukeyHSD(
  anova_model
)

tukey_results
```

# 19. Plotting Tukey intervals

```{r tukey-plot}
plot(
  tukey_results,
  las = 1
)
```

# 20. Extracting Tukey results

```{r extract-tukey-results}
tukey_table <- as.data.frame(
  tukey_results$programme
)

tukey_table$comparison <-
  rownames(
    tukey_table
  )

rownames(
  tukey_table
) <- NULL

tukey_table
```

# 21. Planned contrasts

Suppose the planned comparison is Statistics versus the average of the other
observed programmes.

The number of contrast weights must equal the number of factor levels. The
weights are therefore constructed dynamically rather than assuming a fixed
number of programmes.

```{r planned-contrast}
student_data$programme_planned <- droplevels(
  student_data$programme
)

programme_levels <- levels(
  student_data$programme_planned
)

number_of_programmes <- length(
  programme_levels
)

if (
  number_of_programmes < 2L
) {
  stop(
    "At least two programme levels are required.",
    call. = FALSE
  )
}

target_programme <- "Statistics"

if (!target_programme %in% programme_levels) {
  stop(
    paste(
      "The target programme",
      shQuote(target_programme),
      "is not present in the data."
    ),
    call. = FALSE
  )
}

planned_weights <- rep(
  -1,
  number_of_programmes
)

names(
  planned_weights
) <- programme_levels

planned_weights[
  target_programme
] <- number_of_programmes - 1

planned_weights
sum(
  planned_weights
)
```

Create a full contrast matrix with one row for every observed programme:

```{r planned-contrast-matrix}
contrast_matrix <- contr.helmert(
  number_of_programmes
)

contrast_matrix[
  ,
  1
] <- planned_weights

rownames(
  contrast_matrix
) <- programme_levels

contrast_matrix
```

Apply the matrix to a separate factor so the original programme coding remains
unchanged:

```{r apply-planned-contrast}
contrasts(
  student_data$programme_planned
) <- contrast_matrix

planned_model <- lm(
  final_score ~ programme_planned,
  data = student_data
)

summary(
  planned_model
)
```

The first coefficient for `programme_planned` represents the planned
Statistics-versus-others comparison. The remaining columns complete the factor
coding and keep the model full rank.

# 22. Welch's ANOVA

```{r welch-anova}
oneway.test(
  final_score ~ programme,
  data = student_data,
  var.equal = FALSE
)
```

Welch's ANOVA is useful when group variances differ.

# 23. Pairwise t-tests

```{r pairwise-t-tests}
pairwise.t.test(
  student_data$final_score,
  student_data$programme,
  p.adjust.method = "holm",
  pool.sd = FALSE
)
```

Multiplicity adjustment is essential when performing several comparisons.

# 24. Omnibus versus post hoc testing

The omnibus ANOVA asks:

> Is there evidence that at least one mean differs?

Post hoc comparisons ask:

> Which group pairs differ?

A significant omnibus test does not identify the specific groups.

# 25. Reusable ANOVA reporting function

```{r anova-report-function}
report_one_way_anova <- function(
  data,
  outcome,
  group
) {
  required <- c(
    outcome,
    group
  )

  if (!all(
    required %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  formula_object <- reformulate(
    group,
    response = outcome
  )

  model <- aov(
    formula_object,
    data = data
  )

  table_result <- summary(
    model
  )[[1]]

  between_ss <-
    table_result[
      1,
      "Sum Sq"
    ]

  residual_ss <-
    table_result[
      2,
      "Sum Sq"
    ]

  total_ss <-
    between_ss +
    residual_ss

  data.frame(
    outcome = outcome,
    group = group,
    df_between =
      table_result[
        1,
        "Df"
      ],
    df_within =
      table_result[
        2,
        "Df"
      ],
    f =
      table_result[
        1,
        "F value"
      ],
    p_value =
      table_result[
        1,
        "Pr(>F)"
      ],
    eta_squared =
      between_ss /
      total_ss,
    stringsAsFactors = FALSE
  )
}

report_one_way_anova(
  student_data,
  "final_score",
  "programme"
)
```

# 26. Reusable group summary function

```{r anova-group-summary-function}
create_anova_group_summary <- function(
  data,
  outcome,
  group
) {
  groups <- split(
    data[[
      outcome
    ]],
    data[[
      group
    ]]
  )

  results <- lapply(
    names(groups),
    function(group_name) {
      x <- groups[[
        group_name
      ]]

      data.frame(
        group = group_name,
        n = sum(
          !is.na(x)
        ),
        mean = mean(
          x,
          na.rm = TRUE
        ),
        standard_deviation = sd(
          x,
          na.rm = TRUE
        ),
        stringsAsFactors = FALSE
      )
    }
  )

  result <- do.call(
    rbind,
    results
  )

  names(result)[1] <- group

  result
}

create_anova_group_summary(
  student_data,
  "final_score",
  "programme"
)
```

# 27. Reporting template

A complete report includes:

- group sample sizes;
- group means and standard deviations;
- F statistic;
- degrees of freedom;
- p-value;
- effect size;
- post hoc results;
- and assumption checks.

# 28. Common mistakes

- conducting multiple unadjusted t-tests;
- interpreting a significant F test as proof that every group differs;
- ignoring group sample sizes;
- using pooled ANOVA under severe variance inequality;
- testing normality of the raw outcome instead of residuals;
- relying only on Bartlett's test;
- omitting effect sizes;
- conducting post hoc tests without a clear purpose;
- confusing planned contrasts and post hoc tests;
- ignoring dependence in repeated observations.

# 29. Independent exercises

1. state ANOVA hypotheses.
2. create group summaries.
3. fit an ANOVA model.
4. extract the ANOVA table.
5. calculate the F ratio manually.
6. inspect residual assumptions.
7. test variance homogeneity.
8. calculate eta squared and omega squared.
9. conduct Tukey comparisons.
10. conduct adjusted pairwise t-tests.
11. conduct Welch's ANOVA.
12. create a planned contrast.
13. create reusable reporting functions.
14. write a complete interpretation.

# 30. Lesson assessment

Create a standalone R Markdown report that:

1. states the research question and hypotheses;
2. reports group summaries;
3. fits a one-way ANOVA;
4. assesses residual and variance assumptions;
5. calculates effect sizes;
6. conducts Tukey post hoc comparisons;
7. compares classical and Welch ANOVA;
8. includes one planned contrast;
9. uses reusable functions; and
10. reports the findings accurately.

# 31. Completion checklist

Confirm that you can:

- explain ANOVA logic;
- state hypotheses;
- fit one-way ANOVA;
- interpret an ANOVA table;
- inspect residuals;
- test homogeneity;
- calculate eta squared;
- calculate omega squared;
- conduct Tukey comparisons;
- conduct Welch's ANOVA;
- use planned contrasts; and
- report ANOVA results.

# Lesson conclusion

ANOVA provides an omnibus test of mean differences across several groups.
Complete analysis combines group summaries, diagnostic checks, effect sizes and
appropriate follow-up comparisons.

\[
\boxed{
\text{group summaries}
\rightarrow
\text{ANOVA}
\rightarrow
\text{diagnostics}
\rightarrow
\text{effect size}
\rightarrow
\text{post hoc interpretation}
}
\]
