---
title: "Lesson 34: Chi-Square Tests for Categorical Data"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Chi-square methods analyse categorical frequencies. They compare observed
counts with counts expected under a null hypothesis.

Common applications include:

- goodness-of-fit tests;
- tests of independence;
- tests of homogeneity;
- and analysis of contingency tables.

This lesson also introduces expected counts, residuals, Fisher's exact test and
Cramer's V.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish goodness-of-fit and independence tests;
2. construct contingency tables;
3. calculate expected frequencies;
4. conduct chi-square goodness-of-fit tests;
5. conduct chi-square tests of independence;
6. inspect assumptions;
7. use Fisher's exact test;
8. calculate standardised residuals;
9. identify influential cells;
10. calculate Cramer's V;
11. interpret row and column percentages;
12. create reusable chi-square reporting functions; and
13. report categorical associations accurately.

# 1. Importing data

```{r import-chi-square-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_data$performance_group <- ifelse(
  student_data$final_score >= 75,
  "75 or above",
  "Below 75"
)

student_data$performance_group <- factor(
  student_data$performance_group,
  levels = c(
    "Below 75",
    "75 or above"
  )
)
```

The cut-off of 75 is used only to create two observed performance categories
for this teaching example. It is not presented as an institutional pass mark.
A chi-square test of independence requires variation in both categorical
variables.

# 2. Goodness-of-fit question

Suppose the null hypothesis states that students are equally distributed across
programmes.

```{r programme-frequency-goodness}
programme_counts <- table(
  student_data$programme
)

programme_counts
```

# 3. Goodness-of-fit test

```{r goodness-of-fit-test}
goodness_test <- chisq.test(
  programme_counts,
  p = rep(
    1 /
      length(
        programme_counts
      ),
    length(
      programme_counts
    )
  )
)

goodness_test
```

# 4. Unequal expected proportions

```{r unequal-goodness-proportions}
number_of_programmes <- length(
  programme_counts
)

expected_proportions <- seq_len(
  number_of_programmes
)

expected_proportions <- expected_proportions /
  sum(
    expected_proportions
  )

names(
  expected_proportions
) <- names(
  programme_counts
)

sum(
  expected_proportions
)

chisq.test(
  programme_counts,
  p = expected_proportions
)
```

# 5. Contingency table

```{r contingency-table}
programme_result_table <- table(
  student_data$programme,
  student_data$performance_group
)

if (
  nrow(
    programme_result_table
  ) < 2L ||
  ncol(
    programme_result_table
  ) < 2L
) {
  stop(
    paste(
      "The contingency table must contain at least two observed rows and",
      "two observed columns."
    ),
    call. = FALSE
  )
}

programme_result_table
```

# 6. Chi-square test of independence

```{r chi-square-independence}
independence_test <- chisq.test(
  programme_result_table,
  correct = FALSE
)

independence_test
```

# 7. Expected counts

```{r expected-counts}
independence_test$expected
```

Expected frequency in cell \(ij\):

\[
E_{ij}
=
\frac{
(\text{row total}_i)
(\text{column total}_j)
}{
\text{grand total}
}.
\]

# 8. Manual expected counts

```{r manual-expected-counts}
row_totals <- rowSums(
  programme_result_table
)

column_totals <- colSums(
  programme_result_table
)

grand_total <- sum(
  programme_result_table
)

manual_expected <- outer(
  row_totals,
  column_totals
) / grand_total

manual_expected
```

# 9. Manual chi-square statistic

\[
\chi^2
=
\sum
\frac{
(O-E)^2
}{
E
}.
\]

```{r manual-chi-square}
manual_chi_square <- sum(
  (
    programme_result_table -
      manual_expected
  ) ^ 2 /
    manual_expected
)

manual_chi_square
```

# 10. Assumption checks

```{r chi-square-assumptions}
expected_values <- as.vector(
  independence_test$expected
)

data.frame(
  minimum_expected =
    min(
      expected_values
    ),
  cells_below_five =
    sum(
      expected_values < 5
    ),
  percentage_below_five =
    mean(
      expected_values < 5
    ) * 100
)
```

A common guideline is:

- no expected count below 1;
- no more than about 20 per cent below 5.

# 11. Fisher's exact test

```{r fisher-exact-test}
if (
  nrow(
    programme_result_table
  ) >= 2L &&
  ncol(
    programme_result_table
  ) >= 2L
) {
  fisher_result <- fisher.test(
    programme_result_table
  )

  fisher_result
} else {
  message(
    paste(
      "Fisher's exact test was not conducted because the table has fewer",
      "than two observed rows or columns."
    )
  )
}
```

Fisher's test is especially useful for small contingency tables.

# 12. Simulated p-values

```{r simulated-chi-square-p}
chisq.test(
  programme_result_table,
  simulate.p.value = TRUE,
  B = 5000
)
```

# 13. Pearson residuals

```{r pearson-residuals}
independence_test$residuals
```

Positive residuals indicate more observations than expected.

Negative residuals indicate fewer observations than expected.

# 14. Standardised residuals

```{r standardised-residuals}
independence_test$stdres
```

Cells with large absolute residuals contribute strongly to the association.

# 15. Residual contribution table

```{r chi-square-contributions}
cell_contributions <- (
  programme_result_table -
    independence_test$expected
) ^ 2 /
  independence_test$expected

cell_contributions
```

# 16. Row percentages

```{r chi-square-row-percentages}
row_percentages <- prop.table(
  programme_result_table,
  margin = 1
) * 100

row_percentages
```

# 17. Column percentages

```{r chi-square-column-percentages}
column_percentages <- prop.table(
  programme_result_table,
  margin = 2
) * 100

column_percentages
```

# 18. Cramer's V

\[
V
=
\sqrt{
\frac{
\chi^2
}{
n
\min(r-1,c-1)
}
}.
\]

```{r cramers-v}
chi_square_value <- unname(
  independence_test$statistic
)

n_total <- sum(
  programme_result_table
)

number_rows <- nrow(
  programme_result_table
)

number_columns <- ncol(
  programme_result_table
)

cramers_v <- sqrt(
  chi_square_value /
    (
      n_total *
      min(
        number_rows - 1,
        number_columns - 1
      )
    )
)

cramers_v
```

# 19. Phi coefficient for 2 by 2 tables

```{r phi-example}
small_table <- matrix(
  c(
    30,
    20,
    15,
    35
  ),
  nrow = 2,
  byrow = TRUE
)

small_test <- chisq.test(
  small_table,
  correct = FALSE
)

phi <- sqrt(
  unname(
    small_test$statistic
  ) /
    sum(
      small_table
    )
)

phi
```

# 20. Yates continuity correction

```{r yates-correction}
chisq.test(
  small_table,
  correct = TRUE
)

chisq.test(
  small_table,
  correct = FALSE
)
```

Yates correction applies to 2 by 2 tables and is conservative.

# 21. Visualising a contingency table

```{r contingency-mosaic}
mosaicplot(
  programme_result_table,
  main = "Programme and Result",
  xlab = "Programme",
  ylab = "Result"
)
```

# 22. Goodness-of-fit reporting function

```{r goodness-report-function}
report_goodness_of_fit <- function(
  observed_counts,
  expected_proportions = NULL
) {
  if (is.null(
    expected_proportions
  )) {
    expected_proportions <- rep(
      1 /
        length(
          observed_counts
        ),
      length(
        observed_counts
      )
    )
  }

  if (abs(
    sum(
      expected_proportions
    ) - 1
  ) > 1e-8) {
    stop(
      "Expected proportions must sum to one.",
      call. = FALSE
    )
  }

  result <- chisq.test(
    observed_counts,
    p = expected_proportions
  )

  data.frame(
    chi_square =
      unname(
        result$statistic
      ),
    degrees_of_freedom =
      unname(
        result$parameter
      ),
    p_value =
      result$p.value,
    minimum_expected =
      min(
        result$expected
      ),
    stringsAsFactors = FALSE
  )
}

report_goodness_of_fit(
  programme_counts
)
```

# 23. Independence reporting function

```{r independence-report-function}
report_chi_square_independence <- function(
  table_object
) {
  if (!is.matrix(
    table_object
  ) &&
      !inherits(
        table_object,
        "table"
      )) {
    stop(
      "`table_object` must be a table or matrix.",
      call. = FALSE
    )
  }

  result <- chisq.test(
    table_object,
    correct = FALSE
  )

  total_n <- sum(
    table_object
  )

  r <- nrow(
    table_object
  )

  c <- ncol(
    table_object
  )

  v <- sqrt(
    unname(
      result$statistic
    ) /
      (
        total_n *
        min(
          r - 1,
          c - 1
        )
      )
  )

  data.frame(
    chi_square =
      unname(
        result$statistic
      ),
    degrees_of_freedom =
      unname(
        result$parameter
      ),
    p_value =
      result$p.value,
    cramers_v =
      v,
    minimum_expected =
      min(
        result$expected
      ),
    cells_below_five =
      sum(
        result$expected < 5
      ),
    stringsAsFactors = FALSE
  )
}

report_chi_square_independence(
  programme_result_table
)
```

# 24. Reporting template

A complete report states:

- variables;
- observed counts;
- percentage basis;
- chi-square statistic;
- degrees of freedom;
- p-value;
- expected-count assessment;
- effect size;
- and influential cells.

# 25. Common mistakes

- applying chi-square tests to percentages rather than counts;
- ignoring expected-frequency assumptions;
- treating association as causation;
- failing to report row or column percentages;
- using goodness-of-fit when independence is required;
- ignoring influential cells;
- reporting significance without Cramer's V;
- combining sparse categories without justification;
- using Fisher's exact test without explaining why;
- analysing non-independent repeated records.

# 26. Independent exercises

1. create a goodness-of-fit table.
2. test equal expected proportions.
3. test unequal expected proportions.
4. create a contingency table.
5. conduct a chi-square independence test.
6. calculate expected counts manually.
7. calculate the chi-square statistic manually.
8. inspect assumptions.
9. conduct Fisher's exact test.
10. inspect residuals and cell contributions.
11. calculate row and column percentages.
12. calculate Cramer's V.
13. create reusable reporting functions.
14. write a complete interpretation.

# 27. Lesson assessment

Create a standalone R Markdown report that:

1. conducts a goodness-of-fit test;
2. conducts an independence test;
3. presents observed and expected counts;
4. checks expected-frequency assumptions;
5. compares chi-square and Fisher methods;
6. identifies influential cells;
7. reports row or column percentages;
8. calculates Cramer's V;
9. uses reusable functions; and
10. interprets the association without causal claims.

# 28. Completion checklist

Confirm that you can:

- distinguish chi-square test types;
- create contingency tables;
- calculate expected counts;
- conduct goodness-of-fit tests;
- conduct independence tests;
- inspect assumptions;
- use Fisher's exact test;
- inspect residuals;
- calculate cell contributions;
- calculate Cramer's V;
- create mosaic plots; and
- report categorical associations.

# Lesson conclusion

Chi-square methods compare observed and expected categorical frequencies.
Reliable analysis requires counts, adequate expected frequencies, effect-size
reporting and careful interpretation.

\[
\boxed{
\text{observed counts}
\rightarrow
\text{expected counts}
\rightarrow
\chi^2
\rightarrow
\text{residuals}
\rightarrow
\text{effect size}
\rightarrow
\text{interpretation}
}
\]
