---
title: "Lesson 35: Non-Parametric Tests and Rank-Based Methods"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Non-parametric and rank-based methods provide alternatives when parametric
assumptions are unsuitable or when the data are ordinal.

These procedures usually test distributional or location differences using
ranks rather than raw values.

This lesson covers:

- sign and Wilcoxon signed-rank tests;
- Mann-Whitney-Wilcoxon rank-sum tests;
- Kruskal-Wallis tests;
- post hoc rank comparisons;
- and Spearman correlation.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain when rank-based methods are appropriate;
2. distinguish independent and paired rank tests;
3. conduct one-sample Wilcoxon signed-rank tests;
4. conduct paired Wilcoxon tests;
5. conduct rank-sum tests;
6. conduct Kruskal-Wallis tests;
7. conduct adjusted post hoc comparisons;
8. calculate effect sizes;
9. interpret medians and distributions cautiously;
10. compare parametric and non-parametric results;
11. use exact and approximate p-values appropriately;
12. create reusable rank-test reporting functions; and
13. report results accurately.

# 1. When to consider rank-based methods

Rank-based methods may be useful when:

- the outcome is ordinal;
- distributions are strongly skewed;
- extreme values dominate parametric procedures;
- sample sizes are small;
- normality assumptions are implausible;
- or the research question concerns ordering rather than means.

They do not automatically solve:

- dependence;
- poor sampling design;
- measurement error;
- or confounding.

# 2. One-sample Wilcoxon signed-rank test

```{r one-sample-wilcoxon-data}
scores <- c(
  62,
  68,
  70,
  74,
  75,
  80,
  82,
  66,
  71,
  77
)

wilcox.test(
  scores,
  mu = 70,
  alternative = "two.sided",
  exact = FALSE
)
```

# 3. Sign test concept

The signed-rank test uses both:

- direction of differences; and
- ranks of absolute differences.

A simple sign test uses only direction.

```{r simple-sign-test}
differences <- scores - 70

positive_count <- sum(
  differences > 0
)

nonzero_count <- sum(
  differences != 0
)

binom.test(
  positive_count,
  nonzero_count,
  p = 0.5
)
```

# 4. Paired Wilcoxon signed-rank test

```{r paired-wilcoxon-data}
before_scores <- c(
  60,
  65,
  70,
  72,
  68,
  75,
  66,
  71
)

after_scores <- c(
  64,
  67,
  74,
  73,
  72,
  80,
  70,
  74
)

paired_wilcoxon <- wilcox.test(
  after_scores,
  before_scores,
  paired = TRUE,
  exact = FALSE
)

paired_wilcoxon
```

# 5. Inspecting paired differences

```{r inspect-paired-rank-differences}
paired_differences <-
  after_scores -
  before_scores

paired_differences
median(paired_differences)
```

# 6. Independent rank-sum test

```{r import-rank-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

two_programmes <- student_data[
  student_data$programme %in%
    c(
      "Statistics",
      "Procurement"
    ),
]
```

```{r rank-sum-test}
rank_sum_test <- wilcox.test(
  final_score ~ programme,
  data = two_programmes,
  exact = FALSE,
  conf.int = TRUE
)

rank_sum_test
```

# 7. Group medians and ranks

```{r rank-group-summaries}
aggregate(
  final_score ~ programme,
  data = two_programmes,
  FUN = function(x) {
    c(
      n = length(x),
      median = median(x),
      iqr = IQR(x)
    )
  }
)
```

# 8. Kruskal-Wallis test

```{r kruskal-wallis-test}
kruskal_result <- kruskal.test(
  final_score ~ programme,
  data = student_data
)

kruskal_result
```

The null hypothesis concerns equality of group distributions. Under similarly
shaped distributions, it is often interpreted as a location comparison.

# 9. Post hoc pairwise Wilcoxon tests

```{r pairwise-wilcoxon}
pairwise.wilcox.test(
  student_data$final_score,
  student_data$programme,
  p.adjust.method = "holm",
  exact = FALSE
)
```

# 10. Comparing ANOVA and Kruskal-Wallis

```{r compare-anova-kruskal}
anova_result <- summary(
  aov(
    final_score ~ programme,
    data = student_data
  )
)[[1]]

comparison_table <- data.frame(
  method = c(
    "One-way ANOVA",
    "Kruskal-Wallis"
  ),
  statistic = c(
    anova_result[
      1,
      "F value"
    ],
    unname(
      kruskal_result$statistic
    )
  ),
  p_value = c(
    anova_result[
      1,
      "Pr(>F)"
    ],
    kruskal_result$p.value
  ),
  stringsAsFactors = FALSE
)

comparison_table
```

The tests address related but not identical hypotheses.

# 11. Effect size for rank-sum test

One common effect size is:

\[
r
=
\frac{
Z
}{
\sqrt{N}
}.
\]

Base R does not directly return \(Z\), so an approximate value can be derived
from the p-value and direction.

```{r rank-sum-effect-size}
group_levels <- unique(
  two_programmes$programme
)

group_one <- two_programmes$final_score[
  two_programmes$programme ==
    group_levels[1]
]

group_two <- two_programmes$final_score[
  two_programmes$programme ==
    group_levels[2]
]

direction <- sign(
  median(
    group_one
  ) -
  median(
    group_two
  )
)

approximate_z <- qnorm(
  rank_sum_test$p.value / 2,
  lower.tail = FALSE
) * direction

rank_effect_r <-
  approximate_z /
  sqrt(
    length(group_one) +
      length(group_two)
  )

rank_effect_r
```

# 12. Epsilon squared for Kruskal-Wallis

\[
\epsilon^2
=
\frac{
H-k+1
}{
n-k
}.
\]

```{r kruskal-effect-size}
h_value <- unname(
  kruskal_result$statistic
)

number_groups <- length(
  unique(
    student_data$programme
  )
)

total_n <- sum(
  complete.cases(
    student_data[
      c(
        "final_score",
        "programme"
      )
    ]
  )
)

epsilon_squared <- (
  h_value -
    number_groups +
    1
) / (
  total_n -
    number_groups
)

epsilon_squared
```

# 13. Spearman rank correlation

```{r rank-spearman}
cor.test(
  student_data$coursework_score,
  student_data$examination_score,
  method = "spearman",
  exact = FALSE
)
```

# 14. Kendall rank correlation

```{r rank-kendall}
cor.test(
  student_data$coursework_score,
  student_data$examination_score,
  method = "kendall",
  exact = FALSE
)
```

# 15. Ties

Ties occur when several observations share the same value.

Rank methods adjust for ties, but many ties can affect exact p-value
calculations.

```{r inspect-rank-ties}
table(
  student_data$final_score
)[
  table(
    student_data$final_score
  ) > 1
]
```

# 16. Exact versus approximate p-values

Exact calculations may be unavailable when:

- sample sizes are large;
- ties occur;
- or zero differences occur.

Use `exact = FALSE` when an asymptotic approximation is needed.

# 17. Visualisation before rank tests

```{r rank-boxplot}
boxplot(
  final_score ~ programme,
  data = student_data,
  main = "Final Scores by Programme",
  xlab = "Programme",
  ylab = "Final score",
  las = 2
)
```

# 18. One-sample Wilcoxon reporting function

```{r wilcoxon-one-sample-report}
report_one_sample_wilcoxon <- function(
  x,
  null_location = 0,
  alternative = c(
    "two.sided",
    "less",
    "greater"
  )
) {
  alternative <- match.arg(
    alternative
  )

  observed <- x[
    !is.na(x)
  ]

  result <- wilcox.test(
    observed,
    mu = null_location,
    alternative =
      alternative,
    exact = FALSE
  )

  data.frame(
    n = length(observed),
    median = median(observed),
    null_location =
      null_location,
    statistic =
      unname(
        result$statistic
      ),
    p_value =
      result$p.value,
    stringsAsFactors = FALSE
  )
}

report_one_sample_wilcoxon(
  scores,
  null_location = 70
)
```

# 19. Independent rank-sum reporting function

```{r rank-sum-report-function}
report_rank_sum <- function(
  data,
  outcome,
  group
) {
  required <- c(
    outcome,
    group
  )

  if (!all(
    required %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  groups <- unique(
    data[[
      group
    ]][
      !is.na(
        data[[
          group
        ]]
      )
    ]
  )

  if (length(groups) != 2L) {
    stop(
      "Exactly two groups are required.",
      call. = FALSE
    )
  }

  formula_object <- reformulate(
    group,
    response = outcome
  )

  result <- wilcox.test(
    formula_object,
    data = data,
    exact = FALSE
  )

  data.frame(
    group_1 = groups[1],
    group_2 = groups[2],
    median_1 = median(
      data[[
        outcome
      ]][
        data[[
          group
        ]] == groups[1]
      ],
      na.rm = TRUE
    ),
    median_2 = median(
      data[[
        outcome
      ]][
        data[[
          group
        ]] == groups[2]
      ],
      na.rm = TRUE
    ),
    statistic =
      unname(
        result$statistic
      ),
    p_value =
      result$p.value,
    stringsAsFactors = FALSE
  )
}

report_rank_sum(
  two_programmes,
  "final_score",
  "programme"
)
```

# 20. Kruskal-Wallis reporting function

```{r kruskal-report-function}
report_kruskal_wallis <- function(
  data,
  outcome,
  group
) {
  formula_object <- reformulate(
    group,
    response = outcome
  )

  result <- kruskal.test(
    formula_object,
    data = data
  )

  complete_data <- data[
    complete.cases(
      data[
        c(
          outcome,
          group
        )
      ]
    ),
  ]

  k <- length(
    unique(
      complete_data[[
        group
      ]]
    )
  )

  n <- nrow(
    complete_data
  )

  h <- unname(
    result$statistic
  )

  data.frame(
    h = h,
    degrees_of_freedom =
      unname(
        result$parameter
      ),
    p_value =
      result$p.value,
    epsilon_squared =
      (
        h -
        k +
        1
      ) /
      (
        n -
        k
      ),
    stringsAsFactors = FALSE
  )
}

report_kruskal_wallis(
  student_data,
  "final_score",
  "programme"
)
```

# 21. Reporting principles

A complete rank-test report includes:

- study design;
- group sample sizes;
- medians and IQRs;
- test statistic;
- p-value;
- effect size where appropriate;
- adjustment method for multiple comparisons;
- and cautious interpretation of the tested hypothesis.

# 22. Common mistakes

- choosing a rank test solely because a normality test is significant;
- interpreting rank tests automatically as median tests;
- using an independent rank test for paired data;
- ignoring ties and zero differences;
- conducting unadjusted post hoc tests;
- omitting medians and IQRs;
- comparing parametric and non-parametric p-values as if they test the same hypothesis;
- treating non-significance as equality;
- ignoring group shape differences;
- failing to report effect sizes.

# 23. Independent exercises

1. explain when rank methods are appropriate.
2. conduct a one-sample signed-rank test.
3. conduct a simple sign test.
4. conduct a paired Wilcoxon test.
5. conduct an independent rank-sum test.
6. create median and IQR summaries.
7. conduct a Kruskal-Wallis test.
8. conduct adjusted pairwise Wilcoxon tests.
9. compare ANOVA and Kruskal-Wallis.
10. calculate rank-based effect sizes.
11. inspect ties.
12. conduct Spearman and Kendall correlations.
13. create reusable rank-test functions.
14. write a complete interpretation.

# 24. Lesson assessment

Create a standalone R Markdown report that:

1. explains the choice of rank-based methods;
2. conducts one-sample and paired signed-rank tests;
3. conducts an independent rank-sum test;
4. conducts a Kruskal-Wallis test;
5. performs adjusted post hoc comparisons;
6. reports medians and IQRs;
7. calculates effect sizes;
8. discusses ties and exact p-values;
9. uses reusable functions; and
10. compares results cautiously with parametric alternatives.

# 25. Completion checklist

Confirm that you can:

- choose rank-based methods appropriately;
- conduct one-sample signed-rank tests;
- conduct paired Wilcoxon tests;
- conduct rank-sum tests;
- conduct Kruskal-Wallis tests;
- conduct adjusted post hoc tests;
- report medians and IQRs;
- calculate rank effect sizes;
- handle ties;
- use Spearman and Kendall correlations;
- create reusable reporting functions; and
- interpret rank tests cautiously.

# Lesson conclusion

Rank-based methods provide robust and ordinal-friendly alternatives, but they
still require correct design, appropriate hypotheses and careful reporting.

\[
\boxed{
\text{design}
\rightarrow
\text{rank-based test}
\rightarrow
\text{adjusted comparisons}
\rightarrow
\text{effect size}
\rightarrow
\text{careful interpretation}
}
\]
