---
title: "Lesson 36: Simple Linear Regression"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Simple linear regression models the mean of a numerical outcome as a linear
function of one explanatory variable.

The basic model is:

\[
Y_i
=
\beta_0
+
\beta_1X_i
+
\varepsilon_i.
\]

The method can be used for:

- description;
- prediction;
- adjustment;
- and formal inference about a linear association.

Regression does not automatically establish causation.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain the simple linear regression model;
2. identify outcome and explanatory variables;
3. create and interpret scatter plots;
4. fit a model with `lm()`;
5. interpret intercept and slope coefficients;
6. calculate fitted values and residuals;
7. explain least squares;
8. conduct inference for the slope;
9. interpret confidence intervals;
10. calculate and interpret \(R^2\);
11. distinguish confidence and prediction intervals;
12. generate predictions;
13. inspect basic model assumptions;
14. create reusable regression-reporting functions; and
15. report results without causal overstatement.

# 1. Importing the data

```{r import-simple-regression-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

head(student_data)
str(student_data)
```

# 2. Research question

Does coursework score linearly predict examination score?

```{r select-regression-variables}
regression_data <- student_data[
  c(
    "student_id",
    "coursework_score",
    "examination_score"
  )
]

summary(regression_data)
```

# 3. Scatter plot

```{r simple-regression-scatter}
plot(
  regression_data$coursework_score,
  regression_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)
```

A scatter plot should be inspected before fitting the model.

# 4. Correlation as preliminary evidence

```{r simple-regression-correlation}
cor(
  regression_data$coursework_score,
  regression_data$examination_score,
  use = "complete.obs"
)
```

Correlation measures linear association but does not provide a prediction
equation.

# 5. Fitting the model

```{r fit-simple-regression}
simple_model <- lm(
  examination_score ~ coursework_score,
  data = regression_data
)

simple_model
```

# 6. Model summary

```{r simple-regression-summary}
summary(
  simple_model
)
```

The summary includes:

- coefficients;
- standard errors;
- t statistics;
- p-values;
- residual standard error;
- \(R^2\);
- adjusted \(R^2\);
- and the model F test.

# 7. Regression equation

```{r simple-regression-coefficients}
model_coefficients <- coef(
  simple_model
)

intercept <- model_coefficients[1]
slope <- model_coefficients[2]

intercept
slope
```

The fitted equation is:

\[
\widehat{Y}
=
\widehat{\beta}_0
+
\widehat{\beta}_1X.
\]

# 8. Interpreting the intercept

The intercept is the predicted examination score when coursework score equals
zero.

It should be interpreted only when zero is meaningful and within a relevant
range.

# 9. Interpreting the slope

The slope is the expected change in examination score associated with a
one-unit increase in coursework score.

```{r slope-interpretation-text}
slope_interpretation <- paste(
  "For each additional coursework mark, the predicted examination score",
  "changes by",
  round(
    slope,
    3
  ),
  "marks on average."
)

slope_interpretation
```

# 10. Adding the fitted line

```{r add-simple-regression-line}
plot(
  regression_data$coursework_score,
  regression_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)

abline(
  simple_model,
  lwd = 2
)
```

# 11. Fitted values

```{r simple-fitted-values}
fitted_values <- fitted(
  simple_model
)

head(fitted_values)
```

# 12. Residuals

```{r simple-residuals}
model_residuals <- residuals(
  simple_model
)

head(model_residuals)
```

A residual is:

\[
e_i
=
Y_i
-
\widehat{Y}_i.
\]

# 13. Observation-level output

```{r simple-observation-output}
observation_output <- data.frame(
  student_id =
    regression_data$student_id,
  observed =
    regression_data$examination_score,
  fitted =
    fitted_values,
  residual =
    model_residuals,
  stringsAsFactors = FALSE
)

head(observation_output)
```

# 14. Least-squares principle

Least squares chooses coefficients that minimise:

\[
\sum_{i=1}^{n}
e_i^2.
\]

```{r residual-sum-of-squares}
residual_sum_of_squares <- sum(
  model_residuals ^ 2
)

residual_sum_of_squares
```

# 15. Manual slope and intercept

```{r manual-regression-coefficients}
x <- regression_data$coursework_score
y <- regression_data$examination_score

manual_slope <- cov(
  x,
  y
) / var(
  x
)

manual_intercept <- mean(y) -
  manual_slope *
  mean(x)

data.frame(
  coefficient = c(
    "Intercept",
    "Slope"
  ),
  manual = c(
    manual_intercept,
    manual_slope
  ),
  lm = unname(
    coef(simple_model)
  ),
  stringsAsFactors = FALSE
)
```

# 16. Inference for the slope

The hypotheses are:

\[
H_0:\beta_1=0
\]

\[
H_1:\beta_1\neq0.
\]

```{r slope-inference}
coefficient_table <- summary(
  simple_model
)$coefficients

coefficient_table
```

# 17. Confidence intervals for coefficients

```{r simple-coefficient-confidence}
confint(
  simple_model,
  level = 0.95
)
```

# 18. Model F test

```{r simple-model-anova}
anova(
  simple_model
)
```

With one predictor, the model F test and slope t test test the same null
hypothesis.

# 19. Coefficient of determination

```{r simple-r-squared}
model_summary <- summary(
  simple_model
)

r_squared <- model_summary$r.squared
adjusted_r_squared <-
  model_summary$adj.r.squared

r_squared
adjusted_r_squared
```

\(R^2\) is the proportion of observed outcome variation explained by the fitted
linear model.

# 20. Relationship between correlation and R-squared

```{r correlation-r-squared}
correlation_value <- cor(
  x,
  y
)

data.frame(
  correlation_squared =
    correlation_value ^ 2,
  regression_r_squared =
    r_squared
)
```

For simple linear regression with an intercept, \(R^2=r^2\).

# 21. Residual standard error

```{r simple-residual-standard-error}
residual_standard_error <-
  model_summary$sigma

residual_standard_error
```

This estimates the typical vertical size of residual variation in outcome
units.

# 22. Predicting at selected values

```{r simple-point-predictions}
new_coursework <- data.frame(
  coursework_score = c(
    20,
    25,
    30,
    35
  )
)

point_predictions <- predict(
  simple_model,
  newdata = new_coursework
)

data.frame(
  new_coursework,
  predicted_examination =
    point_predictions
)
```

# 23. Confidence interval for the mean response

```{r simple-mean-confidence-prediction}
mean_response_interval <- predict(
  simple_model,
  newdata = new_coursework,
  interval = "confidence",
  level = 0.95
)

cbind(
  new_coursework,
  mean_response_interval
)
```

# 24. Prediction interval for an individual response

```{r simple-individual-prediction}
individual_prediction_interval <- predict(
  simple_model,
  newdata = new_coursework,
  interval = "prediction",
  level = 0.95
)

cbind(
  new_coursework,
  individual_prediction_interval
)
```

Prediction intervals are wider because they include individual outcome
variation.

# 25. Avoiding extrapolation

```{r simple-predictor-range}
range(
  regression_data$coursework_score
)
```

Predictions far outside this range may be unreliable.

# 26. Residuals versus fitted values

```{r simple-residual-fitted-plot}
plot(
  fitted_values,
  model_residuals,
  main = "Residuals versus Fitted Values",
  xlab = "Fitted examination score",
  ylab = "Residual",
  pch = 19
)

abline(
  h = 0,
  lty = 2
)
```

A random horizontal band is broadly consistent with linearity and constant
variance.

# 27. Normal Q-Q plot

```{r simple-residual-qqplot}
qqnorm(
  model_residuals,
  main = "Normal Q-Q Plot of Residuals"
)

qqline(
  model_residuals
)
```

# 28. Standard diagnostic plots

```{r simple-standard-diagnostics}
old_parameters <- par(
  mfrow = c(
    2,
    2
  )
)

plot(
  simple_model
)

par(
  old_parameters
)
```

# 29. Shapiro-Wilk test of residuals

```{r simple-shapiro-residuals}
shapiro.test(
  model_residuals
)
```

This should not replace graphical assessment.

# 30. Observation influence

```{r simple-cooks-distance}
cooks_values <- cooks.distance(
  simple_model
)

head(cooks_values)

which(
  cooks_values >
    4 /
    nrow(regression_data)
)
```

A large Cook's distance identifies observations requiring investigation, not
automatic deletion.

# 31. Standardised residuals

```{r simple-standardised-residuals}
standardised_residuals <- rstandard(
  simple_model
)

which(
  abs(
    standardised_residuals
  ) > 2
)
```

# 32. Reusable model-reporting function

```{r simple-model-report-function}
report_simple_regression <- function(
  data,
  outcome,
  predictor
) {
  required <- c(
    outcome,
    predictor
  )

  if (!all(
    required %in%
      names(data)
  )) {
    stop(
      "Required variables are missing.",
      call. = FALSE
    )
  }

  formula_object <- reformulate(
    predictor,
    response = outcome
  )

  model <- lm(
    formula_object,
    data = data
  )

  model_summary <- summary(
    model
  )

  coefficient_table <-
    model_summary$coefficients

  slope_row <- coefficient_table[
    predictor,
  ]

  slope_interval <- confint(
    model,
    predictor
  )

  data.frame(
    outcome = outcome,
    predictor = predictor,
    n = nobs(model),
    intercept =
      unname(
        coef(model)[1]
      ),
    slope =
      unname(
        coef(model)[2]
      ),
    slope_standard_error =
      slope_row[
        "Std. Error"
      ],
    t =
      slope_row[
        "t value"
      ],
    p_value =
      slope_row[
        "Pr(>|t|)"
      ],
    confidence_lower =
      slope_interval[1],
    confidence_upper =
      slope_interval[2],
    r_squared =
      model_summary$r.squared,
    adjusted_r_squared =
      model_summary$adj.r.squared,
    residual_standard_error =
      model_summary$sigma,
    stringsAsFactors = FALSE
  )
}

report_simple_regression(
  regression_data,
  "examination_score",
  "coursework_score"
)
```

# 33. Reporting language

A complete report includes:

- the sample size;
- coefficient estimate;
- standard error;
- t statistic;
- degrees of freedom;
- p-value;
- confidence interval;
- \(R^2\);
- and a contextual, non-causal interpretation.

# 34. Common mistakes

- fitting a model without inspecting a scatter plot;
- interpreting association as causation;
- interpreting the intercept outside the observed range;
- reporting \(R^2\) as a percentage of individuals correctly predicted;
- using confidence and prediction intervals interchangeably;
- extrapolating beyond the observed predictor range;
- deleting influential observations automatically;
- ignoring non-linearity;
- reporting only p-values;
- failing to retain the model object.

# 35. Independent exercises

1. import and inspect the data.
2. create a scatter plot.
3. fit a simple linear regression.
4. interpret intercept and slope.
5. extract fitted values and residuals.
6. verify coefficients manually.
7. conduct slope inference.
8. calculate confidence intervals.
9. interpret \(R^2\).
10. create point predictions.
11. create confidence and prediction intervals.
12. inspect diagnostic plots.
13. identify potentially influential records.
14. create a reusable reporting function.
15. write a non-causal interpretation.

# 36. Lesson assessment

Create a standalone R Markdown report that:

1. defines a regression question;
2. creates an appropriate scatter plot;
3. fits a simple regression model;
4. reports the regression equation;
5. explains least squares;
6. conducts coefficient inference;
7. reports model fit;
8. creates predictions and both interval types;
9. assesses diagnostics;
10. identifies influential records;
11. uses a reusable reporting function; and
12. provides a complete interpretation.

# 37. Completion checklist

Confirm that you can:

- specify a simple regression model;
- create a scatter plot;
- fit a model with `lm()`;
- interpret coefficients;
- calculate fitted values;
- calculate residuals;
- explain least squares;
- conduct slope inference;
- interpret \(R^2\);
- generate predictions;
- distinguish interval types;
- inspect diagnostics;
- identify influence; and
- report regression results.

# Lesson conclusion

Simple linear regression describes and predicts a numerical outcome using one
explanatory variable. Sound analysis combines model estimation, uncertainty,
diagnostics and cautious interpretation.

\[
\boxed{
\text{question}
\rightarrow
\text{scatter plot}
\rightarrow
\text{fit model}
\rightarrow
\text{interpret coefficients}
\rightarrow
\text{diagnose}
\rightarrow
\text{predict}
}
\]
