---
title: "Lesson 37: Multiple Linear Regression"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Multiple linear regression models a numerical outcome using two or more
explanatory variables.

The general model is:

\[
Y_i
=
\beta_0
+
\beta_1X_{1i}
+
\beta_2X_{2i}
+
\cdots
+
\beta_kX_{ki}
+
\varepsilon_i.
\]

Each coefficient is interpreted while holding the other model variables
constant.

This lesson develops adjusted interpretation, categorical predictors,
interactions, model comparison and multicollinearity assessment.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain multiple linear regression;
2. specify a model with several predictors;
3. interpret adjusted coefficients;
4. include categorical predictors;
5. choose reference categories;
6. create interaction terms;
7. compare nested models;
8. interpret \(R^2\) and adjusted \(R^2\);
9. assess multicollinearity;
10. calculate variance inflation factors;
11. generate adjusted predictions;
12. distinguish confounding, mediation and interaction conceptually;
13. create reusable multiple-regression reports; and
14. report model results accurately.

# 1. Importing data

```{r import-multiple-regression-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_data$programme <- factor(
  student_data$programme
)

str(student_data)
```

# 2. Basic multiple regression

Model final score using coursework and examination scores.

```{r fit-basic-multiple-regression}
multiple_model <- lm(
  final_score ~
    coursework_score +
    examination_score,
  data = student_data
)

summary(
  multiple_model
)
```

Because final score is the arithmetic sum of the two components in this
teaching dataset, this model demonstrates exact structural dependence and
should not be treated as a substantive predictive model.

# 3. A more suitable teaching model

Use examination score as the outcome and include coursework score and
programme.

```{r fit-adjusted-regression}
adjusted_model <- lm(
  examination_score ~
    coursework_score +
    programme,
  data = student_data
)

summary(
  adjusted_model
)
```

# 4. Design matrix

```{r inspect-design-matrix}
head(
  model.matrix(
    adjusted_model
  )
)
```

Categorical predictors are represented through indicator variables.

# 5. Reference category

```{r multiple-reference-category}
levels(
  student_data$programme
)
```

The first factor level is the default reference category under treatment
coding.

# 6. Changing the reference category

```{r relevel-programme}
student_data$programme_statistics_reference <- relevel(
  student_data$programme,
  ref = "Statistics"
)

reference_model <- lm(
  examination_score ~
    coursework_score +
    programme_statistics_reference,
  data = student_data
)

summary(
  reference_model
)
```

Changing the reference category changes coefficient interpretation but not
fitted values or overall fit.

# 7. Adjusted coefficient interpretation

The coursework coefficient estimates the expected examination-score change for
a one-unit coursework increase, comparing students with the same programme.

Programme coefficients estimate adjusted differences relative to the reference
programme.

# 8. Confidence intervals

```{r multiple-confidence-intervals}
confint(
  reference_model
)
```

# 9. Overall model test

```{r multiple-overall-test}
anova(
  reference_model
)
```

```{r multiple-model-summary}
reference_summary <- summary(
  reference_model
)

reference_summary$r.squared
reference_summary$adj.r.squared
```

# 10. Adjusted R-squared

Adjusted \(R^2\) penalises model complexity and may decline when unhelpful
predictors are added.

# 11. Nested model comparison

```{r nested-models}
reduced_model <- lm(
  examination_score ~
    coursework_score,
  data = student_data
)

full_model <- lm(
  examination_score ~
    coursework_score +
    programme_statistics_reference,
  data = student_data
)

anova(
  reduced_model,
  full_model
)
```

This compares whether programme terms jointly improve model fit.

# 12. Interaction model

```{r fit-interaction-model}
interaction_model <- lm(
  examination_score ~
    coursework_score *
    programme_statistics_reference,
  data = student_data
)

summary(
  interaction_model
)
```

The `*` expands to:

- both main effects;
- plus their interaction.

# 13. Interaction interpretation

An interaction indicates that the coursework slope differs by programme.

Main effects in an interaction model are conditional:

- the coursework main effect applies to the reference programme;
- programme main effects apply when coursework score equals zero.

# 14. Centring a continuous predictor

```{r centre-coursework}
student_data$coursework_centred <-
  student_data$coursework_score -
  mean(
    student_data$coursework_score
  )

centred_interaction_model <- lm(
  examination_score ~
    coursework_centred *
    programme_statistics_reference,
  data = student_data
)

summary(
  centred_interaction_model
)
```

Centring makes programme main effects apply at the mean coursework score.

# 15. Comparing interaction and additive models

```{r compare-interaction-model}
anova(
  full_model,
  interaction_model
)
```

# 16. Predictions from the additive model

```{r adjusted-prediction-grid}
prediction_grid <- expand.grid(
  coursework_score = c(
    20,
    30,
    40
  ),
  programme_statistics_reference =
    levels(
      student_data$programme_statistics_reference
    )
)

prediction_grid$predicted_examination <- predict(
  reference_model,
  newdata = prediction_grid
)

prediction_grid
```

# 17. Prediction intervals

```{r multiple-prediction-intervals}
prediction_output <- predict(
  reference_model,
  newdata = prediction_grid,
  interval = "prediction"
)

cbind(
  prediction_grid,
  prediction_output
)
```

# 18. Coefficient table extraction

```{r multiple-coefficient-table}
multiple_coefficient_table <- summary(
  reference_model
)$coefficients

multiple_coefficient_table
```

# 19. Standardised coefficients

```{r standardised-regression}
standardised_data <- student_data

standardised_data$coursework_z <- as.numeric(
  scale(
    standardised_data$coursework_score
  )
)

standardised_data$examination_z <- as.numeric(
  scale(
    standardised_data$examination_score
  )
)

standardised_model <- lm(
  examination_z ~ coursework_z,
  data = standardised_data
)

summary(
  standardised_model
)
```

Standardised coefficients express changes in standard-deviation units.

# 20. Multicollinearity

Multicollinearity occurs when predictors contain overlapping information.

Consequences may include:

- unstable coefficients;
- inflated standard errors;
- sign changes;
- and difficult interpretation.

# 21. Manual variance inflation factors

```{r manual-vif-function}
calculate_vif <- function(
  model
) {
  design <- model.matrix(
    model
  )

  design <- design[
    ,
    colnames(design) !=
      "(Intercept)",
    drop = FALSE
  ]

  if (ncol(design) < 2L) {
    return(
      setNames(
        1,
        colnames(design)
      )
    )
  }

  vif_values <- vapply(
    seq_len(
      ncol(design)
    ),
    function(column_index) {
      response <- design[
        ,
        column_index
      ]

      predictors <- design[
        ,
        -column_index,
        drop = FALSE
      ]

      auxiliary_model <- lm(
        response ~ predictors
      )

      1 / (
        1 -
          summary(
            auxiliary_model
          )$r.squared
      )
    },
    FUN.VALUE = numeric(1)
  )

  names(vif_values) <- colnames(
    design
  )

  vif_values
}

calculate_vif(
  reference_model
)
```

For multi-level factors, ordinary dummy-variable VIFs require cautious
interpretation.

# 22. Correlation among numerical predictors

```{r predictor-correlation}
numeric_predictors <- student_data[
  c(
    "coursework_score",
    "examination_score"
  )
]

cor(
  numeric_predictors,
  use = "complete.obs"
)
```

# 23. Confounding concept

A variable may confound an association when it is associated with both the
predictor and outcome and is not a consequence of the predictor.

Regression adjustment alone does not prove confounding control is sufficient.

# 24. Mediation concept

A mediator lies on a causal pathway between exposure and outcome.

Including a mediator changes the meaning of coefficients and requires a clear
causal framework.

# 25. Interaction concept

An interaction means the association between one predictor and outcome depends
on another predictor.

It concerns effect modification, not merely statistical significance in
subgroups.

# 26. Model comparison criteria

```{r multiple-model-comparison}
data.frame(
  model = c(
    "Reduced",
    "Additive full",
    "Interaction"
  ),
  residual_degrees_of_freedom = c(
    df.residual(
      reduced_model
    ),
    df.residual(
      full_model
    ),
    df.residual(
      interaction_model
    )
  ),
  r_squared = c(
    summary(
      reduced_model
    )$r.squared,
    summary(
      full_model
    )$r.squared,
    summary(
      interaction_model
    )$r.squared
  ),
  adjusted_r_squared = c(
    summary(
      reduced_model
    )$adj.r.squared,
    summary(
      full_model
    )$adj.r.squared,
    summary(
      interaction_model
    )$adj.r.squared
  ),
  AIC = c(
    AIC(reduced_model),
    AIC(full_model),
    AIC(interaction_model)
  )
)
```

AIC is useful for relative model comparison under a consistent likelihood
framework.

# 27. Reusable coefficient-reporting function

```{r multiple-report-function}
report_multiple_regression <- function(
  model
) {
  if (!inherits(
    model,
    "lm"
  )) {
    stop(
      "`model` must be an lm object.",
      call. = FALSE
    )
  }

  model_summary <- summary(
    model
  )

  coefficient_table <- as.data.frame(
    model_summary$coefficients
  )

  coefficient_table$term <- rownames(
    coefficient_table
  )

  rownames(
    coefficient_table
  ) <- NULL

  confidence_intervals <- as.data.frame(
    confint(model)
  )

  confidence_intervals$term <- rownames(
    confidence_intervals
  )

  rownames(
    confidence_intervals
  ) <- NULL

  names(
    confidence_intervals
  )[1:2] <- c(
    "confidence_lower",
    "confidence_upper"
  )

  merged <- merge(
    coefficient_table,
    confidence_intervals,
    by = "term",
    sort = FALSE
  )

  names(merged)[
    names(merged) ==
      "Estimate"
  ] <- "estimate"

  names(merged)[
    names(merged) ==
      "Std. Error"
  ] <- "standard_error"

  names(merged)[
    names(merged) ==
      "t value"
  ] <- "t"

  names(merged)[
    names(merged) ==
      "Pr(>|t|)"
  ] <- "p_value"

  merged
}

report_multiple_regression(
  reference_model
)
```

# 28. Reusable model-fit summary

```{r multiple-fit-summary-function}
summarise_model_fit <- function(
  model
) {
  model_summary <- summary(
    model
  )

  f_statistic <- model_summary$fstatistic

  data.frame(
    n = nobs(model),
    predictors =
      length(
        coef(model)
      ) - 1,
    r_squared =
      model_summary$r.squared,
    adjusted_r_squared =
      model_summary$adj.r.squared,
    residual_standard_error =
      model_summary$sigma,
    f =
      unname(
        f_statistic[
          "value"
        ]
      ),
    df_model =
      unname(
        f_statistic[
          "numdf"
        ]
      ),
    df_residual =
      unname(
        f_statistic[
          "dendf"
        ]
      ),
    p_value = pf(
      f_statistic[
        "value"
      ],
      f_statistic[
        "numdf"
      ],
      f_statistic[
        "dendf"
      ],
      lower.tail = FALSE
    ),
    AIC = AIC(model),
    BIC = BIC(model)
  )
}

summarise_model_fit(
  reference_model
)
```

# 29. Common mistakes

- interpreting adjusted coefficients as unadjusted differences;
- using arbitrary control variables without a conceptual basis;
- interpreting factor coefficients without stating the reference category;
- omitting main effects from an interaction model;
- interpreting main effects in interactions without conditioning;
- using \(R^2\) alone for model selection;
- treating high VIF as proof that a predictor is invalid;
- changing model specifications only to obtain significance;
- confusing adjustment with causal identification;
- presenting coefficients without uncertainty.

# 30. Independent exercises

1. fit a model with two numerical predictors.
2. explain exact structural dependence in the teaching data.
3. fit a model with a categorical predictor.
4. change the reference category.
5. interpret adjusted coefficients.
6. compare nested models.
7. fit and interpret an interaction.
8. centre a continuous predictor.
9. create adjusted predictions.
10. calculate manual VIFs.
11. compare model-fit criteria.
12. explain confounding, mediation and interaction.
13. create reusable reporting functions.
14. write a complete model interpretation.

# 31. Lesson assessment

Create a standalone R Markdown report that:

1. states a multivariable research question;
2. fits an additive regression model;
3. includes a categorical predictor;
4. states the reference category;
5. reports adjusted coefficients and intervals;
6. compares nested models;
7. assesses an interaction;
8. centres a continuous variable;
9. assesses multicollinearity;
10. produces adjusted predictions;
11. uses reusable reporting functions; and
12. explains why the results do or do not support causal interpretation.

# 32. Completion checklist

Confirm that you can:

- fit multiple regression models;
- interpret adjusted coefficients;
- include categorical variables;
- select reference categories;
- fit interaction terms;
- centre predictors;
- compare nested models;
- interpret adjusted \(R^2\);
- calculate VIFs;
- generate adjusted predictions;
- distinguish confounding, mediation and interaction; and
- report multiple-regression results.

# Lesson conclusion

Multiple regression estimates conditional associations among several variables.
Good modelling requires clear variable roles, careful interpretation, model
comparison and awareness that adjustment alone does not establish causality.

\[
\boxed{
\text{conceptual model}
\rightarrow
\text{specification}
\rightarrow
\text{adjusted estimates}
\rightarrow
\text{interaction and comparison}
\rightarrow
\text{diagnostics}
\rightarrow
\text{interpretation}
}
\]
