---
title: "Lesson 38: Regression Diagnostics, Model Validation and Reporting"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

A fitted regression model is not complete until its assumptions, stability and
predictive performance have been assessed.

Diagnostics examine whether:

- the functional form is adequate;
- residual variance is reasonably constant;
- residual distribution is suitable for inference;
- observations are independent;
- individual records exert excessive influence;
- predictors are collinear;
- and predictions generalise beyond the fitted sample.

This lesson combines residual diagnostics, influence analysis, validation and
reporting.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. distinguish residuals, leverage and influence;
2. inspect linearity;
3. assess constant variance;
4. assess residual normality;
5. identify unusual residuals;
6. calculate leverage;
7. calculate Cook's distance;
8. inspect DFBETAs;
9. assess multicollinearity;
10. compare ordinary and robust uncertainty conceptually;
11. split data for validation;
12. calculate prediction-error measures;
13. perform k-fold cross-validation manually;
14. assess model stability;
15. create a diagnostic report; and
16. write a complete regression-results section.

# 1. Importing data and fitting a model

```{r import-diagnostic-data}
student_data <- read.csv(
  "data/student_scores.csv",
  stringsAsFactors = FALSE
)

student_data$programme <- factor(
  student_data$programme
)

diagnostic_model <- lm(
  examination_score ~
    coursework_score +
    programme,
  data = student_data
)

summary(
  diagnostic_model
)
```

# 2. Core diagnostic quantities

```{r extract-diagnostic-quantities}
fitted_values <- fitted(
  diagnostic_model
)

raw_residuals <- residuals(
  diagnostic_model
)

standardised_residuals <- rstandard(
  diagnostic_model
)

studentised_residuals <- rstudent(
  diagnostic_model
)

leverage_values <- hatvalues(
  diagnostic_model
)

cooks_values <- cooks.distance(
  diagnostic_model
)
```

# 3. Diagnostic data frame

```{r diagnostic-data-frame}
diagnostic_output <- data.frame(
  observation =
    seq_len(
      nobs(
        diagnostic_model
      )
    ),
  fitted =
    fitted_values,
  residual =
    raw_residuals,
  standardised_residual =
    standardised_residuals,
  studentised_residual =
    studentised_residuals,
  leverage =
    leverage_values,
  cooks_distance =
    cooks_values
)

head(diagnostic_output)
```

# 4. Linearity assessment

```{r diagnostic-linearity}
plot(
  fitted_values,
  raw_residuals,
  main = "Residuals versus Fitted Values",
  xlab = "Fitted value",
  ylab = "Residual",
  pch = 19
)

abline(
  h = 0,
  lty = 2
)

lines(
  lowess(
    fitted_values,
    raw_residuals
  ),
  lwd = 2
)
```

Systematic curvature suggests functional-form inadequacy.

# 5. Component-plus-residual concept

For a continuous predictor, a partial-residual plot helps examine its adjusted
functional form.

```{r partial-residual-manual}
coursework_coefficient <- coef(
  diagnostic_model
)[
  "coursework_score"
]

partial_residual <-
  raw_residuals +
  coursework_coefficient *
  student_data$coursework_score

plot(
  student_data$coursework_score,
  partial_residual,
  main = "Partial Residual Plot for Coursework",
  xlab = "Coursework score",
  ylab = "Partial residual",
  pch = 19
)

lines(
  lowess(
    student_data$coursework_score,
    partial_residual
  ),
  lwd = 2
)
```

# 6. Constant-variance assessment

```{r diagnostic-scale-location}
plot(
  fitted_values,
  sqrt(
    abs(
      standardised_residuals
    )
  ),
  main = "Scale-Location Plot",
  xlab = "Fitted value",
  ylab = "Square root of absolute standardised residual",
  pch = 19
)

lines(
  lowess(
    fitted_values,
    sqrt(
      abs(
        standardised_residuals
      )
    )
  ),
  lwd = 2
)
```

A widening or narrowing pattern may indicate heteroskedasticity.

# 7. Breusch-Pagan concept through auxiliary regression

A simple teaching version regresses squared residuals on fitted values.

```{r auxiliary-heteroskedasticity-test}
auxiliary_model <- lm(
  I(
    raw_residuals ^ 2
  ) ~ fitted_values
)

auxiliary_r_squared <- summary(
  auxiliary_model
)$r.squared

lagrange_multiplier <- nobs(
  diagnostic_model
) * auxiliary_r_squared

bp_p_value <- pchisq(
  lagrange_multiplier,
  df = 1,
  lower.tail = FALSE
)

data.frame(
  statistic =
    lagrange_multiplier,
  degrees_of_freedom = 1,
  p_value =
    bp_p_value
)
```

This is pedagogical and not a complete replacement for established package
implementations.

# 8. Residual normality

```{r diagnostic-residual-normality}
hist(
  raw_residuals,
  main = "Residual Distribution",
  xlab = "Residual"
)

qqnorm(
  raw_residuals,
  main = "Normal Q-Q Plot of Residuals"
)

qqline(
  raw_residuals
)

shapiro.test(
  raw_residuals
)
```

Normality is most relevant to small-sample coefficient inference, not to the
existence of the fitted line itself.

# 9. Unusual residuals

```{r unusual-residuals}
unusual_residual_rows <- which(
  abs(
    studentised_residuals
  ) > 2
)

diagnostic_output[
  unusual_residual_rows,
]
```

# 10. Leverage

Leverage reflects unusual predictor combinations.

A common screening value is:

\[
\frac{2p}{n},
\]

where \(p\) includes the intercept.

```{r leverage-screen}
number_parameters <- length(
  coef(
    diagnostic_model
  )
)

leverage_cutoff <- 2 *
  number_parameters /
  nobs(
    diagnostic_model
  )

high_leverage_rows <- which(
  leverage_values >
    leverage_cutoff
)

leverage_cutoff

diagnostic_output[
  high_leverage_rows,
]
```

# 11. Cook's distance

```{r cooks-screen}
cooks_cutoff <- 4 /
  nobs(
    diagnostic_model
  )

influential_rows <- which(
  cooks_values >
    cooks_cutoff
)

cooks_cutoff

diagnostic_output[
  influential_rows,
]
```

# 12. Influence plot

```{r influence-plot-base}
plot(
  leverage_values,
  studentised_residuals,
  main = "Leverage and Studentised Residuals",
  xlab = "Leverage",
  ylab = "Studentised residual",
  pch = 19,
  cex = 1 +
    5 *
    sqrt(
      cooks_values /
      max(
        cooks_values
      )
    )
)

abline(
  h = c(
    -2,
    2
  ),
  lty = 2
)

abline(
  v = leverage_cutoff,
  lty = 2
)
```

# 13. DFBETAs

DFBETAs measure how much each coefficient changes when an observation is
removed.

```{r diagnostic-dfbetas}
dfbeta_values <- dfbetas(
  diagnostic_model
)

head(
  dfbeta_values
)
```

Screening:

```{r dfbetas-screen}
dfbeta_cutoff <- 2 /
  sqrt(
    nobs(
      diagnostic_model
    )
  )

dfbeta_flags <- apply(
  abs(
    dfbeta_values
  ) >
    dfbeta_cutoff,
  1,
  any
)

which(
  dfbeta_flags
)
```

# 14. Multicollinearity function

```{r diagnostic-vif-function}
calculate_vif <- function(
  model
) {
  design <- model.matrix(
    model
  )

  design <- design[
    ,
    colnames(design) !=
      "(Intercept)",
    drop = FALSE
  ]

  if (ncol(design) == 1L) {
    return(
      setNames(
        1,
        colnames(design)
      )
    )
  }

  result <- vapply(
    seq_len(
      ncol(design)
    ),
    function(index) {
      auxiliary <- lm(
        design[
          ,
          index
        ] ~
          design[
            ,
            -index,
            drop = FALSE
          ]
      )

      1 / (
        1 -
          summary(
            auxiliary
          )$r.squared
      )
    },
    FUN.VALUE = numeric(1)
  )

  names(result) <- colnames(
    design
  )

  result
}

calculate_vif(
  diagnostic_model
)
```

# 15. Standard model diagnostic panel

```{r standard-diagnostic-panel}
old_parameters <- par(
  mfrow = c(
    2,
    2
  )
)

plot(
  diagnostic_model
)

par(
  old_parameters
)
```

# 16. Sensitivity analysis excluding flagged observations

```{r sensitivity-model}
rows_to_review <- sort(
  unique(
    c(
      unusual_residual_rows,
      high_leverage_rows,
      influential_rows,
      which(
        dfbeta_flags
      )
    )
  )
)

sensitivity_data <- if (
  length(rows_to_review) > 0L
) {
  student_data[
    -rows_to_review,
  ]
} else {
  student_data
}

sensitivity_model <- lm(
  examination_score ~
    coursework_score +
    programme,
  data = sensitivity_data
)

coefficient_comparison <- cbind(
  original = coef(
    diagnostic_model
  ),
  sensitivity = coef(
    sensitivity_model
  )
)

coefficient_comparison
```

Sensitivity analysis should be reported transparently and should not be used to
hide inconvenient observations.

# 17. Stratified training and test split

A completely random split can omit a small programme from the training data.
The fitted model would then be unable to predict observations from that unseen
factor level.

Because this teaching dataset contains only two observations per programme, one
observation from each programme is assigned to training and the other to
testing.

```{r train-test-split}
set.seed(20260714)

rows_by_programme <- split(
  seq_len(
    nrow(
      student_data
    )
  ),
  student_data$programme
)

programme_counts <- lengths(
  rows_by_programme
)

if (any(
  programme_counts < 2L
)) {
  stop(
    paste(
      "Each programme requires at least two observations for this stratified",
      "training-test demonstration."
    ),
    call. = FALSE
  )
}

training_rows <- unlist(
  lapply(
    rows_by_programme,
    function(rows) {
      sample(
        rows,
        size = 1L
      )
    }
  ),
  use.names = FALSE
)

training_data <- droplevels(
  student_data[
    training_rows,
  ]
)

test_data <- student_data[
  -training_rows,
]

test_data$programme <- factor(
  test_data$programme,
  levels = levels(
    training_data$programme
  )
)

nrow(training_data)
nrow(test_data)

table(
  training_data$programme
)

table(
  test_data$programme
)

stopifnot(
  all(
    levels(
      test_data$programme
    ) %in%
      levels(
        training_data$programme
      )
  )
)
```

# 18. Fit on training data

```{r fit-training-model}
training_model <- lm(
  examination_score ~
    coursework_score +
    programme,
  data = training_data
)

summary(
  training_model
)
```

# 19. Test predictions

```{r test-set-predictions}
test_predictions <- predict(
  training_model,
  newdata = test_data
)

prediction_errors <-
  test_data$examination_score -
  test_predictions
```

# 20. Prediction error measures

```{r prediction-error-measures}
mean_absolute_error <- mean(
  abs(
    prediction_errors
  )
)

root_mean_squared_error <- sqrt(
  mean(
    prediction_errors ^ 2
  )
)

test_r_squared <- 1 -
  sum(
    prediction_errors ^ 2
  ) /
  sum(
    (
      test_data$examination_score -
        mean(
          test_data$examination_score
        )
    ) ^ 2
  )

data.frame(
  MAE =
    mean_absolute_error,
  RMSE =
    root_mean_squared_error,
  test_R_squared =
    test_r_squared
)
```

# 21. Stratified k-fold cross-validation

For a model containing a categorical predictor, each training fold must contain
every factor level that may appear in its validation fold.

The maximum feasible number of stratified folds is limited by the smallest
programme size. Since each programme contains two observations, this example
uses two folds.

```{r k-fold-cross-validation-function}
cross_validate_lm <- function(
  data,
  formula,
  group_variable,
  k = 2,
  seed = 20260714
) {
  required_variables <- c(
    all.vars(
      formula
    ),
    group_variable
  )

  missing_variables <- setdiff(
    required_variables,
    names(data)
  )

  if (length(
    missing_variables
  ) > 0L) {
    stop(
      paste(
        "Variables not found:",
        paste(
          missing_variables,
          collapse = ", "
        )
      ),
      call. = FALSE
    )
  }

  group_counts <- table(
    data[[
      group_variable
    ]]
  )

  maximum_folds <- min(
    group_counts
  )

  if (
    k < 2L ||
    k > maximum_folds
  ) {
    stop(
      paste(
        "`k` must be between 2 and",
        maximum_folds,
        "for stratified validation with the available group sizes."
      ),
      call. = FALSE
    )
  }

  set.seed(seed)

  fold_id <- integer(
    nrow(data)
  )

  rows_by_group <- split(
    seq_len(
      nrow(data)
    ),
    data[[
      group_variable
    ]]
  )

  for (rows in rows_by_group) {
    shuffled_rows <- sample(
      rows
    )

    fold_id[
      shuffled_rows
    ] <- rep(
      seq_len(k),
      length.out = length(
        shuffled_rows
      )
    )
  }

  outcome_name <- all.vars(
    formula
  )[1]

  fold_results <- lapply(
    seq_len(k),
    function(current_fold) {
      training <- droplevels(
        data[
          fold_id != current_fold,
        ]
      )

      validation <- data[
        fold_id == current_fold,
      ]

      factor_variables <- names(
        training
      )[
        vapply(
          training,
          is.factor,
          FUN.VALUE = logical(1)
        )
      ]

      for (variable in factor_variables) {
        validation[[
          variable
        ]] <- factor(
          validation[[
            variable
          ]],
          levels = levels(
            training[[
              variable
            ]]
          )
        )
      }

      if (any(
        !complete.cases(
          validation[
            factor_variables
          ]
        )
      )) {
        stop(
          paste(
            "A validation fold contains a factor level absent from its",
            "training fold."
          ),
          call. = FALSE
        )
      }

      model <- lm(
        formula,
        data = training
      )

      predictions <- predict(
        model,
        newdata = validation
      )

      errors <- validation[[
        outcome_name
      ]] - predictions

      data.frame(
        fold = current_fold,
        n = nrow(validation),
        MAE = mean(
          abs(errors)
        ),
        RMSE = sqrt(
          mean(
            errors ^ 2
          )
        )
      )
    }
  )

  do.call(
    rbind,
    fold_results
  )
}

cross_validation_results <- cross_validate_lm(
  student_data,
  examination_score ~
    coursework_score +
    programme,
  group_variable = "programme",
  k = 2
)

cross_validation_results
```

# 22. Summarising cross-validation

```{r summarise-cross-validation}
data.frame(
  mean_MAE =
    mean(
      cross_validation_results$MAE
    ),
  mean_RMSE =
    mean(
      cross_validation_results$RMSE
    ),
  standard_deviation_RMSE =
    sd(
      cross_validation_results$RMSE
    )
)
```

# 23. Model stability across folds

```{r coefficient-stability-function}
cross_validate_coefficients <- function(
  data,
  formula,
  group_variable,
  k = 2,
  seed = 20260714
) {
  group_counts <- table(
    data[[
      group_variable
    ]]
  )

  maximum_folds <- min(
    group_counts
  )

  if (
    k < 2L ||
    k > maximum_folds
  ) {
    stop(
      paste(
        "`k` must be between 2 and",
        maximum_folds,
        "for the available group sizes."
      ),
      call. = FALSE
    )
  }

  set.seed(seed)

  fold_id <- integer(
    nrow(data)
  )

  rows_by_group <- split(
    seq_len(
      nrow(data)
    ),
    data[[
      group_variable
    ]]
  )

  for (rows in rows_by_group) {
    shuffled_rows <- sample(
      rows
    )

    fold_id[
      shuffled_rows
    ] <- rep(
      seq_len(k),
      length.out = length(
        shuffled_rows
      )
    )
  }

  coefficient_list <- lapply(
    seq_len(k),
    function(current_fold) {
      training <- droplevels(
        data[
          fold_id != current_fold,
        ]
      )

      model <- lm(
        formula,
        data = training
      )

      coefficients <- coef(
        model
      )

      data.frame(
        fold = current_fold,
        term = names(
          coefficients
        ),
        estimate =
          unname(
            coefficients
          ),
        stringsAsFactors = FALSE
      )
    }
  )

  do.call(
    rbind,
    coefficient_list
  )
}

coefficient_stability <- cross_validate_coefficients(
  student_data,
  examination_score ~
    coursework_score +
    programme,
  group_variable = "programme",
  k = 2
)

coefficient_stability
```

# 24. Robust standard errors concept

Ordinary least-squares coefficients remain the same under heteroskedasticity,
but conventional standard errors may be inappropriate.

Robust standard errors can be calculated with packages such as `sandwich` and
`lmtest`.

```{r robust-standard-errors-template, eval=FALSE}
if (
  requireNamespace(
    "sandwich",
    quietly = TRUE
  ) &&
  requireNamespace(
    "lmtest",
    quietly = TRUE
  )
) {
  lmtest::coeftest(
    diagnostic_model,
    vcov. =
      sandwich::vcovHC(
        diagnostic_model,
        type = "HC3"
      )
  )
}
```

# 25. Diagnostic-reporting function

```{r diagnostic-report-function}
create_regression_diagnostic_report <- function(
  model
) {
  if (!inherits(
    model,
    "lm"
  )) {
    stop(
      "`model` must be an lm object.",
      call. = FALSE
    )
  }

  n <- nobs(model)
  p <- length(
    coef(model)
  )

  studentised <- rstudent(
    model
  )

  leverage <- hatvalues(
    model
  )

  cooks <- cooks.distance(
    model
  )

  dfbeta_matrix <- dfbetas(
    model
  )

  data.frame(
    n = n,
    parameters = p,
    maximum_absolute_studentised_residual =
      max(
        abs(
          studentised
        )
      ),
    residuals_above_two =
      sum(
        abs(
          studentised
        ) > 2
      ),
    leverage_cutoff =
      2 *
      p /
      n,
    high_leverage_count =
      sum(
        leverage >
          2 *
          p /
          n
      ),
    cooks_cutoff =
      4 /
      n,
    influential_count =
      sum(
        cooks >
          4 /
          n
      ),
    dfbeta_cutoff =
      2 /
      sqrt(n),
    dfbeta_flagged_count =
      sum(
        apply(
          abs(
            dfbeta_matrix
          ) >
            2 /
            sqrt(n),
          1,
          any
        )
      ),
    residual_shapiro_p =
      shapiro.test(
        residuals(model)
      )$p.value,
    stringsAsFactors = FALSE
  )
}

create_regression_diagnostic_report(
  diagnostic_model
)
```

# 26. Full model-reporting function

```{r full-regression-report-function}
create_full_regression_report <- function(
  model
) {
  model_summary <- summary(
    model
  )

  coefficients <- as.data.frame(
    model_summary$coefficients
  )

  coefficients$term <- rownames(
    coefficients
  )

  rownames(coefficients) <- NULL

  intervals <- as.data.frame(
    confint(model)
  )

  intervals$term <- rownames(
    intervals
  )

  rownames(intervals) <- NULL

  names(intervals)[1:2] <- c(
    "confidence_lower",
    "confidence_upper"
  )

  coefficients <- merge(
    coefficients,
    intervals,
    by = "term",
    sort = FALSE
  )

  fit <- data.frame(
    n = nobs(model),
    r_squared =
      model_summary$r.squared,
    adjusted_r_squared =
      model_summary$adj.r.squared,
    residual_standard_error =
      model_summary$sigma,
    AIC = AIC(model),
    BIC = BIC(model)
  )

  list(
    coefficients = coefficients,
    fit = fit,
    diagnostics =
      create_regression_diagnostic_report(
        model
      )
  )
}

full_report <- create_full_regression_report(
  diagnostic_model
)

full_report
```

# 27. Reporting structure

A clear regression report should present:

1. research question and model specification;
2. descriptive information;
3. coefficient estimates and confidence intervals;
4. model fit;
5. diagnostic findings;
6. sensitivity analysis;
7. validation performance;
8. limitations;
9. and non-causal interpretation unless design supports causality.

# 28. Common mistakes

- treating residual normality as the only assumption;
- deleting observations solely because they are flagged;
- confusing leverage with influence;
- using test-set results to repeatedly modify the model;
- reporting training \(R^2\) as predictive performance;
- conducting validation after selecting the best result on the same data;
- ignoring factor levels absent from training data;
- relying on one random split only;
- reporting robust standard errors without explaining why;
- hiding sensitivity-analysis decisions.

# 29. Independent exercises

1. extract residual and influence measures.
2. create a diagnostic data frame.
3. assess linearity.
4. assess constant variance.
5. assess residual normality.
6. identify unusual residuals.
7. calculate leverage cutoffs.
8. identify influential observations.
9. inspect DFBETAs.
10. calculate VIFs.
11. conduct a transparent sensitivity analysis.
12. create a train-test split.
13. calculate MAE, RMSE and test \(R^2\).
14. conduct manual k-fold cross-validation.
15. assess coefficient stability.
16. create complete diagnostic and model reports.

# 30. Lesson assessment

Create a standalone R Markdown report that:

1. fits a multiple regression model;
2. produces all standard diagnostic plots;
3. assesses functional form and variance;
4. assesses residual distribution;
5. identifies unusual, high-leverage and influential records;
6. conducts a transparent sensitivity analysis;
7. assesses multicollinearity;
8. performs a training-test validation;
9. performs k-fold cross-validation;
10. compares prediction-error measures;
11. creates reusable diagnostic functions; and
12. presents a complete regression report.

# 31. Completion checklist

Confirm that you can:

- distinguish residuals, leverage and influence;
- assess linearity;
- assess constant variance;
- assess residual normality;
- identify unusual residuals;
- calculate leverage;
- calculate Cook's distance;
- inspect DFBETAs;
- assess multicollinearity;
- conduct sensitivity analysis;
- create train-test splits;
- calculate MAE and RMSE;
- conduct k-fold cross-validation;
- assess coefficient stability; and
- report diagnostics transparently.

# Lesson conclusion

Regression diagnostics and validation determine whether a model is credible,
stable and useful beyond the fitted sample. Flagged observations should be
investigated, not automatically removed.

\[
\boxed{
\text{fit}
\rightarrow
\text{diagnose}
\rightarrow
\text{investigate}
\rightarrow
\text{validate}
\rightarrow
\text{report transparently}
}
\]
