---
title: "Lesson 39: Reproducible Statistical Analysis with R Markdown, Projects and Version Control"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

Reproducibility means that another analyst, or the same analyst at a later
date, can repeat the analytical workflow and obtain the same results from the
same data and code.

A reproducible statistical project should clearly separate:

- raw data;
- cleaned data;
- analytical code;
- functions;
- outputs;
- documentation;
- and version history.

This lesson develops a disciplined R workflow using R projects, R Markdown,
relative paths, code organisation, dependency recording, session information
and Git-based version control.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. explain reproducibility and replicability;
2. organise a statistical project;
3. use an RStudio project;
4. use relative paths;
5. separate raw and processed data;
6. create analysis scripts and reusable functions;
7. structure R Markdown documents;
8. control chunk behaviour;
9. parameterise reports;
10. record package and session information;
11. create project metadata;
12. use Git for version control;
13. write useful commit messages;
14. create a suitable `.gitignore`;
15. create a reproducibility checklist; and
16. package a project for handover.

# 1. Reproducibility and replicability

**Reproducibility** means obtaining the same result using the same data and
analysis workflow.

**Replicability** means obtaining substantively similar findings using new data
or an independently repeated study.

Both depend on transparency, but they are not the same.

# 2. Recommended project structure

A useful structure is:

```text
learning-r-project/
├── data/
│   ├── raw/
│   └── processed/
├── R/
├── reports/
├── outputs/
│   ├── figures/
│   └── tables/
├── scripts/
├── README.md
├── .gitignore
└── learning-r-project.Rproj
```

# 3. Why use an R project?

An R project provides a stable working context.

Benefits include:

- predictable relative paths;
- project-specific history;
- project-specific settings;
- easier version control;
- and reduced dependence on `setwd()`.

# 4. Avoiding hard-coded working directories

Avoid:

```r
setwd("C:/Users/Name/Documents/Project")
```

Prefer relative paths from the project root.

```{r inspect-current-directory}
getwd()
```

# 5. Relative paths

```{r relative-path-example}
student_path <- file.path(
  "data",
  "student_scores.csv"
)

file.exists(
  student_path
)
```

# 6. Importing with a relative path

```{r reproducible-import}
student_data <- read.csv(
  student_path,
  stringsAsFactors = FALSE
)

head(student_data)
```

# 7. Creating output folders safely

```{r create-output-folders}
output_directories <- c(
  file.path(
    "outputs",
    "figures"
  ),
  file.path(
    "outputs",
    "tables"
  )
)

for (directory in output_directories) {
  if (!dir.exists(directory)) {
    dir.create(
      directory,
      recursive = TRUE
    )
  }
}

directory_status <- logical(
  length(
    output_directories
  )
)

names(
  directory_status
) <- output_directories

for (index in seq_along(
  output_directories
)) {
  directory_status[index] <- dir.exists(
    output_directories[index]
  )
}

directory_status
```

# 8. Separating raw and processed data

Raw data should normally remain unchanged.

```{r raw-and-processed}
raw_student_data <- student_data

processed_student_data <- raw_student_data

processed_student_data$programme <- trimws(
  processed_student_data$programme
)

processed_student_data$passed <-
  processed_student_data$final_score >= 50

head(processed_student_data)
```

# 9. Recording transformation decisions

```{r transformation-log}
transformation_log <- data.frame(
  step = c(
    1,
    2
  ),
  variable = c(
    "programme",
    "passed"
  ),
  action = c(
    "Trim surrounding whitespace",
    "Create binary pass indicator"
  ),
  rule = c(
    "trimws(programme)",
    "final_score >= 50"
  ),
  stringsAsFactors = FALSE
)

transformation_log
```

# 10. Saving processed data

```{r save-processed-data}
processed_output_path <- file.path(
  "outputs",
  "tables",
  "processed_student_data.csv"
)

write.csv(
  processed_student_data,
  processed_output_path,
  row.names = FALSE
)

file.exists(
  processed_output_path
)
```

# 11. Script separation

A well-organised project may use:

- `01-import.R`;
- `02-clean.R`;
- `03-describe.R`;
- `04-model.R`;
- `05-report.R`.

The numerical prefixes make the intended order visible.

# 12. Reusable functions

```{r reproducible-summary-function}
summarise_numeric <- function(
  x
) {
  observed <- x[
    !is.na(x)
  ]

  data.frame(
    n = length(observed),
    missing = sum(
      is.na(x)
    ),
    mean = mean(observed),
    median = median(observed),
    standard_deviation = sd(observed),
    minimum = min(observed),
    maximum = max(observed)
  )
}

summarise_numeric(
  processed_student_data$final_score
)
```

Functions should be stored in the `R/` folder when they are reused across
scripts or reports.

# 13. Sourcing function files

```{r source-functions-template, eval=FALSE}
source(
  file.path(
    "R",
    "summary_functions.R"
  )
)
```

# 14. R Markdown structure

A reproducible R Markdown report usually contains:

1. title and metadata;
2. research question;
3. data source;
4. data preparation;
5. descriptive analysis;
6. inferential analysis;
7. diagnostics;
8. results;
9. limitations;
10. reproducibility information.

# 15. Chunk options

Common chunk options include:

- `echo`;
- `eval`;
- `include`;
- `message`;
- `warning`;
- `fig.width`;
- `fig.height`;
- `cache`.

Example:

````text
```{r example, echo=TRUE, warning=FALSE, message=FALSE}
summary(student_data)
```
````

# 16. Hiding setup code

The setup chunk often uses:

````text
```{r project-setup-example, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```
````

# 17. Inline R code

Inline code keeps prose and results synchronised.

For example:

```text
The mean score was `r round(mean(student_data$final_score), 2)`.
```

# 18. Captions and figure files

```{r reproducible-figure}
figure_path <- file.path(
  "outputs",
  "figures",
  "final_score_histogram.png"
)

png(
  figure_path,
  width = 1200,
  height = 800,
  res = 150
)

hist(
  processed_student_data$final_score,
  main = "Distribution of Final Scores",
  xlab = "Final score",
  ylab = "Frequency"
)

dev.off()

file.exists(
  figure_path
)
```

# 19. Exporting analysis tables

```{r export-summary-table}
summary_table <- summarise_numeric(
  processed_student_data$final_score
)

summary_output_path <- file.path(
  "outputs",
  "tables",
  "final_score_summary.csv"
)

write.csv(
  summary_table,
  summary_output_path,
  row.names = FALSE
)

summary_table
```

# 20. Parameterised reports

Parameters allow the same report to be rendered for different groups or
settings.

Example YAML:

```yaml
params:
  selected_programme: "Statistics"
```

Access the value with:

```r
params$selected_programme
```

# 21. Parameterised analysis example

```{r parameterised-analysis-example}
selected_programme <- "Statistics"

selected_data <- processed_student_data[
  processed_student_data$programme ==
    selected_programme,
]

summarise_numeric(
  selected_data$final_score
)
```

# 22. Recording package versions

```{r session-information}
sessionInfo()
```

Session information records:

- R version;
- operating system;
- attached packages;
- loaded namespaces;
- and package versions.

# 23. Package dependency inventory

```{r package-dependency-inventory}
required_packages <- c(
  "knitr",
  "rmarkdown",
  "ggplot2"
)

package_installed <- logical(
  length(
    required_packages
  )
)

package_version <- rep(
  NA_character_,
  length(
    required_packages
  )
)

for (index in seq_along(
  required_packages
)) {
  package_name <- required_packages[index]

  package_installed[index] <- base::requireNamespace(
    package_name,
    quietly = TRUE
  )

  if (package_installed[index]) {
    package_version[index] <- as.character(
      utils::packageVersion(
        package_name
      )
    )
  }
}

dependency_status <- data.frame(
  package =
    required_packages,
  installed =
    package_installed,
  version =
    package_version,
  stringsAsFactors = FALSE
)

dependency_status
```

# 24. Random seeds

```{r reproducible-randomness}
set.seed(20260714)

first_draw <- rnorm(5)

set.seed(20260714)

second_draw <- rnorm(5)

identical(
  first_draw,
  second_draw
)
```

A seed makes pseudo-random results reproducible.

# 25. Deterministic versus non-deterministic operations

Deterministic operations return the same output for the same input.

Random sampling, model initialisation and simulation require explicit control
of randomness.

# 26. Project README

A useful README should state:

- project purpose;
- required software;
- folder structure;
- data sources;
- execution order;
- expected outputs;
- contact or ownership information;
- and known limitations.

# 27. Basic README generator

```{r readme-template}
readme_lines <- c(
  "# Statistical Analysis Project",
  "",
  "## Purpose",
  "This project demonstrates a reproducible student-score analysis.",
  "",
  "## Main steps",
  "1. Import data",
  "2. Clean and validate data",
  "3. Produce descriptive statistics",
  "4. Fit statistical models",
  "5. Render the report",
  "",
  "## Main input",
  "- data/student_scores.csv",
  "",
  "## Main outputs",
  "- outputs/tables/",
  "- outputs/figures/"
)

readme_lines
```

# 28. Version control with Git

Git records changes to project files over time.

A basic workflow is:

```bash
git status
git add .
git commit -m "Add descriptive analysis"
git push
```

# 29. Useful commit messages

Good commit messages are:

- specific;
- concise;
- action-oriented;
- and connected to one logical change.

Examples:

```text
Add missing-value validation
Correct programme recoding rule
Create regression diagnostics report
Update README execution steps
```

Avoid:

```text
Update
Changes
Work
Final final
```

# 30. `.gitignore`

A suitable `.gitignore` may contain:

```text
.Rhistory
.RData
.Ruserdata
.Rproj.user/
*.html
*.pdf
outputs/temp/
```

Do not ignore files that are essential for reproducibility.

# 31. Files that should not enter version control

Examples may include:

- passwords;
- API keys;
- confidential raw data;
- personal identifiers;
- very large temporary files;
- local cache folders.

# 32. Data confidentiality

Reproducibility does not require public release of confidential data.

Alternatives include:

- synthetic data;
- de-identified extracts;
- controlled-access repositories;
- data dictionaries;
- reproducible code that runs in a secure environment.

# 33. Handover package

A professional handover package should include:

- project files;
- code;
- data dictionary;
- README;
- dependency information;
- output examples;
- licence information;
- and execution instructions.

# 34. Reproducibility audit function

```{r reproducibility-audit-function}
audit_project_paths <- function(
  required_paths
) {
  data.frame(
    path = required_paths,
    exists = file.exists(
      required_paths
    ) |
      dir.exists(
        required_paths
      ),
    type = {
      path_type <- rep(
        "missing",
        length(
          required_paths
        )
      )

      for (index in seq_along(
        required_paths
      )) {
        current_path <- required_paths[index]

        if (dir.exists(
          current_path
        )) {
          path_type[index] <- "directory"
        } else if (file.exists(
          current_path
        )) {
          path_type[index] <- "file"
        }
      }

      path_type
    },
    stringsAsFactors = FALSE
  )
}

required_project_paths <- c(
  "data",
  "outputs",
  file.path(
    "data",
    "student_scores.csv"
  ),
  "learning-r.Rproj"
)

audit_project_paths(
  required_project_paths
)
```

# 35. Reproducibility checklist

```{r reproducibility-checklist}
reproducibility_checklist <- data.frame(
  item = c(
    "Uses relative paths",
    "Raw data preserved",
    "Transformations documented",
    "Random seed recorded",
    "Package versions recorded",
    "README available",
    "Outputs generated by code",
    "Confidential files excluded",
    "Version history maintained"
  ),
  status = c(
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    FALSE,
    TRUE,
    TRUE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

reproducibility_checklist
```

# 36. Cleaning temporary outputs

```{r cleanup-reproducibility-output}
unlink(
  "outputs",
  recursive = TRUE,
  force = TRUE
)
```

# 37. Common mistakes

- using absolute paths;
- editing raw data manually;
- copying results into reports by hand;
- omitting package versions;
- forgetting random seeds;
- storing passwords in scripts;
- committing confidential data;
- using vague commit messages;
- relying on one undocumented script;
- creating outputs that cannot be regenerated.

# 38. Independent exercises

1. create a project folder structure.
2. use relative paths.
3. import data from the project root.
4. preserve raw data.
5. document transformations.
6. export processed data.
7. create reusable functions.
8. structure an R Markdown report.
9. use inline code.
10. create a parameterised analysis.
11. record package versions.
12. record session information.
13. set and verify random seeds.
14. write a README.
15. write a `.gitignore`.
16. create a reproducibility audit.

# 39. Lesson assessment

Create a reproducible project that:

1. uses an R project;
2. follows a clear folder structure;
3. uses relative paths;
4. preserves raw data;
5. records data-processing decisions;
6. stores reusable functions separately;
7. produces a rendered R Markdown report;
8. records session information;
9. includes a README;
10. includes a `.gitignore`;
11. uses Git commits; and
12. passes a reproducibility checklist.

# 40. Completion checklist

Confirm that you can:

- explain reproducibility;
- organise an R project;
- use relative paths;
- preserve raw data;
- separate scripts and functions;
- create R Markdown reports;
- use parameters;
- record dependencies;
- control randomness;
- write README files;
- use Git;
- write useful commits;
- protect confidential data; and
- package a project for handover.

# Lesson conclusion

Reproducibility is not an optional formatting step. It is part of statistical
quality. A transparent project allows another analyst to understand, repeat and
audit the work.

\[
\boxed{
\text{organise}
\rightarrow
\text{document}
\rightarrow
\text{automate}
\rightarrow
\text{version}
\rightarrow
\text{reproduce}
}
\]
