---
title: "Lesson 40: Integrated Capstone Analysis and Professional Reporting"
author: "Dr. Apam"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
lang: en-GB
---

```{r setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5
)
```


# Lesson overview

The final core lesson integrates the complete analytical workflow developed
across the course.

The capstone moves from a research question to a professional statistical
report through:

- project setup;
- data import;
- validation;
- cleaning;
- descriptive analysis;
- visualisation;
- inferential analysis;
- modelling;
- diagnostics;
- interpretation;
- and reproducibility.

The emphasis is not on running every available test. It is on selecting methods
that answer the stated objectives.

## Learning outcomes

After completing this lesson, the learner should be able to:

1. translate a research problem into analytical objectives;
2. define variables and measurement scales;
3. create an analysis plan;
4. import and validate data;
5. clean data transparently;
6. produce descriptive tables;
7. create analytical visualisations;
8. select inferential methods;
9. fit and diagnose regression models;
10. interpret effect sizes and uncertainty;
11. write a professional results section;
12. distinguish results from discussion;
13. produce reproducible outputs;
14. conduct a final quality audit;
15. package analysis files for assessment or handover; and
16. reflect on limitations and ethical reporting.

# 1. Capstone scenario

A university wishes to examine student performance across programmes.

The available variables include:

- student identifier;
- programme;
- coursework score;
- examination score;
- and final score.

Possible objectives are:

1. describe student performance;
2. compare performance across programmes;
3. examine the relationship between coursework and examination performance;
4. model examination score using coursework and programme;
5. identify data-quality and model-diagnostic concerns.

# 2. Importing the data

```{r capstone-import}
student_data <- read.csv(
  file.path(
    "data",
    "student_scores.csv"
  ),
  stringsAsFactors = FALSE
)

head(student_data)
str(student_data)
```

# 3. Data dictionary

```{r capstone-data-dictionary}
data_dictionary <- data.frame(
  variable = c(
    "student_id",
    "programme",
    "coursework_score",
    "examination_score",
    "final_score"
  ),
  role = c(
    "Identifier",
    "Grouping variable",
    "Numerical predictor",
    "Numerical outcome",
    "Derived numerical outcome"
  ),
  measurement = c(
    "Nominal",
    "Nominal",
    "Ratio",
    "Ratio",
    "Ratio"
  ),
  valid_rule = c(
    "Unique non-missing code",
    "Recognised programme category",
    "0 to 40",
    "0 to 60",
    "0 to 100 and equals component sum"
  ),
  stringsAsFactors = FALSE
)

data_dictionary
```

# 4. Analysis plan

```{r capstone-analysis-plan}
analysis_plan <- data.frame(
  objective = c(
    "Describe performance",
    "Compare programmes",
    "Assess association",
    "Fit adjusted model"
  ),
  outcome = c(
    "Final score",
    "Final score",
    "Examination score",
    "Examination score"
  ),
  predictor = c(
    "None",
    "Programme",
    "Coursework score",
    "Coursework score and programme"
  ),
  method = c(
    "Descriptive statistics",
    "ANOVA or Kruskal-Wallis",
    "Correlation and simple regression",
    "Multiple linear regression"
  ),
  stringsAsFactors = FALSE
)

analysis_plan
```

# 5. Structural validation

```{r capstone-structural-validation}
required_variables <- c(
  "student_id",
  "programme",
  "coursework_score",
  "examination_score",
  "final_score"
)

missing_required_variables <- setdiff(
  required_variables,
  names(
    student_data
  )
)

missing_required_variables

stopifnot(
  length(
    missing_required_variables
  ) == 0L
)
```

# 6. Identifier validation

```{r capstone-identifier-validation}
identifier_missing <- is.na(
  student_data$student_id
) |
  student_data$student_id == ""

identifier_duplicate <- duplicated(
  student_data$student_id
) |
  duplicated(
    student_data$student_id,
    fromLast = TRUE
  )

data.frame(
  missing_identifiers =
    sum(
      identifier_missing
    ),
  duplicated_identifier_rows =
    sum(
      identifier_duplicate
    )
)
```

# 7. Range validation

```{r capstone-range-validation}
coursework_invalid <-
  student_data$coursework_score < 0 |
  student_data$coursework_score > 40

examination_invalid <-
  student_data$examination_score < 0 |
  student_data$examination_score > 60

final_invalid <-
  student_data$final_score < 0 |
  student_data$final_score > 100

data.frame(
  variable = c(
    "coursework_score",
    "examination_score",
    "final_score"
  ),
  invalid_count = c(
    sum(
      coursework_invalid,
      na.rm = TRUE
    ),
    sum(
      examination_invalid,
      na.rm = TRUE
    ),
    sum(
      final_invalid,
      na.rm = TRUE
    )
  )
)
```

# 8. Arithmetic validation

```{r capstone-arithmetic-validation}
component_total <-
  student_data$coursework_score +
  student_data$examination_score

arithmetic_invalid <-
  student_data$final_score !=
  component_total

sum(
  arithmetic_invalid,
  na.rm = TRUE
)
```

# 9. Missingness profile

```{r capstone-missingness}
missingness_report <- data.frame(
  variable = names(
    student_data
  ),
  missing = vapply(
    student_data,
    function(x) {
      sum(
        is.na(x)
      )
    },
    FUN.VALUE = integer(1)
  ),
  missing_percentage = vapply(
    student_data,
    function(x) {
      mean(
        is.na(x)
      ) * 100
    },
    FUN.VALUE = numeric(1)
  ),
  stringsAsFactors = FALSE
)

missingness_report
```

# 10. Creating the analysis dataset

```{r capstone-analysis-dataset}
analysis_data <- student_data

analysis_data$programme <- factor(
  trimws(
    analysis_data$programme
  )
)

analysis_data$passed <-
  analysis_data$final_score >= 50

analysis_data$grade <- cut(
  analysis_data$final_score,
  breaks = c(
    -Inf,
    49,
    59,
    69,
    79,
    Inf
  ),
  labels = c(
    "F",
    "D",
    "C",
    "B",
    "A"
  ),
  ordered_result = TRUE
)

head(analysis_data)
```

# 11. Descriptive numerical summary function

```{r capstone-numerical-summary-function}
summarise_numeric <- function(
  x,
  variable_name
) {
  observed <- x[
    !is.na(x)
  ]

  q <- quantile(
    observed,
    c(
      0.25,
      0.50,
      0.75
    )
  )

  data.frame(
    variable = variable_name,
    n = length(observed),
    missing = sum(
      is.na(x)
    ),
    mean = mean(observed),
    median = median(observed),
    standard_deviation = sd(observed),
    minimum = min(observed),
    first_quartile = q[1],
    third_quartile = q[3],
    maximum = max(observed),
    stringsAsFactors = FALSE,
    row.names = NULL
  )
}
```

# 12. Descriptive table

```{r capstone-descriptive-table}
numerical_variables <- c(
  "coursework_score",
  "examination_score",
  "final_score"
)

descriptive_table <- do.call(
  rbind,
  lapply(
    numerical_variables,
    function(variable_name) {
      summarise_numeric(
        analysis_data[[
          variable_name
        ]],
        variable_name
      )
    }
  )
)

descriptive_table
```

# 13. Categorical summary

```{r capstone-categorical-summary}
programme_counts <- table(
  analysis_data$programme,
  useNA = "ifany"
)

programme_table <- data.frame(
  programme = names(
    programme_counts
  ),
  frequency =
    as.integer(
      programme_counts
    ),
  percentage =
    as.numeric(
      programme_counts /
      sum(
        programme_counts
      ) *
      100
    ),
  stringsAsFactors = FALSE
)

programme_table
```

# 14. Grade distribution

```{r capstone-grade-distribution}
grade_counts <- table(
  analysis_data$grade,
  useNA = "ifany"
)

data.frame(
  grade = names(
    grade_counts
  ),
  frequency =
    as.integer(
      grade_counts
    ),
  percentage =
    as.numeric(
      grade_counts /
      sum(
        grade_counts
      ) *
      100
    ),
  stringsAsFactors = FALSE
)
```

# 15. Programme group summaries

```{r capstone-programme-summary}
programme_summary <- aggregate(
  final_score ~ programme,
  data = analysis_data,
  FUN = function(x) {
    c(
      n = length(x),
      mean = mean(x),
      median = median(x),
      sd = sd(x),
      iqr = IQR(x)
    )
  }
)

programme_summary
```

# 16. Distribution visualisation

```{r capstone-histogram}
hist(
  analysis_data$final_score,
  breaks = 10,
  main = "Distribution of Final Scores",
  xlab = "Final score",
  ylab = "Frequency"
)
```

# 17. Group comparison visualisation

```{r capstone-boxplot}
boxplot(
  final_score ~ programme,
  data = analysis_data,
  main = "Final Scores by Programme",
  xlab = "Programme",
  ylab = "Final score",
  las = 2
)
```

# 18. Association visualisation

```{r capstone-scatter}
plot(
  analysis_data$coursework_score,
  analysis_data$examination_score,
  main = "Coursework and Examination Scores",
  xlab = "Coursework score",
  ylab = "Examination score",
  pch = 19
)
```

# 19. Programme comparison test

```{r capstone-anova}
anova_model <- aov(
  final_score ~ programme,
  data = analysis_data
)

summary(
  anova_model
)
```

# 20. ANOVA diagnostics

```{r capstone-anova-diagnostics}
anova_residuals <- residuals(
  anova_model
)

qqnorm(
  anova_residuals,
  main = "ANOVA Residual Q-Q Plot"
)

qqline(
  anova_residuals
)

fligner.test(
  final_score ~ programme,
  data = analysis_data
)
```

# 21. Non-parametric alternative

```{r capstone-kruskal}
kruskal.test(
  final_score ~ programme,
  data = analysis_data
)
```

# 22. Correlation analysis

```{r capstone-correlation}
pearson_result <- cor.test(
  analysis_data$coursework_score,
  analysis_data$examination_score,
  method = "pearson"
)

spearman_result <- cor.test(
  analysis_data$coursework_score,
  analysis_data$examination_score,
  method = "spearman",
  exact = FALSE
)

pearson_result
spearman_result
```

# 23. Simple regression

```{r capstone-simple-regression}
simple_model <- lm(
  examination_score ~
    coursework_score,
  data = analysis_data
)

summary(
  simple_model
)
```

# 24. Multiple regression

```{r capstone-multiple-regression}
analysis_data$programme <- relevel(
  analysis_data$programme,
  ref = "Statistics"
)

multiple_model <- lm(
  examination_score ~
    coursework_score +
    programme,
  data = analysis_data
)

summary(
  multiple_model
)
```

# 25. Regression coefficient intervals

```{r capstone-regression-intervals}
confint(
  multiple_model
)
```

# 26. Regression diagnostics

```{r capstone-regression-diagnostics}
old_parameters <- par(
  mfrow = c(
    2,
    2
  )
)

plot(
  multiple_model
)

par(
  old_parameters
)
```

# 27. Influence screening

```{r capstone-influence}
cooks_values <- cooks.distance(
  multiple_model
)

influential_rows <- which(
  cooks_values >
    4 /
    nobs(
      multiple_model
    )
)

influential_rows
```

# 28. Adjusted predictions

```{r capstone-prediction-grid}
prediction_grid <- expand.grid(
  coursework_score = c(
    20,
    30,
    40
  ),
  programme =
    levels(
      analysis_data$programme
    )
)

prediction_grid$predicted_examination <- predict(
  multiple_model,
  newdata = prediction_grid
)

prediction_grid
```

# 29. Model-reporting function

```{r capstone-model-report-function}
report_lm <- function(
  model
) {
  model_summary <- summary(
    model
  )

  coefficient_table <- as.data.frame(
    model_summary$coefficients
  )

  coefficient_table$term <- rownames(
    coefficient_table
  )

  rownames(
    coefficient_table
  ) <- NULL

  intervals <- as.data.frame(
    confint(model)
  )

  intervals$term <- rownames(
    intervals
  )

  rownames(
    intervals
  ) <- NULL

  names(
    intervals
  )[1:2] <- c(
    "confidence_lower",
    "confidence_upper"
  )

  coefficient_table <- merge(
    coefficient_table,
    intervals,
    by = "term",
    sort = FALSE
  )

  fit_table <- data.frame(
    n = nobs(model),
    r_squared =
      model_summary$r.squared,
    adjusted_r_squared =
      model_summary$adj.r.squared,
    residual_standard_error =
      model_summary$sigma,
    AIC = AIC(model),
    BIC = BIC(model)
  )

  list(
    coefficients =
      coefficient_table,
    fit =
      fit_table
  )
}

report_lm(
  multiple_model
)
```

# 30. Results writing framework

A professional results section should:

1. begin with sample characteristics;
2. present descriptive results;
3. report inferential results by objective;
4. include effect sizes and confidence intervals;
5. reference tables and figures;
6. avoid repeating every table entry;
7. distinguish evidence from interpretation;
8. avoid causal language unless justified.

# 31. Results versus discussion

**Results** state what the analysis found.

**Discussion** explains:

- what the findings mean;
- how they compare with previous evidence;
- possible mechanisms;
- implications;
- limitations;
- and future research directions.

# 32. Example descriptive result

```{r capstone-descriptive-text}
overall_mean <- mean(
  analysis_data$final_score,
  na.rm = TRUE
)

overall_sd <- sd(
  analysis_data$final_score,
  na.rm = TRUE
)

descriptive_text <- paste(
  "The mean final score was",
  round(
    overall_mean,
    2
  ),
  "with a standard deviation of",
  round(
    overall_sd,
    2
  ),
  "."
)

descriptive_text
```

# 33. Example regression result

```{r capstone-regression-text}
model_summary <- summary(
  multiple_model
)

coursework_row <- model_summary$coefficients[
  "coursework_score",
]

regression_text <- paste(
  "After adjustment for programme, a one-mark increase in coursework score",
  "was associated with an estimated",
  round(
    coursework_row[
      "Estimate"
    ],
    3
  ),
  "mark change in examination score, p =",
  format.pval(
    coursework_row[
      "Pr(>|t|)"
    ],
    digits = 3,
    eps = 0.001
  ),
  "."
)

regression_text
```

# 34. Ethical statistical reporting

Ethical reporting requires:

- reporting all planned analyses;
- avoiding selective p-value reporting;
- documenting exclusions;
- preserving confidentiality;
- acknowledging uncertainty;
- avoiding fabricated precision;
- and stating limitations honestly.

# 35. Final quality audit

```{r capstone-quality-audit}
quality_audit <- data.frame(
  check = c(
    "Required variables present",
    "Identifiers checked",
    "Ranges checked",
    "Arithmetic consistency checked",
    "Missingness reported",
    "Descriptive statistics produced",
    "Visualisations produced",
    "Inferential methods justified",
    "Diagnostics assessed",
    "Effect estimates and intervals reported",
    "Limitations stated",
    "Reproducibility information recorded"
  ),
  completed = c(
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

quality_audit
```

# 36. Exporting final tables

```{r capstone-export-tables}
capstone_output_directory <- file.path(
  "capstone_output"
)

if (!dir.exists(
  capstone_output_directory
)) {
  dir.create(
    capstone_output_directory
  )
}

write.csv(
  descriptive_table,
  file.path(
    capstone_output_directory,
    "descriptive_statistics.csv"
  ),
  row.names = FALSE
)

write.csv(
  programme_table,
  file.path(
    capstone_output_directory,
    "programme_distribution.csv"
  ),
  row.names = FALSE
)
```

# 37. Exporting a final figure

```{r capstone-export-figure}
png(
  file.path(
    capstone_output_directory,
    "programme_boxplot.png"
  ),
  width = 1200,
  height = 800,
  res = 150
)

boxplot(
  final_score ~ programme,
  data = analysis_data,
  main = "Final Scores by Programme",
  xlab = "Programme",
  ylab = "Final score",
  las = 2
)

dev.off()
```

# 38. Capstone handover contents

A final submission should contain:

```text
capstone-project/
├── data/
├── R/
├── scripts/
├── reports/
├── outputs/
├── README.md
├── data_dictionary.csv
├── analysis_plan.csv
├── session-info.txt
└── project.Rproj
```

# 39. Cleaning temporary outputs

```{r capstone-cleanup}
unlink(
  capstone_output_directory,
  recursive = TRUE,
  force = TRUE
)
```

# 40. Common mistakes

- starting analysis without objectives;
- selecting tests after seeing results;
- failing to validate identifiers and ranges;
- presenting descriptive and inferential results inconsistently;
- reporting p-values without estimates or intervals;
- ignoring model assumptions;
- using causal language from cross-sectional data;
- copying outputs manually into reports;
- failing to document exclusions;
- submitting a project that another analyst cannot run.

# 41. Independent capstone tasks

1. define a research problem.
2. write four measurable objectives.
3. create a data dictionary.
4. create an analysis plan.
5. validate identifiers and ranges.
6. document missingness.
7. create an analysis dataset.
8. produce descriptive tables.
9. produce three analytical graphs.
10. conduct a group-comparison test.
11. conduct a correlation analysis.
12. fit simple and multiple regression models.
13. assess diagnostics.
14. produce adjusted predictions.
15. write results paragraphs.
16. conduct a quality audit.
17. package the project for handover.

# 42. Final core-course assessment

The learner must submit a complete reproducible project that includes:

1. an R project;
2. a README;
3. a data dictionary;
4. an analysis plan;
5. raw and processed data folders;
6. documented cleaning code;
7. descriptive tables;
8. publication-ready graphs;
9. at least two inferential analyses;
10. a regression model;
11. diagnostic evidence;
12. a rendered report;
13. session information;
14. a reproducibility checklist; and
15. a reflective note on limitations and ethical reporting.

# 43. Completion checklist

Confirm that you can:

- move from a research question to an analysis plan;
- define variable roles;
- validate and clean data;
- create descriptive summaries;
- create suitable visualisations;
- choose inferential methods;
- fit regression models;
- assess diagnostics;
- report uncertainty;
- write results and discussion separately;
- document limitations;
- export outputs;
- conduct a quality audit;
- package a reproducible project; and
- hand over a professional analysis.

# Lesson conclusion

The capstone demonstrates that statistical programming is not a collection of
isolated commands. It is a disciplined workflow that links research questions,
data quality, appropriate methods, transparent interpretation and
reproducibility.

\[
\boxed{
\text{question}
\rightarrow
\text{plan}
\rightarrow
\text{validate}
\rightarrow
\text{analyse}
\rightarrow
\text{diagnose}
\rightarrow
\text{report}
\rightarrow
\text{reproduce}
}
\]
