---
title: "Learning R"
subtitle: "From First Principles to Advanced Statistical Computing"
author: "Benjamin Apam (PhD)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
documentclass: book
classoption:
  - openany
fontsize: 11pt
geometry:
  - margin=1in
lang: en-GB
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    fig_caption: true
    highlight: tango
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{float}
  - \usepackage{caption}
  - \usepackage{setspace}
  - \usepackage{fancyhdr}
  - \usepackage{titlesec}
  - \usepackage{xcolor}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue}
  - \setstretch{1.15}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[LE,RO]{Learning R}
  - \fancyhead[RE,LO]{\leftmark}
  - \fancyfoot[C]{\thepage}
  - \setlength{\headheight}{14pt}
  - \setcounter{secnumdepth}{3}
  - \setcounter{tocdepth}{2}
---

```{r book-setup, include=FALSE}
options(
  digits = 4,
  scipen = 999,
  width = 90,
  encoding = "UTF-8"
)

Sys.setenv(
  LANG = "en_US.UTF-8",
  LC_ALL = "en_US.UTF-8"
)

set.seed(20260714)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  error = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "centre",
  fig.width = 7,
  fig.height = 5,
  fig.pos = "H"
)

# Permit the master document to handle labels after prefixing them.
options(
  knitr.duplicate.label = "allow"
)

lesson_files <- c(
  "00-before-you-start.Rmd",
  "01-understanding-r.Rmd",
  "02-setting-up-r.Rmd",
  "03-first-r-session.Rmd",
  "04-objects-and-assignment.Rmd",
  "05-data-types.Rmd",
  "06-vectors.Rmd",
  "07-matrices-and-arrays.Rmd",
  "08-lists-data-frames-and-tibbles.Rmd",
  "09-indexing-and-subsetting.Rmd",
  "10-operators-and-logical-conditions.Rmd",
  "11-conditional-execution.Rmd",
  "12-loops-and-repetition.Rmd",
  "13-writing-functions.Rmd",
  "14-function-scope-and-environments.Rmd",
  "15-importing-delimited-files.Rmd",
  "16-importing-excel-spss-stata-sas.Rmd",
  "17-data-inspection-and-validation.Rmd",
  "18-missing-values-and-incomplete-data.Rmd",
  "19-character-data-and-regular-expressions.Rmd",
  "20-dates-times-and-durations.Rmd",
  "21-factors-and-categorical-data.Rmd",
  "22-sorting-filtering-and-recoding.Rmd",
  "23-combining-joining-and-reshaping-data.Rmd",
  "24-descriptive-statistics.Rmd",
  "25-frequency-tables-and-grouped-summaries.Rmd",
  "26-base-r-data-visualisation.Rmd",
  "27-ggplot2-data-visualisation.Rmd",
  "28-exploratory-data-analysis.Rmd",
  "29-correlation-and-visual-association.Rmd",
  "30-sampling-distributions-and-standard-errors.Rmd",
  "31-confidence-intervals-and-estimation.Rmd",
  "32-hypothesis-testing-and-t-tests.Rmd",
  "33-analysis-of-variance-and-post-hoc-tests.Rmd",
  "34-chi-square-tests-for-categorical-data.Rmd",
  "35-non-parametric-and-rank-based-tests.Rmd",
  "36-simple-linear-regression.Rmd",
  "37-multiple-linear-regression.Rmd",
  "38-regression-diagnostics-and-validation.Rmd",
  "39-reproducible-analysis-and-version-control.Rmd",
  "40-integrated-capstone-analysis.Rmd"
)

include_solutions <- TRUE

# Keep FALSE for the book build. When FALSE, solution scripts are displayed
# as syntax-highlighted R code but are not executed.
evaluate_solutions <- TRUE

solution_files <- c(
  "01" = file.path(
    "solutions",
    "01-understanding-r-solutions.R"
  ),
  "02" = file.path(
    "solutions",
    "02-setting-up-r-solutions.R"
  ),
  "03" = file.path(
    "solutions",
    "03-first-r-session-solutions.R"
  ),
  "04" = file.path(
    "solutions",
    "04-objects-and-assignment-solutions.R"
  ),
  "05" = file.path(
    "solutions",
    "05-data-types-solutions.R"
  ),
  "06" = file.path(
    "solutions",
    "06-vectors-solutions.R"
  ),
  "07" = file.path(
    "solutions",
    "07-matrices-and-arrays-solutions.R"
  ),
  "08" = file.path(
    "solutions",
    "08-lists-data-frames-and-tibbles-solutions.R"
  ),
  "09" = file.path(
    "solutions",
    "09-indexing-and-subsetting-solutions.R"
  ),
  "10" = file.path(
    "solutions",
    "10-operators-and-logical-conditions-solutions.R"
  ),
  "11" = file.path(
    "solutions",
    "11-conditional-execution-solutions.R"
  ),
  "12" = file.path(
    "solutions",
    "12-loops-and-repetition-solutions.R"
  ),
  "13" = file.path(
    "solutions",
    "13-writing-functions-solutions.R"
  ),
  "14" = file.path(
    "solutions",
    "14-function-scope-and-environments-solutions.R"
  ),
  "15" = file.path(
    "solutions",
    "15-importing-delimited-files-solutions.R"
  ),
  "16" = file.path(
    "solutions",
    "16-importing-excel-spss-stata-sas-solutions.R"
  ),
  "17" = file.path(
    "solutions",
    "17-data-inspection-and-validation-solutions.R"
  ),
  "18" = file.path(
    "solutions",
    "18-missing-values-and-incomplete-data-solutions.R"
  ),
  "19" = file.path(
    "solutions",
    "19-character-data-and-regular-expressions-solutions.R"
  ),
  "20" = file.path(
    "solutions",
    "20-dates-times-and-durations-solutions.R"
  ),
  "21" = file.path(
    "solutions",
    "21-factors-and-categorical-data-solutions.R"
  ),
  "22" = file.path(
    "solutions",
    "22-sorting-filtering-and-recoding-solutions.R"
  ),
  "23" = file.path(
    "solutions",
    "23-combining-joining-and-reshaping-data-solutions.R"
  ),
  "24" = file.path(
    "solutions",
    "24-descriptive-statistics-solutions.R"
  ),
  "25" = file.path(
    "solutions",
    "25-frequency-tables-and-grouped-summaries-solutions.R"
  ),
  "26" = file.path(
    "solutions",
    "26-base-r-data-visualisation-solutions.R"
  ),
  "27" = file.path(
    "solutions",
    "27-ggplot2-data-visualisation-solutions.R"
  ),
  "28" = file.path(
    "solutions",
    "28-exploratory-data-analysis-solutions.R"
  ),
  "29" = file.path(
    "solutions",
    "29-correlation-and-visual-association-solutions.R"
  ),
  "30" = file.path(
    "solutions",
    "30-sampling-distributions-and-standard-errors-solutions.R"
  ),
  "31" = file.path(
    "solutions",
    "31-confidence-intervals-and-estimation-solutions.R"
  ),
  "32" = file.path(
    "solutions",
    "32-hypothesis-testing-and-t-tests-solutions.R"
  ),
  "33" = file.path(
    "solutions",
    "33-analysis-of-variance-and-post-hoc-tests-solutions.R"
  ),
  "34" = file.path(
    "solutions",
    "34-chi-square-tests-for-categorical-data-solutions.R"
  ),
  "35" = file.path(
    "solutions",
    "35-non-parametric-and-rank-based-tests-solutions.R"
  ),
  "36" = file.path(
    "solutions",
    "36-simple-linear-regression-solutions.R"
  ),
  "37" = file.path(
    "solutions",
    "37-multiple-linear-regression-solutions.R"
  ),
  "38" = file.path(
    "solutions",
    "38-regression-diagnostics-and-validation-solutions.R"
  ),
  "39" = file.path(
    "solutions",
    "39-reproducible-analysis-and-version-control-solutions.R"
  ),
  "40" = file.path(
    "solutions",
    "40-integrated-capstone-analysis-solutions.R"
  )
)

if (include_solutions) {
  missing_solutions <- solution_files[
    !file.exists(
      solution_files
    )
  ]

  if (length(missing_solutions) > 0L) {
    stop(
      paste0(
        "The following solution files are missing:\n",
        paste(
          missing_solutions,
          collapse = "\n"
        )
      ),
      call. = FALSE
    )
  }
}


missing_lessons <- lesson_files[
  !file.exists(
    lesson_files
  )
]

if (length(missing_lessons) > 0L) {
  stop(
    paste0(
      "The following lesson files are missing from the project root:\n",
      paste(
        missing_lessons,
        collapse = "\n"
      )
    ),
    call. = FALSE
  )
}

if (!dir.exists("data")) {
  stop(
    "The `data/` directory is missing from the project root.",
    call. = FALSE
  )
}

extract_yaml_title <- function(
  lines,
  fallback
) {
  title_line <- grep(
    "^title:[[:space:]]*",
    lines,
    value = TRUE
  )

  if (length(title_line) == 0L) {
    return(fallback)
  }

  title <- sub(
    "^title:[[:space:]]*",
    "",
    title_line[1]
  )

  title <- sub(
    '^"(.*)"$',
    "\\1",
    title
  )

  title <- sub(
    "^'(.*)'$",
    "\\1",
    title
  )

  title
}

remove_yaml_header <- function(
  lines
) {
  if (
    length(lines) >= 3L &&
    trimws(lines[1]) == "---"
  ) {
    closing_positions <- which(
      trimws(lines) == "---"
    )

    if (length(closing_positions) >= 2L) {
      closing <- closing_positions[2]

      if (closing < length(lines)) {
        return(
          lines[
            seq.int(
              closing + 1L,
              length(lines)
            )
          ]
        )
      }

      return(character(0))
    }
  }

  lines
}

remove_setup_chunk <- function(
  lines
) {
  output_lines <- character(0)
  inside_removed_chunk <- FALSE

  for (line in lines) {
    if (!inside_removed_chunk) {
      is_setup_header <- grepl(
        "^```\\{r[[:space:]]+setup([,}[:space:]])",
        line
      )

      if (is_setup_header) {
        inside_removed_chunk <- TRUE
      } else {
        output_lines <- c(
          output_lines,
          line
        )
      }
    } else if (trimws(line) == "```") {
      inside_removed_chunk <- FALSE
    }
  }

  output_lines
}

prefix_chunk_labels <- function(
  lines,
  prefix
) {
  vapply(
    lines,
    function(line) {
      if (!grepl(
        "^```\\{r",
        line
      )) {
        return(line)
      }

      # An unnamed chunk starts with ```{r} or ```{r, ...}.
      if (grepl(
        "^```\\{r[},]",
        line
      )) {
        return(line)
      }

      sub(
        "^```\\{r[[:space:]]+([^,}[:space:]]+)",
        paste0(
          "```{r ",
          prefix,
          "-\\1"
        ),
        line
      )
    },
    FUN.VALUE = character(1),
    USE.NAMES = FALSE
  )
}

demote_markdown_headings <- function(
  lines
) {
  output_lines <- character(
    length(lines)
  )

  inside_fence <- FALSE
  fence_marker <- NULL

  for (index in seq_along(lines)) {
    line <- lines[index]
    trimmed <- trimws(line)

    if (grepl(
      "^(```|~~~)",
      trimmed
    )) {
      marker <- substr(
        trimmed,
        1,
        3
      )

      if (!inside_fence) {
        inside_fence <- TRUE
        fence_marker <- marker
      } else if (marker == fence_marker) {
        inside_fence <- FALSE
        fence_marker <- NULL
      }

      output_lines[index] <- line
      next
    }

    if (
      !inside_fence &&
      grepl(
        "^#{1,5}[[:space:]]",
        line
      )
    ) {
      output_lines[index] <- paste0(
        "#",
        line
      )
    } else {
      output_lines[index] <- line
    }
  }

  output_lines
}


normalise_utf8 <- function(
  lines
) {
  converted <- iconv(
    lines,
    from = "",
    to = "UTF-8",
    sub = ""
  )

  converted[
    is.na(converted)
  ] <- ""

  converted
}

disable_environment_check_chunks <- function(
  lines
) {
  environment_check_labels <- c(
    "check-pandoc",
    "check-pandoc-available",
    "check-xelatex",
    "check-r-version",
    "session-information"
  )

  vapply(
    lines,
    function(line) {
      if (!grepl(
        "^```\\{r",
        line
      )) {
        return(line)
      }

      matching_label <- environment_check_labels[
        vapply(
          environment_check_labels,
          function(label) {
            grepl(
              paste0(
                "^```\\{r[[:space:]]+",
                label,
                "([,}[:space:]])"
              ),
              line
            )
          },
          FUN.VALUE = logical(1)
        )
      ]

      if (
        length(matching_label) == 0L ||
        grepl(
          "(^|[,[:space:]])eval[[:space:]]*=",
          line
        )
      ) {
        return(line)
      }

      sub(
        "\\}[[:space:]]*$",
        ", eval=FALSE}",
        line
      )
    },
    FUN.VALUE = character(1),
    USE.NAMES = FALSE
  )
}

prepare_lesson <- function(
  file_path,
  lesson_prefix
) {
  original_lines <- readLines(
    file_path,
    warn = FALSE,
    encoding = "UTF-8",
    skipNul = TRUE
  )

  original_lines <- normalise_utf8(
    original_lines
  )

  title <- extract_yaml_title(
    original_lines,
    fallback = tools::file_path_sans_ext(
      basename(file_path)
    )
  )

  body_lines <- remove_yaml_header(
    original_lines
  )

  body_lines <- remove_setup_chunk(
    body_lines
  )

  body_lines <- disable_environment_check_chunks(
    body_lines
  )

  body_lines <- prefix_chunk_labels(
    body_lines,
    lesson_prefix
  )

  body_lines <- demote_markdown_headings(
    body_lines
  )

  # Remove excessive blank lines at the beginning and end.
  while (
    length(body_lines) > 0L &&
    trimws(body_lines[1]) == ""
  ) {
    body_lines <- body_lines[-1]
  }

  while (
    length(body_lines) > 0L &&
    trimws(
      body_lines[
        length(body_lines)
      ]
    ) == ""
  ) {
    body_lines <- body_lines[
      -length(body_lines)
    ]
  }

  list(
    title = title,
    body = body_lines
  )
}


escape_code_fence <- function(
  lines
) {
  # A four-backtick outer fence safely contains ordinary three-backtick text.
  c(
    "````r",
    lines,
    "````"
  )
}

prepare_solution <- function(
  file_path
) {
  solution_lines <- readLines(
    file_path,
    warn = FALSE,
    encoding = "UTF-8",
    skipNul = TRUE
  )

  solution_lines <- normalise_utf8(
    solution_lines
  )

  while (
    length(solution_lines) > 0L &&
    trimws(solution_lines[1]) == ""
  ) {
    solution_lines <- solution_lines[-1]
  }

  while (
    length(solution_lines) > 0L &&
    trimws(
      solution_lines[
        length(solution_lines)
      ]
    ) == ""
  ) {
    solution_lines <- solution_lines[
      -length(solution_lines)
    ]
  }

  solution_lines
}

render_solution <- function(
  solution_file,
  lesson_number,
  lines_per_block = 60L
) {
  if (!file.exists(
    solution_file
  )) {
    stop(
      paste(
        "Solution file not found:",
        solution_file
      ),
      call. = FALSE
    )
  }

  solution_lines <- readLines(
    solution_file,
    warn = FALSE,
    encoding = "UTF-8",
    skipNul = TRUE
  )

  solution_lines <- iconv(
    solution_lines,
    from = "",
    to = "UTF-8",
    sub = ""
  )

  solution_lines[
    is.na(
      solution_lines
    )
  ] <- ""

  cat(
    "\n\n## Suggested Solutions {-}\n\n"
  )

  cat(
    "> Attempt the lesson exercises before consulting these suggested ",
    "solutions. Alternative correct approaches are possible.\n\n",
    sep = ""
  )

  if (
    length(
      solution_lines
    ) == 0L
  ) {
    cat(
      "_No solution code was found in this file._\n\n"
    )

    return(
      invisible(NULL)
    )
  }

  block_number <- ceiling(
    seq_along(
      solution_lines
    ) /
      lines_per_block
  )

  solution_blocks <- split(
    solution_lines,
    block_number
  )

  total_blocks <- length(
    solution_blocks
  )

  for (current_block in seq_along(
    solution_blocks
  )) {
    cat(
      "```r\n"
    )

    cat(
      paste(
        solution_blocks[[
          current_block
        ]],
        collapse = "\n"
      )
    )

    cat(
      "\n```\n\n"
    )

    if (
      current_block <
        total_blocks
    ) {
      cat(
        "<!-- Solution code continues -->\n\n"
      )
    }
  }

  invisible(NULL)
}

part_titles <- c(
  "01" = "R Foundations",
  "09" = "Accessing and Controlling Data",
  "15" = "Importing, Cleaning and Transforming Data",
  "24" = "Descriptive Analysis and Data Visualisation",
  "30" = "Statistical Inference and Hypothesis Testing",
  "36" = "Regression Modelling and Reproducible Statistical Workflows"
)
```

# Preface {-}

This book combines the complete **Learning R** core course into one
reproducible document. It contains a preparatory chapter and 40 progressively
structured lessons, beginning with the foundations of R and ending with
regression modelling, reproducibility and an integrated capstone analysis.

The individual lesson files remain the authoritative source documents. This
master file reads them in numerical order, removes their separate document
headers, gives every code chunk a unique book-level label and compiles the
content into one PDF.

When `include_solutions` is set to `TRUE`, the corresponding solution script is
placed after each lesson. By default, solutions are printed as
syntax-highlighted R code without being executed during the book build.

\newpage

```{r assemble-book, results='asis', echo=FALSE}
for (lesson_file in lesson_files) {
  lesson_number <- substr(
    basename(lesson_file),
    1,
    2
  )

  if (
    lesson_number %in%
      names(
        part_titles
      )
  ) {
    cat(
      "\n\n\\part{",
      part_titles[
        lesson_number
      ],
      "}\n\n",
      sep = ""
    )
  }

  prepared <- prepare_lesson(
    lesson_file,
    lesson_prefix = paste0(
      "lesson",
      lesson_number
    )
  )

  if (
    lesson_number == "00"
  ) {
    cat(
      "\n\n# ",
      prepared$title,
      " {-}\n\n",
      sep = ""
    )
  } else {
    cat(
      "\n\n# ",
      prepared$title,
      "\n\n",
      sep = ""
    )
  }

  lesson_environment <- new.env(
    parent = globalenv()
  )

  knitted_lesson <- knitr::knit_child(
    text = prepared$body,
    envir = lesson_environment,
    quiet = TRUE,
    encoding = "UTF-8"
  )

  cat(
    paste(
      knitted_lesson,
      collapse = "\n"
    )
  )

  if (
    include_solutions &&
    lesson_number != "00"
  ) {
    render_solution(
      solution_file =
        solution_files[
          lesson_number
        ],
      lesson_number =
        lesson_number
    )
  }

  cat(
    "\n\n\\clearpage\n\n"
  )
}
```
